### 2019-01-03 Paper/Paper.tex
Added my local paper writings.

%\documentclass[twocolumn]{article}
\documentclass{article}
\usepackage{outline}
\usepackage{float}
\usepackage{graphicx}
\usepackage{color}
\usepackage{amsmath}

%% The following should allow me to make subsubsubsections with the command \paragraph{}
\usepackage{titlesec}

\setcounter{secnumdepth}{4}

\titleformat{\paragraph}
{\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
\titlespacing*{\paragraph}
{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
%%

\graphicspath{ {Figs/} }

\begin{document}

 \title{Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer}
 \maketitle

\section{Abstract}
%% Basic Introduction (1-2 sentences)
During tumorigenesis and throughout the lifetime of a tumor, the immune system is constantly engaged in complex interactions in and around the tumor.
%% More Detailed Background (2-3 sentences)
These interactions are influenced by the tumor microenvironment (TME) and in particular by the cellular process of epithelial-to-mesenchymal transition (EMT) in an as-yet unexplained manner.
In turn, both the tumor and the immune system exert substantial influence over the TME in competition to influence the fate of the host.
%% General Problem (1 sentence)
Determining the critical factors and mechanisms involved remains a crucial task for cancer immunotherapy.
%% Summarizing Main Result (1 sentence)
Here, we seek to understand how interplay between the tumor, the immune system, and EMT can determine the outcome of cancer.
%% The Main Result (2-3 sentences)
We developed a multiscale agent-based model that describes tumor-immune interactions and how EMT modulates these.
Exploration of the model revealed that increasing mesenchymal cells' quiescence in some circumstances results in a more tumor-friendly environment and sometimes less.
In contrast, enhancing mesenchymal cells' immune evasiveness always leads to a more tumor-friendly microenvironment.
Moreover, encouraging the TME towards a pro-EMT state also benefits the tumor.
%% Putting into a More General Context (1-2 sentences)
Together, these results indicate that the joint regulation of the TME by the immune system and the tumor itself provides possible targets for future cancer therapies, in particular cancers such as pancreatic cancer in which EMT is speculated to play a role.

%%%%%%%%%%%%%%%%%%%%%%%
%                    INTRODUCTION                    %
%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
Immunotherapy is revolutionizing cancer therapy.
Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival.
This is an exciting and fast-paced time in the field and many believe it just the beginning.
There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer.
While the direct cell-cell interactions between immune cells and cancer cells is an unsurprising line of inquiry, how immune cells interact with and modulate the tumor microenvironment (TME) is also a promising lead.

This is because immune cells are best understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context.
Understanding each step in this immune cascade in the context of the TME will be paramount in fully exploring and exploiting all options for immunotherapy.

One aspect of the TME long suspected of playing a vital role in tumor initiation, progression, and metastasis is the epithelial-to-mesenchymal transition (EMT). 
This is the process by which rigidly bound epithelial cells leave a layer of cells and gain motility as mesenchymal cells, among other properties.
As mesenchymal cells often express stem-like properties, their existence has been tied to the initiation and progression of cancer.
Also, due to their motility, some have viewed them as a natural precursor to metastasis.

Thus, herein we explore the interconnectedness of these three features: the tumor itself, the immune system, and the EMT process.
In addition to the interactions already mentioned above, the immune system also affects and is affected by EMT.
The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME.
In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$).
Conversely, the EMTed cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell.

Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
Each of these three pieces will be briefly explored in turn for how they will influence one another.

\subsection{The Tumor}\label{TheTumor}
The tumor influences the immune system by its presence and by the cellular processes it engages in that result in shaping the TME.
On the one hand, the mutated DNA of tumor cells results in these cells being recognized as foreign by members of the innate immune system such as Natural Killer cells and Dendritic cells.
It is then these cells that initiate the cascade of cytokines that result in the full activation of the body's immune system to fight the non-self cells.
On the other hand, the tumor cells actively work to shape the immune response that the body launches.
This is accomplished through its own set of cytokines, such as TGF-$\beta$, that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Treg cells.

As mentioned above, TGF-$\beta$ is one driver of EMT, and so it is through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells that the tumor encourages itself and the surrounding tissue towards a more mesenchymal profile.

\subsection{The Immune System}\label{TheImmuneSystem}
The immune system has several competing interactions with the tumor.
While the cytotoxic immune cells, NKs and CTLs, seek out and lyse tumor cells, the regulatory branch of the immune system, Tregs, slows down the effector functions of these other cells.

These Tregs also drive the EMT process by releasing TGF-$\beta$ upon arriving at the tumor site.

\subsection{EMT}\label{EMT}
There are many characteristics distinctive of mesenchymal cells and still much research being done in further quantifying these and determining others.
Of interest here, the literature shows two qualities of mesenchymal cells that are of interest particularly in the context of cancer and the immune system.
The first is that mesenchymal cells are less susceptible to immune clearance.
The second is that mesenchymal cells proliferate at a slower rate.

The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.
Once cytotoxic immune cells, for example natural killer cells and cytotoxic T cells, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse.
However, due to down-regulated surface markers on mesenchymal cells, they are refractory to this immunological task thus rendering them less susceptible to immune clearance.
In line with known motility of mesenchymal cells, scientists have referred to this condition as mesenchymal cells being ``slippery.''
This quality of mesenchymal cells is then quantified in our model by the parameter, MIE, for mesenchymal immune evasion.

The other facet of mesenchymal cells of importance in this research is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions.
While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate.
In quantifying this property, we use the model parameter, MQ, for mesenchymal quiescence.
This choice is less about associating this property with quiescence and more about creating a simple name that easily, if not roughly, captures the essence of what is being modeled.
For example, while mesenchymal proliferation reduction might be a more accurate term, thinking in terms of a negative formulation adds one additional twist to accurately interpreting results which we felt was more detrimental than the possible confusion of invoking ``quiescence.''

%%%%%%%%%%%%%%%%%%%%%%%
%                         METHODS                        %
%%%%%%%%%%%%%%%%%%%%%%%

\section{Methods}
To address the questions around the relationships among cancer, the immune system, and EMT, we built a multiscale, agent-based model to explore.
The agents in the model are the tissue cells that can either be healthy or have mutations.
It is worth noting that unlike many agent-based models, the spatial arrangement of these cells is not part of the model.
Thus, it is assumed that all the cells are in effect well-mixed as are all the relevant cytokines and other molecules acting on the system.
The immune cells are present in aggregate, i.e. only the total quantity of them present in the tumor microenvironment (TME) is used.
Finally, it is the tissue cells that can be labeled either epithelial or mesenchymal, and this labeling is fluid and depends both on the TME and the internal workings of each individual cell.

\subsection{Tissue Cells}\label{TissueCells}
The tissue cells at all times have associated with them two important quantities: a logical vector indicating pathway mutations and an EMT score.
The model has three pathways that have the potential to mutate over the course of a simulation.
There is a proliferation pathway that when mutated increases the probability of the cell proliferating.
There is an apoptosis pathway that when mutated decreases the probability of the cell undergoing apoptosis.
There is an immune evasion pathway that when mutated decreases the probability that an immune cell can effectively clear the mutated cell.
% It might be nice to have a table to summarize this information?
The EMT score will be exposited in Section \ref{EMT}.

Add differential equations? % Adam suggested this, but I'm not sure what this means.
\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: Natural Killer cells (NKs), cytotoxic T cells (CTLs), and T regulatory cells (Tregs).
The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
Should they clear a cell, they themselves are then deactivated and removed from the immune population.
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear mutated cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

\subsection{EMT}\label{EMT}
The EMT score indicates how likely the cell is begin the next cell cycle as an epithelial or a mesenchymal cell.\
The score is between $0$ and $1$ with higher values being connected to higher probability of being mesenchymal.
For each patient, there is a fixed mesenchymal threshold.
If the EMT score ends the cycle above this threshold, the cell is mesenchymal, otherwise it is epithelial for the next cycle.
For the purposes of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have some of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear mutated cells, the increased immune evasion only benefits mutated mesenchymal cells.

\subsection{Running the Model}

\subsubsection{Initializing the Model}
Every model run starts with a fixed amount of cells, $N_0$.
However, different parameter values will lead to different steady states for the population of tissue cells.
In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.
Thus, the only immune cell present is NKs.
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a discrete, random jump in their probability to mutate.
From then on, a cell which proliferates either mutates or increases its probability of mutating later on.

\subsubsection{Cell Fate}
During each cell cycle, every cell randomly is assigned a cell fate from the following options:
\begin{itemize}
\item proliferation
\item apoptosis
\item immune clearance (by NKs or CTLs)
\item quiescence
\end{itemize}

For each cell, a weight is chosen for each option and these are normalized to probabilities which then are used to randomly determine what each cell does during the cell cycle.

\paragraph{Proliferation}
There are four factors that contribute to a cell's probability to weight.
The first is a base proliferation weight that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway, then the weight for proliferation is proportionally increased by $\text{prolif}_{\text{up}}$.
Third, if the cell is mesenchymal, then the weight for proliferation is proportionally decreased by $m_Q$, which stands for mesenchymal quiescence.
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor.
In total, the weight for proliferation is given by

$$ p_{\text{prolif}} = p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})(1-\delta_{MES}m_Q)\frac{N_0}{N_0+N} $$

\paragraph{Apoptosis}
There are two factors that contribute to a cell's weight for undergoing apoptosis.
There is a basal apoptosis rate that all cells experience, $d$ for death.
Second, if the cell has a mutation in the apoptosis pathway, then the weight for undergoing apoptosis is proportionally decreased by $\text{apop}_{\text{down}}$.
In total, the weight for apoptosis is given by

$$ p_{\text{apop}} = d(1-\delta_{\text{AM}}\text{apop}_{\text{down}}) $$

\paragraph{Immune Clearance}
For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
First of all, the cell needs to be mutated, $\delta_M$.
Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell.
Third, NKs and CTLs have their own efficacy parameters which can be understood as their probability of clearing a mutated cell given that they have interacted with it.
Fourth, there is a decreasing Hill factor based on the number of Treg cells present.
Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation or if it is mesenchymal.
In total, the weight of NK clearance is given by

$$ p_{\text{NKed}} =  \delta_M \frac{\text{NK}}{N/N_{00}+\text{NK}} E_{\text{NK}} \frac{1}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} $$ $$ (1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}})$$

A similar formula holds for CTLs with only the number of CTLs and their efficacy being different from the above equation.

\paragraph{Quiescence}
The weight associated with quiescence is taken as 1 except in the case of Mesenchymal cells.
For Mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
Hence, the weight of quiescence is given by

$$ p_Q = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_Q\frac{N_0}{N_0+N} $$

The reason for adding that term is due to the understanding that mesenchymal cells are not proliferating as fast because they are instead remaining quiescent, hence naming the parameter $m_Q$.

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
This represents the fact that individual immune cells have a limited amount of efficacy.

Also, all proliferating cells have a probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
If they do not mutate, then their probability of mutating increases.

Then, the EMT values for each cell is updated.
This depends on how much TGF-$\beta$ is currently in the system.
Each cell receives an amount of TGF-$\beta$ given by

$$ \text{TGFB}_i = \text{TGFB}/N + X_i, \quad X_i \sim N(0,\sigma^2)  $$

where $\sigma$ is a parameter in the model.
This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
If $\text{TGFB}_i>2-\text{EMT}_i$, the cell is EMTed and will have a higher EMT score.
If $\text{TGFB}_i<2-\text{EMT}_i$, the cell is METed and will have a lower EMT score.
The reason for the expression $2-\text{EMT}_i$ can be understood this way: 
the bounds on the EMT score guarantee $2-\text{EMT}_i\in[1,2]$.
Hence, there either needs to be a lot of TGF-$\beta$ in the system or a large amount of randomness for a cell to undergo EMT.
In addition, a cell having a higher EMT score will have a higher chance of continuing in the EMT process whereas a cell low on the EMT spectrum will have a higher chance of moving towards a more epithelial phenotype.

Next, the amount of TGF-$\beta$ for the next cell cycle is determined by the number of mutated cells and the number of Treg cells, each one producing a fixed amount of TGF-$\beta$. It is given by

$$ \text{TGFB} = \text{tgfb}_{\text{mut}}N_{\text{mut}} + \text{tgfb}_{\text{Treg}}\text{Treg}$$

Finally, the immune populations are updated.
For the NKs, they obey the following differential equation:

$$ \text{NK}' = \sigma_{\text{NK}} - d_{\text{NK}}\text{NK} $$

which is discretized to

$$ \text{NK}_{i+1} = \left (\text{NK}_i-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} $$

For CTLs and Tregs, however, they rely on antigen presenting cells (APCs) to be activated, and these APCs in turn rely on mutated cells being cleared so that they can bind antigens.
In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function.
Thus, let $\text{APC}$ represent the number of mutated cells cleared by the immune system during a given cycle.
We choose the following differential equations to govern the CTL and Treg populations:

\begin{align*}
\text{CTL}' & = \sigma_{\text{CTL}}\text{APC} - d_{\text{CTL}}\text{CTL} \\
\text{Treg}' & = \sigma_{\text{Treg}}\text{APC} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}- d_{\text{Treg}}\text{Treg}
\end{align*}

Discretized, these are:

\begin{align*}
\text{CTL}_{i+1} & = & \left (\text{CTL}_i-\sigma_{\text{CTL}}\text{APC}/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}\text{APC}/d_{\text{CTL}}\\
\text{Treg}_{i+1} & = & \left (\text{Treg}_i-\frac{\sigma_{\text{Treg}}\text{APC}}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}\right )\exp(-d_{\text{Treg}}\text{Treg}\Delta t)+\\
& & \frac{\sigma_{\text{Treg}}\text{APC}}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}
\end{align*}

\subsubsection{Cancer Onset}
At the end of each cell cycle, the proportion of tissue cells that are mutated is checked, and if it is above a certain threshold the patient is determined to have developed cancer and a Time to Cancer is recorded.
If not, the process continues until cancer is detected or a preset number of cycles has been completed.
It is also possible with the right combination of parameters, to have patients run out of cells entirely.
The model does not distinguish between these patients and those that have a cancer diagnosis.
The reason is that these patients have succumb to some combination of inflammatory and mutagenic events.

%%%%%%%%%%%%%%%%%%%%%%%
%                          RESULTS                         %
%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}

\begin{figure}[H]
\center
\frame{\includegraphics[width=\columnwidth]{Figure1/Figure1.pdf}}
\caption{A. A cartoon of the cell cycle with the immune system interactions as an inset.
B. Global sensitivity analysis.
C. Example survival curve for one cohort.
D. Example of one simulated patient's dynamics. The warmup period for the model is plotted in red.}
\label{fig:ModelIntro}
\end{figure}

\subsection{Exploring the Model}\label{ExplModel}
Refer to Figure \ref{fig:ModelIntro}A for a cartoon of the model.
For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}D.
Notice that at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541.
Also of interest, as the patient's cells are all rapidly mutating right around this time, most of them are also transitioning to a mesenchymal phenotype with all of them being in a mesenchymal phenotype for a short time during the time period that cancer onset occurs.
Finally, considering a cohort of $N=500$ patients, we can generate a survival curve for these patients tracking how many have survived cancer-free up to a give day.
See Figure \ref{fig:ModelIntro}C for such a survival curve.
In this curve, we see that all patients survive for some minimum amount of time before a time period a nearly constant cancer onset rate roughly between $T = 100$ and $T = 300$.

\subsection{Sensitivity Analysis}\label{SensAnalysis}
To assess the sensitivity to the many model parameters, a Morris OAT algorithm was implemented.
The parameters were all given Gaussian distributions centered at their initial value and a large standard deviation.
Those that needed bounds were truncated.
The results of the Morris OAT can be found in Figure \ref{fig:ModelIntro}B.
Notice that the following parameters have the most global influence over the system.
While it makes sense for such and such parameters to have large influences over patient outcomes, these fall outside of the scope of the current paper and instead it is preferable to focus on the following subset.

\subsection{Mesenchymal Cell Properties Dramatically Change Survival Outcomes}\label{MesPars}
The two quantities that change when a cell transitions to a mesenchymal state are its immune evasion and its increased quiescence, MIE and MQ, respectively.
Also involved in the EMT process, is the cytokine TGF-$\beta$, which upregulates EMT.
By the Morris OAT analysis presented in Section \ref{SensAnalysis}, these three parameters do have large impacts on survival outcomes and each warrant exploration.

First, as mesenchymal immune evasion (MIE) increases, Time to Cancer decreases.
This result holds universally among all parameter sets and indicates the simple relationship that mesenchymal immune evasion has with time to cancer.
As this subpopulation of the tumor becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus more dangerous.
These results are summarized by Figure \ref{fig:FirstSurvivalCurves}A.

\begin{figure}[H]
\center
\frame{\includegraphics[width=\columnwidth]{Figure2/Figure2.pdf}}
\caption{A. Increasing MIE decreases Time to Cancer. 
B. Increasing MQ increases Time to Cancer.
C. Increasing Treg TGF-$\beta$ production results in decreased Time to Cancer.
%% {\color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.}\newline
\newline
}
\label{fig:FirstSurvivalCurves}
\end{figure}

Second, as mesenchymal quiescence (MQ) increases, Time to Cancer increases.
This result will prove to be more interesting as this pattern is dependent on the other parameters in the model.
For now, what can be seen in Figure \ref{fig:FirstSurvivalCurves}B is that a slower proliferation for mesenchymal cells slows down cancer growth.
This is not as intuitive as it may sound because this decreased proliferation affects both mutated and non-mutated cells.
% It may be interesting to track how much MQ affects mutated vs. non-mutated cells throughout the time.
These results are summarized by Figure \ref{fig:FirstSurvivalCurves}B.

Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
In Figure \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production is shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
This is expected as the two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing cells to a mesenchymal phenotype.
Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease Time to Cancer.
Mesenchymal cells are more likely to evade the immune system, so pushing the system towards an overall more mesenchymal phenotype will better protect the cancer and decrease the Time to Cancer.
% But what happens when MIE is really high to begin with so becoming mesenchymal mostly just makes you proliferate slower? Perhaps that can go in a supplementary figure?

\subsection{Key EMT regime maximizes cancer-free survival time under chronic inflammation}\label{KeyEMT}
Given the relevance of the inflammatory state of the tumor microenvironment (TME), we explore the effect of varying the inflammation state of the patient on survival.
In particular, some cohorts are assigned permanently low inflammation, others permanently high, and still others non-constant inflammatory schemes.
For those cohorts with a permanently high inflammatory state, the relationship between the mesenchymal parameters and the Time to Cancer is monotonic.
However, when there are periods of a low inflammatory state, then intriguingly the connection between MQ and Time to Cancer becomes concave down with a local maximum around 0.3.
What this indicates is that if the quiescence of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer and the time to death.
Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
See Figure \ref{fig:VaryINFL_and_MesPars}C and Figure \ref{fig:VaryINFL_and_MesPars}D for a summary.

\begin{figure}[H]
\center
\frame{\includegraphics[width=\columnwidth]{Figure3/Figure3.pdf}}
\caption{A.Survival curves for various inflammation cycling schemes. In the top two plots, the Time to Cancer decreases as MIE increases. In the bottom plot, there is no significant difference in survival.
B. The average Time to Cancer for each cohort.
C. Survival curves for various inflammation cycling schemes. In the top two plots, the Time to Cancer reaches a maximum value somewhere near MQ = 0.3. In the bottom plot, there is no significant difference in survival.
D. The average Time to Cancer for each cohort.}
\label{fig:VaryINFL_and_MesPars}
\end{figure}

Contrast this with what happens when MIE is varied for different inflammation cycling schemes.
Regardless of the cycling scheme, the relationship between Time to Cancer and MIE is never non-monotonic.
In the case that the inflammatory state is permanently high, MIE has no significant effect on Time to Cancer.
On the other hand, when the patient experiences at least some time in a low inflammatory state, an increase in MIE results in a decrease in Time to Cancer.
This indicates that reducing the inflammatory state of the patient is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
See Figure \ref{fig:VaryINFL_and_MesPars}A and Figure \ref{fig:VaryINFL_and_MesPars}B for a summary.

\subsection{Heat Map of MIE vs MQ}
To further draw out the contrast between these two mesenchymal properties, we compare Time to Cancer for many different pairs of values for MIE and MQ.
The heat map in Figure \ref{fig:MIEvsMQ} concisely encodes the relationships we have been observing in the previous sections.
Regardless of the value of MQ, increasing MIE leads to decreased Time to Cancer.
On the other hand, for any value of MIE, then there is an optimal MQ value in the sense that it results in the longest Time to Cancer.
It appears based on the plot that this value is around 0.28.

\begin{figure}[H]
\center
\frame{\includegraphics[width=\columnwidth]{Figure4/MIEvsMQ.jpg}}
\caption{Heat Map contrasting the effects of MIE and MQ on Time to Cancer. Increasing MIE always decreases Time to Cancer, but MQ is non-monotonically linked to Time to Cancer.}
\label{fig:MIEvsMQ}
\end{figure}

\section{Discussion}\label{Discussion}
This might actually be where I talk about the implication of the results from Section \ref{KeyEMT}.
In addition, I could discuss the implications from the TGF-$\beta$ section.
Any discussion on this topic, however, would need to acknowledge that regulation of TGF-$\beta$ might also have other, serious side effects, especially if the treatment is not local.
However, that might also be true for the inflammatory results.
In light of this, it might be good to include a new section on how TGF-$\beta$ production's influence on Time to Cancer is influenced by the inflammatory cycling scheme.

%%%%%%%%%%%%%%%%%%%%%%%
%                     CONCLUSION                      %
%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}\label{Conclusion}

\end{document}

### 2019-01-03 Paper/Figs/Figure2/ChangVaryMIEColors.m

@@ -1,0 +1,20 @@
+close all; clear all;
+
+openfig('VaryMIE_BigChange.fig');
+h=gca;
+
+for i = 1:length(h.Children)
+    ht = h.Children(i);
+    if endsWith(class(ht),'Patch')
+        col = ht.FaceColor;
+        col(1:2) = col(2:-1:1);
+        set(ht,'FaceColor',col);
+    elseif endsWith(class(ht),'Stair')
+        col = ht.Color;
+        col(1:2) = col(2:-1:1);
+        set(ht,'Color',col);
+    end
+end
+
+savefig('VaryMIE_BigChange.fig')
+print('VaryMIE_BigChange.jpg','-djpeg')

### 2019-01-11 Paper/Paper.tex
Fixed Fig 1 Mut Example
The mutant example on the right was shifted too far down. I fixed this.

@@
 Refer to Figure \ref{fig:ModelIntro}A for a cartoon of the model.
+Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
+If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the example cell on the right.
+If any component of a cell has a pathway mutation, that cell is then counted as a mutated cell and contributes towards the total number of cancer cells.
+Each cell also has a color along the EMT spectrum seen on the left of the cell block. 
+Biologically, cells can be anywhere along this continuum with research indicating there are several equilibria along the spectrum.
+In our model, this is actually just a binary and that is why the cells shown are at one extreme or the other of the spectrum.
+In the right half of Figure \ref{fig:ModelIntro}A one can see the cell cycle choices that each cell can make each cell cycle.
+These are the four thick, gray arrows.
+Each of the 
 For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}D.

### 2019-01-11 Paper/Paper.tex
Edited section 4.1 and redid single patient
Single patient has INFL cycling scheme more in line with other figs.
Before it was H/L = 60/30, now it is 30/30.

@@
 In our model, this is actually just a binary and that is why the cells shown are at one extreme or the other of the spectrum.
+
 In the right half of Figure \ref{fig:ModelIntro}A one can see the cell cycle choices that each cell can make each cell cycle.
 These are the four thick, gray arrows.
-Each of the 
-For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}D.
+Each of the cells will choose which arrow to follow each cell cycle, and this decision is influenced by the properties of the cell as well as external factors like the immune system and the TME.
+The inset on the left gives an overview of how the immune system works.
+The dashed arrows indicate that the action is probabilistic rather than deterministic.
+In other words, the cytotoxic cells will randomly clear some subset of mutated cells and TGF-$\beta$ increases the probability of cell transitioning or staying in a mesenchymal state.
+
+For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}B.
 Notice that at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541.
+Once the Time to Cancer has been determined for a patient, most simulations shown later stop.
+We do this because once such a high proportion of cells have mutated, the mutant cells quickly take over and make up 100\% of all cells.
 Also of interest, as the patient's cells are all rapidly mutating right around this time, most of them are also transitioning to a mesenchymal phenotype with all of them being in a mesenchymal phenotype for a short time during the time period that cancer onset occurs.
+
+Considering the immune populations, notice that the NK population basically stays constant while the lymphocyte populations quickly dwarf the NK population once cells begin to mutate.
+These lymphocyte populations exhibit oscillatory behavior but this is an artifact of the inflammation scheme this patient is undergoing.
+Every 30 cell cycles the inflammation changes from High to Low or vice versa.
+In these roughly month-long periods, the lymphocyte populations increase or decrease rapidly in accordance with this inflammation.
+
 Finally, considering a cohort of $N=500$ patients, we can generate a survival curve for these patients tracking how many have survived cancer-free up to a give day.

### 2019-01-11 Paper/Paper.tex
Updated Morris_OAT Fig
Used better priors for Morris_OAT (gaussian) and singled out a few
parameters
Updated section 4.2
Fleshed out the analysis of the Morris OAT figure.

@@
 Those that needed bounds were truncated.
-The results of the Morris OAT can be found in Figure \ref{fig:ModelIntro}B.
-Notice that the following parameters have the most global influence over the system.
-While it makes sense for such and such parameters to have large influences over patient outcomes, these fall outside of the scope of the current paper and instead it is preferable to focus on the following subset.
+The results of the Morris OAT can be found in Figure \ref{fig:ModelIntro}D.
+Clearly, a certain subset of parameters have an outsized effect on the model while others do relatively little.
+The two most influential according to this analysis are related to Treg cells: their death rate and recruitment.
+This makes sense as Treg cells both suppress cytotoxic effects of other immune cells and also secrete significant quantities of TGF-$\beta$ which drives EMT.
+Thus, Treg cells are intimately tied to all three aspects of the model and so their presence or lack thereof significantly effects the system.
+However, we want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MQ are of interest.
+In addition, inflammation parameters dictating the cycling scheme are interesting.
+What we can see from this plot is that the model also has some high sensitivity to these parameters.
+In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and so we will also look at that parameter.
 

### 2019-01-11 Paper/Paper.tex
MQ --> MGA
Replaced MQ with MGA in the text. Maybe later will update figs.

@@
 While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate.
-In quantifying this property, we use the model parameter, MQ, for mesenchymal quiescence.
+In quantifying this property, we use the model parameter, MGA, for mesenchymal quiescence.
 This choice is less about associating this property with quiescence and more about creating a simple name that easily, if not roughly, captures the essence of what is being modeled.
@@
 Thus, Treg cells are intimately tied to all three aspects of the model and so their presence or lack thereof significantly effects the system.
-However, we want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MQ are of interest.
+However, we want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MGA are of interest.
 In addition, inflammation parameters dictating the cycling scheme are interesting.
@@
 \subsection{Mesenchymal Cell Properties Dramatically Change Survival Outcomes}\label{MesPars}
-The two quantities that change when a cell transitions to a mesenchymal state are its immune evasion and its increased quiescence, MIE and MQ, respectively.
+The two quantities that change when a cell transitions to a mesenchymal state are its immune evasion and its increased quiescence, MIE and MGA, respectively.
 Also involved in the EMT process, is the cytokine TGF-$\beta$, which upregulates EMT.
@@
 \caption{A. Increasing MIE decreases Time to Cancer. 
-B. Increasing MQ increases Time to Cancer.
+B. Increasing MGA increases Time to Cancer.
 C. Increasing Treg TGF-$\beta$ production results in decreased Time to Cancer.
@@
 
-Second, as mesenchymal quiescence (MQ) increases, Time to Cancer increases.
+Second, as mesenchymal growth arrest (MGA) increases, Time to Cancer increases.
 This result will prove to be more interesting as this pattern is dependent on the other parameters in the model.
@@
 This is not as intuitive as it may sound because this decreased proliferation affects both mutated and non-mutated cells.
-% It may be interesting to track how much MQ affects mutated vs. non-mutated cells throughout the time.
+% It may be interesting to track how much MGA affects mutated vs. non-mutated cells throughout the time.
 These results are summarized by Figure \ref{fig:FirstSurvivalCurves}B.
@@
 For those cohorts with a permanently high inflammatory state, the relationship between the mesenchymal parameters and the Time to Cancer is monotonic.
-However, when there are periods of a low inflammatory state, then intriguingly the connection between MQ and Time to Cancer becomes concave down with a local maximum around 0.3.
+However, when there are periods of a low inflammatory state, then intriguingly the connection between MGA and Time to Cancer becomes concave down with a local maximum around 0.3.
 What this indicates is that if the quiescence of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer and the time to death.
@@
 Survival curves for various inflammation cycling schemes.
-In the top two plots, the Time to Cancer reaches a maximum value somewhere near MQ = 0.3.
+In the top two plots, the Time to Cancer reaches a maximum value somewhere near MGA = 0.3.
 In the bottom plot, there is no significant difference in survival.
@@
 
-\subsection{Heat Map of MIE vs MQ}
-To further draw out the contrast between these two mesenchymal properties, we compare Time to Cancer for many different pairs of values for MIE and MQ.
-The heat map in Figure \ref{fig:MIEvsMQ} concisely encodes the relationships we have been observing in the previous sections.
-Regardless of the value of MQ, increasing MIE leads to decreased Time to Cancer.
-On the other hand, for any value of MIE, then there is an optimal MQ value in the sense that it results in the longest Time to Cancer.
+\subsection{Heat Map of MIE vs MGA}
+To further draw out the contrast between these two mesenchymal properties, we compare Time to Cancer for many different pairs of values for MIE and MGA.
+The heat map in Figure \ref{fig:MIEvsMGA} concisely encodes the relationships we have been observing in the previous sections.
+Regardless of the value of MGA, increasing MIE leads to decreased Time to Cancer.
+On the other hand, for any value of MIE, then there is an optimal MGA value in the sense that it results in the longest Time to Cancer.
 It appears based on the plot that this value is around 0.28.
@@
 \frame{\includegraphics[width=\columnwidth]{Figure4/MIEvsMQ.jpg}}
-\caption{Heat Map contrasting the effects of MIE and MQ on Time to Cancer. Increasing MIE always decreases Time to Cancer, but MQ is non-monotonically linked to Time to Cancer.}
-\label{fig:MIEvsMQ}
+\caption{Heat Map contrasting the effects of MIE and MGA on Time to Cancer. Increasing MIE always decreases Time to Cancer, but MGA is non-monotonically linked to Time to Cancer.}
+\label{fig:MIEvsMGA}
 \end{figure}

### 2019-01-11 Paper/Paper.tex
edits up to Results S2

--- 2019-01-11 Paper/Paper.tex
@@
 \setcounter{secnumdepth}{4}
-
 \titleformat{\paragraph}
@@
 
-\section{Abstract}
+\section*{Abstract}
 %% Basic Introduction (1-2 sentences)
@@
 \section{Introduction}
-Immunotherapy is revolutionizing cancer therapy.
-Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival.
-This is an exciting and fast-paced time in the field and many believe it just the beginning.
-There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer.
-While the direct cell-cell interactions between immune cells and cancer cells is an unsurprising line of inquiry, how immune cells interact with and modulate the tumor microenvironment (TME) is also a promising lead.
-
-This is because immune cells are best understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context.
+Immunotherapy is slowly revolutionizing how we treat cancer.
+Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival. Ref  [examples] 
+There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area.
+Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context.
+Ref
 Understanding each step in this immune cascade in the context of the TME will be paramount in fully exploring and exploiting all options for immunotherapy.
 
-One aspect of the TME long suspected of playing a vital role in tumor initiation, progression, and metastasis is the epithelial-to-mesenchymal transition (EMT). 
-This is the process by which rigidly bound epithelial cells leave a layer of cells and gain motility as mesenchymal cells, among other properties.
-As mesenchymal cells often express stem-like properties, their existence has been tied to the initiation and progression of cancer.
-Also, due to their motility, some have viewed them as a natural precursor to metastasis.
+Recent work on cancer-immune interactions [briefly describe Guo paper]...
 
-Thus, herein we explore the interconnectedness of these three features: the tumor itself, the immune system, and the EMT process.
-In addition to the interactions already mentioned above, the immune system also affects and is affected by EMT.
-The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME.
-In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$).
+Epithelial-to-mesenchymal transition (EMT) describes the (reversible) cell plasticity process by which epithelial cells transition into mesenchymal cells.
+EMT has an important yet complicated role in tumor initiation, progression, and metastasis.
+Ref
+As mesenchymal cells often exhibit stem-like properties, their existence has been tied to the initiation and progression of cancer. Ref
+Also, due to their motility, some have viewed them as a natural precursor to metastasis. Ref
+
+Herein, we explore the interconnectedness of these three key systems: the tumor, the immune system, and EMT.
+In addition to the interactions already discussed, the immune system affects and is affected by EMT.
+The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME. Ref 
+In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$). Ref 
 Conversely, the EMTed cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell.
-
 Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
@@
 
+[No subsections headings within Introduction]
 \subsection{The Tumor}\label{TheTumor}
 The tumor influences the immune system by its presence and by the cellular processes it engages in that result in shaping the TME.
-On the one hand, the mutated DNA of tumor cells results in these cells being recognized as foreign by members of the innate immune system such as Natural Killer cells and Dendritic cells.
+On the one hand, the mutated DNA of tumor cells results in these cells being recognized as foreign by members of the innate immune system such as natural killer (NK) cells and dendritic cells.
 It is then these cells that initiate the cascade of cytokines that result in the full activation of the body's immune system to fight the non-self cells.
 On the other hand, the tumor cells actively work to shape the immune response that the body launches.
-This is accomplished through its own set of cytokines, such as TGF-$\beta$, that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Treg cells.
-
-As mentioned above, TGF-$\beta$ is one driver of EMT, and so it is through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells that the tumor encourages itself and the surrounding tissue towards a more mesenchymal profile.
+This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of T regulatory (Treg) cells.
+Since TGF-$\beta$ is also a driver of EMT, then through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells, cells of the tumor and the surrounding tissue are driven towards a more mesenchymal profile.
 
@@
 The immune system has several competing interactions with the tumor.
-While the cytotoxic immune cells, NKs and CTLs, seek out and lyse tumor cells, the regulatory branch of the immune system, Tregs, slows down the effector functions of these other cells.
+While the cytotoxic immune cells, NKs and [define!] CTLs, seek out and lyse tumor cells, the regulatory branch of the immune system, Tregs, slows down the effector functions of these other cells.
 
@@
 Of interest here, the literature shows two qualities of mesenchymal cells that are of interest particularly in the context of cancer and the immune system.
-The first is that mesenchymal cells are less susceptible to immune clearance.
-The second is that mesenchymal cells proliferate at a slower rate.
+The first is that mesenchymal cells are less susceptible to immune clearance. Ref 
+The second is that mesenchymal cells proliferate at a slower rate. Ref
 
@@
 Once cytotoxic immune cells, for example natural killer cells and cytotoxic T cells, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse.
-However, due to down-regulated surface markers on mesenchymal cells, they are refractory to this immunological task thus rendering them less susceptible to immune clearance.
+Ref
+However, due to down-regulated surface markers on mesenchymal cells, they are refractory to this immunological task thus rendering them less susceptible to immune clearance. Ref 
 In line with known motility of mesenchymal cells, scientists have referred to this condition as mesenchymal cells being ``slippery.''
@@
 While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate.
-In quantifying this property, we use the model parameter, MQ, for mesenchymal quiescence.
+In quantifying this property, we use the model parameter `mesenchymal growth arrest', MGA.
 This choice is less about associating this property with quiescence and more about creating a simple name that easily, if not roughly, captures the essence of what is being modeled.
 For example, while mesenchymal proliferation reduction might be a more accurate term, thinking in terms of a negative formulation adds one additional twist to accurately interpreting results which we felt was more detrimental than the possible confusion of invoking ``quiescence.''
+
+In the following sections of the paper we first introduce the model modularly, which is appropriate diven its multiscale and hybrid nature.
+We then perform sensitivity analysis and identify parameters that are crucial for the progression to tumorigenesis.
+We analyze the competing effects of EMT and of the immune system on tumorigenesis and find that EMT intricately regulates tumorigenesis: under certain regimes a careful balance of EMT- and immune-driven process can significantly prolong cancer-free survival.
 
@@
 \section{Methods}
-To address the questions around the relationships among cancer, the immune system, and EMT, we built a multiscale, agent-based model to explore.
-The agents in the model are the tissue cells that can either be healthy or have mutations.
-It is worth noting that unlike many agent-based models, the spatial arrangement of these cells is not part of the model.
-Thus, it is assumed that all the cells are in effect well-mixed as are all the relevant cytokines and other molecules acting on the system.
-The immune cells are present in aggregate, i.e. only the total quantity of them present in the tumor microenvironment (TME) is used.
-Finally, it is the tissue cells that can be labeled either epithelial or mesenchymal, and this labeling is fluid and depends both on the TME and the internal workings of each individual cell.
+[In this section it is important to say what is old (from previous model), i.e. `following the model of XX, we....' and what is new/our addition.]
+
+To address the questions around the relationships among cancer, the immune system, and EMT, we built a multiscale, agent-based model.
+The agents in the model are the tissue cells with tumorigenic potential, that can either be mutation-free, or have mutations.
+It is worth noting that we do not consider spatial dependencies between agents, i.e. we assume that cells are well-mixed as are all the relevant cytokines and other molecules acting on the system.
+We model the effects of immune cell types and cytokines as continuous fields within the tumor microenvironment (TME), since here single-cell resolution is not required. 
+Finally, tissue cells that can be labeled either epithelial or mesenchymal, and this labeling is fluid and depends both on the TME and the internal workings of each individual cell.
 
@@
 Add differential equations? % Adam suggested this, but I'm not sure what this means.
+% I think this comment applied to the next section: add the equations describing immune system
+
 \subsection{The Immune System}\label{ImmuneSystem}
@@
 
-\subsection{Running the Model}
+\subsection{Model Simulation}
 
-\subsubsection{Initializing the Model}
-Every model run starts with a fixed amount of cells, $N_0$.
+\subsubsection{Initial conditions} Every model run starts with a fixed amount of cells, $N_0$.
 However, different parameter values will lead to different steady states for the population of tissue cells.
@@
 
-\subsubsection{Cell Fate}
+\subsubsection{Tissue cell fate}
 During each cell cycle, every cell randomly is assigned a cell fate from the following options:
@@
 \paragraph{Proliferation}
-There are four factors that contribute to a cell's probability to weight.
-The first is a base proliferation weight that all cells have, $p$.
+There are four factors that contribute to a cell's probability of proliferation.
+% NB try to avoid posessive apostrophe. I.e. replace this with
+There are four factors that contribute to the probability that a cell will proliferate.
+% 
+The first is a base proliferation probability that all cells have, $p$.
 Second, if the cell has a mutation in the proliferation pathway, then the weight for proliferation is proportionally increased by $\text{prolif}_{\text{up}}$.
@@
 If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
-If they do not mutate, then their probability of mutating increases.
+If they do not mutate, then their probability of mutating increases. [??]
 
@@
 
-\subsubsection{Cancer Onset}
-At the end of each cell cycle, the proportion of tissue cells that are mutated is checked, and if it is above a certain threshold the patient is determined to have developed cancer and a Time to Cancer is recorded.
-If not, the process continues until cancer is detected or a preset number of cycles has been completed.
-It is also possible with the right combination of parameters, to have patients run out of cells entirely.
-The model does not distinguish between these patients and those that have a cancer diagnosis.
-The reason is that these patients have succumb to some combination of inflammatory and mutagenic events.
+\subsubsection{Progression and tumorigenesis}
+At the end of each cell cycle, the proportion of tissue cells that are mutated is calculated, and if it is above a certain threshold, a tumorigenic event is recorded and the patient is determined to have cancer (using a similar strategy to Guo).
+The Time to Cancer is given as the time from the start of the simulation until tumorigenesis.
+Simulations run until either a tumorigenic event is recorded or until the maximum number of cycles has been reached.
+%It is also possible with the right combination of parameters, to have patients run out of cells entirely.
+%The model does not distinguish between these patients and those that have a cancer diagnosis.
+%The reason is that these patients have succumb to some combination of inflammatory and mutagenic events.
 
@@
 \frame{\includegraphics[width=\columnwidth]{Figure1/Figure1.pdf}}
-\caption{A. A cartoon of the cell cycle with the immune system interactions as an inset.
-B. Global sensitivity analysis.
-C. Example survival curve for one cohort.
-D. Example of one simulated patient's dynamics. The warmup period for the model is plotted in red.}
+\caption{A.
+Schematic depiction of the model with blue/red denoting non-mutated/mutated cells in the agent-based model.
+In each cycle, the fate of a cell is regulated by the EMT state of the cell and by any pathway mutations harbored.
+Inset shows major processes of immune system regulation on the tissue cells.
+B. Example simulation of the dynamics of a one patient; warmup period is shown in red.
+C. Example survival curve for one cohort of patients with parameters XXXXX.
+D. Global sensitivity analysis of XXX parameters using XXXXX.}
 \label{fig:ModelIntro}
@@
 \subsection{Exploring the Model}\label{ExplModel}
-Refer to Figure \ref{fig:ModelIntro}A for a cartoon of the model.
+In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
 Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
-If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the example cell on the right.
-If any component of a cell has a pathway mutation, that cell is then counted as a mutated cell and contributes towards the total number of cancer cells.
-Each cell also has a color along the EMT spectrum seen on the left of the cell block. 
-Biologically, cells can be anywhere along this continuum with research indicating there are several equilibria along the spectrum.
+If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the annotation to the right of the cell block.
+If any component [???] of a cell has a pathway mutation, that cell is then counted as a mutated cell and contributes towards the total number of cancer cells.
+The EMT score of each cell (denoting position along the EMT spectrum) is associated with a color given by the legend to the left of the cell block. 
+In the model the EMT spectrum is continuous.
+While biologically, there may be only a few discrete stable cell states available on the EMT spectrum, recent studies highlight a growing number of states accessible, with research indicating there are several equilibria along the spectrum.
+Ref [Tian and Levine lab]
 In our model, this is actually just a binary and that is why the cells shown are at one extreme or the other of the spectrum.
@@
 For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}B.
-Notice that at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541.
-Once the Time to Cancer has been determined for a patient, most simulations shown later stop.
+Notice that at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541. [translate into days?]
+Once the Time to Cancer has been determined for a patient, most simulations shown later stop. [??? mean to say: we stop the simulation?]
 We do this because once such a high proportion of cells have mutated, the mutant cells quickly take over and make up 100\% of all cells.
@@
 Considering the immune populations, notice that the NK population basically stays constant while the lymphocyte populations quickly dwarf the NK population once cells begin to mutate.
-These lymphocyte populations exhibit oscillatory behavior but this is an artifact of the inflammation scheme this patient is undergoing.
-Every 30 cell cycles the inflammation changes from High to Low or vice versa.
-In these roughly month-long periods, the lymphocyte populations increase or decrease rapidly in accordance with this inflammation.
+These lymphocyte populations exhibit oscillatory behavior, however this is not due to intrinsic dynamics but rather due to the inflammation scheme that this patient is undergoing: every 30 cell cycles, the inflammation level changes from high to low, or vice versa.
+In each of these periods periods, the lymphocyte populations increase or decrease rapidly in accordance with the inflammation.
 

### 2019-01-12 Paper/Paper.tex
comments in results sections

@@
 \subsection{Sensitivity Analysis}\label{SensAnalysis}
-To assess the sensitivity to the many model parameters, a Morris OAT algorithm was implemented.
+To assess the sensitivity to the many model parameters [how many?], a Morris OAT algorithm was implemented. [need to define] Ref 
 The parameters were all given Gaussian distributions centered at their initial value and a large standard deviation.
@@
 The results of the Morris OAT can be found in Figure \ref{fig:ModelIntro}B.
-Notice that the following parameters have the most global influence over the system.
-While it makes sense for such and such parameters to have large influences over patient outcomes, these fall outside of the scope of the current paper and instead it is preferable to focus on the following subset.
+Notice that the following parameters have the most global influence over the system. [which parameters?]
+While it makes sense for such and such [??] parameters to have large influences over patient outcomes, these fall outside of the scope of the current paper and instead it is preferable to focus on the following subset.
 
@@
 However, when there are periods of a low inflammatory state, then intriguingly the connection between MQ and Time to Cancer becomes concave down with a local maximum around 0.3.
+[what does this value mean] [refer to figure]
 What this indicates is that if the quiescence of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer and the time to death.

### 2019-01-12 Paper/Paper.tex
My first fixes to Adam's

@@
 In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$). Ref 
-Conversely, the EMTed cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell.
+Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell.
 Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
@@
 On the other hand, the tumor cells actively work to shape the immune response that the body launches.
-This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of T regulatory (Treg) cells.
+This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Tregs.
 Since TGF-$\beta$ is also a driver of EMT, then through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells, cells of the tumor and the surrounding tissue are driven towards a more mesenchymal profile.
@@
 The immune system has several competing interactions with the tumor.
-While the cytotoxic immune cells, NKs and [define!] CTLs, seek out and lyse tumor cells, the regulatory branch of the immune system, Tregs, slows down the effector functions of these other cells.
+While the cytotoxic immune cells, NKs and cytotoxic T cells (CTLs), seek out and lyse tumor cells, the regulatory branch of the immune system, Tregs, slows down the effector functions of these other cells.
 
@@
 \subsection{The Immune System}\label{ImmuneSystem}
-The immune system is modeled as having three active immune cell types: Natural Killer cells (NKs), cytotoxic T cells (CTLs), and T regulatory cells (Tregs).
+The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
 The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
@@
 Also, all proliferating cells have a probability of undergoing a pathway mutation.
-If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
-If they do not mutate, then their probability of mutating increases. [??]
+If they do, one pathway is randomly chosen among the three and is mutated.
+If the cell does not undergo a mutation, then its probability of mutating during subsequent cell increases. [**]
 
@@
 This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
-If $\text{TGFB}_i>2-\text{EMT}_i$, the cell is EMTed and will have a higher EMT score.
-If $\text{TGFB}_i<2-\text{EMT}_i$, the cell is METed and will have a lower EMT score.
+If $\text{TGFB}_i>2-\text{EMT}_i$, the cell becomes more mesenchymal and will have a higher EMT score.
+If $\text{TGFB}_i<2-\text{EMT}_i$, the cell becomes more epithelial and will have a lower EMT score.
 The reason for the expression $2-\text{EMT}_i$ can be understood this way: 

### 2019-01-12 Paper/Paper.tex
Included Adam's Edits and Began My Responses
an ad hoc merging of Adam’s edits and the beginnings of my responses

@@
 On the other hand, the tumor cells actively work to shape the immune response that the body launches.
-This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Tregs.
+This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Treg cells.
 Since TGF-$\beta$ is also a driver of EMT, then through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells, cells of the tumor and the surrounding tissue are driven towards a more mesenchymal profile.
@@
 The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.
-Once cytotoxic immune cells, for example natural killer cells and cytotoxic T cells, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse.
-Ref
+Once cytotoxic immune cells, for example NKs and CTLs, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse. Ref 
 However, due to down-regulated surface markers on mesenchymal cells, they are refractory to this immunological task thus rendering them less susceptible to immune clearance. Ref
@@
 Also, all proliferating cells have a probability of undergoing a pathway mutation.
+If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
 If they do, one pathway is randomly chosen among the three and is mutated.
@@
 Those that needed bounds were truncated.
-The results of the Morris OAT can be found in Figure \ref{fig:ModelIntro}B.
-Notice that the following parameters have the most global influence over the system. [which parameters?]
-While it makes sense for such and such [??] parameters to have large influences over patient outcomes, these fall outside of the scope of the current paper and instead it is preferable to focus on the following subset.
+The results of the Morris OAT can be found in Figure \ref{fig:ModelIntro}D.
+Clearly, a certain subset of parameters have an outsized effect on the model while others do relatively little.
+The two most influential according to this analysis are related to Treg cells: their death rate and recruitment.
+This makes sense as Treg cells both suppress cytotoxic effects of other immune cells and also secrete significant quantities of TGF-$\beta$ which drives EMT.
+Thus, Treg cells are intimately tied to all three aspects of the model and so their presence or lack thereof significantly effects the system.
+However, we want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MGA are of interest.
+In addition, inflammation parameters dictating the cycling scheme are interesting.
+What we can see from this plot is that the model also has some high sensitivity to these parameters.
+In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and so we will also look at that parameter.
 
 \subsection{Mesenchymal Cell Properties Dramatically Change Survival Outcomes}\label{MesPars}
-The two quantities that change when a cell transitions to a mesenchymal state are its immune evasion and its increased quiescence, MIE and MQ, respectively.
+The two quantities that change when a cell transitions to a mesenchymal state are its immune evasion and its increased quiescence, MIE and MGA, respectively.
 Also involved in the EMT process, is the cytokine TGF-$\beta$, which upregulates EMT.

### 2019-01-13 Paper/Paper.tex
Added description of Guo's model

@@
 
-Recent work on cancer-immune interactions [briefly describe Guo paper]...
+Recent work on cancer-immune interactions has shown that the inflammatory conditions of the TME can play large and sometimes surprising roles in tumor growth.
+Guo describes a model in which inflammatory state of the TME affects the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations.
+Guo uses the model to show that while inflammation does have an oncogenic effect, it can also have an onco-protective role as periods of mild inflammation are better at preventing cancer than periods of no inflammation in patients that have intermittent bouts of severe inflammation.
+However, the aspect of the model that included the immune system was minimal: regardless of the inflammatory environment, any cells with sufficient pathway mutations (2 in their model) had a fixed probability of being cleared by the metabolism-immune balance response.
+In this paper, we hope to improve the model by including a more robust immune module and exploring how this axis will improve our understanding of inflammatory effects on the tumor.
+In addition, we explore an extra component of Guo's stem cells: where they lie on the epithelial-mesenchymal axis.
 

### 2019-01-14 Paper/Bibliography.tex
References and other edits
Began adding refs and creating a bibtex file. Other light edits, too.
Still need to find out how to compile the bibtex file correctly.

@article{restifo2012adoptive,
  title={Adoptive immunotherapy for cancer: harnessing the T cell response},
  author={Restifo, Nicholas P and Dudley, Mark E and Rosenberg, Steven A},
  journal={Nature Reviews Immunology},
  volume={12},
  number={4},
  pages={269},
  year={2012},
  publisher={Nature Publishing Group}
}

@article{de2006paradoxical,
  title={Paradoxical roles of the immune system during cancer development},
  author={De Visser, Karin E and Eichten, Alexandra and Coussens, Lisa M},
  journal={Nature reviews cancer},
  volume={6},
  number={1},
  pages={24},
  year={2006},
  publisher={Nature Publishing Group}
}

@article{nieto2016emt,
  title={EMT: 2016},
  author={Nieto, M Angela and Huang, Ruby Yun-Ju and Jackson, Rebecca A and Thiery, Jean Paul},
  journal={Cell},
  volume={166},
  number={1},
  pages={21--45},
  year={2016},
  publisher={Elsevier}
}

@article{alderton2012metastasis,
  title={Metastasis: Epithelial to mesenchymal and back again},
  author={Alderton, Gemma K},
  journal={Nature Reviews Cancer},
  volume={13},
  number={1},
  pages={3},
  year={2012},
  publisher={Nature Publishing Group}
}

@article{terry2017new,
  title={New insights into the role of EMT in tumor immune escape},
  author={Terry, St{\'e}phane and Savagner, Pierre and Ortiz-Cuaran, Sandra and Mahjoubi, Linda and Saintigny, Pierre and Thiery, Jean-Paul and Chouaib, Salem},
  journal={Molecular oncology},
  volume={11},
  number={7},
  pages={824--846},
  year={2017},
  publisher={Wiley Online Library}
}

@article{hong2015ovol2,
  title={An Ovol2-Zeb1 mutual inhibitory circuit governs bidirectional and multi-step transition between epithelial and mesenchymal states},
  author={Hong, Tian and Watanabe, Kazuhide and Ta, Catherine Ha and Villarreal-Ponce, Alvaro and Nie, Qing and Dai, Xing},
  journal={PLoS computational biology},
  volume={11},
  number={11},
  pages={e1004569},
  year={2015},
  publisher={Public Library of Science}
}

### 2019-01-14 Paper/Paper.tex
References and other edits
Began adding refs and creating a bibtex file. Other light edits, too.
Still need to find out how to compile the bibtex file correctly.

@@
 Immunotherapy is slowly revolutionizing how we treat cancer.
-Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival. Ref  [examples] 
+Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival. (Restifo 2012)  [examples] 
 There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area.
 Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context.
-Ref
+(De Visser 2006)
 Understanding each step in this immune cascade in the context of the TME will be paramount in fully exploring and exploiting all options for immunotherapy.
@@
 Epithelial-to-mesenchymal transition (EMT) describes the (reversible) cell plasticity process by which epithelial cells transition into mesenchymal cells.
-EMT has an important yet complicated role in tumor initiation, progression, and metastasis.
-Ref
-As mesenchymal cells often exhibit stem-like properties, their existence has been tied to the initiation and progression of cancer. Ref
-Also, due to their motility, some have viewed them as a natural precursor to metastasis. Ref
+EMT has an important yet complicated role in tumor initiation, progression, and metastasis. (Nieto 2016) % References more specific to this claim might be in this paper's references. 
+As mesenchymal cells often exhibit stem-like properties, their existence has been tied to the initiation and progression of cancer.
+(Nieto 2016) % References more specific to this claim might be in this paper's references.
+Also, due to their motility, some have viewed them as a natural precursor to metastasis. (Alderton 2012)
 
@@
 In addition to the interactions already discussed, the immune system affects and is affected by EMT.
-The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME. Ref 
-In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$). Ref 
+The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME. (Terry 2017) 
+In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$). (Terry 2017) 
 Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell.
@@
 Of interest here, the literature shows two qualities of mesenchymal cells that are of interest particularly in the context of cancer and the immune system.
-The first is that mesenchymal cells are less susceptible to immune clearance. Ref 
-The second is that mesenchymal cells proliferate at a slower rate. Ref
+The first is that mesenchymal cells are less susceptible to immune clearance. (Terry 2017) 
+The second is that mesenchymal cells proliferate at a slower rate. Ref % Adam said he had a reference in mind for this.
 
 The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.
-Once cytotoxic immune cells, for example NKs and CTLs, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse. Ref 
-However, due to down-regulated surface markers on mesenchymal cells, they are refractory to this immunological task thus rendering them less susceptible to immune clearance. Ref
+Once cytotoxic immune cells, for example NKs and CTLs, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse. (Terry 2017) 
+However, there is preliminary evidence that due to down-regulated surface markers on mesenchymal cells among other structural changes, they are refractory to this immunological task thus rendering them less susceptible to immune clearance.
+(Terry 2017) %
 In line with known motility of mesenchymal cells, scientists have referred to this condition as mesenchymal cells being ``slippery.''
@@
 
-In the following sections of the paper we first introduce the model modularly, which is appropriate diven its multiscale and hybrid nature.
+In the following sections of the paper we first introduce the model modularly, which is appropriate given its multiscale and hybrid nature.
 We then perform sensitivity analysis and identify parameters that are crucial for the progression to tumorigenesis.
@@
 B. Example simulation of the dynamics of a one patient; warmup period is shown in red.
-C. Example survival curve for one cohort of patients with parameters XXXXX.
-D. Global sensitivity analysis of XXX parameters using XXXXX.}
+C. Example survival curve for one cohort of patients with parameters XXXXX. % I should list out all parameter values?
+D. Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
 \label{fig:ModelIntro}
@@
 If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the annotation to the right of the cell block.
-If any component [???] of a cell has a pathway mutation, that cell is then counted as a mutated cell and contributes towards the total number of cancer cells.
+Any cell that has a pathway mutation is counted as a mutated cell and contributes towards the total number of cancer cells.
 The EMT score of each cell (denoting position along the EMT spectrum) is associated with a color given by the legend to the left of the cell block. 
@@
 While biologically, there may be only a few discrete stable cell states available on the EMT spectrum, recent studies highlight a growing number of states accessible, with research indicating there are several equilibria along the spectrum.
-Ref [Tian and Levine lab]
+(Hong 2015)
 In our model, this is actually just a binary and that is why the cells shown are at one extreme or the other of the spectrum.
@@
 For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}B.
-Notice that at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541. [translate into days?]
-Once the Time to Cancer has been determined for a patient, most simulations shown later stop. [??? mean to say: we stop the simulation?]
-We do this because once such a high proportion of cells have mutated, the mutant cells quickly take over and make up 100\% of all cells.
-Also of interest, as the patient's cells are all rapidly mutating right around this time, most of them are also transitioning to a mesenchymal phenotype with all of them being in a mesenchymal phenotype for a short time during the time period that cancer onset occurs.
+Notice that at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, which translates to 405.75 days. % translate to days like this?
+At the same time as mutated cells are taking over, most cells are transitioning to  a mesenchymal phenotype.
+After the Time to Cancer, the mutated cells continue their rapid growth and soon make up 100\% of the cells.
+Meanwhile, all the cells become mesenchymal for a short duration before most transition back to an epithelial state.% Once the Time to Cancer has been determined for a patient, most simulations shown later stop.
+[???
+mean to say: we stop the simulation?]
+%We do this because once such a high proportion of cells have mutated, the mutant cells quickly take over and make up 100\% of all cells.
 

### 2019-01-14 Paper/Paper.tex
Properly formatted citations and bibliography

@@
 \usepackage{amsmath}
+\usepackage{cite}
 
@@
 Immunotherapy is slowly revolutionizing how we treat cancer.
-Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival. (Restifo 2012)  [examples] 
+Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival\cite{restifo2012adoptive}.  [examples] 
 There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area.
-Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context.
-(De Visser 2006)
+Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context\cite{de2006paradoxical}.
 Understanding each step in this immune cascade in the context of the TME will be paramount in fully exploring and exploiting all options for immunotherapy.
@@
 Epithelial-to-mesenchymal transition (EMT) describes the (reversible) cell plasticity process by which epithelial cells transition into mesenchymal cells.
-EMT has an important yet complicated role in tumor initiation, progression, and metastasis. (Nieto 2016) % References more specific to this claim might be in this paper's references. 
-As mesenchymal cells often exhibit stem-like properties, their existence has been tied to the initiation and progression of cancer.
-(Nieto 2016) % References more specific to this claim might be in this paper's references.
-Also, due to their motility, some have viewed them as a natural precursor to metastasis. (Alderton 2012)
+EMT has an important yet complicated role in tumor initiation, progression, and metastasis\cite{nieto2016emt}. % References more specific to this claim might be in this paper's references. 
+As mesenchymal cells often exhibit stem-like properties, their existence has been tied to the initiation and progression of cancer\cite{nieto2016emt}.
+% References more specific to this claim might be in this paper's references.
+Also, due to their motility, some have viewed them as a natural precursor to metastasis\cite{alderton2012metastasis}.
 
@@
 In addition to the interactions already discussed, the immune system affects and is affected by EMT.
-The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME. (Terry 2017) 
-In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$). (Terry 2017) 
+The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME \cite{terry2017new}. 
+In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$)\cite{terry2017new}. 
 Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell.
@@
 Of interest here, the literature shows two qualities of mesenchymal cells that are of interest particularly in the context of cancer and the immune system.
-The first is that mesenchymal cells are less susceptible to immune clearance. (Terry 2017) 
+The first is that mesenchymal cells are less susceptible to immune clearance\cite{terry2017new}. 
 The second is that mesenchymal cells proliferate at a slower rate. Ref % Adam said he had a reference in mind for this.
@@
 The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.
-Once cytotoxic immune cells, for example NKs and CTLs, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse. (Terry 2017) 
-However, there is preliminary evidence that due to down-regulated surface markers on mesenchymal cells among other structural changes, they are refractory to this immunological task thus rendering them less susceptible to immune clearance.
-(Terry 2017) %
+Once cytotoxic immune cells, for example NKs and CTLs, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse\cite{terry2017new}.
+However, there is preliminary evidence that due to down-regulated surface markers on mesenchymal cells among other structural changes, they are refractory to this immunological task thus rendering them less susceptible to immune clearance\cite{terry2017new}.
+%
 In line with known motility of mesenchymal cells, scientists have referred to this condition as mesenchymal cells being ``slippery.''
@@
 In the model the EMT spectrum is continuous.
-While biologically, there may be only a few discrete stable cell states available on the EMT spectrum, recent studies highlight a growing number of states accessible, with research indicating there are several equilibria along the spectrum.
-(Hong 2015)
+While biologically, there may be only a few discrete stable cell states available on the EMT spectrum, recent studies highlight a growing number of states accessible, with research indicating there are several equilibria along the spectrum\cite{hong2015ovol2}.
 In our model, this is actually just a binary and that is why the cells shown are at one extreme or the other of the spectrum.
@@
 
+\bibliography{mybib}{}
+\bibliographystyle{siam}
+
 \end{document}

### 2019-01-21 Paper/Paper.tex
Lots of edits
Lots of edits. Comments for Adam and I to consider. Got INFL cycling
schematic in Figure 1. Probably other stuff, too.

@@
 Immunotherapy is slowly revolutionizing how we treat cancer.
-Year-by-year, more breakthroughs in the field produce effective new drugs and therapies that are having a significant impact on patient health and survival\cite{restifo2012adoptive}.  [examples] 
+Year-by-year, new breakthroughs in the field produce new, effective drugs and therapies that have a significant impact on patient health and survival\cite{pardoll2012blockade}\cite{restifo2012adoptive}.
+% More examples?
+Reviews or non-reviews?
+Explain some of the examples?
 There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area.
 Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context\cite{de2006paradoxical}.
-Understanding each step in this immune cascade in the context of the TME will be paramount in fully exploring and exploiting all options for immunotherapy.
+Understanding each step in this immune cascade in the context of the TME will be necessary in order to fully explore and exploit all options of immunotherapy.
 
 Recent work on cancer-immune interactions has shown that the inflammatory conditions of the TME can play large and sometimes surprising roles in tumor growth.
-Guo describes a model in which inflammatory state of the TME affects the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations.
+Guo describes a model in which the inflammatory state of the TME affects the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations\cite{guo2017multiscale}.
 Guo uses the model to show that while inflammation does have an oncogenic effect, it can also have an onco-protective role as periods of mild inflammation are better at preventing cancer than periods of no inflammation in patients that have intermittent bouts of severe inflammation.
@@
 In addition, we explore an extra component of Guo's stem cells: where they lie on the epithelial-mesenchymal axis.
+So, in total, we will explore the joint interactions of the tumor, the immune system, and the epithelial-mesenchymal spectrum.
+We will also study the effects of the TME on these interactions.
+
+The tumor interacts with the immune system by initiating the sequence of events that results in the activation of the adaptive immune response.
+It accomplishes this by presenting antigens that innate immune cells recognize and carry to lymph nodes to activate, among other things, T cells.
+Also, tumor cells engage in certain cellular processes that shape the TME and thus impact the immune system.
+In particular, tumor cells release transforming growth factor beta (TGF-$\beta$) that creates a tumor-supportive environment by shifting the immune balance to being immunosuppressive via enhancing activation of Treg cells.
+
+This particular cytokine is also a driver of the epithelial-to-mesenchymal transition (EMT) and through generating TGF-$\beta$ the tumor creates a more mesenchymal phenotype within the TME.
+In this way, the tumor shifts the balance of local environment along the EMT spectrum.
+
+The immune system has several competing interactions with the tumor.
+While the cytotoxic immune cells, NKs and cytotoxic T cells (CTLs), seek out and lyse tumor cells, the regulatory branch of the immune system, which will be accounted for by Tregs in our model, slows down the effector functions of these other cells.
+
+In relation to EMT, the Tregs of the immune system drive the EMT process by releasing TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
+In this way, just as the tumor cells push the TME towards a mesenchymal phenotype, so too do the Tregs.
 
 Epithelial-to-mesenchymal transition (EMT) describes the (reversible) cell plasticity process by which epithelial cells transition into mesenchymal cells.
-EMT has an important yet complicated role in tumor initiation, progression, and metastasis\cite{nieto2016emt}. % References more specific to this claim might be in this paper's references. 
-As mesenchymal cells often exhibit stem-like properties, their existence has been tied to the initiation and progression of cancer\cite{nieto2016emt}.
-% References more specific to this claim might be in this paper's references.
-Also, due to their motility, some have viewed them as a natural precursor to metastasis\cite{alderton2012metastasis}.
-
-Herein, we explore the interconnectedness of these three key systems: the tumor, the immune system, and EMT.
-In addition to the interactions already discussed, the immune system affects and is affected by EMT.
-The immune system affects EMT by releasing a flurry of cytokines upon arrival in the TME \cite{terry2017new}. 
-In particular, regulatory T cells (Tregs) release copious quantities of the EMT-driving transforming growth factor beta (TGF-$\beta$)\cite{terry2017new}. 
-Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell.
-Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
-Each of these three pieces will be briefly explored in turn for how they will influence one another.
-
-[No subsections headings within Introduction]
-\subsection{The Tumor}\label{TheTumor}
-The tumor influences the immune system by its presence and by the cellular processes it engages in that result in shaping the TME.
-On the one hand, the mutated DNA of tumor cells results in these cells being recognized as foreign by members of the innate immune system such as natural killer (NK) cells and dendritic cells.
-It is then these cells that initiate the cascade of cytokines that result in the full activation of the body's immune system to fight the non-self cells.
-On the other hand, the tumor cells actively work to shape the immune response that the body launches.
-This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Treg cells.
-Since TGF-$\beta$ is also a driver of EMT, then through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells, cells of the tumor and the surrounding tissue are driven towards a more mesenchymal profile.
-
-\subsection{The Immune System}\label{TheImmuneSystem}
-The immune system has several competing interactions with the tumor.
-While the cytotoxic immune cells, NKs and cytotoxic T cells (CTLs), seek out and lyse tumor cells, the regulatory branch of the immune system, Tregs, slows down the effector functions of these other cells.
-
-These Tregs also drive the EMT process by releasing TGF-$\beta$ upon arriving at the tumor site.
-
-\subsection{EMT}\label{EMT}
-There are many characteristics distinctive of mesenchymal cells and still much research being done in further quantifying these and determining others.
-Of interest here, the literature shows two qualities of mesenchymal cells that are of interest particularly in the context of cancer and the immune system.
+While the epithelial cells often stay bound together in layers, mesenchymal cells have gained motility and even stem-like properties\cite{nieto2016emt}.
+They have thus become a subject of great interest in cancer research having been connected to tumor initiation, progression, and metastasis\cite{nieto2016emt}.
+There are many other characteristics distinctive of mesenchymal cells and still much research being done in further quantifying these and determining others.
+Of interest to us, the literature reveals two qualities of mesenchymal cells that are particularly relevant in the context of cancer and the immune system.
 The first is that mesenchymal cells are less susceptible to immune clearance\cite{terry2017new}. 
@@
 The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.
-Once cytotoxic immune cells, for example NKs and CTLs, have targeted a cell to lyse, they necessarily must make a physical connection with that cell, termed an immunological synapse\cite{terry2017new}.
-However, there is preliminary evidence that due to down-regulated surface markers on mesenchymal cells among other structural changes, they are refractory to this immunological task thus rendering them less susceptible to immune clearance\cite{terry2017new}.
-%
-In line with known motility of mesenchymal cells, scientists have referred to this condition as mesenchymal cells being ``slippery.''
-This quality of mesenchymal cells is then quantified in our model by the parameter, MIE, for mesenchymal immune evasion.
+As a cell is being targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be created in order to precipitate the lysing event.
+This immunological synapse, however, is dependent upon the surface proteins of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form\cite{terry2017new}.
+There are also other structural changes that take place in mesenchymal cells that could contribute to this refractory nature towards immune clearance.
+Regardless, the evidence currently demonstrates a ``slipperiness'' of mesenchymal cells in regards to immune lysing that in our model we will refer to as mesenchymal immune evasion, or MIE.
 
@@
 Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions.
-While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate.
-In quantifying this property, we use the model parameter `mesenchymal growth arrest', MGA.
-This choice is less about associating this property with quiescence and more about creating a simple name that easily, if not roughly, captures the essence of what is being modeled.
-For example, while mesenchymal proliferation reduction might be a more accurate term, thinking in terms of a negative formulation adds one additional twist to accurately interpreting results which we felt was more detrimental than the possible confusion of invoking ``quiescence.''
+Ref
+While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate. Ref
+Cancer cells are well-known for achieving unregulated proliferation, so exploring a subset of cancer cells with decreased proliferation could provide intriguing new results.
+In our model, we refer to this property as mesenchymal growth arrest, or MGA.
 
@@
 We analyze the competing effects of EMT and of the immune system on tumorigenesis and find that EMT intricately regulates tumorigenesis: under certain regimes a careful balance of EMT- and immune-driven process can significantly prolong cancer-free survival.
+
+%%%%%%%%%%%%%  INTRO PARAGRAPH GRAVEYARD %%%%%%%%%%%%%%%%
+
+%They release cytokines that alter the extracellular matrix, promote angiogenesis, and create an inflammatory environment.
+%This last one, in particular, results in the impairment of cytotoxic capabilities of immune cells.
+%Through 
+%
+% its presence and by the cellular processes it engages in that result in shaping the TME.
+%On the one hand, the mutated DNA of tumor cells results in these cells being recognized as foreign by members of the innate immune system such as natural killer (NK) cells and dendritic cells.
+%It is then these cells that initiate the cascade of cytokines that result in the full activation of the body's immune system to fight the non-self cells.
+%On the other hand, the tumor cells actively work to shape the immune response that the body launches.
+%This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Treg cells.
+Since TGF-$\beta$ is also a driver of EMT, then through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells, cells of the tumor and the surrounding tissue are driven towards a more mesenchymal profile.
+
+%In quantifying this property, we use the model parameter `mesenchymal growth arrest', MGA.
+%This choice is less about associating this property with quiescence and more about creating a simple name that easily, if not roughly, captures the essence of what is being modeled.
+%For example, while mesenchymal proliferation reduction might be a more accurate term, thinking in terms of a negative formulation adds one additional twist to accurately interpreting results which we felt was more detrimental than the possible confusion of invoking ``quiescence.''
+
+%This plays an important yet complicated role in tumor initiation, progression, and metastasis\cite{nieto2016emt}. % References more specific to this claim might be in this paper's references. 
+%As mesenchymal cells often exhibit stem-like properties, their existence in a tumor has been tied to the initiation and progression of cancer\cite{nieto2016emt}.
+% References more specific to this claim might be in this paper's references.
+%Also, due to their motility, some have viewed them as a natural precursor to metastasis\cite{alderton2012metastasis}. 
+%
+%Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell\cite{terry2017new}.
+%Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
+%Each of these three pieces will be briefly explored in turn for how they will influence one another.
 
@@
 \section{Methods}
-[In this section it is important to say what is old (from previous model), i.e. `following the model of XX, we....' and what is new/our addition.]
-
-To address the questions around the relationships among cancer, the immune system, and EMT, we built a multiscale, agent-based model.
-The agents in the model are the tissue cells with tumorigenic potential, that can either be mutation-free, or have mutations.
-It is worth noting that we do not consider spatial dependencies between agents, i.e. we assume that cells are well-mixed as are all the relevant cytokines and other molecules acting on the system.
-We model the effects of immune cell types and cytokines as continuous fields within the tumor microenvironment (TME), since here single-cell resolution is not required. 
-Finally, tissue cells that can be labeled either epithelial or mesenchymal, and this labeling is fluid and depends both on the TME and the internal workings of each individual cell.
+To address the questions around the relationships among cancer, the immune system, and EMT, we extended the multiscale, agent-based model of Guo to include an immune module and an EMT axis for the tissue cells.
+The agents in the model are the tissue cells with tumorigenic potential.
+They can either be mutation-free, or have any combination of three possible pathway mutations.
+We model the immune cells as discrete variables but as one collective, in other words they are not agents themselves.
+The cytokine TGF-$\beta$ is a continuous variables within the tumor microenvironment (TME).
+It is worth noting that there is no spatial aspect to this model.
+In doing this, we are assuming that cells are well-mixed as are all the relevant cytokines and other molecules acting on the system.
+Finally, tissue cells can be labeled either epithelial or mesenchymal, and this labeling can change over time and depends both on the TME and the internal workings of each individual cell.
 
 \subsection{Tissue Cells}\label{TissueCells}
-The tissue cells at all times have associated with them two important quantities: a logical vector indicating pathway mutations and an EMT score.
+The tissue cells at all times have associated with them two important quantities: which combination of mutations they carry and an EMT score.
 The model has three pathways that have the potential to mutate over the course of a simulation.
@@
 % It might be nice to have a table to summarize this information?
-The EMT score will be exposited in Section \ref{EMT}.
+The EMT score will be exposited in Section \ref{EMT}. % make sure this is really the best decision
 
-Add differential equations? % Adam suggested this, but I'm not sure what this means.
-% I think this comment applied to the next section: add the equations describing immune system
+%Add differential equations? % Should this be done here or where I have it later?
 
@@
 The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
-Should they clear a cell, they themselves are then deactivated and removed from the immune population.
+Upon successful clearance, they themselves are then deactivated and removed from the immune population. % Should this be justified here or mentioned in the intro?
 Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear mutated cells.
@@
 \subsection{EMT}\label{EMT}
-The EMT score indicates how likely the cell is begin the next cell cycle as an epithelial or a mesenchymal cell.\
-The score is between $0$ and $1$ with higher values being connected to higher probability of being mesenchymal.
-For each patient, there is a fixed mesenchymal threshold.
-If the EMT score ends the cycle above this threshold, the cell is mesenchymal, otherwise it is epithelial for the next cycle.
+Each tissue cell has an EMT score between 0 and 1.
+When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
+Otherwise, it is below the threshold and labeled as epithelial.
 For the purposes of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have some of its parameters updated.
@@
 Thus, the only immune cell present is NKs.
-After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a discrete, random jump in their probability to mutate.
+After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
 From then on, a cell which proliferates either mutates or increases its probability of mutating later on.
@@
 \item immune clearance (by NKs or CTLs)
-\item quiescence
+\item rest/$G_0$ % use both here?
 \end{itemize}
@@
 \paragraph{Proliferation}
-There are four factors that contribute to a cell's probability of proliferation.
-% NB try to avoid posessive apostrophe. I.e. replace this with
 There are four factors that contribute to the probability that a cell will proliferate.
@@
 Second, if the cell has a mutation in the proliferation pathway, then the weight for proliferation is proportionally increased by $\text{prolif}_{\text{up}}$.
-Third, if the cell is mesenchymal, then the weight for proliferation is proportionally decreased by $m_Q$, which stands for mesenchymal quiescence.
+Third, if the cell is mesenchymal, then the weight for proliferation is proportionally decreased by $m_{GA}$, which stands for mesenchymal growth arrest.
 Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor.
@@
 
-$$ p_{\text{prolif}} = p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})(1-\delta_{MES}m_Q)\frac{N_0}{N_0+N} $$
+$$ p_{\text{prolif}} = p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})(1-\delta_{MES}m_{GA})\frac{N_0}{N_0+N} $$
 
@@
 Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell.
-Third, NKs and CTLs have their own efficacy parameters which can be understood as their probability of clearing a mutated cell given that they have interacted with it.
+Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the weight for immune clearance given an immune cell has found the mutated cell.
 Fourth, there is a decreasing Hill factor based on the number of Treg cells present.
@@
 
-$$ p_{\text{NKed}} =  \delta_M \frac{\text{NK}}{N/N_{00}+\text{NK}} E_{\text{NK}} \frac{1}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} $$ $$ (1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}})$$
+$$ p_{\text{NK}} =  \delta_M \frac{\text{NK}}{N/N_{1}+\text{NK}} E_{\text{NK}} \frac{1}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} \cdot $$
+$$ \qquad(1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}})$$
+% Should I worry about the order in which these appear? Rearranging would make the formula look better, but the order I have introduced them makes the most sense.
+% Also, any thoughts on NK, CTL, Treg not being italicized? While other capital letters are?
 
@@
 
-\paragraph{Quiescence}
+\paragraph{Rest/$G_0$} %use both for this section heading?
 The weight associated with quiescence is taken as 1 except in the case of Mesenchymal cells.
@@
 
-$$ p_Q = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_Q\frac{N_0}{N_0+N} $$
+$$ p_{G0} = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_{GA}\frac{N_0}{N_0+N} $$ %p_{G_0},p_{rest},p_{\text{rest}},p_{GA}???
 
-The reason for adding that term is due to the understanding that mesenchymal cells are not proliferating as fast because they are instead remaining quiescent, hence naming the parameter $m_Q$.
+The reason for adding that term is due to the understanding that mesenchymal cells are not proliferating as fast because they are instead remaining in the $G_0$ phase.
+% is this the right way to word this?
 
@@
 First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
-This represents the fact that individual immune cells have a limited amount of efficacy.
+This represents the fact that individual immune cells lose efficacy as they carry out their effector functions.
+% Similar to previous comment: should I cite reference for this here or put this up in the intro?
 
@@
 If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
-If they do, one pathway is randomly chosen among the three and is mutated.
-If the cell does not undergo a mutation, then its probability of mutating during subsequent cell increases. [**]
+If the cell does not undergo a mutation, then its probability of mutating during subsequent cell increases. % Updated this after Adam had question. Does this make sense as is?
 
 Then, the EMT values for each cell is updated.
-This depends on how much TGF-$\beta$ is currently in the system.
+This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
 Each cell receives an amount of TGF-$\beta$ given by
@@
 where $\sigma$ is a parameter in the model.
+If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
+If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
+Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
+
+%%%%%%%%%	I think this next section is too much detail that just confuses the reader. I don't think I need to explain my implementation, just the process. Thoughts?	%%%%%%%%%%%%%%%%%%%%
 This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
@@
 In addition, a cell having a higher EMT score will have a higher chance of continuing in the EMT process whereas a cell low on the EMT spectrum will have a higher chance of moving towards a more epithelial phenotype.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-$$ \text{TGFB} = \text{tgfb}_{\text{mut}}N_{\text{mut}} + \text{tgfb}_{\text{Treg}}\text{Treg}$$
+$$ \text{TGFB}_k = \text{tgfb}_{\text{mut}}M_k + \text{tgfb}_{\text{Treg}}\text{Treg}_k$$ % clear that M is number of mutants, right?
 
@@
 
-$$ \text{NK}_{i+1} = \left (\text{NK}_i-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} $$
+%%% I like the first of these two options. You set aside the amount you know you will gain (sigma/d) and then let the rest decay exponentially
+$$ \text{NK}_{k+1} = \left (\text{NK}_k-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} $$
 
+%%% For this one, the initial decays exponentially and then you add some fraction of sigma/d depending on how big your time step is
+$$ \text{NK}_{k+1} = \text{NK}_k\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \left(1-\exp(-d_{\text{NK}}\Delta t)\right)$$
+%%% The first one is better though in my opinion.
+
+% For all 3 discretizations of immune cell dynamics, I could create a quantity R_* (for recruitment? maybe better letters exist) so the formula would be 
+% *_{k+1} = (*_k-R_*)\exp(-d_*\Delta t)+R_*,\quad R_*=\frac{\sigma_*}{d_*}
+%% Option 1b from above:
+$$\text{NK}_{k+1} = (\text{NK}_k-R_\text{NK})\exp(-d_\text{NK}\Delta t)+R_\text{NK},\quad R_\text{NK}=\frac{\sigma_\text{NK}}{d_\text{NK}}$$
+
+%%%% Which of these two paragraphs is better? Currently the intro does not mention APCs and I don't think I need to go into that, especially when APC_k = \tilde{M}_k
+For CTLs and Tregs, they rely on mutant cells being cleared before they can be activated.
+Let $\tilde{M}_k$ represent the number of mutant cells cleared by the immune system during cell cycle $k$.
+In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function.
+We choose the following differential equations to govern the CTL and Treg populations:
+
+\begin{align*}
+\text{CTL}' & = \sigma_{\text{CTL}}\tilde{M} - d_{\text{CTL}}\text{CTL} \\
+\text{Treg}' & = \sigma_{\text{Treg}}\tilde{M} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}- d_{\text{Treg}}\text{Treg}
+\end{align*}
+
+Discretized, these are:
+
+\begin{align*}
+\text{CTL}_{k+1} & = & \left (\text{CTL}_k-\sigma_{\text{CTL}}\tilde{M}_k/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}\tilde{M}_k/d_{\text{CTL}}\\
+\text{Treg}_{k+1} & = & \left (\text{Treg}_k-\frac{\sigma_{\text{Treg}}\tilde{M}_k}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}\right )\exp(-d_{\text{Treg}}\Delta t)+\\
+& & \frac{\sigma_{\text{Treg}}\tilde{M}_k}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}
+\end{align*}
+
+%% Or using R_*
+\begin{align*}
+\text{CTL}_{k+1} & =  \left (\text{CTL}_k-\tilde{M}_kR_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \tilde{M}_kR_{\text{CTL}}\\
+\text{Treg}_{k+1} & =  \left (\text{Treg}_k-\tilde{M}_kR_{\text{Treg}}\right )\exp(-d_{\text{Treg}}\Delta t)+\tilde{M}_kR_{\text{Treg}}
+\end{align*}
+where
+\begin{align*}
+R_{\text{CTL}} &= \sigma_{\text{CTL}}/d_{\text{CTL}}\\
+R_{\text{Treg}} &=\frac{\sigma_{\text{Treg}}}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}
+\end{align*}
+%% End of option 1
+
+%% Beginning of option 2
 For CTLs and Tregs, however, they rely on antigen presenting cells (APCs) to be activated, and these APCs in turn rely on mutated cells being cleared so that they can bind antigens.
@@
 \begin{align*}
-\text{CTL}_{i+1} & = & \left (\text{CTL}_i-\sigma_{\text{CTL}}\text{APC}/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}\text{APC}/d_{\text{CTL}}\\
-\text{Treg}_{i+1} & = & \left (\text{Treg}_i-\frac{\sigma_{\text{Treg}}\text{APC}}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}\right )\exp(-d_{\text{Treg}}\text{Treg}\Delta t)+\\
-& & \frac{\sigma_{\text{Treg}}\text{APC}}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}
+\text{CTL}_{k+1} & = & \left (\text{CTL}_k-\sigma_{\text{CTL}}\text{APC}_k/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}\text{APC}_k/d_{\text{CTL}}\\
+\text{Treg}_{k+1} & = & \left (\text{Treg}_k-\frac{\sigma_{\text{Treg}}\text{APC}_k}{d_{\text{Treg}}} \frac{\text{TGFB}_k}{1+\text{TGFB}_k/\text{EC50}}\right )\exp(-d_{\text{Treg}}\Delta t)+\\
+& & \frac{\sigma_{\text{Treg}}\text{APC}_k}{d_{\text{Treg}}} \frac{\text{TGFB}_k}{1+\text{TGFB}_k/\text{EC50}}
 \end{align*}
+%% End of option 2
 
@@
 At the end of each cell cycle, the proportion of tissue cells that are mutated is calculated, and if it is above a certain threshold, a tumorigenic event is recorded and the patient is determined to have cancer (using a similar strategy to Guo).
-The Time to Cancer is given as the time from the start of the simulation until tumorigenesis.
+The Time to Cancer is given as the time from the start of the simulation, not including warmup, until tumorigenesis.
 Simulations run until either a tumorigenic event is recorded or until the maximum number of cycles has been reached.
@@
 B. Example simulation of the dynamics of a one patient; warmup period is shown in red.
-C. Example survival curve for one cohort of patients with parameters XXXXX. % I should list out all parameter values?
+C. Example survival curve for one cohort of patients with basal parameter values. % I should list out all parameter values?
 D. Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
@@
 After the Time to Cancer, the mutated cells continue their rapid growth and soon make up 100\% of the cells.
-Meanwhile, all the cells become mesenchymal for a short duration before most transition back to an epithelial state.% Once the Time to Cancer has been determined for a patient, most simulations shown later stop.
-[???
-mean to say: we stop the simulation?]
+Meanwhile, all the cells become mesenchymal for a short duration before most transition back to an epithelial state.
+% Once the Time to Cancer has been determined for a patient, most simulations shown later stop. [??? mean to say: we stop the simulation?]
 %We do this because once such a high proportion of cells have mutated, the mutant cells quickly take over and make up 100\% of all cells.
@@
 Considering the immune populations, notice that the NK population basically stays constant while the lymphocyte populations quickly dwarf the NK population once cells begin to mutate.
-These lymphocyte populations exhibit oscillatory behavior, however this is not due to intrinsic dynamics but rather due to the inflammation scheme that this patient is undergoing: every 30 cell cycles, the inflammation level changes from high to low, or vice versa.
+These lymphocyte populations appear to exhibit oscillatory behavior, however this is not due to intrinsic dynamics but rather due to the inflammation scheme that this patient is undergoing: every 30 cell cycles, the inflammation level changes from high to low, or vice versa.
 In each of these periods periods, the lymphocyte populations increase or decrease rapidly in accordance with the inflammation.

### 2019-01-25 Paper/Paper.tex
Completed Discussion and Conclusion
Finished a round of edits and got a discussion and conclusion.

@@
 Year-by-year, new breakthroughs in the field produce new, effective drugs and therapies that have a significant impact on patient health and survival\cite{pardoll2012blockade}\cite{restifo2012adoptive}.
-% More examples?
-Reviews or non-reviews?
-Explain some of the examples?
+% More examples? Reviews or non-reviews? Explain some of the examples?
 There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area.
@@
 
-$$ p_{\text{NK}} =  \delta_M \frac{\text{NK}}{N/N_{1}+\text{NK}} E_{\text{NK}} \frac{1}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} \cdot $$
+$$ p_{\text{NK}} =\delta_M \frac{\text{NK}}{N/N_{1}+\text{NK}} E_{\text{NK}} \frac{1}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} \cdot $$
 $$ \qquad(1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}})$$
@@
 
-$$ \text{TGFB}_i = \text{TGFB}/N + X_i, \quad X_i \sim N(0,\sigma^2)  $$
+$$ \text{TGFB}_i = \text{TGFB}/N + X_i, \quad X_i \sim N(0,\sigma^2)$$
 
@@
 \begin{align*}
-\text{CTL}_{k+1} & =  \left (\text{CTL}_k-\tilde{M}_kR_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \tilde{M}_kR_{\text{CTL}}\\
-\text{Treg}_{k+1} & =  \left (\text{Treg}_k-\tilde{M}_kR_{\text{Treg}}\right )\exp(-d_{\text{Treg}}\Delta t)+\tilde{M}_kR_{\text{Treg}}
+\text{CTL}_{k+1} & =\left (\text{CTL}_k-\tilde{M}_kR_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \tilde{M}_kR_{\text{CTL}}\\
+\text{Treg}_{k+1} & =\left (\text{Treg}_k-\tilde{M}_kR_{\text{Treg}}\right )\exp(-d_{\text{Treg}}\Delta t)+\tilde{M}_kR_{\text{Treg}}
 \end{align*}
@@
 B. Example simulation of the dynamics of a one patient; warmup period is shown in red.
-C. Example survival curve for one cohort of patients with basal parameter values. % I should list out all parameter values?
+The inflammation cycling scheme is graphically represented above the patient dynamics.
+C. Example survival curve for one cohort of patients with basal parameter values.
+% Should I list out all parameter values? A subset? Or refer to supplement for base parameter values?
 D. Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
@@
 In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
+% It seems the rest of this paragraph belongs in the Methods section, yeah?
 Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
@@
 The EMT score of each cell (denoting position along the EMT spectrum) is associated with a color given by the legend to the left of the cell block. 
-In the model the EMT spectrum is continuous.
-While biologically, there may be only a few discrete stable cell states available on the EMT spectrum, recent studies highlight a growing number of states accessible, with research indicating there are several equilibria along the spectrum\cite{hong2015ovol2}.
-In our model, this is actually just a binary and that is why the cells shown are at one extreme or the other of the spectrum.
+In the model the EMT score is continuous but the labeling of epithelial or mesenchymal is a binary determined by being below or above a threshold.
 
-In the right half of Figure \ref{fig:ModelIntro}A one can see the cell cycle choices that each cell can make each cell cycle.
+% This next commented out paragraph seems to belong somewhere, but where? I would think in the Intro.
+
+%While biologically, there may be only a few discrete stable cell states available on the EMT spectrum, recent studies highlight a growing number of states accessible, with research indicating there are several equilibria along the spectrum\cite{hong2015ovol2}.
+%In our model, this is actually just a binary and that is why the cells shown are at one extreme or the other of the spectrum.
+
+In the right half of Figure \ref{fig:ModelIntro}A one can see the cell cycle choices that each cell will make each cell cycle.
 These are the four thick, gray arrows.
@@
 For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}B.
-Notice that at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, which translates to 405.75 days. % translate to days like this?
+The inflammation cycling scheme shown at the top of Figure \ref{fig:ModelIntro}B is repeated throughout.
+We will return to this shortly.
+It is at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, which translates to 405.75 days. % translate to days like this?
 At the same time as mutated cells are taking over, most cells are transitioning to a mesenchymal phenotype.
@@
 Considering the immune populations, notice that the NK population basically stays constant while the lymphocyte populations quickly dwarf the NK population once cells begin to mutate.
-These lymphocyte populations appear to exhibit oscillatory behavior, however this is not due to intrinsic dynamics but rather due to the inflammation scheme that this patient is undergoing: every 30 cell cycles, the inflammation level changes from high to low, or vice versa.
+These lymphocyte populations appear to exhibit oscillatory behavior, however this is not due to intrinsic dynamics but rather due to the inflammation scheme that this patient is undergoing: 
+the patient alternates between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
 In each of these periods periods, the lymphocyte populations increase or decrease rapidly in accordance with the inflammation.
@@
 \subsection{Sensitivity Analysis}\label{SensAnalysis}
-To assess the sensitivity to the many model parameters [how many?], a Morris OAT algorithm was implemented. [need to define] Ref 
+% 
+To assess the sensitivity of the model to the 31 model parameters, a Morris OAT algorithm was implemented.
+The algorithm works by running the model many successive times with exactly one parameter changing between each run\cite{morris1991factorial}\cite{sohier2014improvement}.
+% is the above description sufficient? The first source is Morris's original paper and the second is the source of the particular implementation I found
 The parameters were all given Gaussian distributions centered at their initial value and a large standard deviation.
@@
 This makes sense as Treg cells both suppress cytotoxic effects of other immune cells and also secrete significant quantities of TGF-$\beta$ which drives EMT.
-Thus, Treg cells are intimately tied to all three aspects of the model and so their presence or lack thereof significantly effects the system.
+Thus, Treg cells are intimately tied to all three aspects of the model and so their presence or lack thereof significantly affects the system.
 However, we want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MGA are of interest.
-In addition, inflammation parameters dictating the cycling scheme are interesting.
+In addition, inflammation parameters dictating the cycling scheme are worthy of study because of their prominence in Guo's work.
 What we can see from this plot is that the model also has some high sensitivity to these parameters.
@@
 First, as mesenchymal immune evasion (MIE) increases, Time to Cancer decreases.
-This result holds universally among all parameter sets and indicates the simple relationship that mesenchymal immune evasion has with time to cancer.
-As this subpopulation of the tumor becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus more dangerous.
+This result holds universally among all parameter sets and indicates the simple relationship that mesenchymal immune evasion has with Time to Cancer.
+As this subpopulation of the tumor becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
 These results are summarized by Figure \ref{fig:FirstSurvivalCurves}A.
@@
 \frame{\includegraphics[width=\columnwidth]{Figure2/Figure2.pdf}}
-\caption{A. Increasing MIE decreases Time to Cancer. 
+\caption{
+A. Increasing MIE decreases Time to Cancer. 
 B. Increasing MGA increases Time to Cancer.
 C. Increasing Treg TGF-$\beta$ production results in decreased Time to Cancer.
-%% {\color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.}\newline
-\newline
+%% {\color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.}
 }
@@
 In Figure \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production is shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
-This is expected as the two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing cells to a mesenchymal phenotype.
+This is expected as the two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tissue cells to a mesenchymal phenotype.
 Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease Time to Cancer.
@@
 For those cohorts with a permanently high inflammatory state, the relationship between the mesenchymal parameters and the Time to Cancer is monotonic.
-However, when there are periods of a low inflammatory state, then intriguingly the connection between MQ and Time to Cancer becomes concave down with a local maximum around 0.3.
-[what does this value mean] [refer to figure]
-What this indicates is that if the quiescence of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer and the time to death.
+However, when there are periods of a low inflammatory state, then intriguingly the connection between MGA and Time to Cancer becomes concave down with a local maximum around 0.3.
+This can be seen in the top two groups of Figure \ref{fig:VaryINFL_and_MesPars}D in which the MGA=0.3 cohorts both took longer to cancer than their neighbors.
+What this indicates is that if the growth arrest of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer.
 Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
+% Is this next sentence necessary?
 See Figure \ref{fig:VaryINFL_and_MesPars}C and Figure \ref{fig:VaryINFL_and_MesPars}D for a summary.
@@
 Contrast this with what happens when MIE is varied for different inflammation cycling schemes.
-Regardless of the cycling scheme, the relationship between Time to Cancer and MIE is never non-monotonic.
+Regardless of the cycling scheme, the relationship between Time to Cancer and MIE is always -monotonic.
 In the case that the inflammatory state is permanently high, MIE has no significant effect on Time to Cancer.
 On the other hand, when the patient experiences at least some time in a low inflammatory state, an increase in MIE results in a decrease in Time to Cancer.
-This indicates that reducing the inflammatory state of the patient is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
+This indicates that mitigating a chronic inflammatory state is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
+% Is this next sentence necessary?
 See Figure \ref{fig:VaryINFL_and_MesPars}A and Figure \ref{fig:VaryINFL_and_MesPars}B for a summary.
@@
 Regardless of the value of MGA, increasing MIE leads to decreased Time to Cancer.
-On the other hand, for any value of MIE, then there is an optimal MGA value in the sense that it results in the longest Time to Cancer.
-It appears based on the plot that this value is around 0.28.
+On the other hand, for any value of MIE, then there is an MGA value that results in the longest Time to Cancer.
+For the model, this optimal value increases with MIE.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%
+%                      DISCUSSION                       %
+%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Discussion}\label{Discussion}
-This might actually be where I talk about the implication of the results from Section \ref{KeyEMT}.
-In addition, I could discuss the implications from the TGF-$\beta$ section.
-Any discussion on this topic, however, would need to acknowledge that regulation of TGF-$\beta$ might also have other, serious side effects, especially if the treatment is not local.
-However, that might also be true for the inflammatory results.
-In light of this, it might be good to include a new section on how TGF-$\beta$ production's influence on Time to Cancer is influenced by the inflammatory cycling scheme.
+In building this model, we set out to better understand how the immune system and the EMT spectrum would affect Time to Cancer results from previous work.
+What we found is that altering many of the processes have predictable results: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, and increasing Treg TGF-$\beta$ production all lead to shorter Times to Cancer.
+However, when the inflammation of the TME is accounted for, some of these changes to the system have less predictable outcomes.
+The most intriguing is that periods of low inflammation lead to mesenchymal growth arrest having an optimal value for maximizing Time to Cancer.
+
+%In addition, I could discuss the implications from the TGF-$\beta$ section.
+%Any discussion on this topic, however, would need to acknowledge that regulation of TGF-$\beta$ might also have other, serious side effects, especially if the treatment is not local.
+%However, that might also be true for the inflammatory results.
+%In light of this, it might be good to include a new section on how TGF-$\beta$ production's influence on Time to Cancer is influenced by the inflammatory cycling scheme.
 
@@
 \section{Conclusion}\label{Conclusion}
+The interactions of the immune system and cancer remain a promising avenue of research.
+Including more diverse aspects of the TME and the tumor itself make the challenge that much harder but also that much more promising.
+As we begin to understand how processes such as EMT factor in to these complex interactions, we can begin to respond clinically to better improve patient outcomes.
+Here, we saw {\it in silico} evidence of an optimal value for a property of mesenchymal cells that would maximize Time to Cancer.
+There is still much work to be done further isolating this effect in computer simulations and further verifying this effect and exploring further ways EMT can influence tumor progression.
+This model explored much more than mesenchymal growth arrest, so a simpler model could make this point clearer.
+On the other hand, models which incorporate more distinctions between epithelial and mesenchymal cells could give better clarity on the relative importance of these and reveal a fuller picture about how a mesenchymal subpopulation affects cancer outcomes.
+
+% Adam, do you know of any?
+The authors are not aware of current treatments that specifically target mesenchymal growth arrest, influencing it in one direction or another.
+%
+Based on these results, this could prove to be a promising direction for research, in particular for cancers in which it is known that mesenchymal cells are present and active within the tumor.
+The greatest challenge in translating the information here to the clinic, apart from creating the drugs, would be discovering the ideal growth rate for mesenchymal cells within each patient and how other patient-specific data will influence this result.
+
+The goal of improving cancer treatment has always proven attainable even despite great difficulty.
+The processes involved are so complex, yet it is those complexities that we can exploit to better control and eradicate cancer.
+Thus, continuing research into every aspect of cancer is the clearest, best path forward.
 

### 2019-01-25 Paper/Paper.tex
Changed comments for Adam
Labeled all comments I’d like responses by %A

@@
 Year-by-year, new breakthroughs in the field produce new, effective drugs and therapies that have a significant impact on patient health and survival\cite{pardoll2012blockade}\cite{restifo2012adoptive}.
-% More examples? Reviews or non-reviews? Explain some of the examples?
+%A More examples? Reviews or non-reviews? Explain some of the examples?
 There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area.
@@
 The first is that mesenchymal cells are less susceptible to immune clearance\cite{terry2017new}. 
-The second is that mesenchymal cells proliferate at a slower rate. Ref % Adam said he had a reference in mind for this.
+%
+The second is that mesenchymal cells proliferate at a slower rate. Ref
+ %A Adam said he had a reference in mind for this.
 
@@
 The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
-Upon successful clearance, they themselves are then deactivated and removed from the immune population. % Should this be justified here or mentioned in the intro?
+%A Should this be justified here or mentioned in the intro?
+Upon successful clearance, they themselves are then deactivated and removed from the immune population.
+%
 Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear mutated cells.
@@
 \item immune clearance (by NKs or CTLs)
-\item rest/$G_0$ % use both here?
+\item rest/$G_0$ %A use both here?
 \end{itemize}
@@
 There are four factors that contribute to the probability that a cell will proliferate.
-% 
 The first is a base proliferation probability that all cells have, $p$.
@@
 
-$$ p_{\text{NK}} =\delta_M \frac{\text{NK}}{N/N_{1}+\text{NK}} E_{\text{NK}} \frac{1}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} \cdot $$
-$$ \qquad(1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}})$$
+$$ p_{\text{NK}} =\delta_M \frac{\text{NK}}{N/N_{1}+\text{NK}}  \frac{E_{\text{NK}}}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} (1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}}) $$
+%$$\qquad\qquad \qquad(1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}})$$
 % Should I worry about the order in which these appear? Rearranging would make the formula look better, but the order I have introduced them makes the most sense.
@@
 
-\paragraph{Rest/$G_0$} %use both for this section heading?
+\paragraph{Rest/$G_0$} %A use both for this section heading?
 The weight associated with quiescence is taken as 1 except in the case of Mesenchymal cells.
@@
 
-$$ p_{G0} = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_{GA}\frac{N_0}{N_0+N} $$ %p_{G_0},p_{rest},p_{\text{rest}},p_{GA}???
+$$ p_{G0} = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_{GA}\frac{N_0}{N_0+N} $$ 
+%A style choice: p_{G_0},p_{rest},p_{\text{rest}},p_{GA}???
 
 The reason for adding that term is due to the understanding that mesenchymal cells are not proliferating as fast because they are instead remaining in the $G_0$ phase.
-% is this the right way to word this?
+%A is this the right way to word this?
 
@@
 First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
+%A Similar to previous comment: should I cite reference for this here or put this up in the intro?
 This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
-% Similar to previous comment: should I cite reference for this here or put this up in the intro?
-
+%
 Also, all proliferating cells have a probability of undergoing a pathway mutation.
 If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
-If the cell does not undergo a mutation, then its probability of mutating during subsequent cell increases. % Updated this after Adam had question. Does this make sense as is?
-
+%A Updated this after Adam had question. Does this make sense as is?
+If the cell does not undergo a mutation, then its probability of mutating during subsequent cell increases. 
+%
 Then, the EMT values for each cell is updated.
@@
 Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
+
+\emph{Adam, if you're just reading the PDF, then know that I'm really not sure if this next paragraph belongs in the paper.}
 
@@
 which is discretized to
+
+\emph{Adam, if you are reading the PDF, you will see the same equations multiple times as I figure out which one I like best.}
 
@@
 C. Example survival curve for one cohort of patients with basal parameter values.
-% Should I list out all parameter values? A subset? Or refer to supplement for base parameter values?
+%A Should I list out all parameter values? A subset? Or refer to supplement for base parameter values?
 D. Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
@@
 In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
-% It seems the rest of this paragraph belongs in the Methods section, yeah?
+%A It seems the rest of this paragraph belongs in the Methods section, yeah?
 Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
@@
 
-% This next commented out paragraph seems to belong somewhere, but where? I would think in the Intro.
+%A This next commented out paragraph seems to belong somewhere, but where? I would think in the Intro.
 
@@
 We will return to this shortly.
-It is at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, which translates to 405.75 days. % translate to days like this?
+It is at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, which translates to 405.75 days. %A translate to days like this?
 At the same time as mutated cells are taking over, most cells are transitioning to a mesenchymal phenotype.
@@
 \subsection{Sensitivity Analysis}\label{SensAnalysis}
-% 
 To assess the sensitivity of the model to the 31 model parameters, a Morris OAT algorithm was implemented.
 The algorithm works by running the model many successive times with exactly one parameter changing between each run\cite{morris1991factorial}\cite{sohier2014improvement}.
-% is the above description sufficient? The first source is Morris's original paper and the second is the source of the particular implementation I found
+%A is the above description sufficient? The first source is Morris's original paper and the second is the source of the particular implementation I found
 The parameters were all given Gaussian distributions centered at their initial value and a large standard deviation.
@@
 Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
-% Is this next sentence necessary?
+%A Is this next sentence necessary?
 See Figure \ref{fig:VaryINFL_and_MesPars}C and Figure \ref{fig:VaryINFL_and_MesPars}D for a summary.
@@
 This indicates that mitigating a chronic inflammatory state is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
-% Is this next sentence necessary?
+%A Is this next sentence necessary?
 See Figure \ref{fig:VaryINFL_and_MesPars}A and Figure \ref{fig:VaryINFL_and_MesPars}B for a summary.
@@
 
-% Adam, do you know of any?
+%A Adam, do you know of any?
 The authors are not aware of current treatments that specifically target mesenchymal growth arrest, influencing it in one direction or another.

### 2019-01-28 Paper/Paper.tex
replies to comments

@@
 %% The following should allow me to make subsubsubsections with the command \paragraph{}
+%% AM: I don't think this is necessary. please use \paragraph (original style) instead
 \usepackage{titlesec}
@@
 Year-by-year, new breakthroughs in the field produce new, effective drugs and therapies that have a significant impact on patient health and survival\cite{pardoll2012blockade}\cite{restifo2012adoptive}.
-%A More examples? Reviews or non-reviews? Explain some of the examples?
+% A More examples? Reviews or non-reviews? Explain some of the examples?
+%AM I will add a few 
 There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area.
@@
 The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
-%A Should this be justified here or mentioned in the intro?
+% A Should this be justified here or mentioned in the intro?
+% AM: justifed no; Intro yes
 Upon successful clearance, they themselves are then deactivated and removed from the immune population.
@@
 \item immune clearance (by NKs or CTLs)
-\item rest/$G_0$ %A use both here?
+\item rest in $G_0$ %A use both here? - % AM: change to 'rest in' everywhere
 \end{itemize}
@@
 
-\paragraph{Rest/$G_0$} %A use both for this section heading?
+\paragraph{Rest in $G_0$} %A use both for this section heading?
 The weight associated with quiescence is taken as 1 except in the case of Mesenchymal cells.
@@
 
-$$ p_{G0} = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_{GA}\frac{N_0}{N_0+N} $$ 
-%A style choice: p_{G_0},p_{rest},p_{\text{rest}},p_{GA}???
+$$ p_{rest} = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_{GA}\frac{N_0}{N_0+N} $$ 
+%YA style choice: p_{G_0},p_{rest},p_{\text{rest}},p_{GA}???
 
-The reason for adding that term is due to the understanding that mesenchymal cells are not proliferating as fast because they are instead remaining in the $G_0$ phase.
-%A is this the right way to word this?
+The reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells remain longer in the $G_0$ phase.
+%YA is this the right way to word this?
 
@@
 First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
-%A Similar to previous comment: should I cite reference for this here or put this up in the intro?
+% A Similar to previous comment: should I cite reference for this here or put this up in the intro?
+% AM no need for extra here: just in intro
 This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
 %
-Also, all proliferating cells have a probability of undergoing a pathway mutation.
+All proliferating cells have a nonzero probability of undergoing a pathway mutation.
 If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
-%A Updated this after Adam had question. Does this make sense as is?
-If the cell does not undergo a mutation, then its probability of mutating during subsequent cell increases. 
+% A Updated this after Adam had question. Does this make sense as is?
+% AM I don't understnad 'remains mutated'
+If the cell does not undergo a mutation, then its probability of mutation during subsequent cell increases. 
 %
@@
 %%%%%%%%%	I think this next section is too much detail that just confuses the reader. I don't think I need to explain my implementation, just the process. Thoughts?	%%%%%%%%%%%%%%%%%%%%
+% AM: yes, remove or shorten to one sentence at end of previous para.
 This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
@@
 %%% I like the first of these two options. You set aside the amount you know you will gain (sigma/d) and then let the rest decay exponentially
+% AM yes first one 
 $$ \text{NK}_{k+1} = \left (\text{NK}_k-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} $$
@@
 In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
-%A It seems the rest of this paragraph belongs in the Methods section, yeah?
+% A It seems the rest of this paragraph belongs in the Methods section, yeah?
+% AM : yes
 Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
@@
 
-%A This next commented out paragraph seems to belong somewhere, but where? I would think in the Intro.
+% A This next commented out paragraph seems to belong somewhere, but where? I would think in the Intro.
+% AM: yes in intro
 
@@
 We will return to this shortly.
-It is at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, which translates to 405.75 days. %A translate to days like this?
+It is at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, or 405.75 days. 
 At the same time as mutated cells are taking over, most cells are transitioning to a mesenchymal phenotype.
@@
 The algorithm works by running the model many successive times with exactly one parameter changing between each run\cite{morris1991factorial}\cite{sohier2014improvement}.
-%A is the above description sufficient? The first source is Morris's original paper and the second is the source of the particular implementation I found
+% A is the above description sufficient? The first source is Morris's original paper and the second is the source of the particular implementation I found
+% AM: yes fine 
 The parameters were all given Gaussian distributions centered at their initial value and a large standard deviation.
@@
 Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
-%A Is this next sentence necessary?
+% A Is this next sentence necessary?
+% AM: can remove it 
 See Figure \ref{fig:VaryINFL_and_MesPars}C and Figure \ref{fig:VaryINFL_and_MesPars}D for a summary.
@@
 This indicates that mitigating a chronic inflammatory state is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
-%A Is this next sentence necessary?
+% A Is this next sentence necessary?
+% AM: remove
 See Figure \ref{fig:VaryINFL_and_MesPars}A and Figure \ref{fig:VaryINFL_and_MesPars}B for a summary.
@@
 
-The goal of improving cancer treatment has always proven attainable even despite great difficulty.
-The processes involved are so complex, yet it is those complexities that we can exploit to better control and eradicate cancer.
-Thus, continuing research into every aspect of cancer is the clearest, best path forward.
+Progress in cancer treatment has always proven attainable even despite great challenges, for which the complexity of the disease is often responsible.
+As we move forwards, it is these very complexities that we will be able to better exploit to eradicate or control the disease.
 

### 2019-01-31 Paper/Paper.tex
Abstract wording update

@@
 We developed a multiscale agent-based model that describes tumor-immune interactions and how EMT modulates these.
-Exploration of the model revealed that increasing mesenchymal cells' quiescence in some circumstances results in a more tumor-friendly environment and sometimes less.
-In contrast, enhancing mesenchymal cells' immune evasiveness always leads to a more tumor-friendly microenvironment.
+Exploration of the model revealed that increasing the growth arrest of mesenchymal cells in some circumstances results in a more tumor-friendly environment and sometimes less.
+In contrast, enhancing the immune evasiveness of mesenchymal cells always leads to a more tumor-friendly microenvironment.
 Moreover, encouraging the TME towards a pro-EMT state also benefits the tumor.

### 2019-01-31 Paper/Paper.tex
Began checking Adam's Updates
Done with Adam's Updates through Results

@@
-%\documentclass[twocolumn]{article}
 \documentclass{article}
@@
 \usepackage{cite}
-
-%% The following should allow me to make subsubsubsections with the command \paragraph{}
-%% AM: I don't think this is necessary. please use \paragraph (original style) instead
-\usepackage{titlesec}
-
-\setcounter{secnumdepth}{4}
-\titleformat{\paragraph}
-{\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
-\titlespacing*{\paragraph}
-{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
-%%
-
 \graphicspath{ {Figs/} }
@@
 While the cytotoxic immune cells, NKs and cytotoxic T cells (CTLs), seek out and lyse tumor cells, the regulatory branch of the immune system, which will be accounted for by Tregs in our model, slows down the effector functions of these other cells.
+% Make sure there's some ref for immune clearing tumor
+% Also, add that as immune cells clear tumor cells, they become deactivated and give source
 
@@
 The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
-% A Should this be justified here or mentioned in the intro?
-% AM: justifed no; Intro yes
 Upon successful clearance, they themselves are then deactivated and removed from the immune population.
-%
 Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear mutated cells.
@@
 \item immune clearance (by NKs or CTLs)
-\item rest in $G_0$ %A use both here? - % AM: change to 'rest in' everywhere
+\item rest in $G_0$
 \end{itemize}
@@
 
-\paragraph{Rest in $G_0$} %A use both for this section heading?
+\paragraph{Rest in $G_0$} 
 The weight associated with quiescence is taken as 1 except in the case of Mesenchymal cells.
@@
 
-The reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells remain longer in the $G_0$ phase.
-%YA is this the right way to word this?
+The reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.
 
@@
 First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
-% A Similar to previous comment: should I cite reference for this here or put this up in the intro?
-% AM no need for extra here: just in intro
 This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
@@
 All proliferating cells have a nonzero probability of undergoing a pathway mutation.
-If they do, one is randomly chosen among the three and that one either mutates or remains mutated.
+If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
 % A Updated this after Adam had question. Does this make sense as is?
 % AM I don't understnad 'remains mutated'
-If the cell does not undergo a mutation, then its probability of mutation during subsequent cell increases. 
+% If the pathway selected was one that already had a mutation, then nothing happens. 
+If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases. 
 %
@@
 Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
-
-\emph{Adam, if you're just reading the PDF, then know that I'm really not sure if this next paragraph belongs in the paper.}
+Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: quantity of TGF-$\beta$ in the system and the current EMT score of the cell.
 
@@
 % AM: yes, remove or shorten to one sentence at end of previous para.
-This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
-If $\text{TGFB}_i>2-\text{EMT}_i$, the cell becomes more mesenchymal and will have a higher EMT score.
-If $\text{TGFB}_i<2-\text{EMT}_i$, the cell becomes more epithelial and will have a lower EMT score.
-The reason for the expression $2-\text{EMT}_i$ can be understood this way: 
-the bounds on the EMT score guarantee $2-\text{EMT}_i\in[1,2]$.
-Hence, there either needs to be a lot of TGF-$\beta$ in the system or a large amount of randomness for a cell to undergo EMT.
-In addition, a cell having a higher EMT score will have a higher chance of continuing in the EMT process whereas a cell low on the EMT spectrum will have a higher chance of moving towards a more epithelial phenotype.
+
+%This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
+%If $\text{TGFB}_i>2-\text{EMT}_i$, the cell becomes more mesenchymal and will have a higher EMT score.
+%If $\text{TGFB}_i<2-\text{EMT}_i$, the cell becomes more epithelial and will have a lower EMT score.
+%The reason for the expression $2-\text{EMT}_i$ can be understood this way: 
+%the bounds on the EMT score guarantee $2-\text{EMT}_i\in[1,2]$.
+%Hence, there either needs to be a lot of TGF-$\beta$ in the system or a large amount of randomness for a cell to undergo EMT.
+%In addition, a cell having a higher EMT score will have a higher chance of continuing in the EMT process whereas a cell low on the EMT spectrum will have a higher chance of moving towards a more epithelial phenotype.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\emph{Adam, if you are reading the PDF, you will see the same equations multiple times as I figure out which one I like best.}
-
 %%% I like the first of these two options. You set aside the amount you know you will gain (sigma/d) and then let the rest decay exponentially
@@
 %%% For this one, the initial decays exponentially and then you add some fraction of sigma/d depending on how big your time step is
-$$ \text{NK}_{k+1} = \text{NK}_k\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \left(1-\exp(-d_{\text{NK}}\Delta t)\right)$$
+%$$ \text{NK}_{k+1} = \text{NK}_k\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \left(1-\exp(-d_{\text{NK}}\Delta t)\right)$$
 %%% The first one is better though in my opinion.
@@
 %% Option 1b from above:
-$$\text{NK}_{k+1} = (\text{NK}_k-R_\text{NK})\exp(-d_\text{NK}\Delta t)+R_\text{NK},\quad R_\text{NK}=\frac{\sigma_\text{NK}}{d_\text{NK}}$$
+%$$\text{NK}_{k+1} = (\text{NK}_k-R_\text{NK})\exp(-d_\text{NK}\Delta t)+R_\text{NK},\quad R_\text{NK}=\frac{\sigma_\text{NK}}{d_\text{NK}}$$
 

### 2019-02-01 Paper/Paper.tex
Finished addressing Adam's comments

@@
 While the epithelial cells often stay bound together in layers, mesenchymal cells have gained motility and even stem-like properties\cite{nieto2016emt}.
+Recent research by Hong has revealed the intriguing possibility that there are four stable equilibria along the EMT spectrum with each state having its own unique properties\cite{hong2015ovol2}.
 They have thus become a subject of great interest in cancer research having been connected to tumor initiation, progression, and metastasis\cite{nieto2016emt}.
@@
 
+% This paragraph needs to find a home in this section somewhere
+Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
+If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the annotation to the right of the cell block.
+Any cell that has a pathway mutation is counted as a mutated cell and contributes towards the total number of cancer cells.
+The EMT score of each cell (denoting position along the EMT spectrum) is associated with a color given by the legend to the left of the cell block. 
+In the model the EMT score is continuous but the labeling of epithelial or mesenchymal is a binary determined by being below or above a threshold.
+
 \section{Methods}
@@
 They can either be mutation-free, or have any combination of three possible pathway mutations.
+These three pathways are the three components within each cell in Figure \ref{fig:ModelIntro}A.
+If the pathway is mutated, the component is colored red and has an `X'.
+A single mutation is sufficient for the cell being counted as a mutant cell.
+
 We model the immune cells as discrete variables but as one collective, in other words they are not agents themselves.
@@
 Third, if the cell is mesenchymal, then the weight for proliferation is proportionally decreased by $m_{GA}$, which stands for mesenchymal growth arrest.
-Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor.
+Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N$.
 In total, the weight for proliferation is given by
 
-$$ p_{\text{prolif}} = p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})(1-\delta_{MES}m_{GA})\frac{N_0}{N_0+N} $$
+$$ p_{prolif} = p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})(1-\delta_{MES}m_{GA})\frac{N_0}{N_0+N} $$
 
@@
 
-$$ p_{\text{apop}} = d(1-\delta_{\text{AM}}\text{apop}_{\text{down}}) $$
+$$ p_{apop} = d(1-\delta_{\text{AM}}\text{apop}_{\text{down}}) $$
 
@@
 $$ p_{\text{NK}} =\delta_M \frac{\text{NK}}{N/N_{1}+\text{NK}}  \frac{E_{\text{NK}}}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} (1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}}) $$
-%$$\qquad\qquad \qquad(1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}})$$
 % Should I worry about the order in which these appear? Rearranging would make the formula look better, but the order I have introduced them makes the most sense.
@@
 $$ p_{rest} = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_{GA}\frac{N_0}{N_0+N} $$
-%YA style choice: p_{G_0},p_{rest},p_{\text{rest}},p_{GA}???
 
@@
 This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
-%
 All proliferating cells have a nonzero probability of undergoing a pathway mutation.
 If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
-% A Updated this after Adam had question. Does this make sense as is?
-% AM I don't understnad 'remains mutated'
-% If the pathway selected was one that already had a mutation, then nothing happens. 
 If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases. 
-%
 Then, the EMT values for each cell is updated.
@@
 
-%%%%%%%%%	I think this next section is too much detail that just confuses the reader. I don't think I need to explain my implementation, just the process. Thoughts?	%%%%%%%%%%%%%%%%%%%%
-% AM: yes, remove or shorten to one sentence at end of previous para.
+Next, the amount of TGF-$\beta$ for the next cell cycle is determined by the number of mutated cells, $M$, and the number of Treg cells, each one producing a fixed amount of TGF-$\beta$.
+It is given by
 
-%This quantity is then compared to $2-\text{EMT}_i$ where $\text{EMT}_i$ is the EMT score of the $i^{th}$ cell on a scale of $[0,1]$ with $0$ representing most epithelial and $1$ representing most mesenchymal.
-%If $\text{TGFB}_i>2-\text{EMT}_i$, the cell becomes more mesenchymal and will have a higher EMT score.
-%If $\text{TGFB}_i<2-\text{EMT}_i$, the cell becomes more epithelial and will have a lower EMT score.
-%The reason for the expression $2-\text{EMT}_i$ can be understood this way: 
-%the bounds on the EMT score guarantee $2-\text{EMT}_i\in[1,2]$.
-%Hence, there either needs to be a lot of TGF-$\beta$ in the system or a large amount of randomness for a cell to undergo EMT.
-%In addition, a cell having a higher EMT score will have a higher chance of continuing in the EMT process whereas a cell low on the EMT spectrum will have a higher chance of moving towards a more epithelial phenotype.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Next, the amount of TGF-$\beta$ for the next cell cycle is determined by the number of mutated cells and the number of Treg cells, each one producing a fixed amount of TGF-$\beta$. It is given by
-
-$$ \text{TGFB}_k = \text{tgfb}_{\text{mut}}M_k + \text{tgfb}_{\text{Treg}}\text{Treg}_k$$ % clear that M is number of mutants, right?
+$$ \text{TGFB} = \text{tgfb}_{\text{mut}}M + \text{tgfb}_{\text{Treg}}\text{Treg}$$ % clear that M is number of mutants, right?
 
@@
 
-%%% I like the first of these two options. You set aside the amount you know you will gain (sigma/d) and then let the rest decay exponentially
-% AM yes first one 
 $$ \text{NK}_{k+1} = \left (\text{NK}_k-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} $$
-
-%%% For this one, the initial decays exponentially and then you add some fraction of sigma/d depending on how big your time step is
-%$$ \text{NK}_{k+1} = \text{NK}_k\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \left(1-\exp(-d_{\text{NK}}\Delta t)\right)$$
-%%% The first one is better though in my opinion.
-
-% For all 3 discretizations of immune cell dynamics, I could create a quantity R_* (for recruitment? maybe better letters exist) so the formula would be 
-% *_{k+1} = (*_k-R_*)\exp(-d_*\Delta t)+R_*,\quad R_*=\frac{\sigma_*}{d_*}
-%% Option 1b from above:
-%$$\text{NK}_{k+1} = (\text{NK}_k-R_\text{NK})\exp(-d_\text{NK}\Delta t)+R_\text{NK},\quad R_\text{NK}=\frac{\sigma_\text{NK}}{d_\text{NK}}$$
 
@@
 
-%% Or using R_*
-\begin{align*}
-\text{CTL}_{k+1} & =\left (\text{CTL}_k-\tilde{M}_kR_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \tilde{M}_kR_{\text{CTL}}\\
-\text{Treg}_{k+1} & =\left (\text{Treg}_k-\tilde{M}_kR_{\text{Treg}}\right )\exp(-d_{\text{Treg}}\Delta t)+\tilde{M}_kR_{\text{Treg}}
-\end{align*}
-where
-\begin{align*}
-R_{\text{CTL}} &= \sigma_{\text{CTL}}/d_{\text{CTL}}\\
-R_{\text{Treg}} &=\frac{\sigma_{\text{Treg}}}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}
-\end{align*}
-%% End of option 1
-
-%% Beginning of option 2
-For CTLs and Tregs, however, they rely on antigen presenting cells (APCs) to be activated, and these APCs in turn rely on mutated cells being cleared so that they can bind antigens.
-In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function.
-Thus, let $\text{APC}$ represent the number of mutated cells cleared by the immune system during a given cycle.
-We choose the following differential equations to govern the CTL and Treg populations:
-
-\begin{align*}
-\text{CTL}' & = \sigma_{\text{CTL}}\text{APC} - d_{\text{CTL}}\text{CTL} \\
-\text{Treg}' & = \sigma_{\text{Treg}}\text{APC} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}- d_{\text{Treg}}\text{Treg}
-\end{align*}
-
-Discretized, these are:
-
-\begin{align*}
-\text{CTL}_{k+1} & = & \left (\text{CTL}_k-\sigma_{\text{CTL}}\text{APC}_k/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}\text{APC}_k/d_{\text{CTL}}\\
-\text{Treg}_{k+1} & = & \left (\text{Treg}_k-\frac{\sigma_{\text{Treg}}\text{APC}_k}{d_{\text{Treg}}} \frac{\text{TGFB}_k}{1+\text{TGFB}_k/\text{EC50}}\right )\exp(-d_{\text{Treg}}\Delta t)+\\
-& & \frac{\sigma_{\text{Treg}}\text{APC}_k}{d_{\text{Treg}}} \frac{\text{TGFB}_k}{1+\text{TGFB}_k/\text{EC50}}
-\end{align*}
-%% End of option 2
-
 \subsubsection{Progression and tumorigenesis}
@@
 Simulations run until either a tumorigenic event is recorded or until the maximum number of cycles has been reached.
-%It is also possible with the right combination of parameters, to have patients run out of cells entirely.
-%The model does not distinguish between these patients and those that have a cancer diagnosis.
-%The reason is that these patients have succumb to some combination of inflammatory and mutagenic events.
 
@@
 In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
-% A It seems the rest of this paragraph belongs in the Methods section, yeah?
-% AM : yes
-Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
-If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the annotation to the right of the cell block.
-Any cell that has a pathway mutation is counted as a mutated cell and contributes towards the total number of cancer cells.
-The EMT score of each cell (denoting position along the EMT spectrum) is associated with a color given by the legend to the left of the cell block. 
-In the model the EMT score is continuous but the labeling of epithelial or mesenchymal is a binary determined by being below or above a threshold.
-
-% A This next commented out paragraph seems to belong somewhere, but where? I would think in the Intro.
-% AM: yes in intro
-
-%While biologically, there may be only a few discrete stable cell states available on the EMT spectrum, recent studies highlight a growing number of states accessible, with research indicating there are several equilibria along the spectrum\cite{hong2015ovol2}.
-%In our model, this is actually just a binary and that is why the cells shown are at one extreme or the other of the spectrum.
-
-In the right half of Figure \ref{fig:ModelIntro}A one can see the cell cycle choices that each cell will make each cell cycle.
-These are the four thick, gray arrows.
-Each of the cells will choose which arrow to follow each cell cycle, and this decision is influenced by the properties of the cell as well as external factors like the immune system and the TME.
+In the right half of Figure \ref{fig:ModelIntro}A one can see the cell cycle choices that each cell will make each cell cycle and how the various components of the model will influence these probabilities.
 The inset on the left gives an overview of how the immune system works.
@@
 Meanwhile, all the cells become mesenchymal for a short duration before most transition back to an epithelial state.
-% Once the Time to Cancer has been determined for a patient, most simulations shown later stop. [??? mean to say: we stop the simulation?]
-%We do this because once such a high proportion of cells have mutated, the mutant cells quickly take over and make up 100\% of all cells.
 
@@
 The algorithm works by running the model many successive times with exactly one parameter changing between each run\cite{morris1991factorial}\cite{sohier2014improvement}.
-% A is the above description sufficient? The first source is Morris's original paper and the second is the source of the particular implementation I found
-% AM: yes fine 
 The parameters were all given Gaussian distributions centered at their initial value and a large standard deviation.
@@
 This is not as intuitive as it may sound because this decreased proliferation affects both mutated and non-mutated cells.
-% It may be interesting to track how much MGA affects mutated vs. non-mutated cells throughout the time.
 These results are summarized by Figure \ref{fig:FirstSurvivalCurves}B.
@@
 Mesenchymal cells are more likely to evade the immune system, so pushing the system towards an overall more mesenchymal phenotype will better protect the cancer and decrease the Time to Cancer.
-% But what happens when MIE is really high to begin with so becoming mesenchymal mostly just makes you proliferate slower? Perhaps that can go in a supplementary figure?
 
@@
 Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
-% A Is this next sentence necessary?
-% AM: can remove it 
-See Figure \ref{fig:VaryINFL_and_MesPars}C and Figure \ref{fig:VaryINFL_and_MesPars}D for a summary.
 
@@
 This indicates that mitigating a chronic inflammatory state is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
-% A Is this next sentence necessary?
-% AM: remove
-See Figure \ref{fig:VaryINFL_and_MesPars}A and Figure \ref{fig:VaryINFL_and_MesPars}B for a summary.
 

### 2019-02-05 Paper/Paper.tex
add tcga figure

@@
-\documentclass{article}
+\documentclass[11pt, a4paper, preprint]{article}
 \usepackage{outline}
@@
 
-\begin{document}
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+\usepackage{authblk}
+\newcommand{\tcr} { \textcolor{red} }
 
 \title{Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer}
+\author{}
+
+\renewcommand\Authands{ and }
+
+\begin{document}
 \maketitle
@@
 %% Basic Introduction (1-2 sentences)
-During tumorigenesis and throughout the lifetime of a tumor, the immune system is constantly engaged in complex interactions in and around the tumor.
+Preceding tumorigenesis and throughout the lifetime of a tumor, the immune system is engaged in complex sets of interactions in and around the tumor.
 %% More Detailed Background (2-3 sentences)
-These interactions are influenced by the tumor microenvironment (TME) and in particular by the cellular process of epithelial-to-mesenchymal transition (EMT) in an as-yet unexplained manner.
+These are regulated by the tumor microenvironment (TME) in myriad ways.
+Moreover,  epithelial cell plasticity impacts disease incidence and progression via processes of epithelial-to-mesenchymal transition (EMT).
 In turn, both the tumor and the immune system exert substantial influence over the TME in competition to influence the fate of the host.
 %% General Problem (1 sentence)
-Determining the critical factors and mechanisms involved remains a crucial task for cancer immunotherapy.
+Determining the factors and mechanisms underpinning these complex interactions remains an important task for cancer immunotherapy.
 %% Summarizing Main Result (1 sentence)
-Here, we seek to understand how interplay between the tumor, the immune system, and EMT can determine the outcome of cancer.
+%Here, we seek to understand how interplay between the tumor, the immune system, and EMT can determine the outcome of cancer.
 %% The Main Result (2-3 sentences)
-We developed a multiscale agent-based model that describes tumor-immune interactions and how EMT modulates these.
-Exploration of the model revealed that increasing the growth arrest of mesenchymal cells in some circumstances results in a more tumor-friendly environment and sometimes less.
-In contrast, enhancing the immune evasiveness of mesenchymal cells always leads to a more tumor-friendly microenvironment.
-Moreover, encouraging the TME towards a pro-EMT state also benefits the tumor.
+We developed a multiscale agent-based model to describe the interplay between three components: the tumor (or pre-tumor cells), the immune system, and EMT.
+The model revealed a non-monotonic relationship between the cell cycle times associated with the mesencymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival.
+In comparison, increasing the immune-evasiveness of the mesenchymal phenotypes always led to poorer prognosis.
+Moreover, encouraging
+%\tcr
+the TME
+towards a pro-EMT state also benefits the tumor.
 %% Putting into a More General Context (1-2 sentences)
@@
 \section{Introduction}
-Immunotherapy is slowly revolutionizing how we treat cancer.
-Year-by-year, new breakthroughs in the field produce new, effective drugs and therapies that have a significant impact on patient health and survival\cite{pardoll2012blockade}\cite{restifo2012adoptive}.
+Immunotherapy is slowly revolutionizing cancer therapy.
+New breakthroughs in the field begin to produce new therapies that have had a significant impact on patient health and survival\cite{pardoll2012blockade}\cite{restifo2012adoptive}.
 % A More examples? Reviews or non-reviews? Explain some of the examples?
 %AM I will add a few 
-There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer; in addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area.
+%There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer;
+In addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area.
 Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context\cite{de2006paradoxical}.
@@
 
-\subsection{Heat Map of MIE vs MGA}
-To further draw out the contrast between these two mesenchymal properties, we compare Time to Cancer for many different pairs of values for MIE and MGA.
-The heat map in Figure \ref{fig:MIEvsMGA} concisely encodes the relationships we have been observing in the previous sections.
-Regardless of the value of MGA, increasing MIE leads to decreased Time to Cancer.
-On the other hand, for any value of MIE, then there is an MGA value that results in the longest Time to Cancer.
-For the model, this optimal value increases with MIE.
+% \subsection{Heat Map of MIE vs MGA}
+\subsection{Analysis of TCGA data in comparison with model predictions suggests that mesenchymal phenotypes reduce cancer-free survival probability}
+
+To further analyze the effects on tumorigenesis of the mesenchymal properties of immune evasion (MIE) and growth arrest (MGA), we study the joint density plot detailing the effects of each of these parameters on the Time to Cancer (Fig.
+\ref{fig:MIEvsMGA}).
+We found that over the full range of values of MGA considered, increasing the MIE decreases the Time to Cancer.
+However, for any given value of mesenchymal immune evasion, there is a value of mesenchymal growth arrest that maximizes the Time to Cancer.
+Moreover, this optimal value increases with MIE.
 
@@
 \label{fig:MIEvsMGA}
+\end{figure}
+
+To compare these model predictions with experimental studies, we performed analysis of data from The Cancer Genome Atlas (TCGA) database to study the effects of immune and EMT interactions on prognosis of cancers for which inflammation is known to play an important role, such as colonic or pancreatic cancers (REFS).
+We selected pancreatic cancer to investigate, and used gene ontologies as a measure of the XXXX of effects of inflammatory or EMT processes on survival.
+In Fig.
+\ref{fig:tcga} we plot Kaplan-Meier curves for
+
+\begin{figure}[H]
+\center
+\frame{\includegraphics[width=\columnwidth]{FigTCGA.pdf}}
+\caption{......r.}
+\label{fig:tcga}
 \end{figure}

### 2019-02-10 Paper/Paper.tex
updates results

@@
 
-\subsection{Exploring the Model}\label{ExplModel}
+\subsection{A multiscale agent-based model of EMT-immune-tissue cell interactions and tumorigenesis}\label{ExplModel}
+We begin by studying the model under baseline conditions to assess how its various components interact and effect the overal probability of tumorigenesis (Time to Cancer).
 In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
-In the right half of Figure \ref{fig:ModelIntro}A one can see the cell cycle choices that each cell will make each cell cycle and how the various components of the model will influence these probabilities.
-The inset on the left gives an overview of how the immune system works.
-The dashed arrows indicate that the action is probabilistic rather than deterministic.
+Within a single cell cycle, different choices are available to a cell and these are influenced by pathway mutations or EMT.
+For example, if a cell undergoes EMT, the probability that is will proliferate is reduced.
+Also shown are the different means by which the immune system acts on the tissue cells (\ref{fig:ModelIntro}A Inset).
+\tcr{TGF-$\beta$ is [describe briefly action of each immune cell type]}
 In other words, the cytotoxic cells will randomly clear some subset of mutated cells and TGF-$\beta$ increases the probability of cell transitioning or staying in a mesenchymal state.
 
-For a typical {\it in silico} patient, see Figure \ref{fig:ModelIntro}B.
+For a typical {\it in silico} patient (see Figure \ref{fig:ModelIntro}B) \tcr{[What happens??]}
 The inflammation cycling scheme shown at the top of Figure \ref{fig:ModelIntro}B is repeated throughout.
 We will return to this shortly.
-It is at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, or 405.75 days. 
-At the same time as mutated cells are taking over, most cells are transitioning to a mesenchymal phenotype.
+
+\tcr{[What is?]} It is at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, or 405.75 days. 
+At the same time as mutated cells are taking over, many cells transition to a mesenchymal phenotype.
 After the Time to Cancer, the mutated cells continue their rapid growth and soon make up 100\% of the cells.
@@
 
-Considering the immune populations, notice that the NK population basically stays constant while the lymphocyte populations quickly dwarf the NK population once cells begin to mutate.
-These lymphocyte populations appear to exhibit oscillatory behavior, however this is not due to intrinsic dynamics but rather due to the inflammation scheme that this patient is undergoing: 
+Considering the immune populations, we see that the NK population is approximately constant, while the lymphocyte populations grow quickly and dwarf the NK population following the accumulation of mutations.
+The lymphocyte populations exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that this patient is undergoing: 
 the patient alternates between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
@@
 
-Finally, considering a cohort of $N=500$ patients, we can generate a survival curve for these patients tracking how many have survived cancer-free up to a give day.
-See Figure \ref{fig:ModelIntro}C for such a survival curve.
-In this curve, we see that all patients survive for some minimum amount of time before a time period a nearly constant cancer onset rate roughly between $T = 100$ and $T = 300$.
+We next consider a cohort of $N=500$ patients, and simulate the survival curve for these patients (Figure \ref{fig:ModelIntro}C).
+We see that all patients survive for approximately 100 cell cycles (75 days), before a time period where the cancer onset rate is roughly constant, between the approximate range of ($T = 100, T = 300$).
 
-\subsection{Sensitivity Analysis}\label{SensAnalysis}
-To assess the sensitivity of the model to the 31 model parameters, a Morris OAT algorithm was implemented.
-The algorithm works by running the model many successive times with exactly one parameter changing between each run\cite{morris1991factorial}\cite{sohier2014improvement}.
-The parameters were all given Gaussian distributions centered at their initial value and a large standard deviation.
-Those that needed bounds were truncated.
-The results of the Morris OAT can be found in Figure \ref{fig:ModelIntro}D.
-Clearly, a certain subset of parameters have an outsized effect on the model while others do relatively little.
-The two most influential according to this analysis are related to Treg cells: their death rate and recruitment.
-This makes sense as Treg cells both suppress cytotoxic effects of other immune cells and also secrete significant quantities of TGF-$\beta$ which drives EMT.
-Thus, Treg cells are intimately tied to all three aspects of the model and so their presence or lack thereof significantly affects the system.
-However, we want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MGA are of interest.
-In addition, inflammation parameters dictating the cycling scheme are worthy of study because of their prominence in Guo's work.
-What we can see from this plot is that the model also has some high sensitivity to these parameters.
-In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and so we will also look at that parameter.
+\subsection{Identification of key model parameters via Morris global sensitivity analysis}\label{SensAnalysis}
+To assess the sensitivity responses produced by the model with respect to the 31 model parameters, a Morris one-step-at-a-time (OAT) algorithm was implemented. \tcr{add paragraph to methods on this.}
+The algorithm works by running the model many successive times, changing exactly one parameter each run \cite{morris1991factorial, sohier2014improvement}.
+Parameters are sampled from Gaussian distributions centered on their initial \tcr{(which is determined how?)} value and a large \tcr{(specify)} standard deviation.
+Where necessary, these distributions were truncated accordingly.
+The results of Morris OAT sensitivity analysis are shown in Figure \ref{fig:ModelIntro}D. A subset of parameters demonstrate much higher levels of sensitivity than others.
+The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
+This ties Treg cells to all three components of the model.
+Since below we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for further analysis below.
+We want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MGA are of interest.
+In addition, inflammation parameters dictating the cycling scheme are important to study, as shown by Guo et al \cite{}: we find that there are also high sensitivities in this model to these parameters.
+In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and we will also study it further.
 

### 2019-02-12 Paper/Paper.tex

@@
 
-\subsection{Mesenchymal Cell Properties Dramatically Change Survival Outcomes}\label{MesPars}
+\subsection{Mesenchymal phenotypic properties dramatically change survival outcomes}\label{MesPars}
 The two quantities that change when a cell transitions to a mesenchymal state are its immune evasion and its increased quiescence, MIE and MGA, respectively.
+\tcr{[briefly redefine what these terms do]}
 Also involved in the EMT process, is the cytokine TGF-$\beta$, which upregulates EMT.
@@
 
-First, as mesenchymal immune evasion (MIE) increases, Time to Cancer decreases.
-This result holds universally among all parameter sets and indicates the simple relationship that mesenchymal immune evasion has with Time to Cancer.
+We found that as the level of mesenchymal immune evasion (MIE) increases, Time to Cancer decreases ( Figure \ref{fig:FirstSurvivalCurves}A).
+This result holds for all parameter sets studied, and indicates the simple relationship that mesenchymal immune evasion has with Time to Cancer.
 As this subpopulation of the tumor becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
@@
 
-Second, as mesenchymal growth arrest (MGA) increases, Time to Cancer increases.
-This result will prove to be more interesting as this pattern is dependent on the other parameters in the model.
-For now, what can be seen in Figure \ref{fig:FirstSurvivalCurves}B is that a slower proliferation for mesenchymal cells slows down cancer growth.
-This is not as intuitive as it may sound because this decreased proliferation affects both mutated and non-mutated cells.
-These results are summarized by Figure \ref{fig:FirstSurvivalCurves}B.
+Second, as the rate of mesenchymal growth arrest (MGA) increases, Time to Cancer increases, i.e.
+lower proliferation rates for mesenchymal cells slow down cancer growth (Figure \ref{fig:FirstSurvivalCurves}B).
+This is not immediately intuitive, since the decreased proliferation rate affects both mutated and non-mutated tissue cells.
 
 Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
-In Figure \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production is shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
-This is expected as the two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tissue cells to a mesenchymal phenotype.
+In Figure \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
+The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tissue cells to a mesenchymal phenotype.
 Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease Time to Cancer.
@@
 
-\subsection{Key EMT regime maximizes cancer-free survival time under chronic inflammation}\label{KeyEMT}
+\subsection{A key EMT regime maximizes cancer-free survival time under chronic inflammation}\label{KeyEMT}
 Given the relevance of the inflammatory state of the tumor microenvironment (TME), we explore the effect of varying the inflammation state of the patient on survival.
-In particular, some cohorts are assigned permanently low inflammation, others permanently high, and still others non-constant inflammatory schemes.
+In simulations, we assign some cohorts permanently low inflammation levels, some permanently high levels, and some variable inflammatory schemes.
 For those cohorts with a permanently high inflammatory state, the relationship between the mesenchymal parameters and the Time to Cancer is monotonic.
 However, when there are periods of a low inflammatory state, then intriguingly the connection between MGA and Time to Cancer becomes concave down with a local maximum around 0.3.
-This can be seen in the top two groups of Figure \ref{fig:VaryINFL_and_MesPars}D in which the MGA=0.3 cohorts both took longer to cancer than their neighbors.
+This can be seen in the top two groups of Figure \ref{fig:VaryINFL_and_MesPars}D: the MGA=0.3 cohorts both took longer to cancer than their neighbors.
 What this indicates is that if the growth arrest of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer.
 Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
+\tcr{what does this mean?
+how?}
 
@@
 \frame{\includegraphics[width=\columnwidth]{Figure3/Figure3.pdf}}
-\caption{A.Survival curves for various inflammation cycling schemes.
+\caption{A.
+Survival curves for various inflammation cycling schemes.
 In the top two plots, the Time to Cancer decreases as MIE increases.
@@
 Contrast this with what happens when MIE is varied for different inflammation cycling schemes.
-Regardless of the cycling scheme, the relationship between Time to Cancer and MIE is always -monotonic.
+Regardless of the cycling scheme, the relationship between Time to Cancer and MIE is monotonic.
 In the case that the inflammatory state is permanently high, MIE has no significant effect on Time to Cancer.
@@
 % \subsection{Heat Map of MIE vs MGA}
-\subsection{Analysis of TCGA data in comparison with model predictions suggests that mesenchymal phenotypes reduce cancer-free survival probability}
+\subsection{Analysis of TCGA data in comparison with model predictions suggests that mesenchymal phenotypes reduce cancer-free survival probability}\label{tcga}
 
@@
 To compare these model predictions with experimental studies, we performed analysis of data from The Cancer Genome Atlas (TCGA) database to study the effects of immune and EMT interactions on prognosis of cancers for which inflammation is known to play an important role, such as colonic or pancreatic cancers (REFS).
-We selected pancreatic cancer to investigate, and used gene ontologies as a measure of the XXXX of effects of inflammatory or EMT processes on survival.
+We selected pancreatic cancer to investigate, and used gene ontologies as a measure of the of effects of inflammatory or EMT processes on survival.
 In Fig.
-\ref{fig:tcga} we plot Kaplan-Meier curves for
+\ref{fig:tcga} we plot survival curves for two cases: either clustering only on EMT, or on both EMT and inflammation signatures.
+We see that in both cases, survival is significantly affected by the gene ontology signature, but importantly, the presence of both EMT and inflammation signatures has a greater impact on survival than the effects of EMT alone.
 
@@
 \frame{\includegraphics[width=\columnwidth]{FigTCGA.pdf}}
-\caption{......r.}
+\caption{A.
+K-means clustering of pancreatic cancers using gene ontology terms indicative of an EMT signature ($k=2$).
+B.
+Survival plots corresponding to the clustering on EMT.
+C.
+K-means clustering of pancreatic cancers using gene ontology terms indicative of EMT and Inflammation signatures ($k=2$).
+D.
+Survival plots corresponding to the clustering on EMT and inflammation.}
 \label{fig:tcga}

### 2019-02-12 Paper/Paper.tex
Beginning to address Adam's comments/questions
Also, fixed single patient fig x axis for 2 of the subplots.

@@
 We developed a multiscale agent-based model to describe the interplay between three components: the tumor (or pre-tumor cells), the immune system, and EMT.
-The model revealed a non-monotonic relationship between the cell cycle times associated with the mesencymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival.
+The model revealed a non-monotonic relationship between the cell cycle times associated with the mesenchymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival.
 In comparison, increasing the immune-evasiveness of the mesenchymal phenotypes always led to poorer prognosis.
@@
 Also shown are the different means by which the immune system acts on the tissue cells (\ref{fig:ModelIntro}A Inset).
-\tcr{TGF-$\beta$ is [describe briefly action of each immune cell type]}
-In other words, the cytotoxic cells will randomly clear some subset of mutated cells and TGF-$\beta$ increases the probability of cell transitioning or staying in a mesenchymal state.
+The NKs and CTLs attempt to clear out mutated tissue cells and deactivate upon successfully carrying out this cytotoxic function.
+The Tregs inhibit this cytotoxic activity.
+In addition, Tregs release TGF-$\beta$ which helps recruit more Tregs and also drives EMT.
+%A \tcr{TGF-$\beta$ is [describe briefly action of each immune cell type]}  --Adam: Is this good?
 
-For a typical {\it in silico} patient (see Figure \ref{fig:ModelIntro}B) \tcr{[What happens??]}
-The inflammation cycling scheme shown at the top of Figure \ref{fig:ModelIntro}B is repeated throughout.
+For a typical {\it in silico} patient (see Figure \ref{fig:ModelIntro}B), after the warmup period, some cells mutate but are cleared before establishing a tumor.
+However, eventually one mutated cell is able to establish a growing tumor before being cleared.
+This happens around cell cycle 1000.
+At this point, the adaptive immune system (CTLs and Tregs) are being heavily recruited and TGF-$\beta$ is peaking.
+The number of mesenchymal cells grows with the increased TGF-$\beta$, but decreases rapidly after the tissue cells are entirely mutants.
+Interestingly, this return to a more epithelial phenotype occurs even as TGF-$\beta$ reaches its maximum value for the simulation.
+%\tcr{[What happens??]}
+The inflammation cycling scheme for this patient is shown at the top of Figure \ref{fig:ModelIntro}B and is repeated throughout, including during the warmup.
 We will return to this shortly.
 
-\tcr{[What is?]} It is at cell cycle number 541 the proportion of mutated cells reaches 50\% and so the patient has a Time to Cancer of 541, or 405.75 days. 
+%\tcr{[What is?]} 
+At cell cycle number 841 the proportion of mutated cells reaches 50\% and so the model marks the patient as having a Time to Cancer of 841, or 630.75 days. 
 At the same time as mutated cells are taking over, many cells transition to a mesenchymal phenotype.

### 2019-02-22 Paper/Paper.tex
Various changes
Updates to figure 1, split fig 1 into 1 and 2 (moved others up in #),
addressed Adam’s \tcr{} comments.

@@
 
-$$ \text{TGFB}_i = \text{TGFB}/N + X_i, \quad X_i \sim N(0,\sigma^2)$$
+%$$ \text{TGFB}_i = H(\text{TGFB}/N) + X_i, \quad X_i \sim N(0,\sigma^2)$$
 
-where $\sigma$ is a parameter in the model.
+$$ \text{TGFB}_i = \frac{\text{TGFB}_{max}}{N}\frac{\text{TGFB}/\gamma_{EC50}}{1+\text{TGFB}/\gamma_{EC50}}+ X_i
+, \quad X_i \sim N(0,\sigma^2)$$
+The summand on the left divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.
+Then, some randomness is introduced by the independent and identically distributed variables $X_i$where $\sigma$ is a parameter in the model.
 If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
@@
 
+\subsection{Parameter Sensitivity Analysis}
+To identify a subset of the parameters to focus on, we used the Morris method, a one-step-at-a-time global sensitivity analysis.
+The algorithm runs the model numerous successive times, changing exactly one parameter each run\cite{morris1991factorial, sohier2014improvement}.
+Each model run involved 1000 patients and the value returned was the area under the survival curve.
+We ran the algorithm by sampling at 30 points in parameter space and then changing each parameter one time at each point.
+In the implementation by Sohier, they recommend using at least 10 points\cite{sohier2014improvement}.
+
+To determine which points in parameter space to sample at, we had to create distributions on all parameter values.
+To do this, we derived many parameters from existing literature and the rest we made reasoned predictions.
+For example, many previous studies have established the parameters of immune population dynamics such as recruitment and death rates\cite{de2014modeling}.
+Because our model only works with on the order of $10^2$ tissue cells whereas an actual tumor would have on the order of $10^9$ cells\cite{de2014modeling}, we scaled the parameters so the immune populations would also be proportionately reduced.
+That is, for instance, how we ended up with NK cells having a steady state population of 10 cells in the model.
+
+Once we had these parameter values, we created Gaussian distributions centered at these values with large variances and truncated at 0.
+Some of the parameters were also truncated to the right because they represented a proportion, such as MIE and MGA, or because we simply wanted to keep the values within a reasonable range, such as the inflammation low and high duration parameters.
+The variances were chosen as twice the mean to get a large spread of values.
 %%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{fig:ModelIntro}
+\end{figure}
+
+\begin{figure}[H]
+\center
+\frame{\includegraphics[width=\columnwidth]{Figure2/Figure2.jpg}}
+\caption{Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
+\label{fig:MOAT}
 \end{figure}
@@
 \subsection{Identification of key model parameters via Morris global sensitivity analysis}\label{SensAnalysis}
-To assess the sensitivity responses produced by the model with respect to the 31 model parameters, a Morris one-step-at-a-time (OAT) algorithm was implemented. \tcr{add paragraph to methods on this.}
-The algorithm works by running the model many successive times, changing exactly one parameter each run \cite{morris1991factorial, sohier2014improvement}.
-Parameters are sampled from Gaussian distributions centered on their initial \tcr{(which is determined how?)} value and a large \tcr{(specify)} standard deviation.
-Where necessary, these distributions were truncated accordingly.
-The results of Morris OAT sensitivity analysis are shown in Figure \ref{fig:ModelIntro}D. A subset of parameters demonstrate much higher levels of sensitivity than others.
+To assess the sensitivity responses produced by the model with respect to the 30 model parameters, a Morris one-step-at-a-time (OAT) algorithm was implemented.
+The results of Morris OAT sensitivity analysis are shown in Figure \ref{fig:MOAT}. A subset of parameters demonstrate much higher levels of sensitivity than others.
 The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
@@
 \subsection{Mesenchymal phenotypic properties dramatically change survival outcomes}\label{MesPars}
-The two quantities that change when a cell transitions to a mesenchymal state are its immune evasion and its increased quiescence, MIE and MGA, respectively.
-\tcr{[briefly redefine what these terms do]}
+The two quantities that change when a cell transitions to a mesenchymal state are its immune evasion and its proliferation rate, MIE and MGA, respectively.
+Both parameters are proportions between 0 and 1 with higher values indicating more extreme expression of mesenchymal properties.
+MIE is the proportional reduction of the probability a mutated mesenchymal cell will undergo immune clearance in a given cycle.
+MGA, which stands for mesenchymal growth arrest, is the proportional reduction in any mesenchymal cell of its chance to proliferate in a given cell cycle. 
+The reduction in proliferation probability corresponds with an equal increase in probability for resting during that cell cycle.  
+%\tcr{[briefly redefine what these terms do]}
 Also involved in the EMT process, is the cytokine TGF-$\beta$, which upregulates EMT.
@@
 \center
-\frame{\includegraphics[width=\columnwidth]{Figure2/Figure2.pdf}}
+\frame{\includegraphics[width=\columnwidth]{Figure3/Figure3.pdf}}
 \caption{
@@
 What this indicates is that if the growth arrest of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer.
-Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
+Even if the patient is presenting a chronically high inflammatory state, anti-inflammatories could be administered which would create the periods of low inflammation our model predicts as necessary for seeing these changes.
+Reducing the inflammatory state of the TME three weeks out of every nine would be sufficient to make the control of mesenchymal proliferation an effective anti-cancer strategy.
+%Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
 \tcr{what does this mean?
@@
 \center
-\frame{\includegraphics[width=\columnwidth]{Figure3/Figure3.pdf}}
+\frame{\includegraphics[width=\columnwidth]{Figure4/Figure4.pdf}}
 \caption{A.
@@
 \center
-\frame{\includegraphics[width=\columnwidth]{Figure4/MIEvsMQ.jpg}}
+\frame{\includegraphics[width=\columnwidth]{Figure5/MIEvsMQ.jpg}}
 \caption{Heat Map contrasting the effects of MIE and MGA on Time to Cancer. Increasing MIE always decreases Time to Cancer, but MGA is non-monotonically linked to Time to Cancer.}

### 2019-02-25 Paper/Paper.tex
Beginning to Address Qing's Comments

@@
 We developed a multiscale agent-based model to describe the interplay between three components: the tumor (or pre-tumor cells), the immune system, and EMT.
-The model revealed a non-monotonic relationship between the cell cycle times associated with the mesenchymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival.
+The model revealed a \tcr{non-monotonic relationship} between the cell cycle times associated with the mesenchymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival.
 In comparison, increasing the immune-evasiveness of the mesenchymal phenotypes always led to poorer prognosis.
@@
 %There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer;
-In addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important research area.
+In addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important \tcr{research} area.
 Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context\cite{de2006paradoxical}.
@@
 
+\tcr{Still need to address Qing's comments on this paragraph.}
 The tumor interacts with the immune system by initiating the sequence of events that results in the activation of the adaptive immune response.
@@
 While the epithelial cells often stay bound together in layers, mesenchymal cells have gained motility and even stem-like properties\cite{nieto2016emt}.
-Recent research by Hong has revealed the intriguing possibility that there are four stable equilibria along the EMT spectrum with each state having its own unique properties\cite{hong2015ovol2}.
-They have thus become a subject of great interest in cancer research having been connected to tumor initiation, progression, and metastasis\cite{nieto2016emt}.
-There are many other characteristics distinctive of mesenchymal cells and still much research being done in further quantifying these and determining others.
-Of interest to us, the literature reveals two qualities of mesenchymal cells that are particularly relevant in the context of cancer and the immune system.
-The first is that mesenchymal cells are less susceptible to immune clearance\cite{terry2017new}. 
+Recent work has revealed the intriguing possibility that there are intermediate cell states along the EMT spectrum with each state having its own unique properties\cite{hong2015ovol2}.
+Such intermediate states have the potential of being connected to tumor initiation, progression, and metastasis\cite{nieto2016emt}.
+While there might be many other characteristics distinctive of mesenchymal cells that need further exploration, two qualities of mesenchymal cells are particularly relevant in the context of cancer and the immune system:
+1) mesenchymal cells are less susceptible to immune clearance\cite{terry2017new} and
 %
-The second is that mesenchymal cells proliferate at a slower rate. Ref
+2) mesenchymal cells proliferate at a slower rate. \tcr{Ref}
  %A Adam said he had a reference in mind for this.
 
+\tcr{Haven't addressed any of Qing's comments below this remark.}
 The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.

### 2019-02-25 Paper/Paper.tex
Next Par

@@
 
-\tcr{Still need to address Qing's comments on this paragraph.}
+\tcr{Still need to address Qing's comments on this paragraph:}
+
 The tumor interacts with the immune system by initiating the sequence of events that results in the activation of the adaptive immune response.
@@
 
-\tcr{Haven't addressed any of Qing's comments below this remark.}
-The research on mesenchymal cells' immune evasiveness is quite conclusive with the research even going so far as establishing functional causes for this phenomenon.
+The immune evasiveness of mesenchymal cells is well-established with even a functional cause for this phenomenon.
 As a cell is being targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be created in order to precipitate the lysing event.
@@
 There are also other structural changes that take place in mesenchymal cells that could contribute to this refractory nature towards immune clearance.
-Regardless, the evidence currently demonstrates a ``slipperiness'' of mesenchymal cells in regards to immune lysing that in our model we will refer to as mesenchymal immune evasion, or MIE.
+Regardless, the evidence currently demonstrates a ``slipperiness'' of mesenchymal cells in regards to immune lysing.
+We will refer to this as mesenchymal immune evasion, or MIE.
+
+\tcr{Haven't addressed any of Qing's comments below this remark.}
 

### 2019-02-25 Paper/Paper.tex
Moved Guo Paragraph

@@
 
-Recent work on cancer-immune interactions has shown that the inflammatory conditions of the TME can play large and sometimes surprising roles in tumor growth.
-Guo describes a model in which the inflammatory state of the TME affects the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations\cite{guo2017multiscale}.
-Guo uses the model to show that while inflammation does have an oncogenic effect, it can also have an onco-protective role as periods of mild inflammation are better at preventing cancer than periods of no inflammation in patients that have intermittent bouts of severe inflammation.
-However, the aspect of the model that included the immune system was minimal: regardless of the inflammatory environment, any cells with sufficient pathway mutations (2 in their model) had a fixed probability of being cleared by the metabolism-immune balance response.
-In this paper, we hope to improve the model by including a more robust immune module and exploring how this axis will improve our understanding of inflammatory effects on the tumor.
-In addition, we explore an extra component of Guo's stem cells: where they lie on the epithelial-mesenchymal axis.
-So, in total, we will explore the joint interactions of the tumor, the immune system, and the epithelial-mesenchymal spectrum.
-We will also study the effects of the TME on these interactions.
-
 \tcr{Still need to address Qing's comments on this paragraph:}
@@
 
+The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
+Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions.
+\tcr{Ref}
+While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate. \tcr{Ref}
+Cancer cells are well-known for achieving unregulated proliferation, so exploring a subset of cancer cells with decreased proliferation could provide intriguing new results.
+In our model, we refer to this property as mesenchymal growth arrest, or MGA.
+
 \tcr{Haven't addressed any of Qing's comments below this remark.}
 
-The other facet of mesenchymal cells of importance in this research is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
-Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions.
-Ref
-While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate. Ref
-Cancer cells are well-known for achieving unregulated proliferation, so exploring a subset of cancer cells with decreased proliferation could provide intriguing new results.
-In our model, we refer to this property as mesenchymal growth arrest, or MGA.
+Recent work on cancer-immune interactions has shown that the inflammatory conditions of the TME can play large and sometimes surprising roles in tumor growth.
+Guo describes a model in which the inflammatory state of the TME affects the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations\cite{guo2017multiscale}.
+Guo uses the model to show that while inflammation does have an oncogenic effect, it can also have an onco-protective role as periods of mild inflammation are better at preventing cancer than periods of no inflammation in patients that have intermittent bouts of severe inflammation.
+However, the aspect of the model that included the immune system was minimal: regardless of the inflammatory environment, any cells with sufficient pathway mutations (2 in their model) had a fixed probability of being cleared by the metabolism-immune balance response.
+In this paper, we hope to improve the model by including a more robust immune module and exploring how this axis will improve our understanding of inflammatory effects on the tumor.
+In addition, we explore an extra component of Guo's stem cells: where they lie on the epithelial-mesenchymal axis.
+So, in total, we will explore the joint interactions of the tumor, the immune system, and the epithelial-mesenchymal spectrum.
+We will also study the effects of the TME on these interactions.
 

### 2019-02-25 Paper/Paper.tex
Up to last par of intro

@@
 
+Recent work on cancer-immune interactions has shown that the inflammatory conditions of the TME can play large and sometimes surprising roles in tumor growth.
+For example, a mathematical model was developed to couple the inflammatory state of the TME with the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations\cite{guo2017multiscale}.
+The work shows that while inflammation does have an oncogenic effect, it can also have an onco-protective role as periods of mild inflammation are better at preventing cancer than periods of no inflammation in patients that have intermittent bouts of severe inflammation.
+However, the aspect of the model that included the immune system is minimal and the connection with EMT is not included.
+%However, the aspect of the model that included the immune system was minimal: regardless of the inflammatory environment, any cells with sufficient pathway mutations (2 in their model) had a fixed probability of being cleared by the metabolism-immune balance response.
+Here, we include a more robust immune module and explore how this immune module will improve our understanding of inflammatory effects on the tumor, and also include the epithelial-mesenchymal axis.
+%In addition, we explore an extra component of Guo's stem cells: where they lie on the epithelial-mesenchymal axis.
+Such additions will allow us to explore the interactions of the tumor, the immune system, and the epithelial-mesenchymal spectrum as well as the effects of the TME on these interactions.
+%We will also study the effects of the TME on these interactions.
+
 \tcr{Haven't addressed any of Qing's comments below this remark.}
-
-Recent work on cancer-immune interactions has shown that the inflammatory conditions of the TME can play large and sometimes surprising roles in tumor growth.
-Guo describes a model in which the inflammatory state of the TME affects the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations\cite{guo2017multiscale}.
-Guo uses the model to show that while inflammation does have an oncogenic effect, it can also have an onco-protective role as periods of mild inflammation are better at preventing cancer than periods of no inflammation in patients that have intermittent bouts of severe inflammation.
-However, the aspect of the model that included the immune system was minimal: regardless of the inflammatory environment, any cells with sufficient pathway mutations (2 in their model) had a fixed probability of being cleared by the metabolism-immune balance response.
-In this paper, we hope to improve the model by including a more robust immune module and exploring how this axis will improve our understanding of inflammatory effects on the tumor.
-In addition, we explore an extra component of Guo's stem cells: where they lie on the epithelial-mesenchymal axis.
-So, in total, we will explore the joint interactions of the tumor, the immune system, and the epithelial-mesenchymal spectrum.
-We will also study the effects of the TME on these interactions.
 

### 2019-02-28 Paper/Paper.tex
Finished Intro (except for 2nd paragraph)
Wrapping up edits for Qing's comments

@@
 This particular cytokine is also a driver of the epithelial-to-mesenchymal transition (EMT) and through generating TGF-$\beta$ the tumor creates a more mesenchymal phenotype within the TME.
+\tcr{I think this is where I would add which types of cancer most likely to be affected by EMT. But which? And what sources?}
 In this way, the tumor shifts the balance of local environment along the EMT spectrum.
@@
 
-\tcr{Haven't addressed any of Qing's comments below this remark.}
-
-In the following sections of the paper we first introduce the model modularly, which is appropriate given its multiscale and hybrid nature.
+Next, we introduce the model modularly, which is appropriate given its multiscale and hybrid nature.
 We then perform sensitivity analysis and identify parameters that are crucial for the progression to tumorigenesis.
@@
 % This paragraph needs to find a home in this section somewhere
-Here, each cell is represented as a column with three components representing the three possible mutations a cell can have.
-If the component is red and has an `X' in the middle, then that indicates the cell has undergone a pathway mutation corresponding to the annotation to the right of the cell block.
-Any cell that has a pathway mutation is counted as a mutated cell and contributes towards the total number of cancer cells.
-The EMT score of each cell (denoting position along the EMT spectrum) is associated with a color given by the legend to the left of the cell block. 
-In the model the EMT score is continuous but the labeling of epithelial or mesenchymal is a binary determined by being below or above a threshold.
 
 \section{Methods}
-To address the questions around the relationships among cancer, the immune system, and EMT, we extended the multiscale, agent-based model of Guo to include an immune module and an EMT axis for the tissue cells.
+To address the questions around the relationships among cancer, the immune system, and EMT, we extended the multiscale, agent-based model \tcr{REF} to include an immune module and an EMT axis for the tissue cells.
 The agents in the model are the tissue cells with tumorigenic potential.
@@
 Finally, tissue cells can be labeled either epithelial or mesenchymal, and this labeling can change over time and depends both on the TME and the internal workings of each individual cell.
+While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
+In Figure \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
+Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.
 
@@
 \subsubsection{Progression and tumorigenesis}
-At the end of each cell cycle, the proportion of tissue cells that are mutated is calculated, and if it is above a certain threshold, a tumorigenic event is recorded and the patient is determined to have cancer (using a similar strategy to Guo).
+At the end of each cell cycle, the proportion of tissue cells that are mutated is calculated, and if it is above a certain threshold, a tumorigenic event is recorded and the patient is determined to have cancer.
 The Time to Cancer is given as the time from the start of the simulation, not including warmup, until tumorigenesis.
@@
 Inset shows major processes of immune system regulation on the tissue cells.
+The parameters used in this model can be found in the Supplementary material.
+Unless otherwise stated, parameter values are assumed to be those found in the Supplement.
+\tcr{Thoughts on the wording here?
+Once I have the table, I will use the name of that table.}
 B. Example simulation of the dynamics of a one patient; warmup period is shown in red.
@@
 \subsection{Identification of key model parameters via Morris global sensitivity analysis}\label{SensAnalysis}
-To assess the sensitivity responses produced by the model with respect to the 30 model parameters, a Morris one-step-at-a-time (OAT) algorithm was implemented.
+To assess the sensitivity responses \tcr{Is this the right phrasing?
+It sounds funny to me, but I see Adam chose this wording.} produced by the model with respect to the 31 model parameters, a Morris one-step-at-a-time (OAT) algorithm was implemented.
 The results of Morris OAT sensitivity analysis are shown in Figure \ref{fig:MOAT}. A subset of parameters demonstrate much higher levels of sensitivity than others.
@@
 Since below we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for further analysis below.
+
+Any of the parameters in the table which include the word ``low'' are parameters that govern the model in a low inflammatory state.
+With the exception of ``prolif up'', all parameters that end in ``up'' or have `U' at the end of their abbreviation, are the factors by which these parameters change when the system switches from low to high inflammation.
+The IHD and ILD parameters determine the duration of high and low inflammation, respectively.
+A cell with mutated pathways will be affected by the ``apop down'', ``prolif up'', and ``immune evasion'' parameters, depending on which pathways are mutated.
+The mesenchymal cells are influenced by the MGA and MIE parameters.
+Finally, $\sigma$, $k_\text{EMT}$, TGFB$_{max}$, and $\gamma_\text{EC50}$ all control how cells transition between the epithelial and mesenchymal states.
+
 We want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MGA are of interest.
-In addition, inflammation parameters dictating the cycling scheme are important to study, as shown by Guo et al \cite{}: we find that there are also high sensitivities in this model to these parameters.
+In addition, inflammation parameters dictating the cycling scheme are important to study\tcr{REF}: we find that there are also high sensitivities in this model to these parameters.
 In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and we will also study it further.
@@
 \caption{
-A. Increasing MIE decreases Time to Cancer. 
-B. Increasing MGA increases Time to Cancer.
-C. Increasing Treg TGF-$\beta$ production results in decreased Time to Cancer.
+A-C. Trajectory of one patient per cohort followed from warmup through 2000 cell cycles.
+\tcr{OK to do A-C like this?}
+D. Increasing MIE decreases Time to Cancer. 
+E. Increasing MGA increases Time to Cancer.
+F. Increasing Treg TGF-$\beta$ production results in decreased Time to Cancer.
 %% {\color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.}

### 2019-03-05 Paper/Paper.tex
add refs

@@
-\documentclass[11pt, a4paper, preprint]{article}
+\documentclass[11pt, preprint]{article}
 \usepackage{outline}
@@
 \usepackage{authblk}
+
+%\usepackage[ ]{natbib} 
 \newcommand{\tcr} { \textcolor{red} }
 
+% A multiscale model of EMT-mediated cancer reveals 
 \title{Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer}
-\author{}
+\author{Daniel R. Bergman$^{1}$,
+Matthew Karikomi\,$^{1}$,
+Qing Nie\,$^{1,2,*}$
+and Adam L. MacLean\,$^{3,*}$
+}
+\affil{
+  $^1$Department of Mathematics, University of California, Irvine, and \\
+  $^2$Department of Cell and Developmental Biology, University of California, Irvine, Irvine, CA 92697, USA \\
+  $^3$Department of Biological Sciences, University of Southern California, Los Angeles, CA 90089, USA \\
+  $^*$Correspondence: macleana@usc.edu (ALM); qnie@uci.edu (QN).
+}
 
@@
 \section{Introduction}
-Immunotherapy is slowly revolutionizing cancer therapy.
-New breakthroughs in the field begin to produce new therapies that have had a significant impact on patient health and survival\cite{pardoll2012blockade}\cite{restifo2012adoptive}.
-% A More examples? Reviews or non-reviews? Explain some of the examples?
-%AM I will add a few 
-%There are myriad untapped avenues by which scientists can further explore the role of the immune system in regulating and combatting cancer;
-In addition to direct cell-cell interactions between immune and cancer cells, the means by which immune cells interact with and modulate the tumor microenvironment (TME) is an important \tcr{research} area.
-Immune cells can be understood as one constituent of the TME: responding to the environment's inflammatory signals, fighting the mutated cells present, and then restoring homeostasis in that local context\cite{de2006paradoxical}.
-Understanding each step in this immune cascade in the context of the TME will be necessary in order to fully explore and exploit all options of immunotherapy.
+The immune system affects tumorigenesis and cancer progression in myriad intricate ways.
+New breakthroughs in the field of immunotherapy are beginning to produce new therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}.
+Beyond direct tumor-immune cell interactions, the immune system also modulates the tumor microenvironment (TME), and indeed, immune cells can be considered as a part of the TME: responding to local inflammatory signals, targeting mutated cells, and potentially eradicating tumor cells and restoring homeostasis in a local context \cite{de2006paradoxical}.
+\par 
+Tumor cells also interact with the immune system by initiating a sequence of events that results in the activation of the adaptive immune response.
+This is accomplished by the presentation of antigens that innate immune cells recognize and carry to lymph nodes to activate T cells (and other components) \cite{schreiber11_cancer}.
+The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing the immunosuppression via Treg cell activation \cite{schreiber11_cancer}.
+\par
+Epithelial-to-mesenchymal transition (EMT) describes reversible process by which cells with an epithelial phenotype transition into cells with a mesenchymal phenotype.
+While epithelial cells are in-part defined by their adhesion to neighbors, mesenchymal cells show much greater ranges of motility, and perhaps stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.
+Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}.
+Ongoing investigations into the plasticity and stability of EMT overlap with studies of discrete vs.
+continuous processes elsewhere, such as in cell differentiation \cite{moris16_transition}.
+Moreover, intermediate states have emerged as a central mechanisms by which cell fates (and the noise inherent within in them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}
 
-\tcr{Still need to address Qing's comments on this paragraph:}
-
-The tumor interacts with the immune system by initiating the sequence of events that results in the activation of the adaptive immune response.
-It accomplishes this by presenting antigens that innate immune cells recognize and carry to lymph nodes to activate, among other things, T cells.
-Also, tumor cells engage in certain cellular processes that shape the TME and thus impact the immune system.
-In particular, tumor cells release transforming growth factor beta (TGF-$\beta$) that creates a tumor-supportive environment by shifting the immune balance to being immunosuppressive via enhancing activation of Treg cells.
+While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation and progression \cite{nieto2016emt}.
+In addition, TGF-$\beta$ is a master regulator of EMT, thus linking it closely with the signaling networks involved in immune responses [REF].
 
@@
 
-Epithelial-to-mesenchymal transition (EMT) describes the (reversible) cell plasticity process by which epithelial cells transition into mesenchymal cells.
-While the epithelial cells often stay bound together in layers, mesenchymal cells have gained motility and even stem-like properties\cite{nieto2016emt}.
-Recent work has revealed the intriguing possibility that there are intermediate cell states along the EMT spectrum with each state having its own unique properties\cite{hong2015ovol2}.
-Such intermediate states have the potential of being connected to tumor initiation, progression, and metastasis\cite{nieto2016emt}.
 While there might be many other characteristics distinctive of mesenchymal cells that need further exploration, two qualities of mesenchymal cells are particularly relevant in the context of cancer and the immune system:
@@
 2) mesenchymal cells proliferate at a slower rate. \tcr{Ref}
- %A Adam said he had a reference in mind for this.
+
+% model cancer-immune 
+%\cite{West19_JCOClinicalCancerInformatics}
 
@@
 
-\bibliography{mybib}{}
-\bibliographystyle{siam}
+\bibliography{mybib,amrefs}
+\bibliographystyle{plos2015}
 

### 2019-03-05 Paper/Paper.tex
update intro

@@
 \section*{Abstract}
-%% Basic Introduction (1-2 sentences)
 Preceding tumorigenesis and throughout the lifetime of a tumor, the immune system is engaged in complex sets of interactions in and around the tumor.
-%% More Detailed Background (2-3 sentences)
 These are regulated by the tumor microenvironment (TME) in myriad ways.
@@
 In turn, both the tumor and the immune system exert substantial influence over the TME in competition to influence the fate of the host.
+Determining the factors and mechanisms underpinning these complex interactions remains an important task for cancer immunotherapy.
+We developed a multiscale agent-based model to describe the interplay between three components: the tumor (or pre-tumor cells), the immune system, and EMT.
+The model revealed a \tcr{non-monotonic relationship} between the cell cycle times associated with the mesenchymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival.
+In comparison, increasing the immune-evasiveness of the mesenchymal phenotypes always led to poorer prognosis.
+Moreover, encouraging the TME
+towards a pro-EMT state also benefits the tumor.
+Together, these results indicate that the joint regulation of the TME by the immune system and the tumor itself provides possible targets for future cancer therapies, in particular cancers such as pancreatic cancer in which EMT is speculated to play a role.
+
+%% Basic Introduction (1-2 sentences)
+%Preceding tumorigenesis and throughout the lifetime of a tumor, the immune system is engaged in complex sets of interactions in and around the tumor.
+%% More Detailed Background (2-3 sentences)
+%These are regulated by the tumor microenvironment (TME) in myriad ways.
+Moreover,  epithelial cell plasticity impacts disease incidence and progression via processes of epithelial-to-mesenchymal transition (EMT).
+%In turn, both the tumor and the immune system exert substantial influence over the TME in competition to influence the fate of the host.
 %% General Problem (1 sentence)
-Determining the factors and mechanisms underpinning these complex interactions remains an important task for cancer immunotherapy.
+%Determining the factors and mechanisms underpinning these complex interactions remains an important task for cancer immunotherapy.
 %% Summarizing Main Result (1 sentence)
@@
 %% The Main Result (2-3 sentences)
-We developed a multiscale agent-based model to describe the interplay between three components: the tumor (or pre-tumor cells), the immune system, and EMT.
-The model revealed a \tcr{non-monotonic relationship} between the cell cycle times associated with the mesenchymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival.
+%We developed a multiscale agent-based model to describe the interplay between three components: the tumor (or pre-tumor cells), the immune system, and EMT.
+%The model revealed a \tcr{non-monotonic relationship} between the cell cycle times associated with the mesenchymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival.
 In comparison, increasing the immune-evasiveness of the mesenchymal phenotypes always led to poorer prognosis.
@@
 %\tcr
-the TME
-towards a pro-EMT state also benefits the tumor.
+%the TME
+%towards a pro-EMT state also benefits the tumor.
 %% Putting into a More General Context (1-2 sentences)
-Together, these results indicate that the joint regulation of the TME by the immune system and the tumor itself provides possible targets for future cancer therapies, in particular cancers such as pancreatic cancer in which EMT is speculated to play a role.
+%Together, these results indicate that the joint regulation of the TME by the immune system and the tumor itself provides possible targets for future cancer therapies, in particular cancers such as pancreatic cancer in which EMT is speculated to play a role.
 
@@
 The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing the immunosuppression via Treg cell activation \cite{schreiber11_cancer}.
-\par
-Epithelial-to-mesenchymal transition (EMT) describes reversible process by which cells with an epithelial phenotype transition into cells with a mesenchymal phenotype.
-While epithelial cells are in-part defined by their adhesion to neighbors, mesenchymal cells show much greater ranges of motility, and perhaps stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.
-Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}.
-Ongoing investigations into the plasticity and stability of EMT overlap with studies of discrete vs.
-continuous processes elsewhere, such as in cell differentiation \cite{moris16_transition}.
-Moreover, intermediate states have emerged as a central mechanisms by which cell fates (and the noise inherent within in them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}
 
-While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation and progression \cite{nieto2016emt}.
-In addition, TGF-$\beta$ is a master regulator of EMT, thus linking it closely with the signaling networks involved in immune responses [REF].
-
-This particular cytokine is also a driver of the epithelial-to-mesenchymal transition (EMT) and through generating TGF-$\beta$ the tumor creates a more mesenchymal phenotype within the TME.
-\tcr{I think this is where I would add which types of cancer most likely to be affected by EMT. But which? And what sources?}
-In this way, the tumor shifts the balance of local environment along the EMT spectrum.
-
+******Key aspects immune system
+%
 The immune system has several competing interactions with the tumor.
@@
 
+\par
+Epithelial-to-mesenchymal transition (EMT) describes reversible process by which cells with an epithelial phenotype transition into cells with a mesenchymal phenotype.
+While epithelial cells are in-part defined by their adhesion to neighbors, mesenchymal cells show much greater ranges of motility, and perhaps stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.
+Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}.
+Ongoing investigations into the plasticity and stability of EMT overlap with discussions elsewhere, e.g.
+of discrete vs.
+continuous processes during cell differentiation \cite{moris16_transition}.
+Intermediate states have emerged as a central mechanisms by which cell fates (and the noise inherent within in them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}.
+\par 
+While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation and progression \cite{nieto2016emt}.
+In addition, TGF-$\beta$ is a master regulator of EMT [REF], thus tying it to the signaling networks involved in tumor-mediated immune responses.
+\tcr{I think this is where I would add which types of cancer most likely to be affected by EMT. But which? And what sources?}
+In this way, the tumor shifts the balance of local environment along the EMT spectrum.
+
 While there might be many other characteristics distinctive of mesenchymal cells that need further exploration, two qualities of mesenchymal cells are particularly relevant in the context of cancer and the immune system:
@@
 
-% model cancer-immune 
+% MODELS of cancer-immune 
 %\cite{West19_JCOClinicalCancerInformatics}
+
+*******summary of MIE and MGA (move later?)
 
@@
 Regardless, the evidence currently demonstrates a ``slipperiness'' of mesenchymal cells in regards to immune lysing.
-We will refer to this as mesenchymal immune evasion, or MIE.
+We will refer to this as mesenchymal immune evasion (MIE).
 
@@
 In our model, we refer to this property as mesenchymal growth arrest, or MGA.
+
+****thisis ?
 

### 2019-03-06 Paper/Paper.tex
intro

@@
 The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing the immunosuppression via Treg cell activation \cite{schreiber11_cancer}.
-
-******Key aspects immune system
-%
-The immune system has several competing interactions with the tumor.
-While the cytotoxic immune cells, NKs and cytotoxic T cells (CTLs), seek out and lyse tumor cells, the regulatory branch of the immune system, which will be accounted for by Tregs in our model, slows down the effector functions of these other cells.
-% Make sure there's some ref for immune clearing tumor
+\par
+The effects of the immune system on a tumor can be (very coarsely) summarized into two modules.
+Cytotoxic immune cells (such as natural killer (NK) cells and cytotoxic T cells (CTLs)) seek out and lyse tumor cells, whereas the regulatory branch of the immune system (which will below be accounted for by T regulatory cells (Tregs)), slows down the \tcr{effector functions} of these other cells.
+[REF]
+%Make sure there's some ref for immune clearing tumor
 % Also, add that as immune cells clear tumor cells, they become deactivated and give source
-
-In relation to EMT, the Tregs of the immune system drive the EMT process by releasing TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
-In this way, just as the tumor cells push the TME towards a mesenchymal phenotype, so too do the Tregs.
-
+In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), particularly in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
+However, recent work raises questions regarding the relationship between inflammation and cancer \cite{guo2017multiscale}.
 \par
@@
 While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation and progression \cite{nieto2016emt}.
-In addition, TGF-$\beta$ is a master regulator of EMT [REF], thus tying it to the signaling networks involved in tumor-mediated immune responses.
+In addition, TGF-$\beta$ is a master regulator of EMT [REF], thus tying it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
+Thus, even by considering only a single signaling factor, we find these three components: the tumor, the immune system, and the EMT process to be inextricably linked.
+It thus strikes us as a priority to derive models to understand how interactions between each of these three components affect cancer incidence and progression.
+\par 
+Below we develop a multiscale model with this goal: to study interactions between the tumor, immune cells and EMT. Many
+
 \tcr{I think this is where I would add which types of cancer most likely to be affected by EMT. But which? And what sources?}

### 2019-03-08 Paper/Paper.tex
results edts

@@
-\documentclass[11pt, preprint]{article}
-\usepackage{outline}
-\usepackage{float}
-\usepackage{graphicx}
-\usepackage{color}
-\usepackage{amsmath}
-\usepackage{cite}
+\documentclass[11pt]{article}
+\usepackage{amsmath, amssymb, graphicx, outline, float, cite, color}
+
+\input{structure.tex} % Specifies the document structure and loads requires packages
+
 \graphicspath{ {Figs/} }
@@
 \affil{
-  $^1$Department of Mathematics, University of California, Irvine, and \\
+  $^1$Department of Mathematics, University of California, Irvine,  Irvine, CA 92697, USA \\
   $^2$Department of Cell and Developmental Biology, University of California, Irvine, Irvine, CA 92697, USA \\
   $^3$Department of Biological Sciences, University of Southern California, Los Angeles, CA 90089, USA \\
-  $^*$Correspondence: macleana@usc.edu (ALM); qnie@uci.edu (QN).
+  $^*$Correspondence:  qnie@uci.edu (QN); macleana@usc.edu (ALM).
 }
@@
 \renewcommand\Authands{ and }
+
+\date{}
 
@@
 \section{Introduction}
-The immune system affects tumorigenesis and cancer progression in myriad intricate ways.
-New breakthroughs in the field of immunotherapy are beginning to produce new therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}.
-Beyond direct tumor-immune cell interactions, the immune system also modulates the tumor microenvironment (TME), and indeed, immune cells can be considered as a part of the TME: responding to local inflammatory signals, targeting mutated cells, and potentially eradicating tumor cells and restoring homeostasis in a local context \cite{de2006paradoxical}.
+Cancer and the immune system interact in myriad intricate ways.
+Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME), indeed, immune cells can be considered as a part of the TME: responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}.
+Recent breakthroughs in the field of immunotherapy are beginning to produce therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}.
 \par 
-Tumor cells also interact with the immune system by initiating a sequence of events that results in the activation of the adaptive immune response.
-This is accomplished by the presentation of antigens that innate immune cells recognize and carry to lymph nodes to activate T cells (and other components) \cite{schreiber11_cancer}.
-The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing the immunosuppression via Treg cell activation \cite{schreiber11_cancer}.
+Tumor cells interact with the immune system by initiating a sequence of events that results in the activation of the adaptive immune response.
+This is accomplished by the presentation of antigens recognized by innate immune cells that are transported  to lymph nodes where T cells (and other components) can be activated \cite{schreiber11_cancer}.
+The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via Treg cell activation \cite{schreiber11_cancer}.
 \par
-The effects of the immune system on a tumor can be (very coarsely) summarized into two modules.
-Cytotoxic immune cells (such as natural killer (NK) cells and cytotoxic T cells (CTLs)) seek out and lyse tumor cells, whereas the regulatory branch of the immune system (which will below be accounted for by T regulatory cells (Tregs)), slows down the \tcr{effector functions} of these other cells.
+The effects of the immune system on a tumor can be broadly summarized into two modules.
+The cytotoxic branch of the immune system (such as natural killer (NK) cells and cytotoxic T cells (CTLs)) seek out and lyse tumor cells.
+The regulatory branch of the immune system (T regulatory cells (Tregs) and other factors), inhibits the effective functioning of the cytotoxic branch.
 [REF]
@@
 In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), particularly in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
-However, recent work raises questions regarding the relationship between inflammation and cancer \cite{guo2017multiscale}.
+Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo2017multiscale}.
 \par
-Epithelial-to-mesenchymal transition (EMT) describes reversible process by which cells with an epithelial phenotype transition into cells with a mesenchymal phenotype.
-While epithelial cells are in-part defined by their adhesion to neighbors, mesenchymal cells show much greater ranges of motility, and perhaps stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.
+Epithelial-to-mesenchymal transition (EMT) describes a reversible process by which cells with an epithelial phenotype transition into cells with a mesenchymal phenotype.
+While epithelial cells are in part defined by their adhesion to neighbors, mesenchymal cells show much greater ranges of motility, and may display stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.
 Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}.
@@
 continuous processes during cell differentiation \cite{moris16_transition}.
-Intermediate states have emerged as a central mechanisms by which cell fates (and the noise inherent within in them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}.
+Intermediate states have emerged as a central mechanism by which cell fates (and the noise inherent within in them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}.
 \par 
 While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation and progression \cite{nieto2016emt}.
-In addition, TGF-$\beta$ is a master regulator of EMT [REF], thus tying it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
-Thus, even by considering only a single signaling factor, we find these three components: the tumor, the immune system, and the EMT process to be inextricably linked.
-It thus strikes us as a priority to derive models to understand how interactions between each of these three components affect cancer incidence and progression.
+TGF-$\beta$ is a master regulator of EMT [REF], linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
+Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
+It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.
 \par
-Below we develop a multiscale model with this goal: to study interactions between the tumor, immune cells and EMT. Many
-
-\tcr{I think this is where I would add which types of cancer most likely to be affected by EMT. But which? And what sources?}
-In this way, the tumor shifts the balance of local environment along the EMT spectrum.
-
-While there might be many other characteristics distinctive of mesenchymal cells that need further exploration, two qualities of mesenchymal cells are particularly relevant in the context of cancer and the immune system:
-1) mesenchymal cells are less susceptible to immune clearance\cite{terry2017new} and
-%
-2) mesenchymal cells proliferate at a slower rate. \tcr{Ref}
-
-% MODELS of cancer-immune 
-%\cite{West19_JCOClinicalCancerInformatics}
-
-*******summary of MIE and MGA (move later?)
-
-The immune evasiveness of mesenchymal cells is well-established with even a functional cause for this phenomenon.
-As a cell is being targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be created in order to precipitate the lysing event.
-This immunological synapse, however, is dependent upon the surface proteins of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form\cite{terry2017new}.
-There are also other structural changes that take place in mesenchymal cells that could contribute to this refractory nature towards immune clearance.
-Regardless, the evidence currently demonstrates a ``slipperiness'' of mesenchymal cells in regards to immune lysing.
-We will refer to this as mesenchymal immune evasion (MIE).
-
+Two features of the mesenchymal phenotype are of particular relevance in the context of cancer and the immune system: i) mesenchymal cells are less susceptible to immune clearance \cite{terry2017new}; and ii) mesenchymal cells proliferate on average at a slower rate than epithelial cells.
+The immune evasiveness of mesenchymal cells is well-established: as a cell is targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be established before the lysing event.
+This immunological synapse is dependent upon the surface proteins [WHICH??] of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form \cite{terry2017new}.
+%There are also other structural changes that take place in mesenchymal cells that could contribute to this refractory nature towards immune clearance.
+%Regardless, the evidence currently demonstrates a ``slipperiness'' of mesenchymal cells in regards to immune lysing.
+Below we refer to this phenotype as mesenchymal immune evasion (MIE).
+\par
 The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
@@
 In our model, we refer to this property as mesenchymal growth arrest, or MGA.
-
-****thisis ?
-
-Recent work on cancer-immune interactions has shown that the inflammatory conditions of the TME can play large and sometimes surprising roles in tumor growth.
-For example, a mathematical model was developed to couple the inflammatory state of the TME with the cell cycles of the stem cells by increasing proliferation, decreasing apoptosis, and enhancing DNA damage that leads to mutations\cite{guo2017multiscale}.
-The work shows that while inflammation does have an oncogenic effect, it can also have an onco-protective role as periods of mild inflammation are better at preventing cancer than periods of no inflammation in patients that have intermittent bouts of severe inflammation.
-However, the aspect of the model that included the immune system is minimal and the connection with EMT is not included.
-%However, the aspect of the model that included the immune system was minimal: regardless of the inflammatory environment, any cells with sufficient pathway mutations (2 in their model) had a fixed probability of being cleared by the metabolism-immune balance response.
-Here, we include a more robust immune module and explore how this immune module will improve our understanding of inflammatory effects on the tumor, and also include the epithelial-mesenchymal axis.
-%In addition, we explore an extra component of Guo's stem cells: where they lie on the epithelial-mesenchymal axis.
-Such additions will allow us to explore the interactions of the tumor, the immune system, and the epithelial-mesenchymal spectrum as well as the effects of the TME on these interactions.
-%We will also study the effects of the TME on these interactions.
-
-Next, we introduce the model modularly, which is appropriate given its multiscale and hybrid nature.
-We then perform sensitivity analysis and identify parameters that are crucial for the progression to tumorigenesis.
-We analyze the competing effects of EMT and of the immune system on tumorigenesis and find that EMT intricately regulates tumorigenesis: under certain regimes a careful balance of EMT- and immune-driven process can significantly prolong cancer-free survival.
+\par
+Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including older \cite{sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{an15_agentbased, serre16_mathematical, louzoun14_mathematical, benzekry17_mathematical, west19_immune, owen11_mathematical-1, guo2017multiscale}.
+These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximise the efficacy of cytotoxicity while minimizing risks to the patient.
+However, to our knowledge no models have addressed how the effects of EMT alter interactions between the immune system and cancer, and the subsequent implications for treatment.
+\par
+Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
+We seek to describe a set of crucial molecular and cellular interactions in (epithelial) tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will occur and, if so, when.
+A recent model of cancer-immune interactions \cite{guo2017multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as the EMT component to the model.
+We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
+In the next section we develop the model modularly to explain the intuition behind each component, and then go on to analyze its behavior.
+Global sensitivity analysis identifies parameters that are crucial for tumorigenesis.
+We then study these in more depth, focusing on the competing effects of EMT and of the immune system on tumorigenesis: we find that EMT intricately regulates tumorigenesis and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
+To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects predicted by the model for patients with pancreatic cancer.
 
@@
 \section{Methods}
-To address the questions around the relationships among cancer, the immune system, and EMT, we extended the multiscale, agent-based model \tcr{REF} to include an immune module and an EMT axis for the tissue cells.
+We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo2017multiscale}. 
 The agents in the model are the tissue cells with tumorigenic potential.
-They can either be mutation-free, or have any combination of three possible pathway mutations.
-These three pathways are the three components within each cell in Figure \ref{fig:ModelIntro}A.
-If the pathway is mutated, the component is colored red and has an `X'.
-A single mutation is sufficient for the cell being counted as a mutant cell.
+These can either be mutation-free, or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).
+If the pathway is mutated, the component is colored red and is marked with `X'.
+A single mutation is sufficient for the cell to be considered as a mutant cell.
 
-We model the immune cells as discrete variables but as one collective, in other words they are not agents themselves.
-The cytokine TGF-$\beta$ is a continuous variables within the tumor microenvironment (TME).
-It is worth noting that there is no spatial aspect to this model.
-In doing this, we are assuming that cells are well-mixed as are all the relevant cytokines and other molecules acting on the system.
-Finally, tissue cells can be labeled either epithelial or mesenchymal, and this labeling can change over time and depends both on the TME and the internal workings of each individual cell.
+We model immune cells as continuous variables, appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.
+The cytokine TGF-$\beta$ is continuous within the tumor microenvironment (TME), i.e. the system is well-mixed.
+Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both on the TME and cell-intrinsic factors.
 While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
@@
 \paragraph{Proliferation}
+
+\tcr{In next few subsections the parameters are not defined.
+Need to define in text and/or make a big table to summarize all.
+On notation: some changes may be needed, e.g.
+$\delta_{MES}$ not very clear.}
+
 There are four factors that contribute to the probability that a cell will proliferate.
@@
 
-\subsubsection{Progression and tumorigenesis}
+\subsubsection{Mutation and tumorigenesis}
 At the end of each cell cycle, the proportion of tissue cells that are mutated is calculated, and if it is above a certain threshold, a tumorigenic event is recorded and the patient is determined to have cancer.
@@
 The variances were chosen as twice the mean to get a large spread of values.
+
 %%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\begin{figure}[H]
+\subsection{A multiscale agent-based model of EMT-immune-tissue cell interactions to study tumorigenesis}\label{ExplModel}
+
+\begin{figure}
 \center
-\frame{\includegraphics[width=\columnwidth]{Figure1/Figure1.pdf}}
+{\includegraphics[width=0.8\textwidth]{Figure1/Figure1.pdf}}
 \caption{A.
@@
 
-\begin{figure}[H]
-\center
-\frame{\includegraphics[width=\columnwidth]{Figure2/Figure2.jpg}}
-\caption{Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
-\label{fig:MOAT}
-\end{figure}
+We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the probability of cancer, and the Time to Cancer.
+Within the cell cycle, cell fate is determined via a set of rules that are influenced by EMT and immune interactions (Fig.
+\ref{fig:ModelIntro}A).
+For example, if a cell undergoes EMT, the probability that is will proliferate is reduced; if a cell gains a mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
+The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset).
+NK cells and CTLs attempt to clear mutated tissue cells, and deactivate upon successfully carrying out their cytotoxic function.
+T regulatory cells (Tregs) inhibit cytotoxic activity.
+Tregs also release TGF-$\beta$ which promotes the recruitment of Tregs, as well as increasing the rate of EMT.
+\par
+The inflammation cycling scheme for a typical {\it in silico} patient consists of alternating high and low regimes (Fig.
+\ref{fig:ModelIntro}B); the inflammation schemes modeled will be discussed in detail in the next section.
+For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor.
+However, after approximately 1000 cell cycles (750 days), one mutated cell exhibits clonal growth, and a tumor is established.
+At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being heavily recruited, and a peak in TGF-$\beta$ expression is observed.
+The number of mesenchymal cells increases initially with increasing TGF-$\beta$, but decreases after the clone takes over the tissue.
+This shifts back towards an epithelial tissue phenotype occurs even though the expression of TGF-$\beta$ remains high.
+After 841 cell cycles, the proportion of mutated cells reaches 50\%: this is modeled as the threshold defining tumorigenesis. Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days. 
+After cancer incidence, the number of mutated cells continues to grow rapidly and soon makes up 100\% of the cell population.
+A peak in mesenchymal phenotypes is observed following cancer incidence, before most transition back to an epithelial state.
+\par 
+Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of oncogenic mutations.
+%Note: NK cells also lymphocytes! 
+The adaptive immune populations also exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
+Within each of these periods, adaptive immune populations increase or decrease rapidly in accordance with the inflammation state.
+\par 
+The {\em in silico} patient described here is given for the purpose of illustrating features the model.
+Given the multiple sources of noise in the model, in order to quantify patient dynamics and cancer-free survival rates, below we will simulate large cohorts of patients.
+In Fig.
+\ref{fig:ModelIntro}C we simulate the survival curve for a cohort of $500$ patients: we see that all the patients survive for approximately 100 cell cycles (75 days).
+Subsequently, in the approximate range of $T= [100,300]$, the cancer onset rate is roughly constant, and after $600$ cell cycles, no patients remain cancer free.
 
-\subsection{A multiscale agent-based model of EMT-immune-tissue cell interactions and tumorigenesis}\label{ExplModel}
-We begin by studying the model under baseline conditions to assess how its various components interact and effect the overal probability of tumorigenesis (Time to Cancer).
-In Figure \ref{fig:ModelIntro}A a schematic description of the model is shown.
-Within a single cell cycle, different choices are available to a cell and these are influenced by pathway mutations or EMT.
-For example, if a cell undergoes EMT, the probability that is will proliferate is reduced.
-Also shown are the different means by which the immune system acts on the tissue cells (\ref{fig:ModelIntro}A Inset).
-The NKs and CTLs attempt to clear out mutated tissue cells and deactivate upon successfully carrying out this cytotoxic function.
-The Tregs inhibit this cytotoxic activity.
-In addition, Tregs release TGF-$\beta$ which helps recruit more Tregs and also drives EMT.
-%A \tcr{TGF-$\beta$ is [describe briefly action of each immune cell type]}  --Adam: Is this good?
-
-For a typical {\it in silico} patient (see Figure \ref{fig:ModelIntro}B), after the warmup period, some cells mutate but are cleared before establishing a tumor.
-However, eventually one mutated cell is able to establish a growing tumor before being cleared.
-This happens around cell cycle 1000.
-At this point, the adaptive immune system (CTLs and Tregs) are being heavily recruited and TGF-$\beta$ is peaking.
-The number of mesenchymal cells grows with the increased TGF-$\beta$, but decreases rapidly after the tissue cells are entirely mutants.
-Interestingly, this return to a more epithelial phenotype occurs even as TGF-$\beta$ reaches its maximum value for the simulation.
-%\tcr{[What happens??]}
-The inflammation cycling scheme for this patient is shown at the top of Figure \ref{fig:ModelIntro}B and is repeated throughout, including during the warmup.
-We will return to this shortly.
-
-%\tcr{[What is?]} 
-At cell cycle number 841 the proportion of mutated cells reaches 50\% and so the model marks the patient as having a Time to Cancer of 841, or 630.75 days. 
-At the same time as mutated cells are taking over, many cells transition to a mesenchymal phenotype.
-After the Time to Cancer, the mutated cells continue their rapid growth and soon make up 100\% of the cells.
-Meanwhile, all the cells become mesenchymal for a short duration before most transition back to an epithelial state.
-
-Considering the immune populations, we see that the NK population is approximately constant, while the lymphocyte populations grow quickly and dwarf the NK population following the accumulation of mutations.
-The lymphocyte populations exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that this patient is undergoing: 
-the patient alternates between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
-In each of these periods periods, the lymphocyte populations increase or decrease rapidly in accordance with the inflammation.
-
-We next consider a cohort of $N=500$ patients, and simulate the survival curve for these patients (Figure \ref{fig:ModelIntro}C).
-We see that all patients survive for approximately 100 cell cycles (75 days), before a time period where the cancer onset rate is roughly constant, between the approximate range of ($T = 100, T = 300$).
-
-\subsection{Identification of key model parameters via Morris global sensitivity analysis}\label{SensAnalysis}
-To assess the sensitivity responses \tcr{Is this the right phrasing?
-It sounds funny to me, but I see Adam chose this wording.} produced by the model with respect to the 31 model parameters, a Morris one-step-at-a-time (OAT) algorithm was implemented.
-The results of Morris OAT sensitivity analysis are shown in Figure \ref{fig:MOAT}. A subset of parameters demonstrate much higher levels of sensitivity than others.
+\subsection{Identification of key regulatory parameters via Morris global sensitivity analysis}\label{SensAnalysis}
+To assess the sensitivity of the model parameters and thus identify those that are most important in determining cancer-free survival times, we applied Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods) to the model.
+% Sentence on how it is 'global' and sentence on what the measured output is (ie y axis) 
+...
+The results of Morris OAT on the 31 model parameters are shown in \ref{fig:MOAT}: a subset of parameters demonstrate much higher levels of sensitivity than others.
 The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
 This ties Treg cells to all three components of the model.
-Since below we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for further analysis below.
-
-Any of the parameters in the table which include the word ``low'' are parameters that govern the model in a low inflammatory state.
+Since we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for detailed analysis below.
+\par
+%% We need to discuss parameter names. Maybe replace 'low' with 'base' and 'up' with 'gain' ?
+Any of the parameters in the \tcr{table} which include the word ``low'' are parameters that govern the model in a low inflammatory state.
 With the exception of ``prolif up'', all parameters that end in ``up'' or have `U' at the end of their abbreviation, are the factors by which these parameters change when the system switches from low to high inflammation.
@@
 
-We want to better isolate the effects of EMT on these dynamics, so parameters such as MIE and MGA are of interest.
+Since one goal of our analysis is to assess the specific effects of EMT on immune-cancer dynamics, the parameters MIE and MGA are of particular interest.
 In addition, inflammation parameters dictating the cycling scheme are important to study\tcr{REF}: we find that there are also high sensitivities in this model to these parameters.
@@
 
-\subsection{Mesenchymal phenotypic properties dramatically change survival outcomes}\label{MesPars}
-The two quantities that change when a cell transitions to a mesenchymal state are its immune evasion and its proliferation rate, MIE and MGA, respectively.
-Both parameters are proportions between 0 and 1 with higher values indicating more extreme expression of mesenchymal properties.
-MIE is the proportional reduction of the probability a mutated mesenchymal cell will undergo immune clearance in a given cycle.
-MGA, which stands for mesenchymal growth arrest, is the proportional reduction in any mesenchymal cell of its chance to proliferate in a given cell cycle. 
-The reduction in proliferation probability corresponds with an equal increase in probability for resting during that cell cycle.  
-%\tcr{[briefly redefine what these terms do]}
-Also involved in the EMT process, is the cytokine TGF-$\beta$, which upregulates EMT.
-By the Morris OAT analysis presented in Section \ref{SensAnalysis}, these three parameters do have large impacts on survival outcomes and each warrant exploration.
+\begin{figure}
+\center
+{\includegraphics[width=0.85\textwidth]{Figure2/Figure2.jpg}}
+\caption{Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
+\label{fig:MOAT}
+\end{figure}
 
-We found that as the level of mesenchymal immune evasion (MIE) increases, Time to Cancer decreases ( Figure \ref{fig:FirstSurvivalCurves}A).
-This result holds for all parameter sets studied, and indicates the simple relationship that mesenchymal immune evasion has with Time to Cancer.
-As this subpopulation of the tumor becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
-These results are summarized by Figure \ref{fig:FirstSurvivalCurves}A.
+\subsection{Mesenchymal phenotypic properties dramatically change cancer-free survival times}\label{MesPars}
 
-\begin{figure}[H]
+\begin{figure}
 \center
-\frame{\includegraphics[width=\columnwidth]{Figure3/Figure3.pdf}}
+{\includegraphics[width=0.85\textwidth]{Figure3/Figure3.pdf}}
 \caption{
-A-C. Trajectory of one patient per cohort followed from warmup through 2000 cell cycles.
-\tcr{OK to do A-C like this?}
+A-C. Trajectory of one patient per cohort followed from warmup through 2000 cell cycles while varying MIE (A), MGA (B) or XXXXXX (C). 
 D. Increasing MIE decreases Time to Cancer. 
@@
 F. Increasing Treg TGF-$\beta$ production results in decreased Time to Cancer.
-%% {\color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.}
+%% \color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.
 }
@@
 \end{figure}
+
+When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
+Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
+MIE is the proportional reduction in the probability that a mutated mesenchymal cell will be subject to immune clearance in a given cycle.
+MGA is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
+The cytokine TGF-$\beta$ is also involved in the EMT process (by increases the probability of EMT).
+All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
+\par 
+As the level of mesenchymal immune evasion (MIE) is increased, the Time to Cancer decreases (Fig.
+\ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of potentially mutated cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
+These results are summarized by Figure \ref{fig:FirstSurvivalCurves}A.
 
@@
 \subsection{A key EMT regime maximizes cancer-free survival time under chronic inflammation}\label{KeyEMT}
+
 Given the relevance of the inflammatory state of the tumor microenvironment (TME), we explore the effect of varying the inflammation state of the patient on survival.
@@
 
-\begin{figure}[H]
+Contrast this with what happens when MIE is varied for different inflammation cycling schemes.
+Regardless of the cycling scheme, the relationship between Time to Cancer and MIE is monotonic.
+In the case that the inflammatory state is permanently high, MIE has no significant effect on Time to Cancer.
+On the other hand, when the patient experiences at least some time in a low inflammatory state, an increase in MIE results in a decrease in Time to Cancer.
+This indicates that mitigating a chronic inflammatory state is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
+
+\begin{figure}
 \center
-\frame{\includegraphics[width=\columnwidth]{Figure4/Figure4.pdf}}
+{\includegraphics[width=0.9\textwidth]{Figure4/Figure4.pdf}}
 \caption{A.
@@
 
-Contrast this with what happens when MIE is varied for different inflammation cycling schemes.
-Regardless of the cycling scheme, the relationship between Time to Cancer and MIE is monotonic.
-In the case that the inflammatory state is permanently high, MIE has no significant effect on Time to Cancer.
-On the other hand, when the patient experiences at least some time in a low inflammatory state, an increase in MIE results in a decrease in Time to Cancer.
-This indicates that mitigating a chronic inflammatory state is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
+\begin{figure}
+\center
+{\includegraphics[width=0.6\textwidth]{Figure5/MIEvsMQ.jpg}}
+\caption{Heat Map contrasting the effects of MIE and MGA on Time to Cancer. Increasing MIE always decreases Time to Cancer, but MGA is non-monotonically linked to Time to Cancer.}
+\label{fig:MIEvsMGA}
+\end{figure}
 
@@
 
-\begin{figure}[H]
-\center
-\frame{\includegraphics[width=\columnwidth]{Figure5/MIEvsMQ.jpg}}
-\caption{Heat Map contrasting the effects of MIE and MGA on Time to Cancer. Increasing MIE always decreases Time to Cancer, but MGA is non-monotonically linked to Time to Cancer.}
-\label{fig:MIEvsMGA}
-\end{figure}
-
 To compare these model predictions with experimental studies, we performed analysis of data from The Cancer Genome Atlas (TCGA) database to study the effects of immune and EMT interactions on prognosis of cancers for which inflammation is known to play an important role, such as colonic or pancreatic cancers (REFS).
@@
 
-\begin{figure}[H]
+\begin{figure}
 \center
-\frame{\includegraphics[width=\columnwidth]{FigTCGA.pdf}}
+{\includegraphics[width=\columnwidth]{FigTCGA.pdf}}
 \caption{A.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%
-%                     CONCLUSION                      %
-%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Conclusion}\label{Conclusion}
+%% AM: no need to separate this I think - combine conclusions into the discussion section. 
+%\section{Conclusion}\label{Conclusion}
 The interactions of the immune system and cancer remain a promising avenue of research.
@@
 
+\section*{Acknowledgements}
+
 \bibliography{mybib,amrefs}

### 2019-03-08 Paper/structure.tex
results edts

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Wenneker Article
% Structure Specification File
% Version 1.0 (28/2/17)
%
% This file originates from:
% http://www.LaTeXTemplates.com
%
% Authors:
% Frits Wenneker
% Vel (vel@LaTeXTemplates.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\usepackage[english]{babel} % English language hyphenation

\usepackage{microtype} % Better typography

\usepackage{amsmath,amsfonts,amsthm} % Math packages for equations

\usepackage[svgnames]{xcolor} % Enabling colors by their 'svgnames'

\usepackage[hang, small, labelfont=bf, up, textfont=it]{caption} % Custom captions under/above tables and figures

\usepackage{booktabs} % Horizontal rules in tables

\usepackage{lastpage} % Used to determine the number of pages in the document (for "Page X of Total")

\usepackage{graphicx} % Required for adding images

\usepackage{enumitem} % Required for customising lists
\setlist{noitemsep} % Remove spacing between bullet/numbered list elements

\usepackage{sectsty} % Enables custom section titles
\allsectionsfont{\usefont{OT1}{phv}{b}{n}} % Change the font of all section commands (Helvetica)

%----------------------------------------------------------------------------------------
%	MARGINS AND SPACING
%----------------------------------------------------------------------------------------

\usepackage{geometry} % Required for adjusting page dimensions

\geometry{
	top=2cm, % Top margin
	bottom=2cm, % Bottom margin
	left=3cm, % Left margin
	right=3cm, % Right margin
	includehead, % Include space for a header
	includefoot, % Include space for a footer
	%showframe, % Uncomment to show how the type block is set on the page
}

\setlength{\columnsep}{7mm} % Column separation width

%----------------------------------------------------------------------------------------
%	FONTS
%----------------------------------------------------------------------------------------

\usepackage[T1]{fontenc} % Output font encoding for international characters
\usepackage[utf8]{inputenc} % Required for inputting international characters

\usepackage{XCharter} % Use the XCharter font

%----------------------------------------------------------------------------------------
%	HEADERS AND FOOTERS
%----------------------------------------------------------------------------------------

\usepackage{fancyhdr} % Needed to define custom headers/footers
\pagestyle{fancy} % Enables the custom headers/footers

\renewcommand{\headrulewidth}{0.0pt} % No header rule
\renewcommand{\footrulewidth}{0.4pt} % Thin footer rule

\renewcommand{\sectionmark}[1]{\markboth{#1}{}} % Removes the section number from the header when \leftmark is used

%\nouppercase\leftmark % Add this to one of the lines below if you want a section title in the header/footer

% Headers
\lhead{} % Left header
\chead{\textit{\thetitle}} % Center header - currently printing the article title
\rhead{} % Right header

% Footers
\lfoot{} % Left footer
\cfoot{} % Center footer
\rfoot{\footnotesize Page \thepage\ of \pageref{LastPage}} % Right footer, "Page 1 of 2"

\fancypagestyle{firstpage}{ % Page style for the first page with the title
	\fancyhf{}
	\renewcommand{\footrulewidth}{0pt} % Suppress footer rule
}

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\authorstyle}[1]{{\large\usefont{OT1}{phv}{b}{n}\color{DarkRed}#1}} % Authors style (Helvetica)

\newcommand{\institution}[1]{{\footnotesize\usefont{OT1}{phv}{m}{sl}\color{Black}#1}} % Institutions style (Helvetica)

\usepackage{titling} % Allows custom title configuration

%\newcommand{\HorRule}{\color{DarkGrey}\rule{\linewidth}{1pt}} % Defines the gold horizontal rule
\newcommand{\HorRule}{\color{DarkGoldenrod}\rule{\linewidth}{1pt}} % Defines the gold horizontal rule

\pretitle{
	\vspace{-30pt} % Move the entire title section up
	\HorRule\vspace{10pt} % Horizontal rule before the title
	\fontsize{20}{18}\usefont{OT1}{phv}{b}{n}\selectfont % Helvetica
	\color{DarkRed} % Text colour for the title and author(s)
}

\posttitle{\par\vskip 15pt} % Whitespace under the title

\preauthor{} % Anything that will appear before \author is printed

\postauthor{ % Anything that will appear after \author is printed
	\vspace{10pt} % Space before the rule
	\par\HorRule % Horizontal rule after the title
	\vspace{20pt} % Space after the title section
}

%----------------------------------------------------------------------------------------
%	ABSTRACT
%----------------------------------------------------------------------------------------

\usepackage{lettrine} % Package to accentuate the first letter of the text (lettrine)
\usepackage{fix-cm}	% Fixes the height of the lettrine

\newcommand{\initial}[1]{ % Defines the command and style for the lettrine
	\lettrine[lines=3,findent=4pt,nindent=0pt]{% Lettrine takes up 3 lines, the text to the right of it is indented 4pt and further indenting of lines 2+ is stopped
		%\color{DarkRed}% Lettrine colour
		\color{DarkGoldenrod}% Lettrine colour
		{#1}% The letter
	}{}%
}

\usepackage{xstring} % Required for string manipulation

\newcommand{\lettrineabstract}[1]{
	\StrLeft{#1}{1}[\firstletter] % Capture the first letter of the abstract for the lettrine
	\initial{\firstletter}\textbf{\StrGobbleLeft{#1}{1}} % Print the abstract with the first letter as a lettrine and the rest in bold
}

%----------------------------------------------------------------------------------------
%	BIBLIOGRAPHY
%----------------------------------------------------------------------------------------

%\usepackage[backend=bibtex,style=authoryear,natbib=true]{biblatex} % Use the bibtex backend with the authoryear citation style (which resembles APA)

%\addbibresource{example.bib} % The filename of the bibliography

%\usepackage[autostyle=true]{csquotes} % Required to generate language-dependent quotes in the bibliography

### 2019-03-08 Paper/Paper.tex
update refs

@@
 \par
-Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including older \cite{sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{an15_agentbased, serre16_mathematical, louzoun14_mathematical, benzekry17_mathematical, west19_immune, owen11_mathematical-1, guo2017multiscale}.
+Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including older \cite{sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{an15_agentbased, serre16_mathematical, louzoun14_mathematical, benzekry17_mathematical, west19_immune, owen11_mathematical-1}.
 These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximise the efficacy of cytotoxicity while minimizing risks to the patient.

### 2019-03-09 Paper/Paper.tex
update results on TCGA

@@
 
+\subsection{Analysis of patient survival data from TCGA database}
+We obtain primary tumor bulk mRNA sequencing and censored survival data for all individuals monitored in the relevant project from the Cancer Genome Atlas (TCGA) from the Genomic Data Commons portal in R, using the package TCGABiolinks.
+Given $n >= 1$ gene set keywords (e.g.
+"inflammatory" and "emt"), the symbolic names of all msigdb gene sets are searched for matches to these keywords, and matched gene sets are grouped by keyword.
+For each element in the product S of all keyword groups, the following analysis is performed.
+\begin{enumerate}
+     \item The first principle component of the geneset-wide expression is obtained for each element of the n-tuple of gene sets.
+     \item Two clusters of n-dimensional patient vectors are obtained by k-means clustering.
+     \item The patients are separated by their cluster identity and Kaplan-Meier curves are fit to their corresponding survival data.
+     \item Under the null hypothesis that there is no difference between the survival of the two groups, a log-rank test is performed.
+     \item The log-rank test p-value for each element in S is placed in an ordered list, the lowest of which defines the element of S whose composite gene sets are most predictive of patient survival in the given TCGA project or tumor type.
+\end{enumerate}
+Below we analyze data gathered for pancreatic cancer, denoted by ``PAAD'' in TCGA.
+
 %%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Given the relevance of the inflammatory state of the tumor microenvironment (TME), we explore the effect of varying the inflammation state of the patient on survival.
-In simulations, we assign some cohorts permanently low inflammation levels, some permanently high levels, and some variable inflammatory schemes.
-For those cohorts with a permanently high inflammatory state, the relationship between the mesenchymal parameters and the Time to Cancer is monotonic.
-However, when there are periods of a low inflammatory state, then intriguingly the connection between MGA and Time to Cancer becomes concave down with a local maximum around 0.3.
-This can be seen in the top two groups of Figure \ref{fig:VaryINFL_and_MesPars}D: the MGA=0.3 cohorts both took longer to cancer than their neighbors.
-What this indicates is that if the growth arrest of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer.
-Even if the patient is presenting a chronically high inflammatory state, anti-inflammatories could be administered which would create the periods of low inflammation our model predicts as necessary for seeing these changes.
-Reducing the inflammatory state of the TME three weeks out of every nine would be sufficient to make the control of mesenchymal proliferation an effective anti-cancer strategy.
-%Furthermore, even if the patient is presenting a permanent high inflammatory state, medications could be prescribed which would lower the inflammatory state temporarily and thus create a similar cycling inflammatory regime which could be taken advantage of in the same way.
-\tcr{what does this mean?
-how?}
-
-Contrast this with what happens when MIE is varied for different inflammation cycling schemes.
-Regardless of the cycling scheme, the relationship between Time to Cancer and MIE is monotonic.
-In the case that the inflammatory state is permanently high, MIE has no significant effect on Time to Cancer.
-On the other hand, when the patient experiences at least some time in a low inflammatory state, an increase in MIE results in a decrease in Time to Cancer.
-This indicates that mitigating a chronic inflammatory state is a necessity for increasing survival, and once that has been achieved, even to a small degree, then therapies which reduce mesenchymal immune evasion will gain efficacy.
+We explored the effects of varying the inflammation state of the patient on cancer-free survival, to investigate competing interactions within the TME and their effect on EMT.
+Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable inflammatory.
+For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and the Time to Cancer is monotonic, i.e.
+increasing either the rate of mesenchymal immune evasion (MIE) (Fig.
+\ref{fig:VaryINFL_and_MesPars}A-B) or of mesenchymal growth arrest (MGA)  (Fig.
+\ref{fig:VaryINFL_and_MesPars}C-D) decreases the Time to Cancer.
+However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the Time to Cancer is found with respect to the relative mesenchymal growth rate (MGA $= 0.3$).
+This is seen both for intermittent or low inflammation (Fig.
+\ref{fig:VaryINFL_and_MesPars}D).
+\par
+These striking differences in the mean Time to Cancer -- extended by up to one year by fine-grained optimization of mesenchymal growth arrest -- have clear therapeutic implications. 
+%What this indicates is that if the growth arrest of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer.
+%For patients presenting with chronically high inflammatory state, anti-inflammatories could be administered which would create the periods of low inflammation our model predicts as necessary for seeing these changes.
+Our model predicts that reducing the inflammatory state of the TME in three weeks out of every nine would be provide means to prolong cancer-free survival in combination with mesenchymal cell therapies to make the control of mesenchymal proliferation an effective anti-cancer strategy.
+\par
+In contrast, when MIE is varied for different inflammation cycling schemes, regardless of the cycling scheme, increasing the MIE leads to decreases in the Time to Cancer (worse survival probabilities), although in the case where the inflammatory state is permanently high, MIE has little effect on the Time to Cancer.
+Thus under any inflammation regime with periods of low inflammation, any reductions in the rate of mesenchymal immune evasion will lead to improved patient outcomes.
 
@@
 
-To further analyze the effects on tumorigenesis of the mesenchymal properties of immune evasion (MIE) and growth arrest (MGA), we study the joint density plot detailing the effects of each of these parameters on the Time to Cancer (Fig.
+To further assess the effects on tumorigenesis of the mesenchymal properties of immune evasion (MIE) and growth arrest (MGA), we study the joint density plot for each of these parameters against the Time to Cancer (Fig.
 \ref{fig:MIEvsMGA}).
@@
 However, for any given value of mesenchymal immune evasion, there is a value of mesenchymal growth arrest that maximizes the Time to Cancer.
-Moreover, this optimal value increases with MIE.
-
-To compare these model predictions with experimental studies, we performed analysis of data from The Cancer Genome Atlas (TCGA) database to study the effects of immune and EMT interactions on prognosis of cancers for which inflammation is known to play an important role, such as colonic or pancreatic cancers (REFS).
-We selected pancreatic cancer to investigate, and used gene ontologies as a measure of the of effects of inflammatory or EMT processes on survival.
-In Fig.
-\ref{fig:tcga} we plot survival curves for two cases: either clustering only on EMT, or on both EMT and inflammation signatures.
-We see that in both cases, survival is significantly affected by the gene ontology signature, but importantly, the presence of both EMT and inflammation signatures has a greater impact on survival than the effects of EMT alone.
+Moreover, this optimal value increases with increasing MIE.
+Together these results show that while reducing the ability of a mutated cell to evade the immune system by any amount will improve the probability of cancer-free survival, an optimal value of MGA will prolong survival.
+\par 
+To compare these model predictions with experimental studies, we analyzed data from The Cancer Genome Atlas (TCGA) database to study the effects of immune and EMT interactions on prognosis of cancers for which inflammation is known to play an important role, such as colonic or pancreatic cancers \cite{hu10_inflammationinduced, balkwill01_inflammation}.
+Using gene ontologies as a measure of the of effects of inflammatory or EMT processes on cancer-free survival, we investigate pancreatic cancer.
+We note that the comparison between simulation and data here is indirect since the model studies processes occurring before incidence, while the data address subsequent events, but analogous processes are at play during both the pre-cancerous clonal dynamics addressed by the model, and the progression of disease that follows cancer incidence.
+\par
+For a cohort of pancreatic cancer patients, we cluster the patients via k-means ($n=2$) either on gene ontologies relating only to EMT (Fig.
+\ref{fig:tcga}A) or on gene ontologies relating to EMT and inflammation (Fig.
+\ref{fig:tcga}C).
+We also plot the corresponding survival curves for each of the clusters obtained (Fig.
+\ref{fig:tcga}B, D).
+We see that in both cases, survival is significantly affected by the gene ontology signature, but, importantly, the presence of both EMT and inflammatory signatures has a greater impact on survival than the effects of EMT alone.
+This suggests that
 

### 2019-03-18 Paper/Paper.tex

@@
 
-% A multiscale model of EMT-mediated cancer reveals 
-\title{Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer}
+% Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer
+\title{
+The competing effects of the immune system and EMT on epithelial cancers
+}
+
 \author{Daniel R. Bergman$^{1}$,
@@
 \section*{Abstract}
-Preceding tumorigenesis and throughout the lifetime of a tumor, the immune system is engaged in complex sets of interactions in and around the tumor.
-These are regulated by the tumor microenvironment (TME) in myriad ways.
-Moreover,  epithelial cell plasticity impacts disease incidence and progression via processes of epithelial-to-mesenchymal transition (EMT).
-In turn, both the tumor and the immune system exert substantial influence over the TME in competition to influence the fate of the host.
-Determining the factors and mechanisms underpinning these complex interactions remains an important task for cancer immunotherapy.
-We developed a multiscale agent-based model to describe the interplay between three components: the tumor (or pre-tumor cells), the immune system, and EMT.
-The model revealed a \tcr{non-monotonic relationship} between the cell cycle times associated with the mesenchymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival.
-In comparison, increasing the immune-evasiveness of the mesenchymal phenotypes always led to poorer prognosis.
-Moreover, encouraging the TME
-towards a pro-EMT state also benefits the tumor.
-Together, these results indicate that the joint regulation of the TME by the immune system and the tumor itself provides possible targets for future cancer therapies, in particular cancers such as pancreatic cancer in which EMT is speculated to play a role.
+Preceding and during cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor incidence and prognosis.
+Inflammation also has significant effects on gastrointestinal and pancreatic tumors.
+% These are regulated by the tumor microenvironment (TME) in myriad ways.
+Moreover, epithelial cell plasticity can significantly alter disease trajectories (for better or worse) via epithelial-to-mesenchymal transition (EMT).
+Several of the same pathways that regulate EMT are involved in tumor-immune interactions.
+%In turn, both the tumor and the immune system exert substantial influence over the TME in competition to influence the fate of the host.
+Yet little is known about the mechanisms and consequences of these regulatory processes.
+Here we introduce a multiscale agent-based model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial tissues.
+Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that increase cancer-free survival.
+Delaying tumorigenesis depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
+We performed analysis of pancreatic cancer data from The Cancer Genome Atlas to compare clinical measurements with model predictions.
+We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer.
+These results offer novel means to delay disease onset by controlling properties of EMT and demonstrate the importance of studying cancer-immune interactions in the light of EMT.
+%The model presented provides a base for future studies into the intricate regulation of tumorigenesis.
 
@@
 \section{Introduction}
-Cancer and the immune system interact in myriad intricate ways.
+Cancer and the immune system interact in myriad ways.
 Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME), indeed, immune cells can be considered as a part of the TME: responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}.
@@
 \ref{fig:tcga}B, D).
-We see that in both cases, survival is significantly affected by the gene ontology signature, but, importantly, the presence of both EMT and inflammatory signatures has a greater impact on survival than the effects of EMT alone.
-This suggests that
+We see that in both cases, survival is affected by the gene ontology signature, but the presence of both EMT and inflammatory signatures has a greater impact on survival than the effects of EMT alone.
+This suggests that when studying interactions between cancer and the immune system, is it important to consider EMT; the effects of which may impact the dynamics and should not be overlooked.
 

### 2019-03-18 Paper/Supplement/Supplement.tex
Various Updates

\documentclass[11pt, a4paper, preprint]{article}
\usepackage{outline}
\usepackage{float}
\usepackage{graphicx}
\usepackage{color}
\usepackage{amsmath}
\usepackage{cite}
\graphicspath{ {Figs/} }
\usepackage{longtable}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}
\newcommand{\tcr} { \textcolor{red} }

\title{Supplementary Material}
\author{}

\renewcommand\Authands{ and }

\begin{document}
\maketitle

 \begin{longtable}{||c | c | c | c | c | c | c | c | c | c ||} 
 \hline
 Name & Base Value & Single Patient & Survival Curve & Vary MIE & Vary MGA & Vary Treg TGFB & Vary MIE and INFL & Vary MGA and INFL & Heat Map\\ [0.5ex] 
 \hline\hline
 RP Cancer Line &  0.5 &  \\ 
 \hline
 MIE  & 0.6 & 0.9 & 0.4-0.9 & 0.7 & 0.5 & 0.9 \\
 \hline
 MGA & 0.2 & & 0.1-0.4 & 0 & 0.2 & 0 \\
 \hline
 INFL High Duration  & 30 & & & & 60 & & & & 60 \\
 \hline
INFL Low Duration  & 30  \\
 \hline
Mes Threshold & 0.7 \\
 \hline
 p  & 0.28 \\ 
 \hline
 d  & 0.14 \\
 \hline
 $N_{00}$  & 8\\
 \hline
 $N_0$  & 80 \\
 \hline
  NK Efficacy Low  & 0.2 & 10 & & 10 & 10 & 10 & 10 & 10 & 10 \\
  \hline
  Treg EC50 Low & 5 & & & & & 2.5 \\
  \hline
  CTL Efficacy Low  & 4 & 200 & & 200 & 200 & 200 & 200 & 200 & 200\\
  \hline
  NK Efficacy Up & 2 & & & & 1.2 & & & &  1.2 \\
  \hline
  CTL Efficacy Up & 3 & & & & 3 & & & &  3\\
  \hline
  Treg Efficacy Up  & 200 & & & & 10& & & &  10\\
  \hline
  NK Source Low  & 1.3 \\ 
  \hline
  CTL Recruitment Low  & 100 \\ 
  \hline
  Treg Recruitment Low  & 200 \\
  \hline
  NK Recruitment Up & 1 & & & & & 3\\
  \hline
  CTL Recruitment Up & 1 & & & & & 3\\
  \hline
   Treg Recruitment Up & 1 & & & & & 5\\
  \hline
  $p_{mutation}$ Start & 1e-2 \\
  \hline
  $p_{mutation}$ & 1e-4 \\
  \hline
  Apoptosis Down & 0.3 \\
  \hline
  Immune Evasion & 0.48 \\
  \hline
  Proliferation Up & 0.36  \\
  \hline
  $k_{EMT}$  & .01 \\
  \hline
  c & 6 \\
  \hline
  TGFB Received EC50 & 200 \\
  \hline
  TGFB Max & 500 & 700 & & & & & &  & 700\\
  \hline 
  Mutant TGFB & 5e-2 & & & & & 0.5\\
  \hline
  Treg TGFB & 5e-1 \\
  \hline
\end{longtable}

\newpage

\begin{center}
 \begin{longtable}{||c | p{10cm} | c||} 
 \hline
 Name & Description & Value  \\ [0.5ex] 
 \hline\hline
 RP Cancer Line & When proportion of mutant cells reaches this number, then cancer is declared & 0.5  \\ 
 \hline
 MIE & (mesenchymal immune evasion) proportional decrease of probability immune cells clear mutated mesenchymal cells & 0.6  \\
 \hline
 MGA & (mesenchymal growth arrest) proportional decrease of probability mesenchymal cells proliferate; this decreased probability is accompanied and an equal and opposite increase in probability of rest & 0.2  \\
 \hline
 INFL High Duration & Number of consecutive cell cycles a patient will experience high inflammation before returning to a low inflammatory state & 30  \\
 \hline
INFL Low Duration & Number of consecutive cell cycles a patient will experience low inflammation before returning to a high inflammatory state & 30  \\
 \hline
Mes Threshold & the threshold EMT score above which a cell is labeled mesenchymal and below which it is labeled epithelial & 0.7 \\
 \hline
 p & weight of a cell proliferating during a cell cycle & 0.28 \\ 
 \hline
 d & weight of a cell dying during a cell cycle & 0.14 \\
 \hline
 $N_{00}$ & EC50 term for Hill functions describing likelihood of cytotoxic immune cells locating mutated cells & 8\\
 \hline
 $N_0$ & EC50 term for Hill function describing negative feedback of total cell population on cell proliferation rates & 80 \\
 \hline
  NK Efficacy Low & weight of NK cell clearing mutated cells during low inflammation & 0.2 \\
  \hline
  Treg EC50 Low & EC50 term for Hill functions describing Treg-mediated inhibition of efficacy of cytotoxic immune cells during low inflammation & 5 \\
  \hline
  CTL Efficacy Low & weight of CTL cell clearing mutated cells during low inflammation & 4 \\
  \hline
  NK Efficacy Up & proportional increase of NK Efficacy Low during high inflammation & 2 \\
  \hline
  CTL Efficacy Up & proportional increase of CTL Efficacy Low during high inflammation & 3 \\
  \hline
  Treg Efficacy Up & during high inflammation, Treg EC50 is Treg EC50 Low / Treg Efficacy Up & 200 \\
  \hline
  NK Source Low & rate at which NK cells enter TME during low inflammation & 1.3 \\ 
  \hline
  CTL Recruitment Low & rate at which CTL cells are recruited after mutated cells are lysed during low inflammation & 100 \\ 
  \hline
  Treg Recruitment Low & rate at which Treg cells are recruited after mutated cells are lysed during low inflammation & 200 \\
  \hline
  NK Recruitment Up & proportional increase of NK source rate during high inflammation & 1 \\
  \hline
  CTL Recruitment Up & proportional increase of CTL recruitment rate during high inflammation & 1 \\
  \hline
   Treg Recruitment Up & proportional increase of Treg recruitment rate during high inflammation & 1 \\
  \hline
  $p_{mutation}$ Start & maximum probability of mutating after the warmup period ends & 1e-2 \\
  \hline
  $p_{mutation}$ & increase in probability of a cell mutating after it proliferates but does not mutate & 1e-4 \\
  \hline
  Apoptosis Down & proportional decrease in apoptosis weight for cells with apoptosis pathway mutated & 0.3 \\
  \hline
  Immune Evasion & proportional decrease in immune clearance weight for cells with immune pathway mutated & 0.48 \\
  \hline
  Proliferation Up & proportional increase in proliferation weight for cells with proliferation pathway mutation & 0.36 \\
  \hline
  $k_{EMT}$ & rate parameter controlling speed of EMT and MET & .01 \\
  \hline
  c & standard deviation of noise on TGF-$\beta$ received by each cell & 6 \\
  \hline
  TGFB Received EC50 & EC50 term for Hill function determining how much TGF-$\beta$ in TME enters a cell & 200 \\
  \hline
  TGFB Max & maximum amount of TGF-$\beta$ that can enter a cell from the TME during a cell cycle & 500 \\
  \hline 
  Mutant TGFB & amount of TGF-$\beta$ produced by each mutated cell each cycle & 5e-2 \\
  \hline
  Treg TGFB & amount of TGF-$\beta$ produced by each Treg cell each cycle & 5e-1 \\
  \hline
\end{longtable}
\end{center}

\end{document}

### 2019-03-19 Paper/Paper.tex
Equation Updates
Changed parameter names as Adam and I discussed

@@
 
-$$ p_{prolif} = p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})(1-\delta_{MES}m_{GA})\frac{N_0}{N_0+N} $$
+\begin{equation}\tag{2.1}
+\rho_P = p(1+\delta_{P}\Delta_P)(1-\zeta \Delta_{MGA})\frac{K_0}{K_0+N_C}
+\end{equation}
 
@@
 
-$$ p_{apop} = d(1-\delta_{\text{AM}}\text{apop}_{\text{down}}) $$
+\begin{equation}\tag{2.2}
+\rho_A = d_C(1-\delta_{A}\Delta_A)
+\end{equation}
 
@@
 
-$$ p_{\text{NK}} =\delta_M \frac{\text{NK}}{N/N_{1}+\text{NK}}  \frac{E_{\text{NK}}}{1+\text{Treg}/\text{Treg}_{\text{EC50}}} (1-\delta_{IEM}\text{IE}_{\text{up}})(1-\delta_{MES}m_{\text{IE}}) $$
+\begin{equation}\tag{2.3}
+\rho_{NK} =\delta_{MUT} \frac{N_{NK}}{N_C/K_{1}+N_{NK}}  \frac{E_{NK}}{1+N_{Treg}/K_2} (1-\delta_{IE}\Delta_{IE})(1-\zeta \Delta_{MIE})
+\end{equation}
 % Should I worry about the order in which these appear? Rearranging would make the formula look better, but the order I have introduced them makes the most sense.
@@
 
-$$ p_{rest} = 1 + \delta_{\text{MES}}p(1+\delta_{\text{PM}}\text{prolif}_{\text{up}})m_{GA}\frac{N_0}{N_0+N} $$
+\begin{equation}\tag{2.4}
+\rho_R = 1 + \zeta p(1+\delta_P\Delta_P)\Delta_{MGA}\frac{K_0}{K_0+N_C}
+\end{equation}
 
@@
 %$$ \text{TGFB}_i = H(\text{TGFB}/N) + X_i, \quad X_i \sim N(0,\sigma^2)$$
+\begin{equation}\tag{2.5}
+\tau_i = \frac{\tau_{max}}{N_C}\frac{1}{1+K_3/\tau}+ X_i
+, \quad X_i \sim N(0,\sigma^2)
+\end{equation}
 
-$$ \text{TGFB}_i = \frac{\text{TGFB}_{max}}{N}\frac{\text{TGFB}/\gamma_{EC50}}{1+\text{TGFB}/\gamma_{EC50}}+ X_i
-, \quad X_i \sim N(0,\sigma^2)$$
 The summand on the left divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.
@@
 
-$$ \text{TGFB} = \text{tgfb}_{\text{mut}}M + \text{tgfb}_{\text{Treg}}\text{Treg}$$ % clear that M is number of mutants, right?
+\begin{equation}\tag{2.6}
+\tau = \tau_{MUT}N_{MUT} + \tau_{Treg}N_{Treg}
+\end{equation}
 
@@
 
-$$ \text{NK}' = \sigma_{\text{NK}} - d_{\text{NK}}\text{NK} $$
+\begin{equation}\tag{2.7}
+N_{NK}' = \sigma_{NK} - d_{NK}N_{NK}
+\end{equation}
 
@@
 
-$$ \text{NK}_{k+1} = \left (\text{NK}_k-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} $$
+ \begin{equation}\tag{2.8}
+N_{\text{NK}}(k+1) = \left (N_{\text{NK}}(k)-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}}
+\end{equation}
 
@@
 For CTLs and Tregs, they rely on mutant cells being cleared before they can be activated.
-Let $\tilde{M}_k$ represent the number of mutant cells cleared by the immune system during cell cycle $k$.
+Let $N_{\text{MUT}}^*(k)$ represent the number of mutant cells cleared by the immune system during cell cycle $k$.
 In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function.
@@
 
-\begin{align*}
-\text{CTL}' & = \sigma_{\text{CTL}}\tilde{M} - d_{\text{CTL}}\text{CTL} \\
-\text{Treg}' & = \sigma_{\text{Treg}}\tilde{M} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}- d_{\text{Treg}}\text{Treg}
-\end{align*}
+\begin{equation}\tag{2.9}
+\begin{split}
+N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}\text{CTL} \\
+\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
+\end{split}
+\end{equation}
 
@@
 
-\begin{align*}
-\text{CTL}_{k+1} & = & \left (\text{CTL}_k-\sigma_{\text{CTL}}\tilde{M}_k/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}\tilde{M}_k/d_{\text{CTL}}\\
-\text{Treg}_{k+1} & = & \left (\text{Treg}_k-\frac{\sigma_{\text{Treg}}\tilde{M}_k}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}\right )\exp(-d_{\text{Treg}}\Delta t)+\\
-& & \frac{\sigma_{\text{Treg}}\tilde{M}_k}{d_{\text{Treg}}} \frac{\text{TGFB}}{1+\text{TGFB}/\text{EC50}}
-\end{align*}
+\begin{equation}\tag{2.10}
+\begin{split}
+N_\text{CTL}(k+1) & =  \left (N_\text{CTL}(k)-\sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\\
+N_\text{Treg}(k+1) & =  \left (N_\text{Treg}(k)-\frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau}{1+\tau/K_4}\right )\exp(-d_{\text{Treg}}\Delta t)\\
+&+ \frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau}{1+\tau/K_4}
+\end{split}
+\end{equation}
 

### 2019-03-20 Paper/Paper.tex
Minor Updates
Added a ref and deleted some extra spaces.

@@
 continuous processes during cell differentiation \cite{moris16_transition}.
-Intermediate states have emerged as a central mechanism by which cell fates (and the noise inherent within in them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}.
+Intermediate states have emerged as a central mechanism by which cell fates (and the noise inherent within them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}.
 \par 
 While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation and progression \cite{nieto2016emt}.
-TGF-$\beta$ is a master regulator of EMT [REF], linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
+TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
 Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
@@
 The immune evasiveness of mesenchymal cells is well-established: as a cell is targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be established before the lysing event.
-This immunological synapse is dependent upon the surface proteins [WHICH??] of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form \cite{terry2017new}.
+This immunological synapse is dependent upon the surface proteins, such as the T-cell receptors of the major histocompatibility complex, of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form \cite{terry2017new}.
 %There are also other structural changes that take place in mesenchymal cells that could contribute to this refractory nature towards immune clearance.

### 2019-03-20 Paper/Paper.tex
Finished Updates from 3/18 Skype
Still need to look into refs we discussed, however

@@
 A single mutation is sufficient for the cell to be considered as a mutant cell.
-
+\par
 We model immune cells as continuous variables, appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.
 The cytokine TGF-$\beta$ is continuous within the tumor microenvironment (TME), i.e. the system is well-mixed.
-Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both on the TME and cell-intrinsic factors.
+Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
 While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
@@
 There are four factors that contribute to the probability that a cell will proliferate.
-The first is a base proliferation probability that all cells have, $p$.
-Second, if the cell has a mutation in the proliferation pathway, then the weight for proliferation is proportionally increased by $\text{prolif}_{\text{up}}$.
-Third, if the cell is mesenchymal, then the weight for proliferation is proportionally decreased by $m_{GA}$, which stands for mesenchymal growth arrest.
-Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N$.
+The first is a base proliferation rate that all cells have, $p$.
+Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
+Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
+Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
 In total, the weight for proliferation is given by
@@
 \begin{equation}\tag{2.1}
-\rho_P = p(1+\delta_{P}\Delta_P)(1-\zeta \Delta_{MGA})\frac{K_0}{K_0+N_C}
+\rho_P = p(1+\delta_{P}\Delta_P)(1-\zeta \Delta_{\text{MGA}})\frac{K_0}{K_0+N_C}
 \end{equation}
@@
 There are two factors that contribute to a cell's weight for undergoing apoptosis.
-There is a basal apoptosis rate that all cells experience, $d$ for death.
-Second, if the cell has a mutation in the apoptosis pathway, then the weight for undergoing apoptosis is proportionally decreased by $\text{apop}_{\text{down}}$.
+There is a basal apoptosis rate that all cells experience, $d_C$ for death.
+Second, if the cell has a mutation in the apoptosis pathway ($\delta_A=1$), then the weight for undergoing apoptosis is proportionally decreased by $\Delta_A$.
 In total, the weight for apoptosis is given by
@@
 For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
-First of all, the cell needs to be mutated, $\delta_M$.
-Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell.
-Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the weight for immune clearance given an immune cell has found the mutated cell.
-Fourth, there is a decreasing Hill factor based on the number of Treg cells present.
-Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation or if it is mesenchymal.
+First of all, the cell needs to be mutated ($\delta_{\text{MUT}}=1$).
+Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
+Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
+Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
+Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
 In total, the weight of NK clearance is given by
@@
 \begin{equation}\tag{2.3}
-\rho_{NK} =\delta_{MUT} \frac{N_{NK}}{N_C/K_{1}+N_{NK}}  \frac{E_{NK}}{1+N_{Treg}/K_2} (1-\delta_{IE}\Delta_{IE})(1-\zeta \Delta_{MIE})
+\rho_{\text{NK}} =\delta_{\text{MUT}} \frac{N_{\text{NK}}}{N_C/K_{1}+N_{\text{NK}}}  \frac{E_{\text{NK}}}{1+N_{\text{Treg}}/K_2} (1-\delta_{\text{IE}}\Delta_{\text{IE}})(1-\zeta \Delta_{\text{MIE}})
 \end{equation}
@@
 \paragraph{Rest in $G_0$} 
-The weight associated with quiescence is taken as 1 except in the case of Mesenchymal cells.
-For Mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
+The weight associated with quiescence is taken as 1 except in the case of mesenchymal cells.
+For mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
 Hence, the weight of quiescence is given by
@@
 \begin{equation}\tag{2.4}
-\rho_R = 1 + \zeta p(1+\delta_P\Delta_P)\Delta_{MGA}\frac{K_0}{K_0+N_C}
+\rho_R = 1 + \zeta p(1+\delta_P\Delta_P)\Delta_{\text{MGA}}\frac{K_0}{K_0+N_C}
 \end{equation}
@@
 This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
-All proliferating cells have a nonzero probability of undergoing a pathway mutation.
+Second, all proliferating cells have a nonzero probability of undergoing a pathway mutation.
 If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
 If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases. 
-Then, the EMT values for each cell is updated.
+Finally, the EMT values for each cell is updated.
 This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
@@
 \begin{equation}\tag{2.5}
-\tau_i = \frac{\tau_{max}}{N_C}\frac{1}{1+K_3/\tau}+ X_i
+\tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{1}{1+K_3/\tau}+ X_i
 , \quad X_i \sim N(0,\sigma^2)
@@
 
-The summand on the left divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.
+The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.
 Then, some randomness is introduced by the independent and identically distributed variables $X_i$where $\sigma$ is a parameter in the model.
@@
 
-Next, the amount of TGF-$\beta$ for the next cell cycle is determined by the number of mutated cells, $M$, and the number of Treg cells, each one producing a fixed amount of TGF-$\beta$.
+Next, the amount of TGF-$\beta$ for the next cell cycle is determined by the number of mutated cells, $N_{\text{MUT}}$, and the number of Treg cells, $N_{\text{Treg}}$, each one producing a fixed amount of TGF-$\beta$.
 It is given by
@@
 \begin{equation}\tag{2.6}
-\tau = \tau_{MUT}N_{MUT} + \tau_{Treg}N_{Treg}
+\tau = \tau_{\text{MUT}}N_{\text{MUT}} + \tau_{\text{Treg}}N_{\text{Treg}}
 \end{equation}
@@
 \begin{equation}\tag{2.7}
-N_{NK}' = \sigma_{NK} - d_{NK}N_{NK}
+N_{\text{NK}}' = \sigma_{\text{NK}} - d_{\text{NK}}N_{\text{NK}}
 \end{equation}
@@
 Let $N_{\text{MUT}}^*(k)$ represent the number of mutant cells cleared by the immune system during cell cycle $k$.
-In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function.
+In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
 We choose the following differential equations to govern the CTL and Treg populations:
@@
 In the implementation by Sohier, they recommend using at least 10 points\cite{sohier2014improvement}.
-
+\par
 To determine which points in parameter space to sample at, we had to create distributions on all parameter values.
@@
 That is, for instance, how we ended up with NK cells having a steady state population of 10 cells in the model.
-
+\par
 Once we had these parameter values, we created Gaussian distributions centered at these values with large variances and truncated at 0.
@@
 The variances were chosen as twice the mean to get a large spread of values.
+\par
+\tcr{Adam: Does this address what the y-axis is in Figure 2?}
+The algorithm outputs the average of the absolute change of area under the curve for each parameter normalized by the corresponding change in the cdf function for that parameter.
+This output, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.
 
@@
 \begin{enumerate}
-     \item The first principle component of the geneset-wide expression is obtained for each element of the n-tuple of gene sets.
+     \item The first principle component of the \tcr{geneset}-wide expression is obtained for each element of the n-tuple of gene sets.
      \item Two clusters of n-dimensional patient vectors are obtained by k-means clustering.
@@
 Once I have the table, I will use the name of that table.}
-B. Example simulation of the dynamics of a one patient; warmup period is shown in red.
+B. Example simulation of the dynamics of a one patient; warmup period ends at the vertical dashed line.
 The inflammation cycling scheme is graphically represented above the patient dynamics.
@@
 \ref{fig:ModelIntro}A).
-For example, if a cell undergoes EMT, the probability that is will proliferate is reduced; if a cell gains a mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
+For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
 The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset).
@@
 %Note: NK cells also lymphocytes! 
-The adaptive immune populations also exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
+The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
 Within each of these periods, adaptive immune populations increase or decrease rapidly in accordance with the inflammation state.
 \par 
-The {\em in silico} patient described here is given for the purpose of illustrating features the model.
+The {\em in silico} patient described here is given for the purpose of illustrating features of the model.
 Given the multiple sources of noise in the model, in order to quantify patient dynamics and cancer-free survival rates, below we will simulate large cohorts of patients.
@@
 To assess the sensitivity of the model parameters and thus identify those that are most important in determining cancer-free survival times, we applied Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods) to the model.
-% Sentence on how it is 'global' and sentence on what the measured output is (ie y axis) 
-...
 The results of Morris OAT on the 31 model parameters are shown in \ref{fig:MOAT}: a subset of parameters demonstrate much higher levels of sensitivity than others.
@@
 %% We need to discuss parameter names. Maybe replace 'low' with 'base' and 'up' with 'gain' ?
-Any of the parameters in the \tcr{table} which include the word ``low'' are parameters that govern the model in a low inflammatory state.
-With the exception of ``prolif up'', all parameters that end in ``up'' or have `U' at the end of their abbreviation, are the factors by which these parameters change when the system switches from low to high inflammation.
-The IHD and ILD parameters determine the duration of high and low inflammation, respectively.
-A cell with mutated pathways will be affected by the ``apop down'', ``prolif up'', and ``immune evasion'' parameters, depending on which pathways are mutated.
+Many parameters in the model change depending on the inflammation state.
+For the sake of a naming convention, any parameter which includes ``low'' in the name represents the parameter value when inflammation is low.
+For when the inflammation is high, these ``low'' parameters get scaled by parameters which include ``up'' in the name.
+The ILD and IHD parameters determine the duration of low and high inflammation, respectively.
+A cell with mutated pathways will be affected by the $\Delta_A$, $\Delta_P$, and $\Delta_{\text{IE}}$ parameters, depending on whether the apoptosis, proliferation, or immune evasion pathways are mutated.
 The mesenchymal cells are influenced by the MGA and MIE parameters.
-Finally, $\sigma$, $k_\text{EMT}$, TGFB$_{max}$, and $\gamma_\text{EC50}$ all control how cells transition between the epithelial and mesenchymal states.
+Finally, $\sigma$, $k_\text{EMT}$, $\tau_\text{max}$, and $K_3$ all control how cells transition between the epithelial and mesenchymal states.
 
@@
 \caption{
-A-C. Trajectory of one patient per cohort followed from warmup through 2000 cell cycles while varying MIE (A), MGA (B) or XXXXXX (C). 
+A-C. Trajectory of one patient per cohort followed from warmup through 2000 cell cycles while varying $\Delta_\text{MIE}$ (A), $\Delta_\text{MGA}$ (B) or $\tau_\text{Treg}$ (C). 
 D. Increasing MIE decreases Time to Cancer. 
@@
 MGA is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
-The cytokine TGF-$\beta$ is also involved in the EMT process (by increases the probability of EMT).
+The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT).
 All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
@@
 We explored the effects of varying the inflammation state of the patient on cancer-free survival, to investigate competing interactions within the TME and their effect on EMT.
-Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable inflammatory.
+Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable inflammation.
 For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and the Time to Cancer is monotonic, i.e.

### 2019-03-25 Paper/Paper.tex
Removed old comments and added refs

@@
 
+%%%% Possible additional refs:
+% Diefenbach 2001
+% They show that when tumor cells express a particular marker, then the immune system gains considerable potency and clears the tumor.
+Furthermore, immunization via tumor cells expressing these ligands creates a host immunity to subsequent tumors even if these tumors lack the ligand.
+
+% Carretero 2015
+% As eosinophils can better carry out their recruitment function of CD8+ T cells, tumor rejection by the immune system is more readily accomplished.
+
+% Ju 2016
+% A cancer vaccine was demonstrated to be effective in reducing colorectal cancer incidence in mice. In addition, efficacy of CTLs was shown to be increased.
+
 %%%%%%%%%%%%%%%%%%%%%%%
@@
 The effects of the immune system on a tumor can be broadly summarized into two modules.
-The cytotoxic branch of the immune system (such as natural killer (NK) cells and cytotoxic T cells (CTLs)) seek out and lyse tumor cells.
-The regulatory branch of the immune system (T regulatory cells (Tregs) and other factors), inhibits the effective functioning of the cytotoxic branch.
-[REF]
+The cytotoxic branch of the immune system (such as natural killer (NK) cells and cytotoxic T cells (CTLs) seek out and lyse tumor cells.
+Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely\tcr{[REF]}.
+The regulatory branch of the immune system (T regulatory cells (Tregs) and other factors), inhibits the effective functioning of the cytotoxic branch\cite{ruffell2010lymphocytes}.
 %Make sure there's some ref for immune clearing tumor
@@
 This immunological synapse is dependent upon the surface proteins, such as the T-cell receptors of the major histocompatibility complex, of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form \cite{terry2017new}.
-%There are also other structural changes that take place in mesenchymal cells that could contribute to this refractory nature towards immune clearance.
-%Regardless, the evidence currently demonstrates a ``slipperiness'' of mesenchymal cells in regards to immune lysing.
 Below we refer to this phenotype as mesenchymal immune evasion (MIE).
@@
 \tcr{Ref}
-While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate. \tcr{Ref}
+While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate\cite{woods2014effects}.
 Cancer cells are well-known for achieving unregulated proliferation, so exploring a subset of cancer cells with decreased proliferation could provide intriguing new results.
@@
 
-% This paragraph needs to find a home in this section somewhere
-
 \section{Methods}
@@
 There is an immune evasion pathway that when mutated decreases the probability that an immune cell can effectively clear the mutated cell.
-% It might be nice to have a table to summarize this information?
-The EMT score will be exposited in Section \ref{EMT}. % make sure this is really the best decision
-
-%Add differential equations? % Should this be done here or where I have it later?
+The EMT score will be exposited in Section \ref{EMT}.
 
@@
 \end{equation}
-% Should I worry about the order in which these appear? Rearranging would make the formula look better, but the order I have introduced them makes the most sense.
-% Also, any thoughts on NK, CTL, Treg not being italicized? While other capital letters are?
 
@@
 
-%%%% Which of these two paragraphs is better? Currently the intro does not mention APCs and I don't think I need to go into that, especially when APC_k = \tilde{M}_k
 For CTLs and Tregs, they rely on mutant cells being cleared before they can be activated.
@@
 C. Example survival curve for one cohort of patients with basal parameter values.
-%A Should I list out all parameter values? A subset? Or refer to supplement for base parameter values?
 D. Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
@@
 Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of oncogenic mutations.
-%Note: NK cells also lymphocytes! 
 The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
@@
 \par
-%% We need to discuss parameter names. Maybe replace 'low' with 'base' and 'up' with 'gain' ?
 Many parameters in the model change depending on the inflammation state.
@@
 Since one goal of our analysis is to assess the specific effects of EMT on immune-cancer dynamics, the parameters MIE and MGA are of particular interest.
-In addition, inflammation parameters dictating the cycling scheme are important to study\tcr{REF}: we find that there are also high sensitivities in this model to these parameters.
+In addition, inflammation parameters dictating the cycling scheme are of interest because they are both highly influential on Time to Cancer and capable of being targeted by therapeutic treatments.
 In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and we will also study it further.
@@
 These striking differences in the mean Time to Cancer -- extended by up to one year by fine-grained optimization of mesenchymal growth arrest -- have clear therapeutic implications. 
-%What this indicates is that if the growth arrest of mesenchymal cells can be controlled, then they could be manipulated in such a way as to increase the Time to Cancer.
-%For patients presenting with chronically high inflammatory state, anti-inflammatories could be administered which would create the periods of low inflammation our model predicts as necessary for seeing these changes.
 Our model predicts that reducing the inflammatory state of the TME in three weeks out of every nine would be provide means to prolong cancer-free survival in combination with mesenchymal cell therapies to make the control of mesenchymal proliferation an effective anti-cancer strategy.
@@
 
-% \subsection{Heat Map of MIE vs MGA}
 \subsection{Analysis of TCGA data in comparison with model predictions suggests that mesenchymal phenotypes reduce cancer-free survival probability}\label{tcga}
@@
 On the other hand, models which incorporate more distinctions between epithelial and mesenchymal cells could give better clarity on the relative importance of these and reveal a fuller picture about how a mesenchymal subpopulation affects cancer outcomes.
-
 %A Adam, do you know of any?
@@
 The greatest challenge in translating the information here to the clinic, apart from creating the drugs, would be discovering the ideal growth rate for mesenchymal cells within each patient and how other patient-specific data will influence this result.
-
+\par
+Also, since the inflammatory state of the TME played such a large role in our simulations, making this a more robust facet of the model by allowing the tumor and immune system to influence the inflammatory state could give the model higher fidelity.
+One intriguing aspect could be the anti-inflammatory function of Tregs as they seek to restore homeostasis.
+\par
+A final future direction would be continuing to explore the heterogeneity of the tumor and how this helps the tumor evade the immune system and grow into a malignancy. 
+As we understand more and more about the diversity within the TME, additional models will be instrumental in determining the scope of these implications and the possible ways to address them.
+\par
 Progress in cancer treatment has always proven attainable even despite great challenges, for which the complexity of the disease is often responsible.

### 2019-03-25 Paper/Paper.tex
remove completed comments

@@
 The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
-Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions.
-\tcr{Ref}
+Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions \tcr{[REF]}.
 While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate\cite{woods2014effects}.
@@
 
-\tcr{In next few subsections the parameters are not defined.
+%\tcr{In next few subsections the parameters are not defined.
 Need to define in text and/or make a big table to summarize all.
@@
 \par
-\tcr{Adam: Does this address what the y-axis is in Figure 2?}
 The algorithm outputs the average of the absolute change of area under the curve for each parameter normalized by the corresponding change in the cdf function for that parameter.
@@
 \begin{enumerate}
-     \item The first principle component of the \tcr{geneset}-wide expression is obtained for each element of the n-tuple of gene sets.
+     \item The first principle component of the expression across the gene set is obtained for each element of the n-tuple of gene sets.
      \item Two clusters of n-dimensional patient vectors are obtained by k-means clustering.

### 2019-03-27 Paper/Supplement/Supplement.tex
Finished Updates from 3/18 Skype
Still need to look into refs we discussed, however
Added authors to supplement

@@
 \title{Supplementary Material}
-\author{}
+\author{Daniel R. Bergman$^{1}$,
+Matthew Karikomi\,$^{1}$,
+Qing Nie\,$^{1,2,*}$
+and Adam L. MacLean\,$^{3,*}$
+}
+\affil{
+  $^1$Department of Mathematics, University of California, Irvine,  Irvine, CA 92697, USA \\
+  $^2$Department of Cell and Developmental Biology, University of California, Irvine, Irvine, CA 92697, USA \\
+  $^3$Department of Biological Sciences, University of Southern California, Los Angeles, CA 90089, USA \\
+  $^*$Correspondence:  qnie@uci.edu (QN); macleana@usc.edu (ALM).
+}
+\renewcommand\Authands{ and }
 
-\renewcommand\Authands{ and }
+%\title{
+%The competing effects of the immune system and EMT on epithelial cancers
+%}
+
+\date{}
 
@@
 \maketitle
-
- \begin{longtable}{||c | c | c | c | c | c | c | c | c | c ||} 
- \hline
- Name & Base Value & Single Patient & Survival Curve & Vary MIE & Vary MGA & Vary Treg TGFB & Vary MIE and INFL & Vary MGA and INFL & Heat Map\\ [0.5ex] 
- \hline\hline
- RP Cancer Line &  0.5 &  \\ 
- \hline
- MIE  & 0.6 & 0.9 & 0.4-0.9 & 0.7 & 0.5 & 0.9 \\
- \hline
- MGA & 0.2 & & 0.1-0.4 & 0 & 0.2 & 0 \\
- \hline
- INFL High Duration  & 30 & & & & 60 & & & & 60 \\
- \hline
-INFL Low Duration  & 30  \\
- \hline
-Mes Threshold & 0.7 \\
- \hline
- p  & 0.28 \\ 
- \hline
- d  & 0.14 \\
- \hline
- $N_{00}$  & 8\\
- \hline
- $N_0$  & 80 \\
- \hline
-  NK Efficacy Low  & 0.2 & 10 & & 10 & 10 & 10 & 10 & 10 & 10 \\
-  \hline
-  Treg EC50 Low & 5 & & & & & 2.5 \\
-  \hline
-  CTL Efficacy Low  & 4 & 200 & & 200 & 200 & 200 & 200 & 200 & 200\\
-  \hline
-  NK Efficacy Up & 2 & & & & 1.2 & & & &  1.2 \\
-  \hline
-  CTL Efficacy Up & 3 & & & & 3 & & & &  3\\
-  \hline
-  Treg Efficacy Up  & 200 & & & & 10& & & &  10\\
-  \hline
-  NK Source Low  & 1.3 \\ 
-  \hline
-  CTL Recruitment Low  & 100 \\ 
-  \hline
-  Treg Recruitment Low  & 200 \\
-  \hline
-  NK Recruitment Up & 1 & & & & & 3\\
-  \hline
-  CTL Recruitment Up & 1 & & & & & 3\\
-  \hline
-   Treg Recruitment Up & 1 & & & & & 5\\
-  \hline
-  $p_{mutation}$ Start & 1e-2 \\
-  \hline
-  $p_{mutation}$ & 1e-4 \\
-  \hline
-  Apoptosis Down & 0.3 \\
-  \hline
-  Immune Evasion & 0.48 \\
-  \hline
-  Proliferation Up & 0.36  \\
-  \hline
-  $k_{EMT}$  & .01 \\
-  \hline
-  c & 6 \\
-  \hline
-  TGFB Received EC50 & 200 \\
-  \hline
-  TGFB Max & 500 & 700 & & & & & &  & 700\\
-  \hline 
-  Mutant TGFB & 5e-2 & & & & & 0.5\\
-  \hline
-  Treg TGFB & 5e-1 \\
-  \hline
-\end{longtable}
-
-\newpage
 
@@
 
+\newpage
+
+ \begin{longtable}{||c | c | c | c | c | c | c | c | c | c ||} 
+ \hline
+ Name & Base Value & Single Patient & Survival Curve & Vary MIE & Vary MGA & Vary Treg TGFB & Vary MIE and INFL & Vary MGA and INFL & Heat Map\\ [0.5ex] 
+ \hline\hline
+ RP Cancer Line &  0.5 &  \\ 
+ \hline
+ MIE  & 0.6 & 0.9 & 0.4-0.9 & 0.7 & 0.5 & 0.9 \\
+ \hline
+ MGA & 0.2 & & 0.1-0.4 & 0 & 0.2 & 0 \\
+ \hline
+ INFL High Duration  & 30 & & & & 60 & & & & 60 \\
+ \hline
+INFL Low Duration  & 30  \\
+ \hline
+Mes Threshold & 0.7 \\
+ \hline
+ p  & 0.28 \\ 
+ \hline
+ d  & 0.14 \\
+ \hline
+ $N_{00}$  & 8\\
+ \hline
+ $N_0$  & 80 \\
+ \hline
+  NK Efficacy Low  & 0.2 & 10 & & 10 & 10 & 10 & 10 & 10 & 10 \\
+  \hline
+  Treg EC50 Low & 5 & & & & & 2.5 \\
+  \hline
+  CTL Efficacy Low  & 4 & 200 & & 200 & 200 & 200 & 200 & 200 & 200\\
+  \hline
+  NK Efficacy Up & 2 & & & & 1.2 & & & &  1.2 \\
+  \hline
+  CTL Efficacy Up & 3 & & & & 3 & & & &  3\\
+  \hline
+  Treg Efficacy Up  & 200 & & & & 10& & & &  10\\
+  \hline
+  NK Source Low  & 1.3 \\ 
+  \hline
+  CTL Recruitment Low  & 100 \\ 
+  \hline
+  Treg Recruitment Low  & 200 \\
+  \hline
+  NK Recruitment Up & 1 & & & & & 3\\
+  \hline
+  CTL Recruitment Up & 1 & & & & & 3\\
+  \hline
+   Treg Recruitment Up & 1 & & & & & 5\\
+  \hline
+  $p_{mutation}$ Start & 1e-2 \\
+  \hline
+  $p_{mutation}$ & 1e-4 \\
+  \hline
+  Apoptosis Down & 0.3 \\
+  \hline
+  Immune Evasion & 0.48 \\
+  \hline
+  Proliferation Up & 0.36  \\
+  \hline
+  $k_{EMT}$  & .01 \\
+  \hline
+  c & 6 \\
+  \hline
+  TGFB Received EC50 & 200 \\
+  \hline
+  TGFB Max & 500 & 700 & & & & & &  & 700\\
+  \hline 
+  Mutant TGFB & 5e-2 & & & & & 0.5\\
+  \hline
+  Treg TGFB & 5e-1 \\
+  \hline
+\end{longtable}
+
 \end{document}

### 2019-03-28 Paper/Paper.tex
Minor edits

--- 2019-03-25 Paper/Paper.tex
@@
 
-\tcr{In next few subsections the parameters are not defined.
-Need to define in text and/or make a big table to summarize all.
-On notation: some changes may be needed, e.g.
-$\delta_{MES}$ not very clear.}
-
 There are four factors that contribute to the probability that a cell will proliferate.
@@
 \begin{enumerate}
-     \item The first principle component of the \tcr{geneset}-wide expression is obtained for each element of the n-tuple of gene sets.
+     \item The first principle component of the geneset-wide expression is obtained for each element of the n-tuple of gene sets.
      \item Two clusters of n-dimensional patient vectors are obtained by k-means clustering.
@@
 \section*{Acknowledgements}
+The work is partially supported by National Institutes of Health grants R01GM123731 and U01AR073159; National Science Foundation grants DMS1562176 and DMS1763272; The Simons Foundation grant [594598 to Q.N.]; and a grant by Jayne Koskinas Ted Giovanis Foundation for Health and Policy joint with Breast Cancer Research Foundation.
 

### 2019-03-28 Paper/Supplement/Supplement.tex
Minor edits

@@
 \maketitle
+
+\begin{center}
+ \begin{longtable}{|| c | c||} 
+ \hline
+ Name & Description  \\ [0.5ex] 
+ \hline\hline
+  & RP Cancer Line   \\ 
+ \hline
+$\Delta_\text{MIE}$ &  \\
+ \hline
+ $\Delta_\text{MGA}$ & MGA    \\
+ \hline
+&  INFL High Duration   \\
+ \hline
+ & INFL Low Duration    \\
+ \hline
+& Mes Threshold   \\
+ \hline
+ $p$ & proliferation rate of tissue cells \\ 
+ \hline
+ $d_C$  & death rate of tissue cells \\
+ \hline
+ $K_1$ & EC50 term for probability of NK cell finding mutant cell\\
+ \hline
+ $K_0$ & EC50 term for negative feedback of tissue cells on own proliferation\\
+ \hline
+ $E_\text{NK}$ & rate of NKs clearing mutants  \\
+  \hline
+  $K_2$ & EC50 term for Treg inhibition of cytotoxic functions  \\
+  \hline
+  $E_\text{CTL}$ & rate of CTLs clearing mutants \\
+  \hline
+  & increased NK efficacy during high inflammation  \\
+  \hline
+  & increased CTL efficacy during high inflammation  \\
+  \hline
+  & increased Treg efficacy during high inflammation  \\
+  \hline
+  $\sigma_\text{NK}$ & NK source rate \\ 
+  \hline
+  $\sigma_\text{CTL}$ & CTL source rate per cleared mutant cell \\ 
+  \hline
+  $\sigma_\text{Treg}$ & Treg source rate per cleared mutant cell \\ 
+  \hline
+   & increased NK source rate during high inflammation  \\
+  \hline
+   & increased CTL source rate during high inflammation  \\
+  \hline
+   & increased Treg source rate during high inflammation  \\
+  \hline
+  & maximum initial mutation damage after warmup  \\
+  \hline
+  & increase in probability to mutate for non-mutating proliferating cells  \\
+  \hline
+  $\Delta_A$ & mutant cells decreased apoptosis  \\
+  \hline
+  $\Delta_\text{IE}$ & mutant cells increased immune evasion  \\
+  \hline
+  $\Delta_P$ & mutant cells increased proliferation  \\
+  \hline
+  $k_\text{EMT}$ & EMT/MET rate  \\
+  \hline
+  $\sigma$ & standard deviation of noise in TGF-$\beta$ each cell receives  \\
+  \hline
+  $K_3$ & EC50 term for how much TGF-$\beta$ each cell has \\
+  \hline
+ $\tau_\text{max}$ & max amount of TGF-$\beta$ any cell can receive \\
+  \hline 
+ $\tau_\text{MUT}$ & rate of TGF-$\beta$ production by mutant cells\\
+  \hline
+ $\tau_\text{Treg}$ & rate of TGF-$\beta$ production by Treg\\
+  \hline
+\end{longtable}
+\end{center}
+
+\newpage
 

### 2019-04-01 Paper/Paper.tex
Changed first enumerate item
update fig captions

@@
 \paragraph{Proliferation}
-
 There are four factors that contribute to the probability that a cell will proliferate.
@@
 \subsection{Parameter Sensitivity Analysis}
-To identify a subset of the parameters to focus on, we used the Morris method, a one-step-at-a-time global sensitivity analysis.
-The algorithm runs the model numerous successive times, changing exactly one parameter each run\cite{morris1991factorial, sohier2014improvement}.
+To study parameter sensitivity, we implemented the Morris method, a one-step-at-a-time global sensitivity analysis algorithm.
+The algorithm runs the model numerous successive times, \tcr{changing exactly one parameter each run} \cite{morris1991factorial, sohier2014improvement}.
 Each model run involved 1000 patients and the value returned was the area under the survival curve.
@@
 \par
-The algorithm outputs the average of the absolute change of area under the curve for each parameter normalized by the corresponding change in the cdf function for that parameter.
+The algorithm \tcr{outputs the average of the absolute change of area under the curve for each parameter normalized by the corresponding change in the cdf function for that parameter.
+[I don't get this: what is y-axis?]}
 This output, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.
@@
 \begin{enumerate}
-     \item The first principle component of the geneset-wide expression is obtained for each element of the n-tuple of gene sets.
+     \item The first principle component of the expression across the gene set is obtained for each element of the n-tuple of gene sets.
      \item Two clusters of n-dimensional patient vectors are obtained by k-means clustering.
@@
 \caption{A.
-Schematic depiction of the model with blue/red denoting non-mutated/mutated cells in the agent-based model.
-In each cycle, the fate of a cell is regulated by the EMT state of the cell and by any pathway mutations harbored.
-Inset shows major processes of immune system regulation on the tissue cells.
-The parameters used in this model can be found in the Supplementary material.
+Schematic depiction of agent-based model components; blue/red denote tissue cells with/without mutations.
+Black arrows depict regulation of the cell fate in each cell cycle.
+Inset depicts major interactions between the immune system and tissue cells.
+ %The parameters used in this model can be found in the Supplementary material.
 Unless otherwise stated, parameter values are assumed to be those found in the Supplement.
@@
 Once I have the table, I will use the name of that table.}
-B. Example simulation of the dynamics of a one patient; warmup period ends at the vertical dashed line.
-The inflammation cycling scheme is graphically represented above the patient dynamics.
-C. Example survival curve for one cohort of patients with basal parameter values.
-D. Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
+B. A representation simulation of one patient. The model parameter values used can be found in XXXXX. 
+The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: mutated cells; Mes: mesenchymal cells. 
+C. Survival curve for one cohort of patients with the parameter values given in Table XXXXX. }
 \label{fig:ModelIntro}
@@
 {\includegraphics[width=0.85\textwidth]{Figure2/Figure2.jpg}}
-\caption{Global sensitivity analysis of model parameters using the Morris one-step-at-a-time method.}
+\caption{Sensitivity analysis of model parameters, determined via the global Morris one-step-at-a-time method. $\mu^*$ denotes the XXXXXX}
 \label{fig:MOAT}
@@
 {\includegraphics[width=0.85\textwidth]{Figure3/Figure3.pdf}}
-\caption{
-A-C. Trajectory of one patient per cohort followed from warmup through 2000 cell cycles while varying $\Delta_\text{MIE}$ (A), $\Delta_\text{MGA}$ (B) or $\tau_\text{Treg}$ (C). 
-D. Increasing MIE decreases Time to Cancer. 
-E. Increasing MGA increases Time to Cancer.
-F. Increasing Treg TGF-$\beta$ production results in decreased Time to Cancer.
+\caption{Effects of mesencymal cell properites on the time to cancer.
+Trajectories of one patient per cohort from warmup period ([-1000,0]) to 2000 cell cycles, for $\Delta_\text{MIE}$ (A); $\Delta_\text{MGA}$ (B); and $\tau_\text{Treg}$, Treg TGF-$\beta$ (C).
+D. Survival curves corresponding to changes in MIE (A) for a patient cohort of size XXXXX; shaded region represents XXXXXX. 
+E. Survival curves for patient cohort corresponding to changes in MGA (B).
+F. Survival curves for patient cohort corresponding to changes in Treg TGF-$\beta$ (C). 
 %% \color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.
@@
 {\includegraphics[width=0.9\textwidth]{Figure4/Figure4.pdf}}
-\caption{A.
-Survival curves for various inflammation cycling schemes.
-In the top two plots, the Time to Cancer decreases as MIE increases.
-In the bottom plot, there is no significant difference in survival.
-B. The average Time to Cancer for each cohort.
-C.
-Survival curves for various inflammation cycling schemes.
-In the top two plots, the Time to Cancer reaches a maximum value somewhere near MGA = 0.3.
-In the bottom plot, there is no significant difference in survival.
-D. The average Time to Cancer for each cohort.}
+\caption{Effects of inflammation on the time to cancer under different cycling schemes.
+A-B.
+As MIE varies, survival curves (each of XXXX patients)  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.
+C-D.
+As MGA varies, survival curves  and corresponding bar plots to summarize the mean time to cancer for each cohort are shown.}
 \label{fig:VaryINFL_and_MesPars}
@@
 {\includegraphics[width=0.6\textwidth]{Figure5/MIEvsMQ.jpg}}
-\caption{Heat Map contrasting the effects of MIE and MGA on Time to Cancer. Increasing MIE always decreases Time to Cancer, but MGA is non-monotonically linked to Time to Cancer.}
+\caption{Summary of the contrasting effects of MIE and MGA on the Time to Cancer.
+  % Increasing MIE always decreases Time to Cancer, but MGA is non-monotonically linked to Time to Cancer.
+}
 \label{fig:MIEvsMGA}

### 2019-04-05 Paper/Paper.tex
Figs 4 and 5 updated
Added condensed table to paper; made edits to MOAT section

@@
 
+\begin{table}[H]
+\begin{center}
+ \begin{tabular}{|| c | c||} 
+ \hline
+ {\bf Name} & {\bf Description}  \\ [0.5ex] 
+ \hline
+ $p$ & proliferation rate of tissue cells \\ 
+ \hline
+ $d_C$  & death rate of tissue cells \\
+ \hline
+$\Delta_\text{MIE}$ &  mesenchymal immune evasion \\
+ \hline
+ $\Delta_\text{MGA}$ & mesenchymal growth arrest    \\
+ \hline
+  $\Delta_A$ & mutant cells decreased apoptosis  \\
+  \hline
+  $\Delta_\text{IE}$ & mutant cells increased immune evasion  \\
+  \hline
+  $\Delta_P$ & mutant cells increased proliferation  \\
+  \hline
+ $K_0$ & EC50 term for negative feedback of tissue cells on own proliferation\\
+ \hline
+ $K_1$ & EC50 term for probability of NK cell finding mutant cell\\
+ \hline
+  $K_2$ & EC50 term for Treg inhibition of cytotoxic functions  \\
+  \hline
+  $K_3$ & EC50 term for how much TGF-$\beta$ each cell has \\
+  \hline
+ $E_\text{NK}$ & rate of NKs clearing mutants  \\
+  \hline
+  $E_\text{CTL}$ & rate of CTLs clearing mutants \\
+  \hline
+  $\sigma_\text{NK}$ & NK source rate \\ 
+  \hline
+  $\sigma_\text{CTL}$ & CTL source rate per cleared mutant cell \\ 
+  \hline
+  $\sigma_\text{Treg}$ & Treg source rate per cleared mutant cell \\ 
+  \hline
+  $k_\text{EMT}$ & EMT/MET rate  \\
+  \hline
+  $\sigma$ & standard deviation of noise in TGF-$\beta$ each cell receives  \\
+  \hline
+ $\tau_\text{max}$ & max amount of TGF-$\beta$ any cell can receive \\
+  \hline 
+ $\tau_\text{MUT}$ & rate of TGF-$\beta$ production by mutant cells\\
+  \hline
+ $\tau_\text{Treg}$ & rate of TGF-$\beta$ production by Treg\\
+  \hline
+\end{tabular}
+  \caption{The model parameter names and descriptions. Note that many of these values are affected by the inflammation state of the system.}
+\end{center}
+\end{table}
+
 \subsubsection{Mutation and tumorigenesis}
@@
 To study parameter sensitivity, we implemented the Morris method, a one-step-at-a-time global sensitivity analysis algorithm.
-The algorithm runs the model numerous successive times, \tcr{changing exactly one parameter each run} \cite{morris1991factorial, sohier2014improvement}.
+%The algorithm runs the model numerous successive times, \tcr{changing exactly one parameter each run} \cite{morris1991factorial, sohier2014improvement}.
+The algorithm runs the model numerous times by picking several base points and then varying one parameter at a time from each base point and tracking the difference\cite{morris1991factorial, sohier2014improvement}.
 Each model run involved 1000 patients and the value returned was the area under the survival curve.
@@
 \par
-The algorithm \tcr{outputs the average of the absolute change of area under the curve for each parameter normalized by the corresponding change in the cdf function for that parameter.
+%The algorithm \tcr{outputs the average of the absolute change of area under the curve for each parameter normalized by the corresponding change in the cdf function for that parameter.
 [I don't get this: what is y-axis?]}
+The algorithm the average of the absolute change of the output, which in our model is the area under the survival curve.
 This output, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.
@@
 \center
-{\includegraphics[width=0.6\textwidth]{Figure5/MIEvsMQ.jpg}}
+{\includegraphics[width=0.6\textwidth]{Figure5/MIEvsMGA_bigcbar.jpg}}
 \caption{Summary of the contrasting effects of MIE and MGA on the Time to Cancer.

### 2019-04-10 Paper/Paper.tex
Supplement Tables and Replace `XXXXX' s

@@
 Once I have the table, I will use the name of that table.}
-B. A representation simulation of one patient. The model parameter values used can be found in XXXXX. 
+B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
 The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: mutated cells; Mes: mesenchymal cells. 
-C. Survival curve for one cohort of patients with the parameter values given in Table XXXXX. }
+C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
 \label{fig:ModelIntro}
@@
 {\includegraphics[width=0.85\textwidth]{Figure2/Figure2.jpg}}
-\caption{Sensitivity analysis of model parameters, determined via the global Morris one-step-at-a-time method. $\mu^*$ denotes the XXXXXX}
+\caption{Sensitivity analysis of model parameters, determined via the global Morris one-step-at-a-time method.
+$\mu^*$ denotes the average absolute change in Time to Cancer when the parameter is varied.}
 \label{fig:MOAT}
@@
 Trajectories of one patient per cohort from warmup period ([-1000,0]) to 2000 cell cycles, for $\Delta_\text{MIE}$ (A); $\Delta_\text{MGA}$ (B); and $\tau_\text{Treg}$, Treg TGF-$\beta$ (C).
-D. Survival curves corresponding to changes in MIE (A) for a patient cohort of size XXXXX; shaded region represents XXXXXX. 
+D. Survival curves corresponding to changes in MIE (A) for a patient cohort of size 1000; shaded region represents the 95\% confidence interval for the evaluated function. 
 E. Survival curves for patient cohort corresponding to changes in MGA (B).
@@
 A-B.
-As MIE varies, survival curves (each of XXXX patients)  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.
+As MIE varies, survival curves (each of 200 patients)  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.
 C-D.

### 2019-04-10 Paper/Supplement/Supplement.tex
Added condensed table to paper; made edits to MOAT section
Supplement Tables and Replace `XXXXX' s

@@
  Name & Description  \\ [0.5ex] 
- \hline\hline
-  & RP Cancer Line   \\ 
- \hline
-$\Delta_\text{MIE}$ &  \\
- \hline
- $\Delta_\text{MGA}$ & MGA    \\
- \hline
-&  INFL High Duration   \\
- \hline
- & INFL Low Duration    \\
- \hline
-& Mes Threshold   \\
  \hline
@@
  \hline
- $K_1$ & EC50 term for probability of NK cell finding mutant cell\\
+$\Delta_\text{MIE}$ &  mesenchymal immune evasion \\
+ \hline
+ $\Delta_\text{MGA}$ & mesenchymal growth arrest    \\
+ \hline
+  $\Delta_A$ & mutant cells decreased apoptosis  \\
+  \hline
+  $\Delta_\text{IE}$ & mutant cells increased immune evasion  \\
+  \hline
+  $\Delta_P$ & mutant cells increased proliferation  \\
   \hline
@@
  \hline
- $E_\text{NK}$ & rate of NKs clearing mutants  \\
+ $K_1$ & EC50 term for probability of NK cell finding mutant cell\\
  \hline
@@
   \hline
+  $K_3$ & EC50 term for how much TGF-$\beta$ each cell has \\
+  \hline
+ $E_\text{NK}$ & rate of NKs clearing mutants  \\
+  \hline
   $E_\text{CTL}$ & rate of CTLs clearing mutants \\
-  \hline
-  & increased NK efficacy during high inflammation  \\
-  \hline
-  & increased CTL efficacy during high inflammation  \\
-  \hline
-  & increased Treg efficacy during high inflammation  \\
   \hline
@@
   \hline
-   & increased NK source rate during high inflammation  \\
-  \hline
-   & increased CTL source rate during high inflammation  \\
-  \hline
-   & increased Treg source rate during high inflammation  \\
-  \hline
-  & maximum initial mutation damage after warmup  \\
-  \hline
-  & increase in probability to mutate for non-mutating proliferating cells  \\
-  \hline
-  $\Delta_A$ & mutant cells decreased apoptosis  \\
-  \hline
-  $\Delta_\text{IE}$ & mutant cells increased immune evasion  \\
-  \hline
-  $\Delta_P$ & mutant cells increased proliferation  \\
-  \hline
   $k_\text{EMT}$ & EMT/MET rate  \\
@@
   $\sigma$ & standard deviation of noise in TGF-$\beta$ each cell receives  \\
-  \hline
-  $K_3$ & EC50 term for how much TGF-$\beta$ each cell has \\
   \hline
@@
 
-\newpage
-
 \begin{center}
- \begin{longtable}{||c | p{10cm} | c||} 
+ \begin{longtable}{|| c | p{7cm} | c | c ||} 
  \hline
- Name & Description & Value  \\ [0.5ex] 
+ Name & Description & INFL Low Value & INFL High Value  \\ [0.5ex] 
  \hline\hline
- RP Cancer Line & When proportion of mutant cells reaches this number, then cancer is declared & 0.5  \\ 
+  $p$ & proliferation rate of tissue cells & 0.28 & \\ 
  \hline
- MIE & (mesenchymal immune evasion) proportional decrease of probability immune cells clear mutated mesenchymal cells & 0.6  \\
+ $d_C$  & death rate of tissue cells & 0.14 & \\
  \hline
- MGA & (mesenchymal growth arrest) proportional decrease of probability mesenchymal cells proliferate; this decreased probability is accompanied and an equal and opposite increase in probability of rest & 0.2  \\
+$\Delta_\text{MIE}$ &  MIE & 0.6 & \\
  \hline
- INFL High Duration & Number of consecutive cell cycles a patient will experience high inflammation before returning to a low inflammatory state & 30  \\
+ $\Delta_\text{MGA}$ & MGA  & 0.2 &  \\
  \hline
-INFL Low Duration & Number of consecutive cell cycles a patient will experience low inflammation before returning to a high inflammatory state & 30  \\
+  $\Delta_A$ & mutant cells decreased apoptosis & 0.3 &  \\
   \hline
-Mes Threshold & the threshold EMT score above which a cell is labeled mesenchymal and below which it is labeled epithelial & 0.7 \\
+  $\Delta_\text{IE}$ & mutant cells increased immune evasion & 0.48 &  \\
   \hline
- p & weight of a cell proliferating during a cell cycle & 0.28 \\ 
+  $\Delta_P$ & mutant cells increased proliferation & 0.36 &  \\
   \hline
- d & weight of a cell dying during a cell cycle & 0.14 \\
+ $K_0$ & EC50 term for negative feedback of tissue cells on own proliferation & 80 & \\
  \hline
- $N_{00}$ & EC50 term for Hill functions describing likelihood of cytotoxic immune cells locating mutated cells & 8\\
+ $K_1$ & EC50 term for probability of NK cell finding mutant cell & 8 & \\
  \hline
- $N_0$ & EC50 term for Hill function describing negative feedback of total cell population on cell proliferation rates & 80 \\
+  $K_2$ & EC50 term for Treg inhibition of cytotoxic functions & 5 & 0.025  \\
   \hline
-  NK Efficacy Low & weight of NK cell clearing mutated cells during low inflammation & 0.2 \\
+  $K_3$ & EC50 term for how much TGF-$\beta$ each cell has & 200 & \\
   \hline
-  Treg EC50 Low & EC50 term for Hill functions describing Treg-mediated inhibition of efficacy of cytotoxic immune cells during low inflammation & 5 \\
+ $E_\text{NK}$ & rate of NKs clearing mutants & 10 & 30  \\
   \hline
-  CTL Efficacy Low & weight of CTL cell clearing mutated cells during low inflammation & 4 \\
+  $E_\text{CTL}$ & rate of CTLs clearing mutants & 200 & 600 \\
   \hline
-  NK Efficacy Up & proportional increase of NK Efficacy Low during high inflammation & 2 \\
+  $\sigma_\text{NK}$ & NK source rate & 1.3 &  \\ 
   \hline
-  CTL Efficacy Up & proportional increase of CTL Efficacy Low during high inflammation & 3 \\
+  $\sigma_\text{CTL}$ & CTL source rate per cleared mutant cell & 100 & \\ 
   \hline
-  Treg Efficacy Up & during high inflammation, Treg EC50 is Treg EC50 Low / Treg Efficacy Up & 200 \\
+  $\sigma_\text{Treg}$ & Treg source rate per cleared mutant cell & 200 & \\ 
   \hline
-  NK Source Low & rate at which NK cells enter TME during low inflammation & 1.3 \\ 
+  $k_\text{EMT}$ & EMT/MET rate & 0.01 & \\
   \hline
-  CTL Recruitment Low & rate at which CTL cells are recruited after mutated cells are lysed during low inflammation & 100 \\ 
+  $\sigma$ & standard deviation of noise in TGF-$\beta$ each cell receives & 6 & \\
   \hline
-  Treg Recruitment Low & rate at which Treg cells are recruited after mutated cells are lysed during low inflammation & 200 \\
+ $\tau_\text{max}$ & max amount of TGF-$\beta$ any cell can receive & 500 &\\
   \hline 
-  NK Recruitment Up & proportional increase of NK source rate during high inflammation & 1 \\
+ $\tau_\text{MUT}$ & rate of TGF-$\beta$ production by mutant cells & 0.05 & \\
   \hline
-  CTL Recruitment Up & proportional increase of CTL recruitment rate during high inflammation & 1 \\
+ $\tau_\text{Treg}$ & rate of TGF-$\beta$ production by Treg & 0.5 & \\
   \hline
-   Treg Recruitment Up & proportional increase of Treg recruitment rate during high inflammation & 1 \\
+  & RP Cancer Line  & 0.5 &  \\ 
  \hline
-  $p_{mutation}$ Start & maximum probability of mutating after the warmup period ends & 1e-2 \\
+&  INFL High Duration & 30&   \\
  \hline
-  $p_{mutation}$ & increase in probability of a cell mutating after it proliferates but does not mutate & 1e-4 \\
+ & INFL Low Duration &30 &   \\
  \hline
-  Apoptosis Down & proportional decrease in apoptosis weight for cells with apoptosis pathway mutated & 0.3 \\
+& Mes Threshold & 0.7 &    \\
  \hline
-  Immune Evasion & proportional decrease in immune clearance weight for cells with immune pathway mutated & 0.48 \\
+  & maximum initial mutation damage after warmup  & 0.01 &\\
   \hline
-  Proliferation Up & proportional increase in proliferation weight for cells with proliferation pathway mutation & 0.36 \\
-  \hline
-  $k_{EMT}$ & rate parameter controlling speed of EMT and MET & .01 \\
-  \hline
-  c & standard deviation of noise on TGF-$\beta$ received by each cell & 6 \\
-  \hline
-  TGFB Received EC50 & EC50 term for Hill function determining how much TGF-$\beta$ in TME enters a cell & 200 \\
-  \hline
-  TGFB Max & maximum amount of TGF-$\beta$ that can enter a cell from the TME during a cell cycle & 500 \\
-  \hline 
-  Mutant TGFB & amount of TGF-$\beta$ produced by each mutated cell each cycle & 5e-2 \\
-  \hline
-  Treg TGFB & amount of TGF-$\beta$ produced by each Treg cell each cycle & 5e-1 \\
+  & increase in probability to mutate for non-mutating proliferating cells & 0.0001 & \\
   \hline
@@
 
-\newpage
-
- \begin{longtable}{||c | c | c | c | c | c | c | c | c | c ||} 
- \hline
- Name & Base Value & Single Patient & Survival Curve & Vary MIE & Vary MGA & Vary Treg TGFB & Vary MIE and INFL & Vary MGA and INFL & Heat Map\\ [0.5ex] 
- \hline\hline
- RP Cancer Line &  0.5 &  \\ 
- \hline
- MIE  & 0.6 & 0.9 & 0.4-0.9 & 0.7 & 0.5 & 0.9 \\
- \hline
- MGA & 0.2 & & 0.1-0.4 & 0 & 0.2 & 0 \\
- \hline
- INFL High Duration  & 30 & & & & 60 & & & & 60 \\
- \hline
-INFL Low Duration  & 30  \\
- \hline
-Mes Threshold & 0.7 \\
- \hline
- p  & 0.28 \\ 
- \hline
- d  & 0.14 \\
- \hline
- $N_{00}$  & 8\\
- \hline
- $N_0$  & 80 \\
- \hline
-  NK Efficacy Low  & 0.2 & 10 & & 10 & 10 & 10 & 10 & 10 & 10 \\
-  \hline
-  Treg EC50 Low & 5 & & & & & 2.5 \\
-  \hline
-  CTL Efficacy Low  & 4 & 200 & & 200 & 200 & 200 & 200 & 200 & 200\\
-  \hline
-  NK Efficacy Up & 2 & & & & 1.2 & & & &  1.2 \\
-  \hline
-  CTL Efficacy Up & 3 & & & & 3 & & & &  3\\
-  \hline
-  Treg Efficacy Up  & 200 & & & & 10& & & &  10\\
-  \hline
-  NK Source Low  & 1.3 \\ 
-  \hline
-  CTL Recruitment Low  & 100 \\ 
-  \hline
-  Treg Recruitment Low  & 200 \\
-  \hline
-  NK Recruitment Up & 1 & & & & & 3\\
-  \hline
-  CTL Recruitment Up & 1 & & & & & 3\\
-  \hline
-   Treg Recruitment Up & 1 & & & & & 5\\
-  \hline
-  $p_{mutation}$ Start & 1e-2 \\
-  \hline
-  $p_{mutation}$ & 1e-4 \\
-  \hline
-  Apoptosis Down & 0.3 \\
-  \hline
-  Immune Evasion & 0.48 \\
-  \hline
-  Proliferation Up & 0.36  \\
-  \hline
-  $k_{EMT}$  & .01 \\
-  \hline
-  c & 6 \\
-  \hline
-  TGFB Received EC50 & 200 \\
-  \hline
-  TGFB Max & 500 & 700 & & & & & &  & 700\\
-  \hline 
-  Mutant TGFB & 5e-2 & & & & & 0.5\\
-  \hline
-  Treg TGFB & 5e-1 \\
-  \hline
-\end{longtable}
+%\newpage
+%
+%
+% \begin{longtable}{||c | c | c | c | c | c | c | c | c | c ||} 
+% \hline
+% Name & Base Value & Single Patient & Survival Curve & Vary MIE & Vary MGA & Vary Treg TGFB & Vary MIE and INFL & Vary MGA and INFL & Heat Map\\ [0.5ex] 
+% \hline\hline
+% RP Cancer Line &  0.5 &  \\ 
+% \hline
+% MIE  & 0.6 & 0.9 & 0.4-0.9 & 0.7 & 0.5 & 0.9 \\
+% \hline
+% MGA & 0.2 & & 0.1-0.4 & 0 & 0.2 & 0 \\
+% \hline
+% INFL High Duration  & 30 & & & & 60 & & & & 60 \\
+% \hline
+%INFL Low Duration  & 30  \\
+% \hline
+%Mes Threshold & 0.7 \\
+% \hline
+% p  & 0.28 \\ 
+% \hline
+% d  & 0.14 \\
+% \hline
+% $N_{00}$  & 8\\
+% \hline
+% $N_0$  & 80 \\
+% \hline
+%  NK Efficacy Low  & 0.2 & 10 & & 10 & 10 & 10 & 10 & 10 & 10 \\
+%  \hline
+%  Treg EC50 Low & 5 & & & & & 2.5 \\
+%  \hline
+%  CTL Efficacy Low  & 4 & 200 & & 200 & 200 & 200 & 200 & 200 & 200\\
+%  \hline
+%  NK Efficacy Up & 2 & & & & 1.2 & & & &  1.2 \\
+%  \hline
+%  CTL Efficacy Up & 3 & & & & 3 & & & &  3\\
+%  \hline
+%  Treg Efficacy Up  & 200 & & & & 10& & & &  10\\
+%  \hline
+%  NK Source Low  & 1.3 \\ 
+%  \hline
+%  CTL Recruitment Low  & 100 \\ 
+%  \hline
+%  Treg Recruitment Low  & 200 \\
+%  \hline
+%  NK Recruitment Up & 1 & & & & & 3\\
+%  \hline
+%  CTL Recruitment Up & 1 & & & & & 3\\
+%  \hline
+%   Treg Recruitment Up & 1 & & & & & 5\\
+%  \hline
+%  $p_{mutation}$ Start & 1e-2 \\
+%  \hline
+%  $p_{mutation}$ & 1e-4 \\
+%  \hline
+%  Apoptosis Down & 0.3 \\
+%  \hline
+%  Immune Evasion & 0.48 \\
+%  \hline
+%  Proliferation Up & 0.36  \\
+%  \hline
+%  $k_{EMT}$  & .01 \\
+%  \hline
+%  c & 6 \\
+%  \hline
+%  TGFB Received EC50 & 200 \\
+%  \hline
+%  TGFB Max & 500 & 700 & & & & & &  & 700\\
+%  \hline 
+%  Mutant TGFB & 5e-2 & & & & & 0.5\\
+%  \hline
+%  Treg TGFB & 5e-1 \\
+%  \hline
+%\end{longtable}
+%
+%
 

### 2019-04-15 Paper/Paper.tex
update discussion and refs

@@
 
+\begin{figure}
+\center
+{\includegraphics[width=0.99\textwidth]{Figure4/Figure4.pdf}}
+\caption{Effects of inflammation on the time to cancer under different cycling schemes.
+A-B.
+As MIE varies, survival curves (each of 200 patients)  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.
+C-D.
+As MGA varies, survival curves  and corresponding bar plots to summarize the mean time to cancer for each cohort are shown.}
+\label{fig:VaryINFL_and_MesPars}
+\end{figure}
+
 \subsection{A key EMT regime maximizes cancer-free survival time under chronic inflammation}\label{KeyEMT}
@@
 \center
-{\includegraphics[width=0.9\textwidth]{Figure4/Figure4.pdf}}
-\caption{Effects of inflammation on the time to cancer under different cycling schemes.
-A-B.
-As MIE varies, survival curves (each of 200 patients)  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.
-C-D.
-As MGA varies, survival curves  and corresponding bar plots to summarize the mean time to cancer for each cohort are shown.}
-\label{fig:VaryINFL_and_MesPars}
-\end{figure}
-
-\begin{figure}
-\center
-{\includegraphics[width=0.6\textwidth]{Figure5/MIEvsMGA_bigcbar.jpg}}
+{\includegraphics[width=0.5
+  \textwidth]{Figure5/MIEvsMGA_bigcbar.jpg}}
 \caption{Summary of the contrasting effects of MIE and MGA on the Time to Cancer.
@@
 \center
-{\includegraphics[width=\columnwidth]{FigTCGA.pdf}}
+{\includegraphics[width=0.95\columnwidth]{FigTCGA.pdf}}
 \caption{A.
@@
 \section{Discussion}\label{Discussion}
-In building this model, we set out to better understand how the immune system and the EMT spectrum would affect Time to Cancer results from previous work.
-What we found is that altering many of the processes have predictable results: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, and increasing Treg TGF-$\beta$ production all lead to shorter Times to Cancer.
-However, when the inflammation of the TME is accounted for, some of these changes to the system have less predictable outcomes.
-The most intriguing is that periods of low inflammation lead to mesenchymal growth arrest having an optimal value for maximizing Time to Cancer.
+Despite the intense interest in cancer and the immune system, and, separately, in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been any model developed that combines all three of these components: cancer, the immune system, and EMT.
+We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$.
+We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).
+\par
+We found that this model recapitulated cancer-free survival dynamics, and via a global (one-at-a-time) parameter sensitivity analysis, we identified parameters exerting key control over model behavior.
+Focussing on these led us to identify that: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, or increasing Treg TGF-$\beta$ production all lead to shorter Times to Cancer.
+However varying the level of inflammation leads to paradoxical effects: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the Time to Cancer.
+\par
+To test these predictions, we performed unsupervised analysis of pancreatic cancer data from The Cancer Genome Atlas, and looked at survival across two groups: we found that the presence of an EMT signature increased differences in survival between groups in combination with inflammation.
+We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model.
+There are, of course, many tissue trajectories that all result in cancer, and analysis of the transient cell dynamics in tissues bound to become cancerous might tease out
 
+%%%% AM: Yes please add a para on this. 
 %In addition, I could discuss the implications from the TGF-$\beta$ section.
@@
 
-%% AM: no need to separate this I think - combine conclusions into the discussion section. 
-%\section{Conclusion}\label{Conclusion}
-The interactions of the immune system and cancer remain a promising avenue of research.
-Including more diverse aspects of the TME and the tumor itself make the challenge that much harder but also that much more promising.
-As we begin to understand how processes such as EMT factor in to these complex interactions, we can begin to respond clinically to better improve patient outcomes.
-Here, we saw {\it in silico} evidence of an optimal value for a property of mesenchymal cells that would maximize Time to Cancer.
-There is still much work to be done further isolating this effect in computer simulations and further verifying this effect and exploring further ways EMT can influence tumor progression.
-This model explored much more than mesenchymal growth arrest, so a simpler model could make this point clearer.
-On the other hand, models which incorporate more distinctions between epithelial and mesenchymal cells could give better clarity on the relative importance of these and reveal a fuller picture about how a mesenchymal subpopulation affects cancer outcomes.
-%A Adam, do you know of any?
-The authors are not aware of current treatments that specifically target mesenchymal growth arrest, influencing it in one direction or another.
-%
-Based on these results, this could prove to be a promising direction for research, in particular for cancers in which it is known that mesenchymal cells are present and active within the tumor.
-The greatest challenge in translating the information here to the clinic, apart from creating the drugs, would be discovering the ideal growth rate for mesenchymal cells within each patient and how other patient-specific data will influence this result.
-\par
+%%% AM: please expand on this. how do we give it a larger role? What does "more robust" mean? how to further explore the anti-inflamm function of the Tregs? 
 Also, since the inflammatory state of the TME played such a large role in our simulations, making this a more robust facet of the model by allowing the tumor and immune system to influence the inflammatory state could give the model higher fidelity.
 One intriguing aspect could be the anti-inflammatory function of Tregs as they seek to restore homeostasis.
-\par
-A final future direction would be continuing to explore the heterogeneity of the tumor and how this helps the tumor evade the immune system and grow into a malignancy. 
-As we understand more and more about the diversity within the TME, additional models will be instrumental in determining the scope of these implications and the possible ways to address them.
-\par
-Progress in cancer treatment has always proven attainable even despite great challenges, for which the complexity of the disease is often responsible.
-As we move forwards, it is these very complexities that we will be able to better exploit to eradicate or control the disease.
+
+Another important point for future work is to further explore (and exploit) the heterogeneity of the tumor to understand how this helps the tumor evade the immune system.
+It is our view that the role of decanalization \cite{gibson09_decanalization} in during cancer incidence and progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour}.
+Progress in cancer therapies has proven to be attainable even despite great challenges, for which the complexity of the disease is often in large part responsible.
+As we move towards a new generation of immunotherapies, it is these very complexities that we must better understand in order to eradicate or control the disease.
 

### 2019-04-18 Paper/Paper.tex
update abstract
Discussion Update

@@
 Inflammation also has significant effects on gastrointestinal and pancreatic tumors.
-% These are regulated by the tumor microenvironment (TME) in myriad ways.
 Moreover, epithelial cell plasticity can significantly alter disease trajectories (for better or worse) via epithelial-to-mesenchymal transition (EMT).
 Several of the same pathways that regulate EMT are involved in tumor-immune interactions.
-%In turn, both the tumor and the immune system exert substantial influence over the TME in competition to influence the fate of the host.
 Yet little is known about the mechanisms and consequences of these regulatory processes.
@@
 We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model.
-There are, of course, many tissue trajectories that all result in cancer, and analysis of the transient cell dynamics in tissues bound to become cancerous might tease out
+There are, of course, many tissue trajectories that all result in cancer, and analysis of the transient cell dynamics in tissues bound to become cancerous might tease out \tcr{which are indicators of cancer.}
+\par
 
@@
 %In light of this, it might be good to include a new section on how TGF-$\beta$ production's influence on Time to Cancer is influenced by the inflammatory cycling scheme.
-
+These promising results lead us to believe that continuing to study and improve the model will lead to further understanding of the competing roles of the immune system and EMT on epithelial cancers.
+Developing the inflammation module of our model stands out as an attractive next step.
+This is because we have seen that the inflammatory state of the TME had some of the most significant effects on Time to Cancer (see Figures \ref{fig:MOAT} and \ref{fig:fig:VaryINFL_and_MesPars}BD).
+The cycling nature of the scheme we chose is meant to synthesize the disparate inputs that affect the local tissue and output a simple and interpretable result.
+However, many of the agents explicitly considered in the model actively contribute to the inflammatory state of the TME, providing a basis for connecting these two pieces of our model so that, for example, as mutant cells start growing, inflammation begins to rise.
+In this way, a more robust inflammatory state could be modeled that would allow for more informative analysis.
+\par
+Coupling nicely with this improvement, the natural anti-inflammatory role of Tregs would give another layer to the competing roles of the immune system.
+One consequence of our current model is that it appears that fewer Tregs means longer Time to Cancer, implying that simply eliminating Tregs is best.
+Adding in this main effector function of Tregs into the model might help change this oversimplified analysis and add depth to the understanding of Tregs we glean from the model.
+%Also, since the inflammatory state of the TME played such a large role in our simulations, making this a more robust facet of the model by allowing the tumor and immune system to influence the inflammatory state could give the model higher fidelity.
+%One intriguing aspect could be the anti-inflammatory function of Tregs as they seek to restore homeostasis.
+\par
+In addition, the role that TGF-$\beta$ plays throughout the TME sets it up as a prime target of research.
+We saw in Figure \ref{fig:FirstSurvivalCurves}E that below some threshold, reduction of TGF-$\beta$ increases Time to Cancer.
+Thus, our model suggests that reducing quantities of TGF-$\beta$ in the TME will benefit patient survival.
+However, it must be noted that TGF-$\beta$ is a powerful cytokine implicated in many cellular processes and attempts to shift the concentrations of TGF-$\beta$ in the TME or throughout the body could have undesirable side effects.
+With an aggressive disease like cancer, though, more aggressive therapies are warranted.
+\par
 %%% AM: please expand on this. how do we give it a larger role? What does "more robust" mean? how to further explore the anti-inflamm function of the Tregs? 
-Also, since the inflammatory state of the TME played such a large role in our simulations, making this a more robust facet of the model by allowing the tumor and immune system to influence the inflammatory state could give the model higher fidelity.
-One intriguing aspect could be the anti-inflammatory function of Tregs as they seek to restore homeostasis.
-
-Another important point for future work is to further explore (and exploit) the heterogeneity of the tumor to understand how this helps the tumor evade the immune system.
+\par
+Finally, Another important point for future work is to further explore (and exploit) the heterogeneity of the tumor to understand how this helps the tumor evade the immune system.
 It is our view that the role of decanalization \cite{gibson09_decanalization} in during cancer incidence and progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour}.

### 2019-04-18 Paper/Paper.tex
update discussion

@@
 While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
-In Figure \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
+In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
 Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.
@@
 \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of potentially mutated cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
-These results are summarized by Figure \ref{fig:FirstSurvivalCurves}A.
+These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.
 
 Second, as the rate of mesenchymal growth arrest (MGA) increases, Time to Cancer increases, i.e.
-lower proliferation rates for mesenchymal cells slow down cancer growth (Figure \ref{fig:FirstSurvivalCurves}B).
+lower proliferation rates for mesenchymal cells slow down cancer growth (Fig.
+\ref{fig:FirstSurvivalCurves}B).
 This is not immediately intuitive, since the decreased proliferation rate affects both mutated and non-mutated tissue cells.
@@
 Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
-In Figure \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
+In Fig. \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
 The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tissue cells to a mesenchymal phenotype.
@@
 \section{Discussion}\label{Discussion}
-Despite the intense interest in cancer and the immune system, and, separately, in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been any model developed that combines all three of these components: cancer, the immune system, and EMT.
+Despite the intense interest in cancer and the immune system, and in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these components: cancer, the immune system, and EMT.
 We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$.
@@
 \par
-We found that this model recapitulated cancer-free survival dynamics, and via a global (one-at-a-time) parameter sensitivity analysis, we identified parameters exerting key control over model behavior.
-Focussing on these led us to identify that: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, or increasing Treg TGF-$\beta$ production all lead to shorter Times to Cancer.
+We found that this model recapitulated cancer-free survival dynamics, and via global parameter sensitivity analysis, we identified parameters exerting key control over model behavior.
+Focusing on these led us to identify that: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, or increasing Treg TGF-$\beta$ production all lead to shorter Times to Cancer.
 However varying the level of inflammation leads to paradoxical effects: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the Time to Cancer.
@@
 We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model.
-There are, of course, many tissue trajectories that all result in cancer, and analysis of the transient cell dynamics in tissues bound to become cancerous might tease out \tcr{which are indicators of cancer.}
+There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in pre-cancerous tissues is a pressing need to shed insight into cellular biomarkers of cancer.
 \par
@@
 %In light of this, it might be good to include a new section on how TGF-$\beta$ production's influence on Time to Cancer is influenced by the inflammatory cycling scheme.
-These promising results lead us to believe that continuing to study and improve the model will lead to further understanding of the competing roles of the immune system and EMT on epithelial cancers.
-Developing the inflammation module of our model stands out as an attractive next step.
-This is because we have seen that the inflammatory state of the TME had some of the most significant effects on Time to Cancer (see Figures \ref{fig:MOAT} and \ref{fig:fig:VaryINFL_and_MesPars}BD).
+
+These results represent promising steps in understanding the competing roles of the immune system and EMT during (pre-) development of epithelial cancers, yet much remains to be done.
+Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs.
+\ref{fig:MOAT} and \ref{fig:fig:VaryINFL_and_MesPars}B-D).
 The cycling nature of the scheme we chose is meant to synthesize the disparate inputs that affect the local tissue and output a simple and interpretable result.
@@
 In addition, the role that TGF-$\beta$ plays throughout the TME sets it up as a prime target of research.
-We saw in Figure \ref{fig:FirstSurvivalCurves}E that below some threshold, reduction of TGF-$\beta$ increases Time to Cancer.
+We saw in Fig. \ref{fig:FirstSurvivalCurves}E that below some threshold, reduction of TGF-$\beta$ increases Time to Cancer.
 Thus, our model suggests that reducing quantities of TGF-$\beta$ in the TME will benefit patient survival.

### 2019-04-19 Paper/Paper.tex
updates to discussion

@@
 Preceding and during cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor incidence and prognosis.
-Inflammation also has significant effects on gastrointestinal and pancreatic tumors.
+Inflammation has particularly significant effects on gastrointestinal and pancreatic tumors.
 Moreover, epithelial cell plasticity can significantly alter disease trajectories (for better or worse) via epithelial-to-mesenchymal transition (EMT).
-Several of the same pathways that regulate EMT are involved in tumor-immune interactions.
-Yet little is known about the mechanisms and consequences of these regulatory processes.
-Here we introduce a multiscale agent-based model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial tissues.
-Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that increase cancer-free survival.
+Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of these regulatory processes.
+Here we introduce a multiscale evolutionary model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial tissues.
+Through {\em in silico} analyses of large patient cohorts, we found controllable regimes that maximise cancer-free survival.
 Delaying tumorigenesis depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
@@
 We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer.
-These results offer novel means to delay disease onset by controlling properties of EMT and demonstrate the importance of studying cancer-immune interactions in the light of EMT.
-%The model presented provides a base for future studies into the intricate regulation of tumorigenesis.
+These results offer novel means to delay disease onset by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in the light of EMT.
 
@@
 \ref{fig:MOAT} and \ref{fig:fig:VaryINFL_and_MesPars}B-D).
-The cycling nature of the scheme we chose is meant to synthesize the disparate inputs that affect the local tissue and output a simple and interpretable result.
-However, many of the agents explicitly considered in the model actively contribute to the inflammatory state of the TME, providing a basis for connecting these two pieces of our model so that, for example, as mutant cells start growing, inflammation begins to rise.
-In this way, a more robust inflammatory state could be modeled that would allow for more informative analysis.
-\par
-Coupling nicely with this improvement, the natural anti-inflammatory role of Tregs would give another layer to the competing roles of the immune system.
-One consequence of our current model is that it appears that fewer Tregs means longer Time to Cancer, implying that simply eliminating Tregs is best.
-Adding in this main effector function of Tregs into the model might help change this oversimplified analysis and add depth to the understanding of Tregs we glean from the model.
+Currently, inflammation is modeled as independently cycling between high and low schemes, however many of the agents considered in the model actively contribute to the inflammatory state, thus it would be interesting to consider connecting these components (for example by letting the level of inflammation be conditional on the number of mutant cells in the tissue).
+Another layer of complexity is revealed by the natural anti-inflammatory role of Tregs.
+One consequence of the current model is that decreasing the number of Tregs increases the cancer-free survival.
+Clearly, there exists a trade-off to be accounted for, and adding to the model the main effector function of Tregs could remedy this and add depth to our understanding of the various roles that Tregs play in and around the tumor.
+
 %Also, since the inflammatory state of the TME played such a large role in our simulations, making this a more robust facet of the model by allowing the tumor and immune system to influence the inflammatory state could give the model higher fidelity.
@@
 \par
-In addition, the role that TGF-$\beta$ plays throughout the TME sets it up as a prime target of research.
-We saw in Fig. \ref{fig:FirstSurvivalCurves}E that below some threshold, reduction of TGF-$\beta$ increases Time to Cancer.
-Thus, our model suggests that reducing quantities of TGF-$\beta$ in the TME will benefit patient survival.
-However, it must be noted that TGF-$\beta$ is a powerful cytokine implicated in many cellular processes and attempts to shift the concentrations of TGF-$\beta$ in the TME or throughout the body could have undesirable side effects.
-With an aggressive disease like cancer, though, more aggressive therapies are warranted.
+The role that TGF-$\beta$ plays throughout the tumor and its microenvironment also requires further research.
+We found that, below a threshold, reduction of TGF-$\beta$ increases the Time to Cancer  (Fig.
+\ref{fig:FirstSurvivalCurves}E), thus reducing expression levels of TGF-$\beta$ in the tumor microenvironment benefits survival.
+However, TGF-$\beta$ is a master regulator implicated in numerous cellular signaling processes, and changing the concentrations of TGF-$\beta$ even in a local tumor microenvironment could have large off-target effects.
+Thus it might be more fruitful to target signaling factors downstream of TGF-$\beta$ that still have the ability to modulate epithelial cell dynamics: this is an area of active and future investigation, we are currently developing a larger TGF-$\beta$ signaling pathway module with appropriate crosstalks to  epithelial/mesenchymal/immune cell functions to be incorporated into the model.
 \par
-%%% AM: please expand on this. how do we give it a larger role? What does "more robust" mean? how to further explore the anti-inflamm function of the Tregs? 
-\par
-Finally, Another important point for future work is to further explore (and exploit) the heterogeneity of the tumor to understand how this helps the tumor evade the immune system.
-It is our view that the role of decanalization \cite{gibson09_decanalization} in during cancer incidence and progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour}.
-Progress in cancer therapies has proven to be attainable even despite great challenges, for which the complexity of the disease is often in large part responsible.
-As we move towards a new generation of immunotherapies, it is these very complexities that we must better understand in order to eradicate or control the disease.
+A further goal for future work is to explore (and exploit) the heterogeneity of tumor evolution in greater depth: this heterogeneity aids the evasion of the tumor from immune effects.
+Studying the consequences of decanalization \cite{gibson09_decanalization} during cancer progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour, punt17_tumour, dagogo-jack18_tumour}.
+Yet despite these challenges, for which the complexity of the disease may be often in large part responsible, great progress has been and continues to be made.
+As we approach a new generation of immunotherapies, it is these very complexities that we must better understand in order to control or eradicate the disease.
 

### 2019-04-19 Paper/Paper.tex
Small Edits and a TGFB Source

--- 2019-04-18 Paper/Paper.tex
@@
 The effects of the immune system on a tumor can be broadly summarized into two modules.
-The cytotoxic branch of the immune system (such as natural killer (NK) cells and cytotoxic T cells (CTLs) seek out and lyse tumor cells.
+The cytotoxic branch of the immune system, such as natural killer cells (NKs) and cytotoxic T cells (CTLs), seek out and lyse tumor cells.
 Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely\tcr{[REF]}.
-The regulatory branch of the immune system (T regulatory cells (Tregs) and other factors), inhibits the effective functioning of the cytotoxic branch\cite{ruffell2010lymphocytes}.
+The regulatory branch of the immune system, such as T regulatory cells (Tregs) and other factors, inhibits the effective functioning of the cytotoxic branch\cite{ruffell2010lymphocytes}.
 %Make sure there's some ref for immune clearing tumor
@@
 The agents in the model are the tissue cells with tumorigenic potential.
-These can either be mutation-free, or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).
+These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).
 If the pathway is mutated, the component is colored red and is marked with `X'.
@@
 We model immune cells as continuous variables, appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.
-The cytokine TGF-$\beta$ is continuous within the tumor microenvironment (TME), i.e. the system is well-mixed.
+The cytokine TGF-$\beta$ is continuous within the TME, i.e. the system is well-mixed.
 Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
@@
 After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
-From then on, a cell which proliferates either mutates or increases its probability of mutating later on.
+From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.
 
@@
 \paragraph{Proliferation}
-There are four factors that contribute to the probability that a cell will proliferate.
+There are four factors that contribute to the weight of a cell to proliferate.
 The first is a base proliferation rate that all cells have, $p$.
@@
 This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
-Second, all proliferating cells have a nonzero probability of undergoing a pathway mutation.
+Second, all proliferating cells have a cell-specific probability of undergoing a pathway mutation.
 If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
@@
 The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.
-Then, some randomness is introduced by the independent and identically distributed variables $X_i$where $\sigma$ is a parameter in the model.
+Then, some randomness is introduced by the independent and identically distributed variables $X_i$ where $\sigma$ is a parameter in the model.
 If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
@@
 Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
-Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: quantity of TGF-$\beta$ in the system and the current EMT score of the cell.
+Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: concentration of TGF-$\beta$ in the system and the current EMT score of the cell.
 
@@
 \begin{split}
-N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}\text{CTL} \\
+N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}N_\text{CTL} \\
 \text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
@@
 N_\text{CTL}(k+1) & =  \left (N_\text{CTL}(k)-\sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\\
-N_\text{Treg}(k+1) & =  \left (N_\text{Treg}(k)-\frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau}{1+\tau/K_4}\right )\exp(-d_{\text{Treg}}\Delta t)\\
-&+ \frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau}{1+\tau/K_4}
+N_\text{Treg}(k+1) & =  \left (N_\text{Treg}(k)-\frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}\right )\exp(-d_{\text{Treg}}\Delta t)\\
+&+ \frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}
 \end{split}
@@
 [I don't get this: what is y-axis?]}
-The algorithm the average of the absolute change of the output, which in our model is the area under the survival curve.
-This output, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.
+The algorithm computes the average of the absolute change of the output, which in our model is the area under the survival curve.
+This output out the algorithm, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.
 
@@
 NK cells and CTLs attempt to clear mutated tissue cells, and deactivate upon successfully carrying out their cytotoxic function.
-T regulatory cells (Tregs) inhibit cytotoxic activity.
+Tregs inhibit cytotoxic activity.
 Tregs also release TGF-$\beta$ which promotes the recruitment of Tregs, as well as increasing the rate of EMT.
@@
 The number of mesenchymal cells increases initially with increasing TGF-$\beta$, but decreases after the clone takes over the tissue.
-This shifts back towards an epithelial tissue phenotype occurs even though the expression of TGF-$\beta$ remains high.
+This shift back towards an epithelial tissue phenotype occurs even though the expression of TGF-$\beta$ remains high.
 After 841 cell cycles, the proportion of mutated cells reaches 50\%: this is modeled as the threshold defining tumorigenesis. Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days. 
@@
 To assess the sensitivity of the model parameters and thus identify those that are most important in determining cancer-free survival times, we applied Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods) to the model.
-The results of Morris OAT on the 31 model parameters are shown in \ref{fig:MOAT}: a subset of parameters demonstrate much higher levels of sensitivity than others.
+The results of Morris OAT on the 31 model parameters are shown in Figure \ref{fig:MOAT}: a subset of parameters demonstrate much higher levels of sensitivity than others.
 The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
@@
 \caption{Effects of mesencymal cell properites on the time to cancer.
-Trajectories of one patient per cohort from warmup period ([-1000,0]) to 2000 cell cycles, for $\Delta_\text{MIE}$ (A); $\Delta_\text{MGA}$ (B); and $\tau_\text{Treg}$, Treg TGF-$\beta$ (C).
+Trajectories of one patient per cohort from warmup period ([-1000,0]) to 2000 cell cycles, for $\Delta_\text{MIE}$ (A); $\Delta_\text{MGA}$ (B); and $\tau_\text{Treg}$ (C).
 D. Survival curves corresponding to changes in MIE (A) for a patient cohort of size 1000; shaded region represents the 95\% confidence interval for the evaluated function. 
@@
 Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
-MIE is the proportional reduction in the probability that a mutated mesenchymal cell will be subject to immune clearance in a given cycle.
-MGA is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
+The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a mutated mesenchymal cell will be subject to immune clearance in a given cycle.
+The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
 The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT).
@@
 \par 
-As the level of mesenchymal immune evasion (MIE) is increased, the Time to Cancer decreases (Fig.
+As MIE is increased, the Time to Cancer decreases (Fig.
 \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of potentially mutated cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
@@
 
-Second, as the rate of mesenchymal growth arrest (MGA) increases, Time to Cancer increases, i.e.
-lower proliferation rates for mesenchymal cells slow down cancer growth (Figure \ref{fig:FirstSurvivalCurves}B).
+Second, as MGA increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Figure \ref{fig:FirstSurvivalCurves}B).
 This is not immediately intuitive, since the decreased proliferation rate affects both mutated and non-mutated tissue cells.
@@
 For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and the Time to Cancer is monotonic, i.e.
-increasing either the rate of mesenchymal immune evasion (MIE) (Fig.
-\ref{fig:VaryINFL_and_MesPars}A-B) or of mesenchymal growth arrest (MGA)  (Fig.
+increasing either MIE (Fig.
+\ref{fig:VaryINFL_and_MesPars}A-B) or MGA (Fig.
 \ref{fig:VaryINFL_and_MesPars}C-D) decreases the Time to Cancer.
-However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the Time to Cancer is found with respect to the relative mesenchymal growth rate (MGA $= 0.3$).
+However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the Time to Cancer is found with respect to MGA ($\Delta_\text{MGA}= 0.3$).
 This is seen both for intermittent or low inflammation (Fig.
@@
 
-To further assess the effects on tumorigenesis of the mesenchymal properties of immune evasion (MIE) and growth arrest (MGA), we study the joint density plot for each of these parameters against the Time to Cancer (Fig.
+To further assess the effects on tumorigenesis of the mesenchymal properties of immune evasion and growth arrest, we study the joint density plot for each of these parameters against the Time to Cancer (Fig.
 \ref{fig:MIEvsMGA}).
-We found that over the full range of values of MGA considered, increasing the MIE decreases the Time to Cancer.
+We found that over the full range of values of $\Delta_\text{MGA}$ considered, increasing $\Delta_\text{MIE}$ decreases the Time to Cancer.
 However, for any given value of mesenchymal immune evasion, there is a value of mesenchymal growth arrest that maximizes the Time to Cancer.
@@
 Developing the inflammation module of our model stands out as an attractive next step.
-This is because we have seen that the inflammatory state of the TME had some of the most significant effects on Time to Cancer (see Figures \ref{fig:MOAT} and \ref{fig:fig:VaryINFL_and_MesPars}BD).
+This is because we have seen that the inflammatory state of the TME had some of the most significant effects on Time to Cancer (see Figures \ref{fig:MOAT} and \ref{fig:VaryINFL_and_MesPars}B, D).
 The cycling nature of the scheme we chose is meant to synthesize the disparate inputs that affect the local tissue and output a simple and interpretable result.
@@
 Thus, our model suggests that reducing quantities of TGF-$\beta$ in the TME will benefit patient survival.
-However, it must be noted that TGF-$\beta$ is a powerful cytokine implicated in many cellular processes and attempts to shift the concentrations of TGF-$\beta$ in the TME or throughout the body could have undesirable side effects.
+However, it must be noted that TGF-$\beta$ is a powerful cytokine implicated in many cellular processes and attempts to shift the concentrations of TGF-$\beta$ in the TME or throughout the body could have undesirable side effects \cite{blobe2000role}.
 With an aggressive disease like cancer, though, more aggressive therapies are warranted.
@@
 Finally, Another important point for future work is to further explore (and exploit) the heterogeneity of the tumor to understand how this helps the tumor evade the immune system.
-It is our view that the role of decanalization \cite{gibson09_decanalization} in during cancer incidence and progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour}.
+It is our view that the role of decanalization \cite{gibson09_decanalization} during cancer incidence and progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour}.
 Progress in cancer therapies has proven to be attainable even despite great challenges, for which the complexity of the disease is often in large part responsible.

### 2019-04-19 Paper/Supplement/Supplement.tex
Names Supplementary Tables

@@
 
+\begin{table}[H]
 \begin{center}
- \begin{longtable}{|| c | c||} 
+ \begin{tabular}{|| c | c||} 
  \hline
@@
   \hline
-\end{longtable}
+\end{tabular}
+  \caption{The model parameter names and descriptions. Note that many of these values are affected by the inflammation state of the system.}
 \end{center}
+\end{table}
 
+\begin{table}[H]
 \begin{center}
- \begin{longtable}{|| c | p{7cm} | c | c ||} 
+ \begin{tabular}{|| c | p{7cm} | c | c ||} 
  \hline
@@
   \hline
-\end{longtable}
+\end{tabular}
+  \caption{The model parameter names, descriptions, and values during both low and high inflammation. Parameters with only one value do not change with the inflammatory state.}
 \end{center}
+\end{table}
 

### 2019-04-19 Paper/Paper.tex
minor changes throughout

--- 2019-04-19 Paper/Paper.tex
@@
 Cancer and the immune system interact in myriad ways.
-Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME), indeed, immune cells can be considered as a part of the TME: responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}.
+Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME): responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}.
 Recent breakthroughs in the field of immunotherapy are beginning to produce therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}.
@@
 This is accomplished by the presentation of antigens recognized by innate immune cells that are transported to lymph nodes where T cells (and other components) can be activated \cite{schreiber11_cancer}.
-The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via Treg cell activation \cite{schreiber11_cancer}.
+The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via activation of T regulatory cells (Tregs)  \cite{schreiber11_cancer}.
 \par
@@
 The cytotoxic branch of the immune system (such as natural killer (NK) cells and cytotoxic T cells (CTLs) seek out and lyse tumor cells.
-Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely\tcr{[REF]}.
-The regulatory branch of the immune system (T regulatory cells (Tregs) and other factors), inhibits the effective functioning of the cytotoxic branch\cite{ruffell2010lymphocytes}.
+Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely \cite{finn12_immunooncology-1}.
+The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch\cite{ruffell2010lymphocytes}.
 %Make sure there's some ref for immune clearing tumor
@@
 \par
-The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
-Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions \tcr{[REF]}.
-While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate\cite{woods2014effects}.
-Cancer cells are well-known for achieving unregulated proliferation, so exploring a subset of cancer cells with decreased proliferation could provide intriguing new results.
-In our model, we refer to this property as mesenchymal growth arrest, or MGA.
+The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculated to exhibit a more ``stemness'' phenotype.
+%Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions \tcr{[REF]}.
+While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate  \cite{woods2014effects} (which, under certain conditions can be called quiescence, though we avoid this term as it can be controversial, and is not a necessary assumption to the model developed below).
+Cancer cells are of course characterized by some degree of unregulated proliferation, thus it is appealing to explore the effects of differential proliferation rates within the cancer cell population.
+In our model, we refer to the reduced proliferative capacity associated with a mesenchymal cell phenotype as mesenchymal growth arrest (MGA).
 \par
@@
 Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
-We seek to describe a set of crucial molecular and cellular interactions in (epithelial) tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will occur and, if so, when.
-A recent model of cancer-immune interactions \cite{guo2017multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as the EMT component to the model.
+We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will occur and, if so, when.
+A recent model of cancer-immune interactions \cite{guo2017multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
 We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
-In the next section we develop the model modularly to explain the intuition behind each component, and then go on to analyze its behavior.
+\par 
+In the next section we develop the model and give explanation of the intuition behind each component.
+We then go on to analyze its behavior.
 Global sensitivity analysis identifies parameters that are crucial for tumorigenesis.
@@
 %                          RESULTS                         %
-%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Results}
+%%%%%%%%%%%%%%%%%%%%%%%\section{Results}
 
@@
 \center 
-{\includegraphics[width=0.8\textwidth]{Figure1/Figure1.pdf}}
+{\includegraphics[width=0.98\textwidth]{Figure1/Figure1.pdf}}
 \caption{A.
@@
 Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs.
-\ref{fig:MOAT} and \ref{fig:fig:VaryINFL_and_MesPars}B-D).
+\ref{fig:MOAT} and \ref{fig:VaryINFL_and_MesPars}B-D).
 Currently, inflammation is modeled as independently cycling between high and low schemes, however many of the agents considered in the model actively contribute to the inflammatory state, thus it would be interesting to consider connecting these components (for example by letting the level of inflammation be conditional on the number of mutant cells in the tissue).

### 2019-04-19 Paper/Paper.tex
First Intro Para's Updates to makes Adam's fixes

--- 2019-04-19 Paper/Paper.tex
@@
 Moreover, epithelial cell plasticity can significantly alter disease trajectories (for better or worse) via epithelial-to-mesenchymal transition (EMT).
-Several of the same pathways that regulate EMT are involved in tumor-immune interactions.
-Yet little is known about the mechanisms and consequences of these regulatory processes.
-Here we introduce a multiscale agent-based model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial tissues.
-Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that increase cancer-free survival.
+Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of these regulatory processes.
+Here we introduce a multiscale evolutionary model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial tissues.
+Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that maximize cancer-free survival.
 Delaying tumorigenesis depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
@@
 We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer.
-These results offer novel means to delay disease onset by controlling properties of EMT and demonstrate the importance of studying cancer-immune interactions in the light of EMT.
+These results offer novel means to delay disease onset by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in the light of EMT.
 %The model presented provides a base for future studies into the intricate regulation of tumorigenesis.

### 2019-04-19 Paper/Paper.tex
Continuing to try match Adam's

@@
 This is accomplished by the presentation of antigens recognized by innate immune cells that are transported to lymph nodes where T cells (and other components) can be activated \cite{schreiber11_cancer}.
-The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via Treg cell activation \cite{schreiber11_cancer}.
+The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via activation of T regulatory cells (Tregs) \cite{schreiber11_cancer}.
 \par
@@
 The cytotoxic branch of the immune system, such as natural killer cells (NKs) and cytotoxic T cells (CTLs), seek out and lyse tumor cells.
-Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely\tcr{[REF]}.
-The regulatory branch of the immune system, such as T regulatory cells (Tregs) and other factors, inhibits the effective functioning of the cytotoxic branch\cite{ruffell2010lymphocytes}.
+Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely \cite{finn12_immunooncology-1}.
+The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
 %Make sure there's some ref for immune clearing tumor

### 2019-04-19 Paper/Paper.tex
Still making Adam's changes...

@@
 \par
-The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells often express a more stem-like phenotype.
-Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions \tcr{[REF]}.
-While many qualities are associated with being stem-like, we choose to focus on a reduced proliferation rate\cite{woods2014effects}.
-Cancer cells are well-known for achieving unregulated proliferation, so exploring a subset of cancer cells with decreased proliferation could provide intriguing new results.
-In our model, we refer to this property as mesenchymal growth arrest, or MGA.
+The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculate to exhibit a more ``stemness'' phenotype.
+%Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions \tcr{[REF]}.
+While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects} (which, under certain conditions can be called quiescence, though we avoid this term as it can be controversial, and is not a necessary assumption to the model developed below).
+Cancer cells are of course characterized by some degree of unregulated proliferation, thus it is appealing to explore the effects of differential proliferation rates within the cancer cell population.
+In our model, we refer to the reduced proliferative capacity associated with a mesenchymal cell phenotype as mesenchymal growth arrest (MGA).
 \par

### 2019-04-19 Paper/Paper.tex
Should be done with manual merge now

@@
 \center
-{\includegraphics[width=0.8\textwidth]{Figure1/Figure1.pdf}}
+{\includegraphics[width=0.98\textwidth]{Figure1/Figure1.pdf}}
 \caption{A.
@@
 \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of potentially mutated cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
-These results are summarized by Figure \ref{fig:FirstSurvivalCurves}A.
+These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.
 
-Second, as MGA increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Figure \ref{fig:FirstSurvivalCurves}B).
+Second, as MGA increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Fig \ref{fig:FirstSurvivalCurves}B).
 This is not immediately intuitive, since the decreased proliferation rate affects both mutated and non-mutated tissue cells.
@@
 Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
-In Figure \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
+In Fig \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
 The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tissue cells to a mesenchymal phenotype.
@@
 %In light of this, it might be good to include a new section on how TGF-$\beta$ production's influence on Time to Cancer is influenced by the inflammatory cycling scheme.
-These promising results lead us to believe that continuing to study and improve the model will lead to further understanding of the competing roles of the immune system and EMT on epithelial cancers.
-Developing the inflammation module of our model stands out as an attractive next step.
-This is because we have seen that the inflammatory state of the TME had some of the most significant effects on Time to Cancer (see Figures \ref{fig:MOAT} and \ref{fig:VaryINFL_and_MesPars}B, D).
-The cycling nature of the scheme we chose is meant to synthesize the disparate inputs that affect the local tissue and output a simple and interpretable result.
-However, many of the agents explicitly considered in the model actively contribute to the inflammatory state of the TME, providing a basis for connecting these two pieces of our model so that, for example, as mutant cells start growing, inflammation begins to rise.
-In this way, a more robust inflammatory state could be modeled that would allow for more informative analysis.
-\par
-Coupling nicely with this improvement, the natural anti-inflammatory role of Tregs would give another layer to the competing roles of the immune system.
-One consequence of our current model is that it appears that fewer Tregs means longer Time to Cancer, implying that simply eliminating Tregs is best.
-Adding in this main effector function of Tregs into the model might help change this oversimplified analysis and add depth to the understanding of Tregs we glean from the model.
+
+These results represent promising steps in understanding the competing roles of the immune system and EMT during (pre-) development of epithelial cancers, yet much remains to be done.
+Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs.
+\ref{fig:MOAT} and \ref{fig:VaryINFL_and_MesPars}B, D).
+Currently, inflammation is modeled as independently cycling between high and low schemes, however many of the agents considered in the model actively contribute to the inflammatory state, thus it would be interesting to consider connecting these components (for example by letting the level of inflammation be conditional on the number of mutant cells in the tissue).
+Another layer of complexity is revealed by the natural anti-inflammatory role of Tregs.
+One consequence of the current model is that decreasing the number of Tregs increases the cancer-free survival.
+Clearly, there exists a trade-off to be accounted for, and adding to the model the main effector function of Tregs could remedy this and add depth to our understanding of the various roles that Tregs play in and around the tumor.
+
 %Also, since the inflammatory state of the TME played such a large role in our simulations, making this a more robust facet of the model by allowing the tumor and immune system to influence the inflammatory state could give the model higher fidelity.
@@
 \par
-In addition, the role that TGF-$\beta$ plays throughout the TME sets it up as a prime target of research.
-We saw in Figure \ref{fig:FirstSurvivalCurves}E that below some threshold, reduction of TGF-$\beta$ increases Time to Cancer.
-Thus, our model suggests that reducing quantities of TGF-$\beta$ in the TME will benefit patient survival.
-However, it must be noted that TGF-$\beta$ is a powerful cytokine implicated in many cellular processes and attempts to shift the concentrations of TGF-$\beta$ in the TME or throughout the body could have undesirable side effects \cite{blobe2000role}.
-With an aggressive disease like cancer, though, more aggressive therapies are warranted.
+The role that TGF-$\beta$ plays throughout the tumor and its microenvironment also requires further research.
+We found that, below a threshold, reduction of TGF-$\beta$ increases the Time to Cancer  (Fig.
+\ref{fig:FirstSurvivalCurves}E), thus reducing expression levels of TGF-$\beta$ in the tumor microenvironment benefits survival.
+However, TGF-$\beta$ is a master regulator implicated in numerous cellular signaling processes, and changing the concentrations of TGF-$\beta$ even in a local tumor microenvironment could have large off-target effects.
+Thus it might be more fruitful to target signaling factors downstream of TGF-$\beta$ that still have the ability to modulate epithelial cell dynamics: this is an area of active and future investigation, we are currently developing a larger TGF-$\beta$ signaling pathway module with appropriate crosstalks to  epithelial/mesenchymal/immune cell functions to be incorporated into the model.
 \par
-%%% AM: please expand on this. how do we give it a larger role? What does "more robust" mean? how to further explore the anti-inflamm function of the Tregs? 
-\par
-Finally, Another important point for future work is to further explore (and exploit) the heterogeneity of the tumor to understand how this helps the tumor evade the immune system.
-It is our view that the role of decanalization \cite{gibson09_decanalization} during cancer incidence and progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour}.
-Progress in cancer therapies has proven to be attainable even despite great challenges, for which the complexity of the disease is often in large part responsible.
-As we move towards a new generation of immunotherapies, it is these very complexities that we must better understand in order to eradicate or control the disease.
+A further goal for future work is to explore (and exploit) the heterogeneity of tumor evolution in greater depth: this heterogeneity aids the evasion of the tumor from immune effects.
+Studying the consequences of decanalization \cite{gibson09_decanalization} during cancer progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour, punt17_tumour, dagogo-jack18_tumour}.
+Yet despite these challenges, for which the complexity of the disease may be often in large part responsible, great progress has been and continues to be made.
+As we approach a new generation of immunotherapies, it is these very complexities that we must better understand in order to control or eradicate the disease.
 

### 2019-04-19 Paper/Paper.tex
maximize

@@
 Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including older \cite{sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{an15_agentbased, serre16_mathematical, louzoun14_mathematical, benzekry17_mathematical, west19_immune, owen11_mathematical-1}.
-These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximise the efficacy of cytotoxicity while minimizing risks to the patient.
+These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximize the efficacy of cytotoxicity while minimizing risks to the patient.
 However, to our knowledge no models have addressed how the effects of EMT alter interactions between the immune system and cancer, and the subsequent implications for treatment.

### 2019-04-22 Paper/Paper.tex
figure adjustments
add refs and update SI format

@@
 \title{
-The competing effects of the immune system and EMT on epithelial cancers
+Modeling the competing effects of the immune system and EMT on epithelial cancers
 }
@@
 We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer.
-These results offer novel means to delay disease onset by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in the light of EMT.
-%The model presented provides a base for future studies into the intricate regulation of tumorigenesis.
+These results offer novel means to delay disease onset by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.
 
@@
 In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), particularly in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
-Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo2017multiscale}.
+Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.
 \par
@@
 \par
-Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including older \cite{sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{an15_agentbased, serre16_mathematical, louzoun14_mathematical, benzekry17_mathematical, west19_immune, owen11_mathematical-1}.
+Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including classic \cite{anderson98_continuous, sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{kim18_cell, gallaher14_bridging, gallaher18_spatial, an15_agentbased, serre16_mathematical, louzoun14_mathematical, briones-orta13_arkadia, lavi13_role, greene15_modeling, greene16_mathematical, cho17_modeling-1,  benzekry17_mathematical, owen11_mathematical, west18_multidrug}.
 These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximize the efficacy of cytotoxicity while minimizing risks to the patient.
@@
 We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will occur and, if so, when.
-A recent model of cancer-immune interactions \cite{guo2017multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
+A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
 We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
@@
 \section{Methods}
-We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo2017multiscale}. 
+We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
 The agents in the model are the tissue cells with tumorigenic potential.
@@
 The algorithm computes the average of the absolute change of the output, which in our model is the area under the survival curve.
-This output out the algorithm, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.
+This output of the algorithm, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.
 
@@
 %                          RESULTS                         %
-%%%%%%%%%%%%%%%%%%%%%%%\section{Results}
+%%%%%%%%%%%%%%%%%%%%%%%
+\section{Results}
 
@@
 \center
-{\includegraphics[width=0.98\textwidth]{Figure1/Figure1.pdf}}
+{\includegraphics[width=0.9\textwidth]{Figure1/Figure1.pdf}}
 \caption{A.
@@
 
-\subsection{Identification of key regulatory parameters via Morris global sensitivity analysis}\label{SensAnalysis}
-To assess the sensitivity of the model parameters and thus identify those that are most important in determining cancer-free survival times, we applied Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods) to the model.
-The results of Morris OAT on the 31 model parameters are shown in Figure \ref{fig:MOAT}: a subset of parameters demonstrate much higher levels of sensitivity than others.
+\subsection{Identifying regulatory parameters via Morris global sensitivity analysis}\label{SensAnalysis}
+Exploring the parameter spaces of models in systems biology is -- in general -- a hard problem.
+Performing Bayesian parameter inference to inform parameter values is advisable wherever possible \cite{kirk13_model}.
+Here, a lack of detailed molecular measurements  (i.e.
+data for tumor growth dynamics abound, but simultaneous data on the immune dynamics are lacking) preclude inference of the full model.
+In addition, while inference schemes for agent-based models are developing \cite{gallaher17_hybrid, warne19_simulation}, simulation times remain a hurdle \cite{lambert18_bayesian}.
+Parameters for aspects of this model that were studied previously can be constrained \cite{guo17_multiscale}, however even for these, the new additions to the model could push it into new behavioral regimes.
+Thus to adequately sample the parameter space of the model and identify those areas of parameter space that exert the most control, we use sensitivity analysis.
+\par
+To assess the sensitivity of the model parameters and thus identify those that are most important in determining cancer-free survival times, we performed Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods).
+The results of Morris OAT on the 31 model parameters are shown in Figure \ref{fig:MOAT}. We see that a subset of parameters demonstrate much higher levels of sensitivity than others.
 The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
@@
 \center
-{\includegraphics[width=0.99\textwidth]{Figure4/Figure4.pdf}}
+{\includegraphics[width=0.95\textwidth]{Figure4/Figure4.pdf}}
 \caption{Effects of inflammation on the time to cancer under different cycling schemes.
@@
 \center
-{\includegraphics[width=0.5
-  \textwidth]{Figure5/MIEvsMGA_bigcbar.jpg}}
+{\includegraphics[width=0.5\textwidth]{Figure5/MIEvsMGA_bigcbar.jpg}}
 \caption{Summary of the contrasting effects of MIE and MGA on the Time to Cancer.
@@
 \center
-{\includegraphics[width=0.95\columnwidth]{FigTCGA.pdf}}
+{\includegraphics[width=0.9\textwidth]{FigTCGA.pdf}}
 \caption{A.
@@
 \par
-
-%%%% AM: Yes please add a para on this. 
-%In addition, I could discuss the implications from the TGF-$\beta$ section.
-%Any discussion on this topic, however, would need to acknowledge that regulation of TGF-$\beta$ might also have other, serious side effects, especially if the treatment is not local.
-%However, that might also be true for the inflammatory results.
-%In light of this, it might be good to include a new section on how TGF-$\beta$ production's influence on Time to Cancer is influenced by the inflammatory cycling scheme.
-
 These results represent promising steps in understanding the competing roles of the immune system and EMT during (pre-) development of epithelial cancers, yet much remains to be done.
@@
 \par
-The role that TGF-$\beta$ plays throughout the tumor and its microenvironment also requires further research.
-We found that, below a threshold, reduction of TGF-$\beta$ increases the Time to Cancer  (Fig.
+The roles that TGF-$\beta$ plays throughout the tumor and its microenvironment also warrant further investigation.
+We found that, below a certain threshold, reduction of TGF-$\beta$ increases the Time to Cancer  (Fig.
 \ref{fig:FirstSurvivalCurves}E), thus reducing expression levels of TGF-$\beta$ in the tumor microenvironment benefits survival.
+Intriguingly, recent experimental work demonstrated that TGF-$\beta$ drives tumor suppression in pancreatic cancer by promoting EMT \cite{david16_tgfv}.
 However, TGF-$\beta$ is a master regulator implicated in numerous cellular signaling processes, and changing the concentrations of TGF-$\beta$ even in a local tumor microenvironment could have large off-target effects.
-Thus it might be more fruitful to target signaling factors downstream of TGF-$\beta$ that still have the ability to modulate epithelial cell dynamics: this is an area of active and future investigation, we are currently developing a larger TGF-$\beta$ signaling pathway module with appropriate crosstalks to  epithelial/mesenchymal/immune cell functions to be incorporated into the model.
+Indeed, it has been showns that TGF-$\beta$ promotes invasion and heterogeneity (although suppresses cell proliferation) in squamous cell carcinoma \cite{oshimori15_tgfv}.
+Future work thus ought to consider the effects of targeting signaling factors downstream of TGF-$\beta$ that still have the ability to modulate epithelial cell dynamics.
+Towards this end, we are currently developing a larger TGF-$\beta$ signaling pathway module with appropriate crosstalks to  epithelial/mesenchymal/immune cell functions to be incorporated into the model.
 \par
@@
 \section*{Acknowledgements}
-The work is partially supported by National Institutes of Health grants R01GM123731 and U01AR073159; National Science Foundation grants DMS1562176 and DMS1763272; The Simons Foundation grant [594598 to Q.N.]; and a grant by Jayne Koskinas Ted Giovanis Foundation for Health and Policy joint with Breast Cancer Research Foundation.
+The work is partially supported by National Institutes of Health grants R01GM123731, U01AR073159, and U54-CA217378; National Science Foundation grants DMS1562176 and DMS1763272; The Simons Foundation grant (594598 to Q.N.) and a grant from the Jayne Koskinas Ted Giovanis Foundation for Health and Policy joint with the Breast Cancer Research Foundation.
 

### 2019-04-22 Paper/Supplement/Supplement.tex
Changed front matter on Supplement
used same structure.tex for supplement

@@
 \documentclass[11pt, a4paper, preprint]{article}
-\usepackage{outline}
-\usepackage{float}
-\usepackage{graphicx}
-\usepackage{color}
-\usepackage{amsmath}
-\usepackage{cite}
-\graphicspath{ {Figs/} }
-\usepackage{longtable}
+%\usepackage{outline}
+%\usepackage{float}
+%\usepackage{graphicx}
+%\usepackage{color}
+%\usepackage{amsmath}
+%\usepackage{cite}
+%\graphicspath{ {Figs/} }
+%\usepackage{longtable}
+
+\usepackage{amsmath, amssymb, graphicx, outline, float, cite, color}
+
+%\input{../structure.tex} % Specifies the document structure and loads requires packages
+
 \usepackage[T1]{fontenc}
@@
 
-\title{Supplementary Material}
+\title{Supplementary Materials for \\
+Modeling the competing effects of the immune system and EMT on epithelial cancers
+}
 \author{Daniel R. Bergman$^{1}$,
@@
 Qing Nie\,$^{1,2,*}$
-and Adam L. MacLean\,$^{3,*}$
+and \\Adam L. MacLean\,$^{3,*}$
 }
@@
   $^3$Department of Biological Sciences, University of Southern California, Los Angeles, CA 90089, USA \\
-  $^*$Correspondence:  qnie@uci.edu (QN); macleana@usc.edu (ALM).
+  $^*$Correspondence:  qnie@uci.edu (Q.N.); macleana@usc.edu (A.L.M.)
 }
@@
 \maketitle
+\tableofcontents
+
+\section{Definition of parameters specifying the model}
 
@@
 \end{table}
+
+\section{Parameter values used for simulation}
 

### 2019-04-26 Paper/Supplement/Supplement.tex
edits to methods and update title
fix fig 2 legend

@@
 \title{Supplementary Materials for \\
-Modeling the competing effects of the immune system and EMT on epithelial cancers
+Modeling the competing effects of the immune system and EMT on tumor development
 }
 \author{Daniel R. Bergman$^{1}$,
-Matthew Karikomi\,$^{1}$,
+Matthew K. Karikomi\,$^{1}$,
 Qing Nie\,$^{1,2,*}$

### 2019-09-16 Paper/Paper.tex
add line numbers
Rewrote after referees got back to us

@@
 \usepackage{amsmath, amssymb, graphicx, outline, float, cite, color}
+\usepackage{lineno}
 
-\input{structure.tex} % Specifies the document structure and loads requires packages
+\input{structure.tex}
 
@@
 
+\usepackage[normalem]{ulem}
 \usepackage[T1]{fontenc}
@@
 
-%\usepackage[ ]{natbib} 
 \newcommand{\tcr} { \textcolor{red} }
+\newcommand{\tcb} { \textcolor{blue} }
 
-% Multiscale Modeling of the Paradoxical Roles of the Immune System in EMT-Mediated Cancer
 \title{
@@
 \section*{Abstract}
-Preceding and during cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor incidence and prognosis.
+\sout{Preceding and d}\tcb{D}uring cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor \sout{incidence and} prognosis.
 Inflammation also has significant effects on gastrointestinal and pancreatic tumors.
@@
 Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that maximize cancer-free survival.
-Delaying tumorigenesis depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
+Delaying \sout{tumorigenesis} \tcb{progression} depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
 We performed analysis of pancreatic cancer data from The Cancer Genome Atlas to compare clinical measurements with model predictions.
 We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer.
-These results offer novel means to delay disease onset by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.
+These results offer novel means to delay disease \sout{onset} \tcb{progression} by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.
 
-%% Basic Introduction (1-2 sentences)
-%Preceding tumorigenesis and throughout the lifetime of a tumor, the immune system is engaged in complex sets of interactions in and around the tumor.
-%% More Detailed Background (2-3 sentences)
-%These are regulated by the tumor microenvironment (TME) in myriad ways.
-Moreover,  epithelial cell plasticity impacts disease incidence and progression via processes of epithelial-to-mesenchymal transition (EMT).
-%In turn, both the tumor and the immune system exert substantial influence over the TME in competition to influence the fate of the host.
-%% General Problem (1 sentence)
-%Determining the factors and mechanisms underpinning these complex interactions remains an important task for cancer immunotherapy.
-%% Summarizing Main Result (1 sentence)
-%Here, we seek to understand how interplay between the tumor, the immune system, and EMT can determine the outcome of cancer.
-%% The Main Result (2-3 sentences)
-%We developed a multiscale agent-based model to describe the interplay between three components: the tumor (or pre-tumor cells), the immune system, and EMT.
-%The model revealed a \tcr{non-monotonic relationship} between the cell cycle times associated with the mesenchymal phenotype and the probability of tumorigenesis, identifying a critical growth rate that maximizes cancer-free survival.
-In comparison, increasing the immune-evasiveness of the mesenchymal phenotypes always led to poorer prognosis.
-Moreover, encouraging
-%\tcr
-%the TME
-%towards a pro-EMT state also benefits the tumor.
-%% Putting into a More General Context (1-2 sentences)
-%Together, these results indicate that the joint regulation of the TME by the immune system and the tumor itself provides possible targets for future cancer therapies, in particular cancers such as pancreatic cancer in which EMT is speculated to play a role.
-
-%%%% Possible additional refs:
-% Diefenbach 2001
-% They show that when tumor cells express a particular marker, then the immune system gains considerable potency and clears the tumor.
-Furthermore, immunization via tumor cells expressing these ligands creates a host immunity to subsequent tumors even if these tumors lack the ligand.
-
-% Carretero 2015
-% As eosinophils can better carry out their recruitment function of CD8+ T cells, tumor rejection by the immune system is more readily accomplished.
-
-% Ju 2016
-% A cancer vaccine was demonstrated to be effective in reducing colorectal cancer incidence in mice. In addition, efficacy of CTLs was shown to be increased.
+\linenumbers
 
@@
 The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
-%Make sure there's some ref for immune clearing tumor
-% Also, add that as immune cells clear tumor cells, they become deactivated and give source
-In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), particularly in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
-Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.
+% Does the following sentence still fit with our new focus on progression? I'm looking specifically at the words "pre-cancerous" and "tumorigenesis".
+\tcr{In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.}
+% Similar concern to above as Guo's paper dealt with initiation, not progression.
+\tcr{Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.}
 \par
@@
 Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}.
-Ongoing investigations into the plasticity and stability of EMT overlap with discussions elsewhere, e.g.
-of discrete vs.
-continuous processes during cell differentiation \cite{moris16_transition}.
+Ongoing investigations into the plasticity and stability of EMT overlap with discussions elsewhere, e.g. of discrete vs. continuous processes during cell differentiation \cite{moris16_transition}.
 Intermediate states have emerged as a central mechanism by which cell fates (and the noise inherent within them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}.
 \par 
-While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation and progression \cite{nieto2016emt}.
+% Should we even mention initiation here? Make sure to combine below end-of-sentence refs into single \cite command.
+While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor \tcr{initiation} \tcb{\cite{puisieux2014oncogenic}} and progression \cite{nieto2016emt}\tcb{\cite{peinado2007snail}}.
 TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
-Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
+\tcb{There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.} Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
 It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.
@@
 The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculate to exhibit a more ``stemness'' phenotype.
-%Indeed, even in healthy tissue, this is readily observed and seen as a useful mechanism for homeostasis as epithelial cells can transition to mesenchymal, stem-like cells under appropriate conditions \tcr{[REF]}.
-While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects} (which, under certain conditions can be called quiescence, though we avoid this term as it can be controversial, and is not a necessary assumption to the model developed below).
+While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects} \sout{(which, under certain conditions can be called quiescence, though we avoid this term as it can be controversial, and is not a necessary assumption to the model developed below)}.
 Cancer cells are of course characterized by some degree of unregulated proliferation, thus it is appealing to explore the effects of differential proliferation rates within the cancer cell population.
@@
 Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
-We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will occur and, if so, when.
+We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will \sout{occur} \tcb{progress} and, if so, when.
 A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
+\tcb{
+In doing so, we shift the focus of the previous model from cancer initiation to cancer progression.
+We do this to reflect the fact that cancer progression hinges on escape from the immune system and the fact that EMT has a more well-defined role during progression and metastasis.
+}
 We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
@@
 We then go on to analyze its behavior.
-Global sensitivity analysis identifies parameters that are crucial for tumorigenesis.
-We then study these in more depth, focusing on the competing effects of EMT and of the immune system on tumorigenesis: we find that EMT intricately regulates tumorigenesis and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
-To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects predicted by the model for patients with pancreatic cancer.
-
-%%%%%%%%%%%%%  INTRO PARAGRAPH GRAVEYARD %%%%%%%%%%%%%%%%
-
-%They release cytokines that alter the extracellular matrix, promote angiogenesis, and create an inflammatory environment.
-%This last one, in particular, results in the impairment of cytotoxic capabilities of immune cells.
-%Through 
-%
-% its presence and by the cellular processes it engages in that result in shaping the TME.
-%On the one hand, the mutated DNA of tumor cells results in these cells being recognized as foreign by members of the innate immune system such as natural killer (NK) cells and dendritic cells.
-%It is then these cells that initiate the cascade of cytokines that result in the full activation of the body's immune system to fight the non-self cells.
-%On the other hand, the tumor cells actively work to shape the immune response that the body launches.
-%This is accomplished through its own set of cytokines, such as transforming growth factor beta (TGF-$\beta$), that work to create a tumor-supportive environment, including shifting the immune balance to be immunosuppressive by enhancing the recruitment of Treg cells.
-Since TGF-$\beta$ is also a driver of EMT, then through direct release of TGF-$\beta$ and recruitment of TGF-$\beta$-secreting Treg cells, cells of the tumor and the surrounding tissue are driven towards a more mesenchymal profile.
-
-%In quantifying this property, we use the model parameter `mesenchymal growth arrest', MGA.
-%This choice is less about associating this property with quiescence and more about creating a simple name that easily, if not roughly, captures the essence of what is being modeled.
-%For example, while mesenchymal proliferation reduction might be a more accurate term, thinking in terms of a negative formulation adds one additional twist to accurately interpreting results which we felt was more detrimental than the possible confusion of invoking ``quiescence.''
-
-%This plays an important yet complicated role in tumor initiation, progression, and metastasis\cite{nieto2016emt}. % References more specific to this claim might be in this paper's references. 
-%As mesenchymal cells often exhibit stem-like properties, their existence in a tumor has been tied to the initiation and progression of cancer\cite{nieto2016emt}.
-% References more specific to this claim might be in this paper's references.
-%Also, due to their motility, some have viewed them as a natural precursor to metastasis\cite{alderton2012metastasis}. 
-%
-%Conversely, the mesenchymal cells in the tumor downregulate certain surface markers making it more difficult for cytotoxic immune cells to form the immune synapse necessary to lyse the cancer cell\cite{terry2017new}.
-%Taking these three aspects together, all influencing the others, we can consider the simple diagram in Figure \ref{fig:3PartFig} and see just how much potential there is for exploring all these relationships.
-%Each of these three pieces will be briefly explored in turn for how they will influence one another.
+Global sensitivity analysis identifies parameters that are crucial for \sout{tumorigenesis} \tcb{progression}.
+We then study these in more depth, focusing on the competing effects of EMT and of the immune system on \sout{tumorigenesis} \tcb{progression}: we find that EMT intricately regulates \sout{tumorigenesis} \tcb{progression} and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
+\sout{To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects predicted by the model for patients with pancreatic cancer.}
+\tcb{To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.}
+% Consider rewriting above
 
@@
 
+\begin{table}[H]
+\begin{tabular}{|p{\textwidth}|}
+\hline
+%%%
+\tcb{
+\begin{itemize}
+\item TGF-$\beta$ is well-mixed within the TME
+\item All cells are on the same ``clock'', i.e. they all decide at the same time what to do next
+\item During a given cell cycle, exactly one event can happen to any one cell: proliferation, apoptosis, immune clearance, or rest
+\item All tissue cells (mutated and non-mutated) are equally likely to encounter immune cells
+\item Other spatial effects are ignored by the model, e.g. angiogenesis, hypoxia, invasive front, and chemotaxis
+\item The initial cells are mutated, but are not able to form an aggressive cancer without further mutations
+\end{itemize}
+}
+\\
+\hline
+\end{tabular}
+\caption{\tcb{Basic assumptions of the model}}
+\label{table:model_assumptions}
+\end{table}
+
 \section{Methods}
 We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
-The agents in the model are the tissue cells with tumorigenic potential.
-These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).
+The agents in the model are the tissue cells \sout{with tumorigenic potential} \tcb{that have already formed a non-aggressive tumor yet have not progressed}.
+\sout{These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).}
+\tcb{
+In the process of the simulation, these cells can acquire new pathway mutations among three possible (shown in Fig. \ref{fig:ModelIntro}A).
+}
 If the pathway is mutated, the component is colored red and is marked with `X'.
-A single mutation is sufficient for the cell to be considered as a mutant cell.
+\sout{A single mutation is sufficient for the cell to be considered as a mutant cell.}
+\tcb{
+% I'm not a fan of "hypermutated", but I'll use it for now so that a quick replace all can change it later.
+We distinguish between cells that have zero of the three pathway mutations and those that have at least one by referring to the latter as hypermutated.
+}
 \par
-We model immune cells as continuous variables, appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.
-The cytokine TGF-$\beta$ is continuous within the TME, i.e. the system is well-mixed.
+We model immune cells as continuous variables, \sout{appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.}
+\tcb{
+effectively assuming that the TME is well-mixed with regards to the immune infiltrate.
+}
+The cytokine TGF-$\beta$ is \sout{\tcr{continuous} within the TME, i.e. the system is well-mixed.}
+\tcb{
+also assumed to be well-mixed in the TME so that all tissue cells have equal access to it.
+}
 Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
-While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
-In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
+\sout{While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
+\tcb{While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
+% Above line replaced
+In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the \sout{color} \tcb{shade of blue} of the cell.
 Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.
@@
 The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
-The NKs and CTLs act on the system by recognizing mutated cells and clearing them.
+% We could justify the immune system not clearing "base" mutated cells by saying the immune surveillance provides a baseline control over these cells accounted for in their apoptosis.
+Or suggesting that we are interested in immune activity above and beyond the baseline requirement to keep this tumor in check.
+The NKs and CTLs act on the system by recognizing \sout{mutated} \tcb{hypermutated} cells and clearing them.
 Upon successful clearance, they themselves are then deactivated and removed from the immune population.
-Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear mutated cells.
+Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear \sout{mutated} \tcb{hypermutated} cells.
 In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.
+
+\tcb{
+\subsection{Inflammation} % How did we not have a section on this before?
+We also include an inflammation score that follows a fixed cycling scheme between low and high inflammatory states.
+This is to simulate the various changes to the TME that can happen over long periods of time (days and weeks).
+For the purpose of the model, we consider low inflammation to be the default state.
+When the inflammation switches to high, then a number of the parameters change values to reflect the additional immune activity.
+\tcr{See Table 2 in the supplement to see which parameters are changed by the inflammation score changing.} % Table 2 or S2? Change label in supplement?
+}
 
@@
 Otherwise, it is below the threshold and labeled as epithelial.
-For the purposes of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have some of its parameters updated.
+For the purpose of \tcb{parameter values} of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have \sout{some} \tcb{a subset} of its parameters updated.
 In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
-Note, that because the immune system can only clear mutated cells, the increased immune evasion only benefits mutated mesenchymal cells.
-
+Note, that because the immune system can only clear \sout{mutated} \tcb{hypermutated} cells, the increased immune evasion only benefits \sout{mutated} \tcb{hypermutated} mesenchymal cells.
+\par
+\tcb{
+In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
+In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
+}
 \subsection{Model Simulation}
 
-\subsubsection{Initial conditions} Every model run starts with a fixed amount of cells, $N_0$.
-However, different parameter values will lead to different steady states for the population of tissue cells.
-In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.
-Thus, the only immune cell present is NKs.
+\subsubsection{Initial conditions} 
+Every model run starts with \sout{a fixed amount} \tcb{an amount} of \tcb{tissue} cells, $N_0$ \tcb{, determined by the set of parameters used for the run.
+A set of warmup cycles are run to ensure arrival to this steady state.
+During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.}
+\sout{However, different parameter values will lead to different steady states for the population of tissue cells.}
+\sout{In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.}
+\sout{Thus, the only immune cell present is NKs.}
 After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
@@
 Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
+% Add: 
+\tcb{This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.}
 Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
@@
 For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
-First of all, the cell needs to be mutated ($\delta_{\text{MUT}}=1$).
+First of all, the cell needs to be \sout{mutated} \tcb{hypermutated} ($\delta_{\text{MUT}}=1$).
 Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
@@
 \paragraph{Rest in $G_0$} 
-The weight associated with quiescence is taken as 1 except in the case of mesenchymal cells.
+\sout{The weight associated with quiescence is taken as 1 except in the case of mesenchymal cells.
 For mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
-Hence, the weight of quiescence is given by
+Hence, the weight of quiescence is given by}
+\tcb{The weight associated with rest is taken as 1 except in the case of mesenchymal cells.
+Recall that mesenchymal cells had their proliferation rate decreased by $1-\zeta\Delta_{\text{MGA}}$ (see Eq. 2.1).
+The biological assumption here is that mesenchymal cells instead of proliferating will instead rest, so this lost proliferation weight is added to the resting weight.
+Hence, the weight of rest is given by}
+% Replace quiescence with rest and explain more above
 
@@
 
-The reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.
+\tcb{Again, t}\sout{T}he reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.
 
@@
 If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases.
+\par
 Finally, the EMT values for each cell is updated.
 This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
-Each cell receives an amount of TGF-$\beta$ given by
+% The following sentence could be altered to reflect that the threshold value for $\tau$ is 0, so not only are negative values possible, but very common.
+So, this is not really so much a TGFB received as it is a change in TGFB signaling.
+\sout{Each cell receives an amount of TGF-$\beta$ given by}
+\tcb{The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
+The saturation effect is to limit the amount of TGF-$\beta$ a cell can absorb in a given time interval.
+This quantity is then divided up randomly among the $N_C$ living cells via a normally distributed noise term to determine how much exogenous TGF-$\beta$ each cell receives during this cell cycle.
+Should this value, $\tau_i$ in Eq. 2.5, be negative, we interpret this as the cell losing TGF-$\beta$ to the TME and thus being more likely to undergo MET.}
 
-%$$ \text{TGFB}_i = H(\text{TGFB}/N) + X_i, \quad X_i \sim N(0,\sigma^2)$$
 \begin{equation}\tag{2.5}
-\tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{1}{1+K_3/\tau}+ X_i
+\tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{\tau/K_3}{1+\tau/K_3} + X_i
 , \quad X_i \sim N(0,\sigma^2)
@@
 
-The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.
-Then, some randomness is introduced by the independent and identically distributed variables $X_i$ where $\sigma$ is a parameter in the model.
+\tcb{
+We then combine $\tau_i$ with the current EMT score of the cell, as a proxy for the endogenous TGF-$\beta$.
+Finally, if this quantity is large enough, \sout{the cell undergoes EMT; otherwise, it undergoes MET.} \tcb{the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.}
+}
+% Need to add better explanation for Hill fn
+\sout{The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.}
+\sout{Then, some randomness is introduced by the independent and identically distributed variables $X_i$ where $\sigma$ is a parameter in the model.
 If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
 If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
+}
 Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
@@
 
-For CTLs and Tregs, they rely on mutant cells being cleared before they can be activated.
-Let $N_{\text{MUT}}^*(k)$ represent the number of mutant cells cleared by the immune system during cell cycle $k$.
+For CTLs and Tregs, they rely on \sout{mutant} \tcb{hypermutated} cells being cleared before they can be activated.
+Let $N_{\text{MUT}}^*(k)$ represent the number of \sout{mutant} \tcb{hypermutated} cells cleared by the immune system during cell cycle $k$.
 In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
@@
 N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}N_\text{CTL} \\
-\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
+N_\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
 \end{split}
@@
  \hline
-  $\Delta_A$ & mutant cells decreased apoptosis  \\
+  $\Delta_A$ & \tcr{mutant cells decreased apoptosis}  \\ % should this and the others like it be more clear in stating this is not for all hypermutated cells, but only those with apoptosis pathway mutation?
   \hline
@@
   \hline
+  $K_4$ & EC50 term for TGF-$\beta$ activation of Tregs \\
+  \hline
  $E_\text{NK}$ & rate of NKs clearing mutants  \\
@@
   \hline
-  $\sigma_\text{CTL}$ & CTL source rate per cleared mutant cell \\ 
+  $\sigma_\text{CTL}$ & CTL source rate per cleared \sout{mutant} \tcb{hypermutated} cell \\ 
   \hline
-  $\sigma_\text{Treg}$ & Treg source rate per cleared mutant cell \\ 
+  $\sigma_\text{Treg}$ & Treg source rate per cleared \sout{mutant} \tcb{hypermutated} cell \\ 
+  \hline
+  $d_\text{NK}$ & NK death rate \\ 
+  \hline
+  $d_\text{CTL}$ & CTL death rate \\ 
+  \hline
+  $d_\text{Treg}$ & Treg death rate \\ 
   \hline
@@
 \subsubsection{Mutation and tumorigenesis}
-At the end of each cell cycle, the proportion of tissue cells that are mutated is calculated, and if it is above a certain threshold, a tumorigenic event is recorded and the patient is determined to have cancer.
-The Time to Cancer is given as the time from the start of the simulation, not including warmup, until tumorigenesis.
-Simulations run until either a tumorigenic event is recorded or until the maximum number of cycles has been reached.
+% 3rd Referee did not like this terminology for a tumorigenic event. Should we change it or better explain it?
+At the end of each cell cycle, the proportion of tissue cells that are \sout{mutated} \tcb{hypermutated} is calculated, and if it is above a certain threshold, \sout{a tumorigenic event is recorded and the patient is determined to have cancer.}
+\tcb{the simulation ends and we declare the tumor to have progressed.}
+The Time to Cancer is given as the time from the start of the simulation, not including warmup, until \sout{tumorigenesis.} \tcb{this timepoint.}
+Simulations run until either a \sout{tumorigenic event is recorded} \tcb{this occurs} or until the \tcb{predetermined} maximum number of cycles has been reached.
 
@@
 To study parameter sensitivity, we implemented the Morris method, a one-step-at-a-time global sensitivity analysis algorithm.
-%The algorithm runs the model numerous successive times, \tcr{changing exactly one parameter each run} \cite{morris1991factorial, sohier2014improvement}.
 The algorithm runs the model numerous times by picking several base points and then varying one parameter at a time from each base point and tracking the difference\cite{morris1991factorial, sohier2014improvement}.
@@
 Some of the parameters were also truncated to the right because they represented a proportion, such as MIE and MGA, or because we simply wanted to keep the values within a reasonable range, such as the inflammation low and high duration parameters.
-The variances were chosen as twice the mean to get a large spread of values.
+The \sout{variances} \tcb{standard deviations} were chosen as twice the mean to get a large spread of values.
 \par
-%The algorithm \tcr{outputs the average of the absolute change of area under the curve for each parameter normalized by the corresponding change in the cdf function for that parameter.
-[I don't get this: what is y-axis?]}
-The algorithm computes the average of the absolute change of the output, which in our model is the area under the survival curve.
+% Referee 2 either missed the following line or doesn't understand what average of absolute change means.
+\tcr{The algorithm computes the average of the absolute change of the output, which in our model is the area under the survival curve.}
 This output of the algorithm, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Results}
@@
 {\includegraphics[width=0.9\textwidth]{Figure1/Figure1.pdf}}
-\caption{A.
-Schematic depiction of agent-based model components; blue/red denote tissue cells with/without mutations.
+\caption{A. Schematic depiction of agent-based model components; \tcb{
+each of the 10 columns represents a single tissue cell divided into three compartments representing the state (mutated or not) of the three pathways with mutagenic potential;
+}\sout{blue/red denote tissue cells with/without mutations} \tcb{red/blue denotes mutated/unmutated pathways} .
 Black arrows depict regulation of the cell fate in each cell cycle.
@@
 B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
-The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: mutated cells; Mes: mesenchymal cells. 
+The inflammation cycling scheme is represented above the patient dynamics.
+The vertical dashed line denotes the end of the warmup period.
+Mut: \sout{mutated} \tcb{hypermutated} cells; Mes: mesenchymal cells.
 C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
@@
 
-We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the probability of cancer, and the Time to Cancer.
+We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the \tcr{probability} of cancer, and the Time to Cancer.
 Within the cell cycle, cell fate is determined via a set of rules that are influenced by EMT and immune interactions (Fig.
@@
 The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset).
-NK cells and CTLs attempt to clear mutated tissue cells, and deactivate upon successfully carrying out their cytotoxic function.
+NK cells and CTLs attempt to clear \sout{mutated} \tcb{hypermutated} tissue cells, and deactivate upon successfully carrying out their cytotoxic function.
 Tregs inhibit cytotoxic activity.
@@
 For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor.
+\sout{
 However, after approximately 1000 cell cycles (750 days), one mutated cell exhibits clonal growth, and a tumor is established.
@@
 After cancer incidence, the number of mutated cells continues to grow rapidly and soon makes up 100\% of the cell population.
-A peak in mesenchymal phenotypes is observed following cancer incidence, before most transition back to an epithelial state.
+A peak in mesenchymal phenotypes is observed following cancer incidence, before most transition back to an epithelial state.}
+\tcb{
+After about 700 cell cycles, the hypermutated cell population begins to grow steadily.
+At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being quickly recruited into the TME, and a peak in TGF-$\beta$ expression is observed.
+After 841 cell cycles, the proportion of hypermutated cells reaches 50\%: the threshold defining progression.
+Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days.
+For this patient, we kept let the simulation go beyond this timepoint, and we se that the number of hypermutated cells grows rapidly and soon makes up 100\% of the cell population.
+A peak in number of mesenchymal cells is observed shortly after this threshold is reached, and following this most cells transition back to epithelial.
+}
 \par 
-Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of oncogenic mutations.
+Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of \sout{oncogenic mutations} \tcb{hypermutations}.
 The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
@@
 F. Survival curves for patient cohort corresponding to changes in Treg TGF-$\beta$ (C). 
-%% \color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.
 }
@@
 Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
-The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a mutated mesenchymal cell will be subject to immune clearance in a given cycle.
+The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a \sout{mutated} \tcb{hypermutated} mesenchymal cell will be subject to immune clearance in a given cycle.
 The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
@@
 As MIE is increased, the Time to Cancer decreases (Fig.
-\ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of potentially mutated cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
+\ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of \sout{potentially mutated} \tcb{hypermutated} cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
 These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.
-
+\par
 Second, as MGA increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Fig \ref{fig:FirstSurvivalCurves}B).
-This is not immediately intuitive, since the decreased proliferation rate affects both mutated and non-mutated tissue cells.
-
+This is not immediately intuitive, since the decreased proliferation rate affects \sout{both mutated and non-mutated tissue cells.} \tcb{all mesenchymal cells, not just hypermutated cells.}
+\par
 Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
@@
 \par
-These striking differences in the mean Time to Cancer -- extended by up to one year by fine-grained optimization of mesenchymal growth arrest -- have clear therapeutic implications. 
+These striking differences in the mean Time to Cancer -- extended by up to one \tcb{{\it in silico}} year by fine-grained optimization of mesenchymal growth arrest -- have clear therapeutic implications.
+\sout{
 Our model predicts that reducing the inflammatory state of the TME in three weeks out of every nine would be provide means to prolong cancer-free survival in combination with mesenchymal cell therapies to make the control of mesenchymal proliferation an effective anti-cancer strategy.
+}
+\tcb{
+Our model predicts that a patient suffering intermittent inflammatory attacks at the tumor site will benefit from controlling the proliferation of the mesenchymal cells.
+}
 \par
@@
 \caption{Summary of the contrasting effects of MIE and MGA on the Time to Cancer.
-  % Increasing MIE always decreases Time to Cancer, but MGA is non-monotonically linked to Time to Cancer.
 }
@@
 \par
+To compare these model predictions with experimental studies, we analyzed data from The Cancer Genome Atlas (TCGA) database to study the effects of immune and EMT interactions on prognosis of cancers for which inflammation is known to play an important role, such as colonic or pancreatic cancers \cite{greten2019inflammation,hu2010inflammation}.
+We note that the comparison between simulation and data here is indirect since the model studies the progression of cancer from an arbitrary starting point, while the data address the progression of cancer following treatment, but analogous processes are at play during both the progression of  clonal dynamics addressed by the model, and the progression of disease after cancer treatment.
+In particular, the plasticity of tumor cells allows them to evade treatment by undergoing post-treatment processes resembling the de-novo appearance of cancer\cite{sanchez2018slow}.
+\par 
+The TCGA Pan-Cancer Clinical Data Resource (TCGA-CDR) provides multiple computed clinical endpoints for PAAD \cite{liu2018integrated}.
+Here, we focus on the disease-free interval (DFI) and the overall survival (OS) endpoints.
+In essence, a tumor where the DFI is short will undergo rapid (post-treatment) initiation.
+In turn, we expect the distance between DFI and OS to be relatively small for a tumor undergoing rapid progression following initial (post-treatment) detection.
+Therefore, recurrent tumors may be roughly categorized as either rapidly progressing or slowly progressing by some arbitrary threshold of DFI/OS divergence.
+We selected three tumor types whose TCGA-CDR profile indicated slow progression (OV, SKCM, LIHC) and three tumor types whose profile indicated rapid progression (PAAD, LUAD, COAD).
+\par
+For each cohort of cancer patients, we cluster the patients via k-means (n = 2) either on gene ontologies relating only to EMT (linked figures) or on gene ontologies relating to EMT and inflammation.
+We also plot the corresponding survival curves for each of the clusters obtained (utilizing the originally reported OS endpoint).
+We see that in both cases, survival is affected by the gene ontology signature, but the presence of both EMT and inflammatory signatures has a greater impact on survival than the effects of EMT alone.
+This suggests that when studying interactions between cancer and the immune system, is it important to consider EMT; the effects of which may impact the dynamics and should not be overlooked.
+
+\par
+
+\tcr{THE REST OF THIS SECTION IS FROM THE FIRST DRAFT AND WILL BE DELETED}
+
+\par 
 To compare these model predictions with experimental studies, we analyzed data from The Cancer Genome Atlas (TCGA) database to study the effects of immune and EMT interactions on prognosis of cancers for which inflammation is known to play an important role, such as colonic or pancreatic cancers \cite{hu10_inflammationinduced, balkwill01_inflammation}.
@@
 \ref{fig:tcga}C).
-We also plot the corresponding survival curves for each of the clusters obtained (Fig.
-\ref{fig:tcga}B, D).
+We also plot the corresponding survival curves for each of the clusters obtained (Fig. \ref{fig:tcga}B, D).
 We see that in both cases, survival is affected by the gene ontology signature, but the presence of both EMT and inflammatory signatures has a greater impact on survival than the effects of EMT alone.
@@
 Clearly, there exists a trade-off to be accounted for, and adding to the model the main effector function of Tregs could remedy this and add depth to our understanding of the various roles that Tregs play in and around the tumor.
-
-%Also, since the inflammatory state of the TME played such a large role in our simulations, making this a more robust facet of the model by allowing the tumor and immune system to influence the inflammatory state could give the model higher fidelity.
-%One intriguing aspect could be the anti-inflammatory function of Tregs as they seek to restore homeostasis.
 \par

### 2019-09-16 Paper/Supplement/Supplement_rewrite.tex
Rewrote after referees got back to us

--- 2019-04-22 Paper/Supplement/Supplement.tex
@@
   $\sigma_\text{Treg}$ & Treg source rate per cleared mutant cell \\ 
+  \hline
+  $d_\text{NK}$ & NK death rate  \\ 
+  \hline
+  $d_\text{CTL}$ & CTL death rate \\ 
+  \hline
+  $d_\text{Treg}$ & Treg death rate \\ 
   \hline
@@
 \section{Parameter values used for simulation}
+
+\begin{table}[H]
+\begin{center}
+ \begin{tabular}{|| c | p{7cm} | c | c ||} 
+ \hline
+ Name & Description & INFL Low Value & INFL High Value  \\ [0.5ex] 
+ \hline\hline
+  $p$ & weight of proliferation for tissue cells & 0.28 & \\ 
+ \hline
+ $d_C$  & weight of apoptosis for tissue cells & 0.14 & \\
+ \hline
+$\Delta_\text{MIE}$ & MIE & 0.6 & \\
+ \hline
+ $\Delta_\text{MGA}$ & MGA  & 0.2 &  \\
+ \hline
+  $\Delta_A$ & proportional decrease to weight of apoptosis for cells with mutated apoptosis pathway & 0.3 &  \\
+  \hline
+  $\Delta_\text{IE}$ & proportional increase to weight of immune evasion for cells with mutated immune evasion pathway & 0.48 &  \\
+  \hline
+  $\Delta_P$ & proportional increase to weight of proliferation for cells with mutated proliferation pathway & 0.36 &  \\
+  \hline
+ $K_0$ & EC50 term for negative feedback of tissue cells on own proliferation & 80 cells& \\
+ \hline
+ $K_1$ & EC50 term for probability of NK cell finding mutant cell & 8 cells& \\
+ \hline
+  $K_2$ & EC50 term for Treg inhibition of cytotoxic functions & 5 cells / volume & 0.025 cells / volume \\
+  \hline
+  $K_3$ & EC50 term for cumulative absorption of TGF-$\beta$ & 200 amount / volume & \\
+  \hline
+  $K_4$ & EC50 term for TGF-$\beta$ activation of Tregs & 50 amount / volume & \\
+  \hline
+ $E_\text{NK}$ & weight of NKs clearing mutants & 10 & 30  \\
+  \hline
+  $E_\text{CTL}$ & weight of CTLs clearing mutants & 200 & 600 \\
+  \hline
+  $\sigma_\text{NK}$ & NK source rate & 1.3 cells / cycle &  \\ 
+  \hline
+  $\sigma_\text{CTL}$ & CTL source rate per cleared mutant cell & 100 cells / (cleared mutants $\times$ cycles)& \\ 
+  \hline
+  $\sigma_\text{Treg}$ & Treg source rate per cleared mutant cell & 200 cells / (cleared mutants $\times$ concentration of TGF-$\beta$ $\times$ cycles)& \\ 
+  \hline
+  $d_\text{NK}$ & NK death rate & 0.13 / cycle &  \\ 
+  \hline
+  $d_\text{CTL}$ & CTL death rate & 0.0260 / cycle& \\ 
+  \hline
+  $d_\text{Treg}$ & Treg death rate & 0.0260 / cycle& \\ 
+  \hline
+  $k_\text{EMT}$ & EMT/MET rate & 0.01 / concentration of TGF-$\beta$ & \\
+  \hline
+  $\sigma$ & standard deviation of noise in TGF-$\beta$ each cell receives & 6 concentration of TGF-$\beta$ & \\
+  \hline
+ $\tau_\text{max}$ & max amount of TGF-$\beta$ any cell can receive & 500 concentration of TGF-$\beta$ &\\
+  \hline 
+ $\tau_\text{MUT}$ & rate of TGF-$\beta$ production by mutant cells & 0.05 concentration of TGF-$\beta$ / cell / cycle& \\
+  \hline
+ $\tau_\text{Treg}$ & rate of TGF-$\beta$ production by Treg & 0.5 concentration of TGF-$\beta$ / cell / cycle& \\
+  \hline
+  & RP Cancer Line  & 0.5 &  \\ 
+ \hline
+&  INFL High Duration & 30 cycles&   \\
+ \hline
+ & INFL Low Duration &30 cycles&   \\
+ \hline
+& Mes Threshold & 0.7 &    \\
+ \hline
+  & maximum initial mutation damage after warmup  & 0.01 &\\
+  \hline
+  & increase in probability to mutate for non-mutating proliferating cells & 0.0001 & \\
+  \hline
+\end{tabular}
+  \caption{The model parameter names, descriptions, and values during both low and high inflammation. Parameters with only one value do not change with the inflammatory state.}
+\end{center}
+\end{table}
+
+\newpage
 
@@
   \hline
+  $K_4$ & EC50 term for TGF-$\beta$ activation of Tregs & 50 & \\
+  \hline
  $E_\text{NK}$ & rate of NKs clearing mutants & 10 & 30  \\
@@
   $\sigma_\text{Treg}$ & Treg source rate per cleared mutant cell & 200 & \\ 
+  \hline
+  $d_\text{NK}$ & NK death rate & 0.13 &  \\ 
+  \hline
+  $d_\text{CTL}$ & CTL death rate & 0.0260 & \\ 
+  \hline
+  $d_\text{Treg}$ & Treg death rate & 0.0260 & \\ 
   \hline

### 2019-09-16 Paper/CleanPaper.tex
Clean paper and premalignant/malignant

--- 2019-09-16 Paper/Paper.tex
@@
 \section*{Abstract}
-\sout{Preceding and d}\tcb{D}uring cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor \sout{incidence and} prognosis.
+During cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor prognosis.
 Inflammation also has significant effects on gastrointestinal and pancreatic tumors.
@@
 Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that maximize cancer-free survival.
-Delaying \sout{tumorigenesis} \tcb{progression} depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
+Delaying progression depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
 We performed analysis of pancreatic cancer data from The Cancer Genome Atlas to compare clinical measurements with model predictions.
 We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer.
-These results offer novel means to delay disease \sout{onset} \tcb{progression} by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.
+These results offer novel means to delay disease progression by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.
 
@@
 % Does the following sentence still fit with our new focus on progression? I'm looking specifically at the words "pre-cancerous" and "tumorigenesis".
-\tcr{In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.}
+In addition, the immune system plays important regulatory roles in premalignant tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), including in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
 % Similar concern to above as Guo's paper dealt with initiation, not progression.
-\tcr{Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.}
+Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.
 \par
@@
 % Should we even mention initiation here? Make sure to combine below end-of-sentence refs into single \cite command.
-While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor \tcr{initiation} \tcb{\cite{puisieux2014oncogenic}} and progression \cite{nieto2016emt}\tcb{\cite{peinado2007snail}}.
+While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation \cite{puisieux2014oncogenic} and progression \cite{nieto2016emt,peinado2007snail}.
 TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
-\tcb{There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.} Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
+There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.
+Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
 It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.
@@
 The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculate to exhibit a more ``stemness'' phenotype.
-While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects} \sout{(which, under certain conditions can be called quiescence, though we avoid this term as it can be controversial, and is not a necessary assumption to the model developed below)}.
+While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects}.
 Cancer cells are of course characterized by some degree of unregulated proliferation, thus it is appealing to explore the effects of differential proliferation rates within the cancer cell population.
@@
 Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
-We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will \sout{occur} \tcb{progress} and, if so, when.
+We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that premalignant tissue will progress and, if so, when.
 A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
-\tcb{
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 In doing so, we shift the focus of the previous model from cancer initiation to cancer progression.
 We do this to reflect the fact that cancer progression hinges on escape from the immune system and the fact that EMT has a more well-defined role during progression and metastasis.
-}
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
@@
 We then go on to analyze its behavior.
-Global sensitivity analysis identifies parameters that are crucial for \sout{tumorigenesis} \tcb{progression}.
-We then study these in more depth, focusing on the competing effects of EMT and of the immune system on \sout{tumorigenesis} \tcb{progression}: we find that EMT intricately regulates \sout{tumorigenesis} \tcb{progression} and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
-\sout{To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects predicted by the model for patients with pancreatic cancer.}
+Global sensitivity analysis identifies parameters that are crucial for progression.
+We then study these in more depth, focusing on the competing effects of EMT and of the immune system on progression: we find that EMT intricately regulates progression and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \tcb{To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.}
@@
 %%%
-\tcb{
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \begin{itemize}
@@
 \end{itemize}
-}
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \\
@@
 We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
-The agents in the model are the tissue cells \sout{with tumorigenic potential} \tcb{that have already formed a non-aggressive tumor yet have not progressed}.
-\sout{These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).}
-\tcb{
-In the process of the simulation, these cells can acquire new pathway mutations among three possible (shown in Fig. \ref{fig:ModelIntro}A).
-}
+% might want to define "driver mutation" in intro. See PreMalignant_Thoughts.txt for definition found in a source
+The agents in the model are the tissue cells that have already formed a premalignant tumor yet lack key driver mutations.
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+In the process of the simulation, these cells can acquire driver mutations in any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 If the pathway is mutated, the component is colored red and is marked with `X'.
-\sout{A single mutation is sufficient for the cell to be considered as a mutant cell.}
-\tcb{
-% I'm not a fan of "hypermutated", but I'll use it for now so that a quick replace all can change it later.
-We distinguish between cells that have zero of the three pathway mutations and those that have at least one by referring to the latter as hypermutated.
-}
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% I'm not a fan of "malignant", but I'll use it for now so that a quick replace all can change it later.
+We distinguish between cells that have zero driver mutations and those with at least one as premalignant and malignant, respectively.
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \par
-We model immune cells as continuous variables, \sout{appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.}
-\tcb{
+We model immune cells as continuous variables, 
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 effectively assuming that the TME is well-mixed with regards to the immune infiltrate.
-}
-The cytokine TGF-$\beta$ is \sout{\tcr{continuous} within the TME, i.e. the system is well-mixed.}
-\tcb{
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+The cytokine TGF-$\beta$ is
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 also assumed to be well-mixed in the TME so that all tissue cells have equal access to it.
-}
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
-\sout{While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
-\tcb{While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
-% Above line replaced
-In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the \sout{color} \tcb{shade of blue} of the cell.
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the shade of blue of the cell.
 Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.
@@
 \subsection{Tissue Cells}\label{TissueCells}
-The tissue cells at all times have associated with them two important quantities: which combination of mutations they carry and an EMT score.
+The tissue cells at all times have associated with them two important quantities: which combination of driver mutations they carry and an EMT score.
 The model has three pathways that have the potential to mutate over the course of a simulation.
@@
 Or suggesting that we are interested in immune activity above and beyond the baseline requirement to keep this tumor in check.
-The NKs and CTLs act on the system by recognizing \sout{mutated} \tcb{hypermutated} cells and clearing them.
+The NKs and CTLs act on the system by recognizing malignant cells and clearing them.
 Upon successful clearance, they themselves are then deactivated and removed from the immune population.
-Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear \sout{mutated} \tcb{hypermutated} cells.
+Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear malignant cells.
 In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.
 
-\tcb{
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \subsection{Inflammation} % How did we not have a section on this before?
@@
 When the inflammation switches to high, then a number of the parameters change values to reflect the additional immune activity.
-\tcr{See Table 2 in the supplement to see which parameters are changed by the inflammation score changing.} % Table 2 or S2? Change label in supplement?
-}
+See Table 2 in the supplement to see which parameters are changed by the inflammation score changing. % Table 2 or S2? Change label in supplement?
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 
@@
 Otherwise, it is below the threshold and labeled as epithelial.
-For the purpose of \tcb{parameter values} of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have \sout{some} \tcb{a subset} of its parameters updated.
+For the purpose of parameter values of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have a subset of its parameters updated.
 In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
-Note, that because the immune system can only clear \sout{mutated} \tcb{hypermutated} cells, the increased immune evasion only benefits \sout{mutated} \tcb{hypermutated} mesenchymal cells.
+Note, that because the immune system can only clear malignant cells, the increased immune evasion only benefits malignant mesenchymal cells.
 \par
-\tcb{
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
 In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
-}
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \subsection{Model Simulation}
@@
 \subsubsection{Initial conditions} 
-Every model run starts with \sout{a fixed amount} \tcb{an amount} of \tcb{tissue} cells, $N_0$ \tcb{, determined by the set of parameters used for the run.
+Every model run starts with an amount of premalignant tissue cells, $N_0$, determined by the set of parameters used for the run.
 A set of warmup cycles are run to ensure arrival to this steady state.
-During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.}
-\sout{However, different parameter values will lead to different steady states for the population of tissue cells.}
-\sout{In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.}
-\sout{Thus, the only immune cell present is NKs.}
+During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
@@
 % Add: 
-\tcb{This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.}
+This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.
 Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
@@
 For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
-First of all, the cell needs to be \sout{mutated} \tcb{hypermutated} ($\delta_{\text{MUT}}=1$).
+First of all, the cell needs to be malignant ($\delta_{\text{MUT}}=1$).
 Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
@@
 \paragraph{Rest in $G_0$} 
-\sout{The weight associated with quiescence is taken as 1 except in the case of mesenchymal cells.
-For mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
-Hence, the weight of quiescence is given by}
-\tcb{The weight associated with rest is taken as 1 except in the case of mesenchymal cells.
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+The weight associated with rest is taken as 1 except in the case of mesenchymal cells.
 Recall that mesenchymal cells had their proliferation rate decreased by $1-\zeta\Delta_{\text{MGA}}$ (see Eq. 2.1).
 The biological assumption here is that mesenchymal cells instead of proliferating will instead rest, so this lost proliferation weight is added to the resting weight.
-Hence, the weight of rest is given by}
+Hence, the weight of rest is given by
 % Replace quiescence with rest and explain more above
@@
 
-\tcb{Again, t}\sout{T}he reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.
+Again, the reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.
 
@@
 So, this is not really so much a TGFB received as it is a change in TGFB signaling.
-\sout{Each cell receives an amount of TGF-$\beta$ given by}
-\tcb{The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
 The saturation effect is to limit the amount of TGF-$\beta$ a cell can absorb in a given time interval.
 This quantity is then divided up randomly among the $N_C$ living cells via a normally distributed noise term to determine how much exogenous TGF-$\beta$ each cell receives during this cell cycle.
-Should this value, $\tau_i$ in Eq. 2.5, be negative, we interpret this as the cell losing TGF-$\beta$ to the TME and thus being more likely to undergo MET.}
+Should this value, $\tau_i$ in Eq. 2.5, be negative, we interpret this as the cell losing TGF-$\beta$ to the TME and thus being more likely to undergo MET.
 
@@
 
-\tcb{
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 We then combine $\tau_i$ with the current EMT score of the cell, as a proxy for the endogenous TGF-$\beta$.
-Finally, if this quantity is large enough, \sout{the cell undergoes EMT; otherwise, it undergoes MET.} \tcb{the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.}
-}
+Finally, if this quantity is large enough, the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 % Need to add better explanation for Hill fn
-\sout{The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.}
-\sout{Then, some randomness is introduced by the independent and identically distributed variables $X_i$ where $\sigma$ is a parameter in the model.
-If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
-If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
-}
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
@@
 
-For CTLs and Tregs, they rely on \sout{mutant} \tcb{hypermutated} cells being cleared before they can be activated.
-Let $N_{\text{MUT}}^*(k)$ represent the number of \sout{mutant} \tcb{hypermutated} cells cleared by the immune system during cell cycle $k$.
+For CTLs and Tregs, they rely on malignant cells being cleared before they can be activated.
+Let $N_{\text{MUT}}^*(k)$ represent the number of malignant cells cleared by the immune system during cell cycle $k$.
 In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
@@
  \hline
-  $\Delta_A$ & \tcr{mutant cells decreased apoptosis}  \\ % should this and the others like it be more clear in stating this is not for all hypermutated cells, but only those with apoptosis pathway mutation?
+  $\Delta_A$ & mutant cells decreased apoptosis  \\ % should this and the others like it be more clear in stating this is not for all malignant cells, but only those with apoptosis pathway mutation?
   \hline
@@
   \hline
-  $\sigma_\text{CTL}$ & CTL source rate per cleared \sout{mutant} \tcb{hypermutated} cell \\ 
+  $\sigma_\text{CTL}$ & CTL source rate per cleared malignant cell \\ 
   \hline
-  $\sigma_\text{Treg}$ & Treg source rate per cleared \sout{mutant} \tcb{hypermutated} cell \\ 
+  $\sigma_\text{Treg}$ & Treg source rate per cleared malignant cell \\ 
   \hline
@@
 % 3rd Referee did not like this terminology for a tumorigenic event. Should we change it or better explain it?
-At the end of each cell cycle, the proportion of tissue cells that are \sout{mutated} \tcb{hypermutated} is calculated, and if it is above a certain threshold, \sout{a tumorigenic event is recorded and the patient is determined to have cancer.}
-\tcb{the simulation ends and we declare the tumor to have progressed.}
-The Time to Cancer is given as the time from the start of the simulation, not including warmup, until \sout{tumorigenesis.} \tcb{this timepoint.}
-Simulations run until either a \sout{tumorigenic event is recorded} \tcb{this occurs} or until the \tcb{predetermined} maximum number of cycles has been reached.
+At the end of each cell cycle, the proportion of tissue cells that are malignant is calculated, and if it is above a certain threshold, 
+the simulation ends and we declare the tumor to have progressed.
+The Time to Cancer is given as the time from the start of the simulation, not including warmup, until this timepoint.
+Simulations run until either a this occurs or until the predetermined maximum number of cycles has been reached.
 
@@
 Some of the parameters were also truncated to the right because they represented a proportion, such as MIE and MGA, or because we simply wanted to keep the values within a reasonable range, such as the inflammation low and high duration parameters.
-The \sout{variances} \tcb{standard deviations} were chosen as twice the mean to get a large spread of values.
+The standard deviations were chosen as twice the mean to get a large spread of values.
 \par
 % Referee 2 either missed the following line or doesn't understand what average of absolute change means.
-\tcr{The algorithm computes the average of the absolute change of the output, which in our model is the area under the survival curve.}
+The algorithm computes the average of the absolute change of the output, which in our model is the area under the survival curve.
 This output of the algorithm, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.
@@
 {\includegraphics[width=0.9\textwidth]{Figure1/Figure1.pdf}}
-\caption{A. Schematic depiction of agent-based model components; \tcb{
+\caption{A. Schematic depiction of agent-based model components;
 each of the 10 columns represents a single tissue cell divided into three compartments representing the state (mutated or not) of the three pathways with mutagenic potential;
-}\sout{blue/red denote tissue cells with/without mutations} \tcb{red/blue denotes mutated/unmutated pathways} .
-Black arrows depict regulation of the cell fate in each cell cycle.
-Inset depicts major interactions between the immune system and tissue cells.
+red/blue denotes mutated/unmutated pathways. Black arrows depict regulation of the cell fate in each cell cycle. Inset depicts major interactions between the immune system and tissue cells.
  %The parameters used in this model can be found in the Supplementary material.
@@
 B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
-The inflammation cycling scheme is represented above the patient dynamics.
-The vertical dashed line denotes the end of the warmup period.
-Mut: \sout{mutated} \tcb{hypermutated} cells; Mes: mesenchymal cells.
+The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: malignant cells; Mes: mesenchymal cells. 
 C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
@@
 
-We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the \tcr{probability} of cancer, and the Time to Cancer.
+We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the probability of progression, and the Time to Cancer.
 Within the cell cycle, cell fate is determined via a set of rules that are influenced by EMT and immune interactions (Fig.
@@
 The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset).
-NK cells and CTLs attempt to clear \sout{mutated} \tcb{hypermutated} tissue cells, and deactivate upon successfully carrying out their cytotoxic function.
+NK cells and CTLs attempt to clear malignant tissue cells, and deactivate upon successfully carrying out their cytotoxic function.
 Tregs inhibit cytotoxic activity.
@@
 For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor.
-\sout{
-However, after approximately 1000 cell cycles (750 days), one mutated cell exhibits clonal growth, and a tumor is established.
-At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being heavily recruited, and a peak in TGF-$\beta$ expression is observed.
-The number of mesenchymal cells increases initially with increasing TGF-$\beta$, but decreases after the clone takes over the tissue.
-This shift back towards an epithelial tissue phenotype occurs even though the expression of TGF-$\beta$ remains high.
-After 841 cell cycles, the proportion of mutated cells reaches 50\%: this is modeled as the threshold defining tumorigenesis. Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days. 
-After cancer incidence, the number of mutated cells continues to grow rapidly and soon makes up 100\% of the cell population.
-A peak in mesenchymal phenotypes is observed following cancer incidence, before most transition back to an epithelial state.}
-\tcb{
-After about 700 cell cycles, the hypermutated cell population begins to grow steadily.
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+After about 700 cell cycles, the malignant cell population begins to grow steadily.
 At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being quickly recruited into the TME, and a peak in TGF-$\beta$ expression is observed.
-After 841 cell cycles, the proportion of hypermutated cells reaches 50\%: the threshold defining progression.
+After 841 cell cycles, the proportion of malignant cells reaches 50\%: the threshold defining progression.
 Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days.
-For this patient, we kept let the simulation go beyond this timepoint, and we se that the number of hypermutated cells grows rapidly and soon makes up 100\% of the cell population.
+For this patient, we kept let the simulation go beyond this timepoint, and we se that the number of malignant cells grows rapidly and soon makes up 100\% of the cell population.
 A peak in number of mesenchymal cells is observed shortly after this threshold is reached, and following this most cells transition back to epithelial.
-}
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \par 
-Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of \sout{oncogenic mutations} \tcb{hypermutations}.
+Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of driver mutations.
 The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
@@
 Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
-The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a \sout{mutated} \tcb{hypermutated} mesenchymal cell will be subject to immune clearance in a given cycle.
+The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a malignant mesenchymal cell will be subject to immune clearance in a given cycle.
 The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
@@
 As MIE is increased, the Time to Cancer decreases (Fig.
-\ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of \sout{potentially mutated} \tcb{hypermutated} cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
+\ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of malignant cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
 These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.
@@
 Second, as MGA increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Fig \ref{fig:FirstSurvivalCurves}B).
-This is not immediately intuitive, since the decreased proliferation rate affects \sout{both mutated and non-mutated tissue cells.} \tcb{all mesenchymal cells, not just hypermutated cells.}
+This is not immediately intuitive, since the decreased proliferation rate affects all mesenchymal cells, not just malignant cells.
 \par
@@
 C-D.
-As MGA varies, survival curves  and corresponding bar plots to summarize the mean time to cancer for each cohort are shown.}
+As MGA varies, survival curves  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.}
 \label{fig:VaryINFL_and_MesPars}
@@
 \par
-These striking differences in the mean Time to Cancer -- extended by up to one \tcb{{\it in silico}} year by fine-grained optimization of mesenchymal growth arrest -- have clear therapeutic implications.
-\sout{
+These striking differences in the mean Time to Cancer -- extended by up to one {\it in silico} year by fine-grained optimization of mesenchymal growth arrest -- have clear therapeutic implications. 
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 Our model predicts that reducing the inflammatory state of the TME in three weeks out of every nine would be provide means to prolong cancer-free survival in combination with mesenchymal cell therapies to make the control of mesenchymal proliferation an effective anti-cancer strategy.
-}
-\tcb{
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 Our model predicts that a patient suffering intermittent inflammatory attacks at the tumor site will benefit from controlling the proliferation of the mesenchymal cells.
-}
+% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \par
@@
 
-\tcr{THE REST OF THIS SECTION IS FROM THE FIRST DRAFT AND WILL BE DELETED}
-
-\par 
-To compare these model predictions with experimental studies, we analyzed data from The Cancer Genome Atlas (TCGA) database to study the effects of immune and EMT interactions on prognosis of cancers for which inflammation is known to play an important role, such as colonic or pancreatic cancers \cite{hu10_inflammationinduced, balkwill01_inflammation}.
-Using gene ontologies as a measure of the of effects of inflammatory or EMT processes on cancer-free survival, we investigate pancreatic cancer.
-We note that the comparison between simulation and data here is indirect since the model studies processes occurring before incidence, while the data address subsequent events, but analogous processes are at play during both the pre-cancerous clonal dynamics addressed by the model, and the progression of disease that follows cancer incidence.
-\par
-For a cohort of pancreatic cancer patients, we cluster the patients via k-means ($n=2$) either on gene ontologies relating only to EMT (Fig.
-\ref{fig:tcga}A) or on gene ontologies relating to EMT and inflammation (Fig.
-\ref{fig:tcga}C).
-We also plot the corresponding survival curves for each of the clusters obtained (Fig. \ref{fig:tcga}B, D).
-We see that in both cases, survival is affected by the gene ontology signature, but the presence of both EMT and inflammatory signatures has a greater impact on survival than the effects of EMT alone.
-This suggests that when studying interactions between cancer and the immune system, is it important to consider EMT; the effects of which may impact the dynamics and should not be overlooked.
-
 \begin{figure}
@@
 \section{Discussion}\label{Discussion}
-Despite the intense interest in cancer and the immune system, and in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these components: cancer, the immune system, and EMT.
+Despite the intense interest in cancer and the immune system, and in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these three components--cancer, the immune system, and EMT--during the progression from a premalignant to a malignant tumor.
 We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$.
@@
 We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model.
-There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in pre-cancerous tissues is a pressing need to shed insight into cellular biomarkers of cancer.
+There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in premalignant tissues is a pressing need to shed insight into cellular biomarkers of cancer.
 \par
-These results represent promising steps in understanding the competing roles of the immune system and EMT during (pre-) development of epithelial cancers, yet much remains to be done.
+These results represent promising steps in understanding the competing roles of the immune system and EMT during progression of epithelial cancers, yet much remains to be done.
 Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs.

### 2019-09-17 Paper/CleanPaper.tex
Added LIHC and PAAD to Paper and CleanPaper

@@
 % commented to keep same line numbers as Paper.txt (as of 9/16/19)
-\tcb{To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.}
+To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.
 % Consider rewriting above
@@
 \end{tabular}
-\caption{\tcb{Basic assumptions of the model}}
+\caption{Basic assumptions of the model}
 \label{table:model_assumptions}
@@
 We selected three tumor types whose TCGA-CDR profile indicated slow progression (OV, SKCM, LIHC) and three tumor types whose profile indicated rapid progression (PAAD, LUAD, COAD).
+See Figures \ref{fig:LIHC} and \ref{fig:PAAD} for LIHC and PAAD. The others will be in the supplement.
 \par
@@
 \center
-{\includegraphics[width=0.9\textwidth]{FigTCGA.pdf}}
-\caption{A.
-K-means clustering of pancreatic cancers using gene ontology terms indicative of an EMT signature ($k=2$).
-B.
-Survival plots corresponding to the clustering on EMT.
-C.
-K-means clustering of pancreatic cancers using gene ontology terms indicative of EMT and Inflammation signatures ($k=2$).
-D.
-Survival plots corresponding to the clustering on EMT and inflammation.}
-\label{fig:tcga}
+{\includegraphics[width=0.65\textwidth]{Figure6/LIHC.pdf}}
+\caption{A. K-means clustering of LIHC using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
+B. Survival plots corresponding to the clustering on EMT and inflammation.
+C. K-means clustering of LIHC using gene ontology terms indicative of an EMT signature ($k=2$).
+D. Survival plots corresponding to the clustering on EMT.
+E. K-means clustering of LIHC using gene ontology terms indicative of inflammation ($k=2$).
+F. Survival plots corresponding to the clustering on inflammation.}
+\label{fig:LIHC}
+\end{figure}
+
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{Figure6/PAAD.pdf}}
+\caption{A. K-means clustering of PAAD using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
+B. Survival plots corresponding to the clustering on EMT and inflammation.
+C. K-means clustering of PAAD using gene ontology terms indicative of an EMT signature ($k=2$).
+D. Survival plots corresponding to the clustering on EMT.
+E. K-means clustering of PAAD using gene ontology terms indicative of inflammation ($k=2$).
+F. Survival plots corresponding to the clustering on inflammation.}
+\label{fig:PAAD}
 \end{figure}

### 2019-09-17 Paper/Paper.tex
Added LIHC and PAAD to Paper and CleanPaper

@@
 % Does the following sentence still fit with our new focus on progression? I'm looking specifically at the words "pre-cancerous" and "tumorigenesis".
-\tcr{In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.}
+In addition, the immune system plays important regulatory roles in \sout{pre-cancerous} \tcb{premalignant} tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
 % Similar concern to above as Guo's paper dealt with initiation, not progression.
@@
 TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
-\tcb{There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.} Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
+\tcb{There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.}
+Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
 It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.
@@
 Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
-We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will \sout{occur} \tcb{progress} and, if so, when.
+We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that \sout{cancer} \tcb{premalignant tissue} will \sout{occur} \tcb{progress} and, if so, when.
 A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
@@
 We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
-The agents in the model are the tissue cells \sout{with tumorigenic potential} \tcb{that have already formed a non-aggressive tumor yet have not progressed}.
+% might want to define "driver mutation" in intro. See PreMalignant_Thoughts.txt for definition found in a source
+The agents in the model are the tissue cells \sout{with tumorigenic potential} \tcb{that have already formed a premalignant tumor yet lack key driver mutations}.
 \sout{These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).}
 \tcb{
-In the process of the simulation, these cells can acquire new pathway mutations among three possible (shown in Fig. \ref{fig:ModelIntro}A).
+In the process of the simulation, these cells can acquire driver mutations in any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
 }
@@
 \tcb{
-% I'm not a fan of "hypermutated", but I'll use it for now so that a quick replace all can change it later.
-We distinguish between cells that have zero of the three pathway mutations and those that have at least one by referring to the latter as hypermutated.
+% I'm not a fan of "malignant", but I'll use it for now so that a quick replace all can change it later.
+We distinguish between cells that have zero driver mutations and those with at least one as premalignant and malignant, respectively.
 }
@@
 \subsection{Tissue Cells}\label{TissueCells}
-The tissue cells at all times have associated with them two important quantities: which combination of mutations they carry and an EMT score.
+The tissue cells at all times have associated with them two important quantities: which combination of \tcb{driver} mutations they carry and an EMT score.
 The model has three pathways that have the potential to mutate over the course of a simulation.
@@
 Or suggesting that we are interested in immune activity above and beyond the baseline requirement to keep this tumor in check.
-The NKs and CTLs act on the system by recognizing \sout{mutated} \tcb{hypermutated} cells and clearing them.
+The NKs and CTLs act on the system by recognizing \sout{mutated} \tcb{malignant} cells and clearing them.
 Upon successful clearance, they themselves are then deactivated and removed from the immune population.
-Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear \sout{mutated} \tcb{hypermutated} cells.
+Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear \sout{mutated} \tcb{malignant} cells.
 In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.
@@
 In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
-Note, that because the immune system can only clear \sout{mutated} \tcb{hypermutated} cells, the increased immune evasion only benefits \sout{mutated} \tcb{hypermutated} mesenchymal cells.
+Note, that because the immune system can only clear \sout{mutated} \tcb{malignant} cells, the increased immune evasion only benefits \sout{mutated} \tcb{malignant} mesenchymal cells.
 \par
@@
 \subsubsection{Initial conditions} 
-Every model run starts with \sout{a fixed amount} \tcb{an amount} of \tcb{tissue} cells, $N_0$ \tcb{, determined by the set of parameters used for the run.
+Every model run starts with \sout{a fixed amount} \tcb{an amount} of \tcb{premalignant tissue} cells, $N_0$ \tcb{, determined by the set of parameters used for the run.
 A set of warmup cycles are run to ensure arrival to this steady state.
@@
 For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
-First of all, the cell needs to be \sout{mutated} \tcb{hypermutated} ($\delta_{\text{MUT}}=1$).
+First of all, the cell needs to be \sout{mutated} \tcb{malignant} ($\delta_{\text{MUT}}=1$).
 Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
@@
 
-For CTLs and Tregs, they rely on \sout{mutant} \tcb{hypermutated} cells being cleared before they can be activated.
-Let $N_{\text{MUT}}^*(k)$ represent the number of \sout{mutant} \tcb{hypermutated} cells cleared by the immune system during cell cycle $k$.
+For CTLs and Tregs, they rely on \sout{mutant} \tcb{malignant} cells being cleared before they can be activated.
+Let $N_{\text{MUT}}^*(k)$ represent the number of \sout{mutant} \tcb{malignant} cells cleared by the immune system during cell cycle $k$.
 In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
@@
  \hline
-  $\Delta_A$ & \tcr{mutant cells decreased apoptosis}  \\ % should this and the others like it be more clear in stating this is not for all hypermutated cells, but only those with apoptosis pathway mutation?
+  $\Delta_A$ & \tcr{mutant cells decreased apoptosis}  \\ % should this and the others like it be more clear in stating this is not for all malignant cells, but only those with apoptosis pathway mutation?
   \hline
@@
   \hline
-  $\sigma_\text{CTL}$ & CTL source rate per cleared \sout{mutant} \tcb{hypermutated} cell \\ 
+  $\sigma_\text{CTL}$ & CTL source rate per cleared \sout{mutant} \tcb{malignant} cell \\ 
   \hline
-  $\sigma_\text{Treg}$ & Treg source rate per cleared \sout{mutant} \tcb{hypermutated} cell \\ 
+  $\sigma_\text{Treg}$ & Treg source rate per cleared \sout{mutant} \tcb{malignant} cell \\ 
   \hline
@@
 % 3rd Referee did not like this terminology for a tumorigenic event. Should we change it or better explain it?
-At the end of each cell cycle, the proportion of tissue cells that are \sout{mutated} \tcb{hypermutated} is calculated, and if it is above a certain threshold, \sout{a tumorigenic event is recorded and the patient is determined to have cancer.}
+At the end of each cell cycle, the proportion of tissue cells that are \sout{mutated} \tcb{malignant} is calculated, and if it is above a certain threshold, \sout{a tumorigenic event is recorded and the patient is determined to have cancer.}
 \tcb{the simulation ends and we declare the tumor to have progressed.}
@@
 The vertical dashed line denotes the end of the warmup period.
-Mut: \sout{mutated} \tcb{hypermutated} cells; Mes: mesenchymal cells.
+Mut: \sout{mutated} \tcb{malignant} cells; Mes: mesenchymal cells.
 C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
@@
 
-We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the \tcr{probability} of cancer, and the Time to Cancer.
+We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the \tcr{probability} of \sout{cancer} \tcb{progression}, and the Time to Cancer.
 Within the cell cycle, cell fate is determined via a set of rules that are influenced by EMT and immune interactions (Fig.
@@
 The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset).
-NK cells and CTLs attempt to clear \sout{mutated} \tcb{hypermutated} tissue cells, and deactivate upon successfully carrying out their cytotoxic function.
+NK cells and CTLs attempt to clear \sout{mutated} \tcb{malignant} tissue cells, and deactivate upon successfully carrying out their cytotoxic function.
 Tregs inhibit cytotoxic activity.
@@
 \tcb{
-After about 700 cell cycles, the hypermutated cell population begins to grow steadily.
+After about 700 cell cycles, the malignant cell population begins to grow steadily.
 At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being quickly recruited into the TME, and a peak in TGF-$\beta$ expression is observed.
-After 841 cell cycles, the proportion of hypermutated cells reaches 50\%: the threshold defining progression.
+After 841 cell cycles, the proportion of malignant cells reaches 50\%: the threshold defining progression.
 Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days.
-For this patient, we kept let the simulation go beyond this timepoint, and we se that the number of hypermutated cells grows rapidly and soon makes up 100\% of the cell population.
+For this patient, we kept let the simulation go beyond this timepoint, and we se that the number of malignant cells grows rapidly and soon makes up 100\% of the cell population.
 A peak in number of mesenchymal cells is observed shortly after this threshold is reached, and following this most cells transition back to epithelial.
@@
 \par 
-Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of \sout{oncogenic mutations} \tcb{hypermutations}.
+Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of \sout{oncogenic mutations} \tcb{driver mutations}.
 The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
@@
 Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
-The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a \sout{mutated} \tcb{hypermutated} mesenchymal cell will be subject to immune clearance in a given cycle.
+The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a \sout{mutated} \tcb{malignant} mesenchymal cell will be subject to immune clearance in a given cycle.
 The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
@@
 As MIE is increased, the Time to Cancer decreases (Fig.
-\ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of \sout{potentially mutated} \tcb{hypermutated} cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
+\ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of \sout{potentially mutated} \tcb{malignant} cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
 These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.
@@
 Second, as MGA increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Fig \ref{fig:FirstSurvivalCurves}B).
-This is not immediately intuitive, since the decreased proliferation rate affects \sout{both mutated and non-mutated tissue cells.} \tcb{all mesenchymal cells, not just hypermutated cells.}
+This is not immediately intuitive, since the decreased proliferation rate affects \sout{both mutated and non-mutated tissue cells.} \tcb{all mesenchymal cells, not just malignant cells.}
 \par
@@
 We selected three tumor types whose TCGA-CDR profile indicated slow progression (OV, SKCM, LIHC) and three tumor types whose profile indicated rapid progression (PAAD, LUAD, COAD).
+\tcb{See Figures \ref{fig:LIHC} and \ref{fig:PAAD} for LIHC and PAAD. The others will be in the supplement.}
 \par
@@
 \section{Discussion}\label{Discussion}
-Despite the intense interest in cancer and the immune system, and in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these components: cancer, the immune system, and EMT.
-We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$.
+Despite the intense interest in cancer and the immune system, and in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these \tcb{three} components\sout{: cancer, the immune system, and EMT.} \tcb{--cancer, the immune system, and EMT--during the progression from a premalignant to a malignant tumor.} We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$.
 We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).
@@
 We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model.
-There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in pre-cancerous tissues is a pressing need to shed insight into cellular biomarkers of cancer.
+There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in \sout{pre-cancerous} \tcb{premalignant} tissues is a pressing need to shed insight into cellular biomarkers of cancer.
 \par
-These results represent promising steps in understanding the competing roles of the immune system and EMT during (pre-) development of epithelial cancers, yet much remains to be done.
+These results represent promising steps in understanding the competing roles of the immune system and EMT during \sout{(pre-) development} \tcb{progression} of epithelial cancers, yet much remains to be done.
 Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs.

### 2019-09-23 Paper/Paper.tex
swapped OV in for LIHC because of better looking clusters in EMT vs Infl plot

@@
 We selected three tumor types whose TCGA-CDR profile indicated slow progression (OV, SKCM, LIHC) and three tumor types whose profile indicated rapid progression (PAAD, LUAD, COAD).
-\tcb{See Figures \ref{fig:LIHC} and \ref{fig:PAAD} for LIHC and PAAD. The others will be in the supplement.}
+\tcb{See Figures \ref{fig:OV} and \ref{fig:PAAD} for OV and PAAD. The others will be in the supplement.}
 \par
@@
 \center
-{\includegraphics[width=0.65\textwidth]{Figure6/LIHC.pdf}}
-\caption{A. K-means clustering of LIHC using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
+{\includegraphics[width=0.65\textwidth]{Figure6/OV.pdf}}
+\caption{A. K-means clustering of OV using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
 B. Survival plots corresponding to the clustering on EMT and inflammation.
-C. K-means clustering of LIHC using gene ontology terms indicative of an EMT signature ($k=2$).
+C. K-means clustering of OV using gene ontology terms indicative of an EMT signature ($k=2$).
 D. Survival plots corresponding to the clustering on EMT.
-E. K-means clustering of LIHC using gene ontology terms indicative of inflammation ($k=2$).
+E. K-means clustering of OV using gene ontology terms indicative of inflammation ($k=2$).
 F. Survival plots corresponding to the clustering on inflammation.}
-\label{fig:LIHC}
+\label{fig:OV}
 \end{figure}

### 2019-09-23 Paper/CleanPaper.tex
updated CleanPaper similarly and put LIHC in Supplement

@@
 We selected three tumor types whose TCGA-CDR profile indicated slow progression (OV, SKCM, LIHC) and three tumor types whose profile indicated rapid progression (PAAD, LUAD, COAD).
-See Figures \ref{fig:LIHC} and \ref{fig:PAAD} for LIHC and PAAD. The others will be in the supplement.
+See Figures \ref{fig:OV} and \ref{fig:PAAD} for OV and PAAD. The others will be in the supplement.
 \par

### 2019-09-23 Paper/Supplement/Supplement_rewrite.tex
updated CleanPaper similarly and put LIHC in Supplement

@@
 
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure6/LIHC.pdf}}
+\caption{A. K-means clustering of LIHC using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
+B. Survival plots corresponding to the clustering on EMT and inflammation.
+C. K-means clustering of LIHC using gene ontology terms indicative of an EMT signature ($k=2$).
+D. Survival plots corresponding to the clustering on EMT.
+E. K-means clustering of LIHC using gene ontology terms indicative of inflammation ($k=2$).
+F. Survival plots corresponding to the clustering on inflammation.}
+\label{fig:LIHC}
+\end{figure}
+
 %\newpage

### 2019-10-02 Paper/Paper.tex
consolidated pathway/driver/hypermutated terminology; created quick guide box; changed supplement figure/table labeling

@@
 
+\noindent\fbox{%
+    \parbox{\textwidth}{%
+    Quick Guide to Equations and Assumptions
+    \newline
+
+    A premalignant tumor relies on driver mutations to alter key pathways that will allow the tumor to achieve greater growth potential and clinically progress to a malignant tumor.
+    At the same time, the immune system is recognizing and adapting to the resulting neoantigens to suppress the tumor.
+    All of this occurs within an environment that is often under the influence of inflammatory signals that can vary over time.
+    \newline
+
+    In our model, we assume that as the premalignant (and later the malignant) tumor cells proliferate, they become increasingly likely to acquire a driver mutation altering a key pathway that unlocks greater growth potential.
+    However, innate immune cells (Natural Killer cells) are present to identify and clear these tumor cells, a process which results in neoantigens being used to prime and activate T cells in local lymph nodes.
+    In turn, these T cells move to the tumor site and begin to either lyse tumor cells (cytotoxic T lymphocytes or CTLs) or suppress cytotoxic activity (T regulatory cells or Tregs).
+    We use the following expression to determine the probability of a tumor cell being lysed by a CTL in the timeframe of a cell cycle (18 hours):
+    $$\rho_{\text{CTL}} =\delta_{\text{MUT}} \frac{N_{\text{CTL}}}{N_C/K_{1}+N_{\text{CTL}}}  \frac{E_{\text{CTL}}}{1+N_{\text{Treg}}/K_2} (1-\delta_{\text{IE}}\Delta_{\text{IE}})(1-\zeta \Delta_{\text{MIE}})$$
+    Here, $\delta_\text{MUT}$ is 1 or 0 depending on whether the cell has a driver mutation or not, respectively.
+    The second term is a hill function similar to the de Pillis-Radunskaya law.
+    The third term is the regulatory effect of Tregs.
+    The final two terms are increased immune evasiveness due to a driver mutation in the immune evasion pathway and the immune evasiveness of mesenchymal cells (discussed below), respectively.
+    The inflammatory state of the tumor microenvironment affects the speed at which immune cells are recruited to the tumor.
+    \newline
+
+    Adding to the complexity of this situation, tumor cells can express heterogeneous cell types by their location along the epithelial-mesenchymal spectrum.
+    Depending on which end of this spectrum a tumor cell is on, its proliferation and its evasion of the immune system will change.
+    Specifically, if a cell is above a certain threshold, it will have its proliferation reduced by a fixed proportion and its immune evasiveness increased by a fixed proportion.
+    These cells also demonstrate cellular plasticity by being able to undergo epithelial-to-mesenchymal transition (EMT) and move along this spectrum as a result of intrinsic and extrinsic factors.
+    Here, we choose TGF-$\beta$ as a master regulator of EMT and connect the TGF-$\beta$ level in the tumor microenvironment (TME) with the propensity of tumor cells to undergo EMT.
+    As TGF-$\beta$ is produced by Tregs as well as by tumor cells, this creates an intriguing set of interlocking interactions that we show here to be capable of shaping tumor and patient outcomes.
+    To update the EMT state of a cell, we determine how much TGF-$\beta$ is present in the TME--based on the number of malignant cells and Tregs--and apply a Hill function to determine whether a cell undergoes EMT or MET (the reverse process).
+    For each cell, this quantity takes the following form:
+    $$
+    \tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{\tau/K_3}{1+\tau/K_3} + X_i, \quad X_i \sim N(0,\sigma^2)
+    $$
+    Here, $\tau$ is the concentration of TGF-$\beta$ in the TME and $\tau_\text{max}$ represents a limit on the amount of TGF-$\beta$ that can be absorbed by all cells.
+    A Gaussian noise term is added with standard deviation $\sigma$.
+    When this quantity is large enough relative to the current EMT score of the cell, the cell will transition towards the mesenchymal phenotype.
+    Otherwise, it transitions towards the epithelial phenotype.
+        }%
+}
+
 \section{Introduction}
-Cancer and the immune system interact in myriad ways.
+\tcb{
+A key question in understanding cancer is how tumors transition from a premalignant state into an aggressive, malignant state.
+This process is commonly referred to as progression \tcr{(citation needed!)} and results in a much worse prognosis for the patient.
+There are many means \tcr{(citation needed!)} by which a tumor can transition from a premalignant state into a malignant state, but one of the main ones is that of driver mutations.
+Driver mutations are those that "directly or indirectly [confer] a selective growth advantage to the cell in which it occurs." \cite{ryan2016hallmarks}.
+There are several identifiable cell processes that can be affected by such driver mutations, including proliferation, apoptosis, and immunogenicity.
+}
+\par
+\sout{Cancer and the immune system} \tcb{When it comes to cancer and the immune system, they} interact in myriad ways.
 Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME): responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}.
@@
 % Does the following sentence still fit with our new focus on progression? I'm looking specifically at the words "pre-cancerous" and "tumorigenesis".
-In addition, the immune system plays important regulatory roles in \sout{pre-cancerous} \tcb{premalignant} tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
+In addition, the immune system plays important regulatory roles in \sout{pre-cancerous} \tcb{premalignant} tissue, where inflammation can increase the probability of \sout{tumorigenesis} progression (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
 % Similar concern to above as Guo's paper dealt with initiation, not progression.
@@
 % Should we even mention initiation here? Make sure to combine below end-of-sentence refs into single \cite command.
-While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor \tcr{initiation} \tcb{\cite{puisieux2014oncogenic}} and progression \cite{nieto2016emt}\tcb{\cite{peinado2007snail}}.
+While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of tumor \sout{initiation} \tcb{\cite{puisieux2014oncogenic}} progression \cite{nieto2016emt}\tcb{\cite{peinado2007snail}}.
 TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
@@
 We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
-% might want to define "driver mutation" in intro. See PreMalignant_Thoughts.txt for definition found in a source
+% might want to define "driver mutation" in intro. See PreMalignant_Thoughts.txt for definition found in a source. Will do this.
 The agents in the model are the tissue cells \sout{with tumorigenic potential} \tcb{that have already formed a premalignant tumor yet lack key driver mutations}.
@@
 \tcb{
-In the process of the simulation, these cells can acquire driver mutations in any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
+In the process of the simulation, these cells can acquire driver mutations altering any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
 }
-If the pathway is mutated, the component is colored red and is marked with `X'.
+If the pathway is altered, the component is colored red and is marked with `X'.
 \sout{A single mutation is sufficient for the cell to be considered as a mutant cell.}
@@
 The tissue cells at all times have associated with them two important quantities: which combination of \tcb{driver} mutations they carry and an EMT score.
-The model has three pathways that have the potential to mutate over the course of a simulation.
-There is a proliferation pathway that when mutated increases the probability of the cell proliferating.
-There is an apoptosis pathway that when mutated decreases the probability of the cell undergoing apoptosis.
-There is an immune evasion pathway that when mutated decreases the probability that an immune cell can effectively clear the mutated cell.
+The model has three pathways that have the potential to become altered over the course of a simulation.
+There is a proliferation pathway that when altered increases the probability of the cell proliferating.
+There is an apoptosis pathway that when altered decreases the probability of the cell undergoing apoptosis.
+There is an immune evasion pathway that when altered decreases the probability that an immune cell can effectively clear the mutated cell.
 The EMT score will be exposited in Section \ref{EMT}.
@@
 When the inflammation switches to high, then a number of the parameters change values to reflect the additional immune activity.
-\tcr{See Table 2 in the supplement to see which parameters are changed by the inflammation score changing.} % Table 2 or S2? Change label in supplement?
+\tcr{See Table \tcb{S}2 in the supplement to see which parameters are changed by the inflammation score changing.}
 }
@@
 This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
-Second, all proliferating cells have a cell-specific probability of undergoing a pathway mutation.
-If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
+Second, all proliferating cells have a cell-specific probability of undergoing a \sout{pathway mutation} \tcb{driver mutation in one of the three pathways}.
+If they do, one is randomly chosen among the three pathways and the pathway in that cell \sout{mutates} \tcb{becomes altered}.
 If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases.
@@
 So, this is not really so much a TGFB received as it is a change in TGFB signaling.
+Try calling it a gradient/change of TGFB in the cell.
 \sout{Each cell receives an amount of TGF-$\beta$ given by}
@@
 }
-% Need to add better explanation for Hill fn
+% Need to add better explanation for Hill fn. Make sure thoroughly explained
 \sout{The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.}
@@
   $\Delta_A$ & \tcr{mutant cells decreased apoptosis}  \\ % should this and the others like it be more clear in stating this is not for all malignant cells, but only those with apoptosis pathway mutation?
+Better to be longer and unambiguous "decrease in apoptosis rate in cells with driver mutation in apoptosis pathway"
   \hline
@@
 
-\subsubsection{Mutation and tumorigenesis}
+\subsubsection{Mutation and \sout{tumorigenesis} \tcb{progression}}
 % 3rd Referee did not like this terminology for a tumorigenic event. Should we change it or better explain it?
@@
 
-\subsection{A multiscale agent-based model of EMT-immune-tissue cell interactions to study tumorigenesis}\label{ExplModel}
+\subsection{A multiscale agent-based model of EMT-immune-tissue cell interactions to study \sout{tumorigenesis} \tcb{tumor progression}}\label{ExplModel}
 
@@
 \caption{A. Schematic depiction of agent-based model components; \tcb{
-each of the 10 columns represents a single tissue cell divided into three compartments representing the state (mutated or not) of the three pathways with mutagenic potential;
-}\sout{blue/red denote tissue cells with/without mutations} \tcb{red/blue denotes mutated/unmutated pathways} .
+each of the 10 columns represents a single tissue cell divided into three compartments representing the state (altered or not) of the three pathways with mutagenic potential;
+}\sout{blue/red denote tissue cells with/without mutations} \tcb{red/blue denotes altered/unaltered pathways} .
 Black arrows depict regulation of the cell fate in each cell cycle.
@@
 Once I have the table, I will use the name of that table.}
-B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
+B. A representative simulation of one patient. The model parameter values used can be found in \sout{Supplementary} Table \tcb{S}2. 
 The inflammation cycling scheme is represented above the patient dynamics.
@@
 Mut: \sout{mutated} \tcb{malignant} cells; Mes: mesenchymal cells.
-C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
+C. Survival curve for one cohort of patients with the parameter values given in \sout{Supplementary} Table {S}2. }
 \label{fig:ModelIntro}
@@
 \ref{fig:ModelIntro}A).
-For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
+For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a \tcb{driver} mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
 The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset).
@@
 This shift back towards an epithelial tissue phenotype occurs even though the expression of TGF-$\beta$ remains high.
-After 841 cell cycles, the proportion of mutated cells reaches 50\%: this is modeled as the threshold defining tumorigenesis. Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days. 
+After 841 cell cycles, the proportion of mutated cells reaches 50\%: this is modeled as the threshold defining \sout{tumorigenesis} \tcb{progression}.
+Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days. 
 After cancer incidence, the number of mutated cells continues to grow rapidly and soon makes up 100\% of the cell population.
@@
 The ILD and IHD parameters determine the duration of low and high inflammation, respectively.
-A cell with mutated pathways will be affected by the $\Delta_A$, $\Delta_P$, and $\Delta_{\text{IE}}$ parameters, depending on whether the apoptosis, proliferation, or immune evasion pathways are mutated.
+A cell with \sout{mutated} \tcb{altered} pathways will be affected by the $\Delta_A$, $\Delta_P$, and $\Delta_{\text{IE}}$ parameters, depending on whether the apoptosis, proliferation, or immune evasion pathways are \sout{mutated} \tcb{altered}.
 The mesenchymal cells are influenced by the MGA and MIE parameters.
@@
 
-To further assess the effects on tumorigenesis of the mesenchymal properties of immune evasion and growth arrest, we study the joint density plot for each of these parameters against the Time to Cancer (Fig.
+To further assess the effects on \sout{tumorigenesis} \tcb{progression} of the mesenchymal properties of immune evasion and growth arrest, we study the joint density plot for each of these parameters against the Time to Cancer (Fig.
 \ref{fig:MIEvsMGA}).
@@
 \par
-For each cohort of cancer patients, we cluster the patients via k-means (n = 2) either on gene ontologies relating only to EMT (linked figures) or on gene ontologies relating to EMT and inflammation.
+For each cohort of cancer patients, we cluster the patients via k-means (n = 2) \sout{either on gene ontologies relating only to EMT (linked figures) or on gene ontologies relating to EMT and inflammation.} \tcb{on gene ontologies relating to only EMT (Figures \ref{fig:OV}C, \ref{fig:PAAD}C), only inflammation (Figures \ref{fig:OV}E, \ref{fig:PAAD}E), and both EMT and inflammation (Figures \ref{fig:OV}A, \ref{fig:PAAD}A).}
 We also plot the corresponding survival curves for each of the clusters obtained (utilizing the originally reported OS endpoint).
@@
 Despite the intense interest in cancer and the immune system, and in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these \tcb{three} components\sout{: cancer, the immune system, and EMT.} \tcb{--cancer, the immune system, and EMT--during the progression from a premalignant to a malignant tumor.} We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$.
-We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).
+We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to \tcb{driver} mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).
 \par

### 2019-10-06 Paper/Paper.tex
finished updating with edits for Adam. Next, will update clean version

@@
 \usepackage{authblk}
+\usepackage{framed}
 
@@
 
-\noindent\fbox{%
-    \parbox{\textwidth}{%
+\begin{framed}
     Quick Guide to Equations and Assumptions
 \newline
-
     A premalignant tumor relies on driver mutations to alter key pathways that will allow the tumor to achieve greater growth potential and clinically progress to a malignant tumor.
@@
 \newline
-
-    In our model, we assume that as the premalignant (and later the malignant) tumor cells proliferate, they become increasingly likely to acquire a driver mutation altering a key pathway that unlocks greater growth potential.
-    However, innate immune cells (Natural Killer cells) are present to identify and clear these tumor cells, a process which results in neoantigens being used to prime and activate T cells in local lymph nodes.
-    In turn, these T cells move to the tumor site and begin to either lyse tumor cells (cytotoxic T lymphocytes or CTLs) or suppress cytotoxic activity (T regulatory cells or Tregs).
-    We use the following expression to determine the probability of a tumor cell being lysed by a CTL in the timeframe of a cell cycle (18 hours):
+    To capture these dynamics, we built a non-spatial agent-based model where the tumor cells are the agents and immune cells are present as homogeneous populations.
+    The model updates at discrete time intervals approximating the duration of a cell cycle.
+    During each cycle, the tumor cells are allowed one of four options: proliferation, apoptosis, immune clearance, and rest.
+    The probability of proliferation decreases with increasing tumor cell population, corresponding to competition for resources, and is also modified by cell intrinsic factors.
+    For apoptosis, the probability depends only on the EMT state of the tumor cell.
+    The probability of immune clearance relies on the number of mutated cells, cytotoxic cells, and regulatory cells as well as cell-intrinsic factors.
+    \newline
+    We assume that as the premalignant (and later the malignant) tumor cells proliferate, they become increasingly likely to acquire a driver mutation altering a key pathway that unlocks greater growth potential.
+    However, innate immune cells (Natural Killer cells) are present to identify and clear these tumor cells, a process which results in neoantigens being used to prime and activate T cells in local lymph nodes and subsequently infiltrate the tumor microenvironment.
+    Once there, these T cells move to the tumor site and begin to either lyse tumor cells (cytotoxic T lymphocytes or CTLs) or suppress cytotoxic activity (T regulatory cells or Tregs).
+    We use the following expression to determine the probability of a tumor cell being lysed by a CTL in one time step of the model:
     $$\rho_{\text{CTL}} =\delta_{\text{MUT}} \frac{N_{\text{CTL}}}{N_C/K_{1}+N_{\text{CTL}}}  \frac{E_{\text{CTL}}}{1+N_{\text{Treg}}/K_2} (1-\delta_{\text{IE}}\Delta_{\text{IE}})(1-\zeta \Delta_{\text{MIE}})$$
@@
     \newline
-
     Adding to the complexity of this situation, tumor cells can express heterogeneous cell types by their location along the epithelial-mesenchymal spectrum.
@@
     Otherwise, it transitions towards the epithelial phenotype.
-        }%
-}
+    \newline
+    What we desire to know is when do the malignant tumor cells reach a particular threshold proportion of all tumor cells.
+    Once the tumor reaches this threshold, we declare that the tumor has progressed and we note how long the process took to reach this stage.
+\end{framed}
 
@@
 }
+\par
+\tcb{
+For the cells that have undergone EMT and are currently labeled in the model as mesenchymal, they experience two modifications to their governing parameters.
+The first is a reduction in proliferation, which we call mesenchymal growth arrest (MGA).
+This causes to mesenchymal cells to be less likely to undergo proliferation and more likely to rest during a cell cycle.
+The second is a decrease to their likelihood of being cleared by the immune system, which we call mesenchymal immune evasion (MIE). 
+This decreases the likelihood of mesenchymal cells being cleared by both NKs and CTLs during a cell cycle.
+Though we have not found clinical data on which to base parameter values for either of these, by their very nature they are proportions on the interval $[0,1]$.
+Thus, it is possible to sample from the entire parameter and in fact we do present results from a large subset of this parameter space and our sensitivity analysis fully explores this space.
+}
 \subsection{Model Simulation}
@@
 \subsubsection{Tissue cell fate}
-During each cell cycle, every cell randomly is assigned a cell fate from the following options:
-\begin{itemize}
-\item proliferation
-\item apoptosis
-\item immune clearance (by NKs or CTLs)
-\item rest in $G_0$
-\end{itemize}
+During each cell cycle, every cell is randomly assigned one of four fates: proliferation, apoptosis, immune clearance, and rest in $G_0$.
+Depending on both cell intrinsic, e.g.
+EMT and mutational state, and cell extrinsic, e.g.
+immune infiltrate and total number of tissue cells, each of the four options is given a probability of occurring for each cell.
+For proliferation, we assume the total number of tissue cells exerts a negative feedback on the proliferation probability for each cell.
+Also, cells that have a driver mutation in the proliferation pathway have an increased probability of proliferating.
+Meanwhile, mesenchymal cells have a decreased probability of proliferating and an increased probability of resting.
+For apoptosis, the probability is constant except for those cells with a driver mutation in the apoptosis pathway which have a decreased probability of apoptosis.
+For immune clearance, we assume that only cells with newly acquired driver mutations are targeted by the immune system, and their probability of being cleared depends on the immune infiltrate.
+This probability can be decreased by a cell having a driver mutation in the immune evasion pathway or by a cell being mesenchymal.
+Further details and equations are given in the supplement.
+%%%%%% MOVED TO SUPPLEMENT %%%%%%%%%%%%%
+%During each cell cycle, every cell randomly is assigned a cell fate from the following options:
+%\begin{itemize}
+%\item proliferation
+%\item apoptosis
+%\item immune clearance (by NKs or CTLs)
+%\item rest in $G_0$
+%\end{itemize}
+%
+%For each cell, a weight is chosen for each option and these are normalized to probabilities which then are used to randomly determine what each cell does during the cell cycle.
+%
+%\paragraph{Proliferation}
+%There are four factors that contribute to the weight of a cell to proliferate.
+%The first is a base proliferation rate that all cells have, $p$.
+%Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
+%Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
+%% Add: 
+%\tcb{This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.}
+%Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
+%In total, the weight for proliferation is given by
+%
+%\begin{equation}\tag{2.1}
+%\rho_P = p(1+\delta_{P}\Delta_P)(1-\zeta \Delta_{\text{MGA}})\frac{K_0}{K_0+N_C}
+%\end{equation}
+%
+%\paragraph{Apoptosis}
+%There are two factors that contribute to a cell's weight for undergoing apoptosis.
+%There is a basal apoptosis rate that all cells experience, $d_C$ for death.
+%Second, if the cell has a mutation in the apoptosis pathway ($\delta_A=1$), then the weight for undergoing apoptosis is proportionally decreased by $\Delta_A$.
+%In total, the weight for apoptosis is given by 
+%
+%\begin{equation}\tag{2.2}
+%\rho_A = d_C(1-\delta_{A}\Delta_A)
+%\end{equation}
+%
+%\paragraph{Immune Clearance}
+%For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
+%First of all, the cell needs to be \sout{mutated} \tcb{malignant} ($\delta_{\text{MUT}}=1$).
+%Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
+%Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
+%Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
+%Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
+%In total, the weight of NK clearance is given by
+%
+%\begin{equation}\tag{2.3}
+%\rho_{\text{NK}} =\delta_{\text{MUT}} \frac{N_{\text{NK}}}{N_C/K_{1}+N_{\text{NK}}}  \frac{E_{\text{NK}}}{1+N_{\text{Treg}}/K_2} (1-\delta_{\text{IE}}\Delta_{\text{IE}})(1-\zeta \Delta_{\text{MIE}})
+%\end{equation}
+%
+%A similar formula holds for CTLs with only the number of CTLs and their efficacy being different from the above equation.
+%
+%
+%\paragraph{Rest in $G_0$} 
+%\sout{The weight associated with quiescence is taken as 1 except in the case of mesenchymal cells.
+%For mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
+%Hence, the weight of quiescence is given by}
+%\tcb{The weight associated with rest is taken as 1 except in the case of mesenchymal cells.
+%Recall that mesenchymal cells had their proliferation rate decreased by $1-\zeta\Delta_{\text{MGA}}$ (see Eq. 2.1).
+%The biological assumption here is that mesenchymal cells instead of proliferating will instead rest, so this lost proliferation weight is added to the resting weight.
+%Hence, the weight of rest is given by}
+%% Replace quiescence with rest and explain more above
+%
+%\begin{equation}\tag{2.4}
+%\rho_R = 1 + \zeta p(1+\delta_P\Delta_P)\Delta_{\text{MGA}}\frac{K_0}{K_0+N_C}
+%\end{equation}
+%
+%\tcb{Again, t}\sout{T}he reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.
 
-For each cell, a weight is chosen for each option and these are normalized to probabilities which then are used to randomly determine what each cell does during the cell cycle.
-
-\paragraph{Proliferation}
-There are four factors that contribute to the weight of a cell to proliferate.
-The first is a base proliferation rate that all cells have, $p$.
-Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
-Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
-% Add: 
-\tcb{This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.}
-Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
-In total, the weight for proliferation is given by
-
-\begin{equation}\tag{2.1}
-\rho_P = p(1+\delta_{P}\Delta_P)(1-\zeta \Delta_{\text{MGA}})\frac{K_0}{K_0+N_C}
-\end{equation}
-
-\paragraph{Apoptosis}
-There are two factors that contribute to a cell's weight for undergoing apoptosis.
-There is a basal apoptosis rate that all cells experience, $d_C$ for death.
-Second, if the cell has a mutation in the apoptosis pathway ($\delta_A=1$), then the weight for undergoing apoptosis is proportionally decreased by $\Delta_A$.
-In total, the weight for apoptosis is given by
-
-\begin{equation}\tag{2.2}
-\rho_A = d_C(1-\delta_{A}\Delta_A)
-\end{equation}
-
-\paragraph{Immune Clearance}
-For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
-First of all, the cell needs to be \sout{mutated} \tcb{malignant} ($\delta_{\text{MUT}}=1$).
-Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
-Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
-Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
-Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
-In total, the weight of NK clearance is given by
-
-\begin{equation}\tag{2.3}
-\rho_{\text{NK}} =\delta_{\text{MUT}} \frac{N_{\text{NK}}}{N_C/K_{1}+N_{\text{NK}}}  \frac{E_{\text{NK}}}{1+N_{\text{Treg}}/K_2} (1-\delta_{\text{IE}}\Delta_{\text{IE}})(1-\zeta \Delta_{\text{MIE}})
-\end{equation}
-
-A similar formula holds for CTLs with only the number of CTLs and their efficacy being different from the above equation.
-
-\paragraph{Rest in $G_0$} 
-\sout{The weight associated with quiescence is taken as 1 except in the case of mesenchymal cells.
-For mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
-Hence, the weight of quiescence is given by}
-\tcb{The weight associated with rest is taken as 1 except in the case of mesenchymal cells.
-Recall that mesenchymal cells had their proliferation rate decreased by $1-\zeta\Delta_{\text{MGA}}$ (see Eq. 2.1).
-The biological assumption here is that mesenchymal cells instead of proliferating will instead rest, so this lost proliferation weight is added to the resting weight.
-Hence, the weight of rest is given by}
-% Replace quiescence with rest and explain more above
-
-\begin{equation}\tag{2.4}
-\rho_R = 1 + \zeta p(1+\delta_P\Delta_P)\Delta_{\text{MGA}}\frac{K_0}{K_0+N_C}
-\end{equation}
-
-\tcb{Again, t}\sout{T}he reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.
+%%%%%% END MOVED TO SUPPLEMENT %%%%%%%%%%%%%
 
 \subsubsection{Completing the Cell Cycle}
-After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
-First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
-This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
-Second, all proliferating cells have a cell-specific probability of undergoing a \sout{pathway mutation} \tcb{driver mutation in one of the three pathways}.
-If they do, one is randomly chosen among the three pathways and the pathway in that cell \sout{mutates} \tcb{becomes altered}.
-If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases.
+Once all the cell fates are carried out, the rest of the components of the model update to account for these changes.
+First, the immune populations are updated in two steps.
+In step one, Immune cells are exhausted based on the number of malignant cells that were cleared by their type.
+So, each clearance by an NK cell results in the NK population decreasing by one, and similarly for CTLs.
+In step two, all three immune populations are updated deterministically based on a simple ODE model of their dynamics.
+NKs have a constant source and death rate while CTLs and Tregs have a source rate that is dependent on the number of malignant cells cleared that cycle and a constant death rate.
+In addition, the concentration of TGF-$\beta$ mediates recruitment of Tregs with higher concentrations of TGF-$\beta$ resulting in a greater influx of Tregs.
 \par
-Finally, the EMT values for each cell is updated.
-This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
-% The following sentence could be altered to reflect that the threshold value for $\tau$ is 0, so not only are negative values possible, but very common.
+Also at this time, new driver mutations have a chance to occur.
+The cells that proliferated are the only ones that have a probability of acquiring a new driver mutation.
+This probability is cell-specific and is increased if no mutation occurs and reset to 0 if a mutation does occur.
+\par
+Finally, the concentration of TGF-$\beta$ is updated as well as the EMT score for each cell is updated.
+Malignant cells and Tregs are the two sources of TGF-$\beta$ each producing a constant amount.
+Tregs are assumed to produce much more TGF-$\beta$ than malignant cells.
+The resulting quantity of TGF-$\beta$ is divided up randomly among the tissue cells.
+This quantity is combined with the EMT score of each cell and the resulting quantity determines which cells undergo EMT and which undergo MET.
+The equations governing these processes can be found in the supplement.
+%After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
+%First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
+%This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
+%Second, all proliferating cells have a cell-specific probability of undergoing a \sout{pathway mutation} \tcb{driver mutation in one of the three pathways}.
+%If they do, one is randomly chosen among the three pathways and the pathway in that cell \sout{mutates} \tcb{becomes altered}.
+%If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases.
+%\par
+%Finally, the EMT values for each cell is updated.
+%This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
+%% The following sentence could be altered to reflect that the threshold value for $\tau$ is 0, so not only are negative values possible, but very common.
 So, this is not really so much a TGFB received as it is a change in TGFB signaling.
 Try calling it a gradient/change of TGFB in the cell.
-\sout{Each cell receives an amount of TGF-$\beta$ given by}
-\tcb{The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
-The saturation effect is to limit the amount of TGF-$\beta$ a cell can absorb in a given time interval.
-This quantity is then divided up randomly among the $N_C$ living cells via a normally distributed noise term to determine how much exogenous TGF-$\beta$ each cell receives during this cell cycle.
-Should this value, $\tau_i$ in Eq. 2.5, be negative, we interpret this as the cell losing TGF-$\beta$ to the TME and thus being more likely to undergo MET.}
-
-\begin{equation}\tag{2.5}
-\tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{\tau/K_3}{1+\tau/K_3} + X_i
-, \quad X_i \sim N(0,\sigma^2)
-\end{equation}
-
-\tcb{
-We then combine $\tau_i$ with the current EMT score of the cell, as a proxy for the endogenous TGF-$\beta$.
-Finally, if this quantity is large enough, \sout{the cell undergoes EMT; otherwise, it undergoes MET.} \tcb{the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.}
-}
-% Need to add better explanation for Hill fn. Make sure thoroughly explained
-\sout{The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.}
-\sout{Then, some randomness is introduced by the independent and identically distributed variables $X_i$ where $\sigma$ is a parameter in the model.
-If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
-If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
-}
-Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
-Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: concentration of TGF-$\beta$ in the system and the current EMT score of the cell.
-
-Next, the amount of TGF-$\beta$ for the next cell cycle is determined by the number of mutated cells, $N_{\text{MUT}}$, and the number of Treg cells, $N_{\text{Treg}}$, each one producing a fixed amount of TGF-$\beta$.
+%\sout{Each cell receives an amount of TGF-$\beta$ given by}
+%\tcb{The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
+%The saturation effect is to limit the amount of TGF-$\beta$ a cell can absorb in a given time interval.
+%This quantity is then divided up randomly among the $N_C$ living cells via a normally distributed noise term to determine how much exogenous TGF-$\beta$ each cell receives during this cell cycle.
+%Should this value, $\tau_i$ in Eq. 2.5, be negative, we interpret this as the cell losing TGF-$\beta$ to the TME and thus being more likely to undergo MET.}	
+%
+%\begin{equation}\tag{2.5}
+%\tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{\tau/K_3}{1+\tau/K_3} + X_i
+%, \quad X_i \sim N(0,\sigma^2)
+%\end{equation}
+%
+%
+%\tcb{
+%We then combine $\tau_i$ with the current EMT score of the cell, as a proxy for the endogenous TGF-$\beta$.
+%Finally, if this quantity is large enough, \sout{the cell undergoes EMT; otherwise, it undergoes MET.} \tcb{the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.}
+%}
+%% Need to add better explanation for Hill fn. Make sure thoroughly explained
+%\sout{The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.}
+%\sout{Then, some randomness is introduced by the independent and identically distributed variables $X_i$ where $\sigma$ is a parameter in the model.
+%If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
+%If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
+%}
+%Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
+%Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: concentration of TGF-$\beta$ in the system and the current EMT score of the cell.
+%
+%Next, the amount of TGF-$\beta$ for the next cell cycle is determined by the number of mutated cells, $N_{\text{MUT}}$, and the number of Treg cells, $N_{\text{Treg}}$, each one producing a fixed amount of TGF-$\beta$.
 It is given by
-
-\begin{equation}\tag{2.6}
-\tau = \tau_{\text{MUT}}N_{\text{MUT}} + \tau_{\text{Treg}}N_{\text{Treg}}
-\end{equation}
-
-Finally, the immune populations are updated.
-For the NKs, they obey the following differential equation:
-
-\begin{equation}\tag{2.7}
-N_{\text{NK}}' = \sigma_{\text{NK}} - d_{\text{NK}}N_{\text{NK}}
-\end{equation}
-
-which is discretized to
-
- \begin{equation}\tag{2.8}
-N_{\text{NK}}(k+1) = \left (N_{\text{NK}}(k)-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}}
-\end{equation}
-
-For CTLs and Tregs, they rely on \sout{mutant} \tcb{malignant} cells being cleared before they can be activated.
-Let $N_{\text{MUT}}^*(k)$ represent the number of \sout{mutant} \tcb{malignant} cells cleared by the immune system during cell cycle $k$.
-In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
-We choose the following differential equations to govern the CTL and Treg populations:
-
-\begin{equation}\tag{2.9}
-\begin{split}
-N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}N_\text{CTL} \\
-N_\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
-\end{split}
-\end{equation}
-
-Discretized, these are:
-
-\begin{equation}\tag{2.10}
-\begin{split}
-N_\text{CTL}(k+1) & =  \left (N_\text{CTL}(k)-\sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\\
-N_\text{Treg}(k+1) & =  \left (N_\text{Treg}(k)-\frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}\right )\exp(-d_{\text{Treg}}\Delta t)\\
-&+ \frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}
-\end{split}
-\end{equation}
+%
+%\begin{equation}\tag{2.6}
+%\tau = \tau_{\text{MUT}}N_{\text{MUT}} + \tau_{\text{Treg}}N_{\text{Treg}}
+%\end{equation}
+%
+%
+%Finally, the immune populations are updated.
+%For the NKs, they obey the following differential equation:
+% 
+%\begin{equation}\tag{2.7}
+%N_{\text{NK}}' = \sigma_{\text{NK}} - d_{\text{NK}}N_{\text{NK}}
+%\end{equation}
+%
+%which is discretized to
+% 
+% \begin{equation}\tag{2.8}
+%N_{\text{NK}}(k+1) = \left (N_{\text{NK}}(k)-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}}
+%\end{equation}
+%
+%For CTLs and Tregs, they rely on \sout{mutant} \tcb{malignant} cells being cleared before they can be activated.
+%Let $N_{\text{MUT}}^*(k)$ represent the number of \sout{mutant} \tcb{malignant} cells cleared by the immune system during cell cycle $k$.
+%In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
+%We choose the following differential equations to govern the CTL and Treg populations:
+%
+%\begin{equation}\tag{2.9}
+%\begin{split}
+%N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}N_\text{CTL} \\
+%N_\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
+%\end{split}
+%\end{equation}
+%
+%Discretized, these are:
+%
+%\begin{equation}\tag{2.10}
+%\begin{split}
+%N_\text{CTL}(k+1) & =  \left (N_\text{CTL}(k)-\sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\\
+%N_\text{Treg}(k+1) & =  \left (N_\text{Treg}(k)-\frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}\right )\exp(-d_{\text{Treg}}\Delta t)\\
+%&+ \frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}
+%\end{split}
+%\end{equation}
 
@@
 \begin{center}
- \begin{tabular}{|| c | c||} 
+ \begin{tabular}{|| c | c ||} 
  \hline
@@
  \hline
-  $\Delta_A$ & \tcr{mutant cells decreased apoptosis}  \\ % should this and the others like it be more clear in stating this is not for all malignant cells, but only those with apoptosis pathway mutation?
+  $\Delta_A$ & \tcb{decrease in apoptosis rate in cells with driver mutation in apoptosis pathway}  \\ % should this and the others like it be more clear in stating this is not for all malignant cells, but only those with apoptosis pathway mutation?
 Better to be longer and unambiguous "decrease in apoptosis rate in cells with driver mutation in apoptosis pathway"
   \hline
-  $\Delta_\text{IE}$ & mutant cells increased immune evasion  \\
+  $\Delta_\text{IE}$ & \tcb{increase in immune evasion in cells with driver mutation in immune evasion pathway} \\
   \hline
-  $\Delta_P$ & mutant cells increased proliferation  \\
+  $\Delta_P$ & \tcb{increase in proliferation in cells with driver mutation in proliferation pathway}\\
   \hline
@@
 \end{tabular}
-  \caption{The model parameter names and descriptions. Note that many of these values are affected by the inflammation state of the system.}
+  \caption{The model parameter names and descriptions. Note that \sout{many} \tcb{some} of these values are affected by the inflammation state of the system.}
 \end{center}

### 2019-10-06 Paper/Supplement/Supplement_rewrite.tex
consolidated pathway/driver/hypermutated terminology; created quick guide box; changed supplement figure/table labeling
finished updating with edits for Adam. Next, will update clean version

@@
 \usepackage{amsmath, amssymb, graphicx, outline, float, cite, color}
+\usepackage{longtable,array}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
 
@@
 
+\section{Model Description}
+
+\subsection{Tissue cell fate}
+During each cell cycle, every cell randomly is assigned a cell fate from the following options:
+\begin{itemize}
+\item proliferation
+\item apoptosis
+\item immune clearance (by NKs or CTLs)
+\item rest in $G_0$
+\end{itemize}
+
+For each cell, a weight is chosen for each option and these are normalized to probabilities which then are used to randomly determine what each cell does during the cell cycle.
+
+\subsubsection{Proliferation}
+There are four factors that contribute to the weight of a cell to proliferate.
+The first is a base proliferation rate that all cells have, $p$.
+Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
+Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
+% Add: 
+This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.
+Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
+In total, the weight for proliferation is given by
+
+\begin{equation}\tag{2.1}
+\rho_P = p(1+\delta_{P}\Delta_P)(1-\zeta \Delta_{\text{MGA}})\frac{K_0}{K_0+N_C}
+\end{equation}
+
+\subsubsection{Apoptosis}
+There are two factors that contribute to a cell's weight for undergoing apoptosis.
+There is a basal apoptosis rate that all cells experience, $d_C$ for death.
+Second, if the cell has a mutation in the apoptosis pathway ($\delta_A=1$), then the weight for undergoing apoptosis is proportionally decreased by $\Delta_A$.
+In total, the weight for apoptosis is given by
+
+\begin{equation}\tag{2.2}
+\rho_A = d_C(1-\delta_{A}\Delta_A)
+\end{equation}
+
+\subsubsection{Immune Clearance}
+For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
+First of all, the cell needs to be malignant ($\delta_{\text{MUT}}=1$).
+Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
+Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
+Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
+Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
+In total, the weight of NK clearance is given by
+
+\begin{equation}\tag{2.3}
+\rho_{\text{NK}} =\delta_{\text{MUT}} \frac{N_{\text{NK}}}{N_C/K_{1}+N_{\text{NK}}}  \frac{E_{\text{NK}}}{1+N_{\text{Treg}}/K_2} (1-\delta_{\text{IE}}\Delta_{\text{IE}})(1-\zeta \Delta_{\text{MIE}})
+\end{equation}
+
+A similar formula holds for CTLs with only the number of CTLs and their efficacy being different from the above equation.
+
+\subsubsection{Rest in $G_0$} 
+The weight associated with rest is taken as 1 except in the case of mesenchymal cells.
+Recall that mesenchymal cells had their proliferation rate decreased by $1-\zeta\Delta_{\text{MGA}}$ (see Eq. 2.1).
+The biological assumption here is that mesenchymal cells instead of proliferating will instead rest, so this lost proliferation weight is added to the resting weight.
+Hence, the weight of rest is given by
+
+\begin{equation}\tag{2.4}
+\rho_R = 1 + \zeta p(1+\delta_P\Delta_P)\Delta_{\text{MGA}}\frac{K_0}{K_0+N_C}
+\end{equation}
+
+Again, the reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.
+
+\subsubsection{Completing the Cell Cycle}
+After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
+First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
+This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
+Second, all proliferating cells have a cell-specific probability of undergoing a driver mutation in one of the three pathways.
+If they do, one is randomly chosen among the three pathways and the pathway in that cell becomes altered.
+If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases.
+\par
+Finally, the EMT values for each cell is updated.
+This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
+% The following sentence could be altered to reflect that the threshold value for $\tau$ is 0, so not only are negative values possible, but very common.
+So, this is not really so much a TGFB received as it is a change in TGFB signaling.
+Try calling it a gradient/change of TGFB in the cell.
+The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
+The saturation effect is to limit the amount of TGF-$\beta$ a cell can absorb in a given time interval.
+This quantity is then divided up randomly among the $N_C$ living cells via a normally distributed noise term to determine how much exogenous TGF-$\beta$ each cell receives during this cell cycle.
+Should this value, $\tau_i$ in Eq. 2.5, be negative, we interpret this as the cell losing TGF-$\beta$ to the TME and thus being more likely to undergo MET.
+
+\begin{equation}\tag{2.5}
+\tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{\tau/K_3}{1+\tau/K_3} + X_i
+, \quad X_i \sim N(0,\sigma^2)
+\end{equation}
+
+We then combine $\tau_i$ with the current EMT score of the cell, as a proxy for the endogenous TGF-$\beta$.
+Finally, if this quantity is large enough, the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.
+% Need to add better explanation for Hill fn. Make sure thoroughly explained
+Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
+Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: concentration of TGF-$\beta$ in the system and the current EMT score of the cell.
+
+Next, the amount of TGF-$\beta$ for the next cell cycle is determined by the number of mutated cells, $N_{\text{MUT}}$, and the number of Treg cells, $N_{\text{Treg}}$, each one producing a fixed amount of TGF-$\beta$.
+It is given by
+
+\begin{equation}\tag{2.6}
+\tau = \tau_{\text{MUT}}N_{\text{MUT}} + \tau_{\text{Treg}}N_{\text{Treg}}
+\end{equation}
+
+Finally, the immune populations are updated.
+For the NKs, they obey the following differential equation:
+
+\begin{equation}\tag{2.7}
+N_{\text{NK}}' = \sigma_{\text{NK}} - d_{\text{NK}}N_{\text{NK}}
+\end{equation}
+
+which is discretized to
+
+ \begin{equation}\tag{2.8}
+N_{\text{NK}}(k+1) = \left (N_{\text{NK}}(k)-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}}
+\end{equation}
+
+For CTLs and Tregs, they rely on malignant cells being cleared before they can be activated.
+Let $N_{\text{MUT}}^*(k)$ represent the number of malignant cells cleared by the immune system during cell cycle $k$.
+In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
+We choose the following differential equations to govern the CTL and Treg populations:
+
+\begin{equation}\tag{2.9}
+\begin{split}
+N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}N_\text{CTL} \\
+N_\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
+\end{split}
+\end{equation}
+
+Discretized, these are:
+
+\begin{equation}\tag{2.10}
+\begin{split}
+N_\text{CTL}(k+1) & =  \left (N_\text{CTL}(k)-\sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\\
+N_\text{Treg}(k+1) & =  \left (N_\text{Treg}(k)-\frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}\right )\exp(-d_{\text{Treg}}\Delta t)\\
+&+ \frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}
+\end{split}
+\end{equation}
+
 \section{Definition of parameters specifying the model}
@@
 \begin{center}
- \begin{tabular}{|| c | c||} 
+ \begin{tabular}{| c | c|} 
  \hline
@@
 
+\newpage
 \section{Parameter values used for simulation}
 
-\begin{table}[H]
-\begin{center}
- \begin{tabular}{|| c | p{7cm} | c | c ||} 
+\begin{longtable}{| c | p{5cm} | m{3cm} | m{3cm} |} 
  \hline
@@
   \hline
-\end{tabular}
   \caption{The model parameter names, descriptions, and values during both low and high inflammation. Parameters with only one value do not change with the inflammatory state.}
-\end{center}
-\end{table}
+\end{longtable}
+
+%\begin{table}[H]
+%\begin{center}
+% \begin{tabular}{|| c | p{5cm} | p{3cm} | p{3cm} ||} 
+% \hline
+% Name & Description & INFL Low Value & INFL High Value  \\ [0.5ex] 
+% \hline\hline
+%  $p$ & weight of proliferation for tissue cells & 0.28 & \\ 
+% \hline
+% $d_C$  & weight of apoptosis for tissue cells & 0.14 & \\
+% \hline
+%$\Delta_\text{MIE}$ & MIE & 0.6 & \\
+% \hline
+% $\Delta_\text{MGA}$ & MGA  & 0.2 &  \\
+% \hline
+%  $\Delta_A$ & proportional decrease to weight of apoptosis for cells with mutated apoptosis pathway & 0.3 &  \\
+%  \hline
+%  $\Delta_\text{IE}$ & proportional increase to weight of immune evasion for cells with mutated immune evasion pathway & 0.48 &  \\
+%  \hline
+%  $\Delta_P$ & proportional increase to weight of proliferation for cells with mutated proliferation pathway & 0.36 &  \\
+%  \hline
+% $K_0$ & EC50 term for negative feedback of tissue cells on own proliferation & 80 cells& \\
+% \hline
+% $K_1$ & EC50 term for probability of NK cell finding mutant cell & 8 cells& \\
+% \hline
+%  $K_2$ & EC50 term for Treg inhibition of cytotoxic functions & 5 cells / volume & 0.025 cells / volume \\
+%  \hline
+%  $K_3$ & EC50 term for cumulative absorption of TGF-$\beta$ & 200 amount / volume & \\
+%  \hline
+%  $K_4$ & EC50 term for TGF-$\beta$ activation of Tregs & 50 amount / volume & \\
+%  \hline
+% $E_\text{NK}$ & weight of NKs clearing mutants & 10 & 30  \\
+%  \hline
+%  $E_\text{CTL}$ & weight of CTLs clearing mutants & 200 & 600 \\
+%  \hline
+%  $\sigma_\text{NK}$ & NK source rate & 1.3 cells / cycle &  \\ 
+%  \hline
+%  $\sigma_\text{CTL}$ & CTL source rate per cleared mutant cell & 100 cells / (cleared mutants $\times$ cycles)& \\ 
+%  \hline
+%  $\sigma_\text{Treg}$ & Treg source rate per cleared mutant cell & 200 cells / (cleared mutants $\times$ concentration of TGF-$\beta$ $\times$ cycles)& \\ 
+%  \hline
+%  $d_\text{NK}$ & NK death rate & 0.13 / cycle &  \\ 
+%  \hline
+%  $d_\text{CTL}$ & CTL death rate & 0.0260 / cycle& \\ 
+%  \hline
+%  $d_\text{Treg}$ & Treg death rate & 0.0260 / cycle& \\ 
+%  \hline
+%  $k_\text{EMT}$ & EMT/MET rate & 0.01 / concentration of TGF-$\beta$ & \\
+%  \hline
+%  $\sigma$ & standard deviation of noise in TGF-$\beta$ each cell receives & 6 concentration of TGF-$\beta$ & \\
+%  \hline
+% $\tau_\text{max}$ & max amount of TGF-$\beta$ any cell can receive & 500 concentration of TGF-$\beta$ &\\
+%  \hline 
+% $\tau_\text{MUT}$ & rate of TGF-$\beta$ production by mutant cells & 0.05 concentration of TGF-$\beta$ / cell / cycle& \\
+%  \hline
+% $\tau_\text{Treg}$ & rate of TGF-$\beta$ production by Treg & 0.5 concentration of TGF-$\beta$ / cell / cycle& \\
+%  \hline
+%  & RP Cancer Line  & 0.5 &  \\ 
+% \hline
+%&  INFL High Duration & 30 cycles&   \\
+% \hline
+% & INFL Low Duration &30 cycles&   \\
+% \hline
+%& Mes Threshold & 0.7 &    \\
+% \hline
+%  & maximum initial mutation damage after warmup  & 0.01 &\\
+%  \hline
+%  & increase in probability to mutate for non-mutating proliferating cells & 0.0001 & \\
+%  \hline
+%\end{tabular}
+%  \caption{The model parameter names, descriptions, and values during both low and high inflammation. Parameters with only one value do not change with the inflammatory state.}
+%\end{center}
+%\end{table}
 

### 2019-10-06 Paper/CleanPaper.tex
cleaned up main text and supplement

@@
 
+\begin{framed}
+    Quick Guide to Equations and Assumptions
+\newline
+    A premalignant tumor relies on driver mutations to alter key pathways that will allow the tumor to achieve greater growth potential and clinically progress to a malignant tumor.
+    At the same time, the immune system is recognizing and adapting to the resulting neoantigens to suppress the tumor.
+    All of this occurs within an environment that is often under the influence of inflammatory signals that can vary over time.
+\newline
+    To capture these dynamics, we built a non-spatial agent-based model where the tumor cells are the agents and immune cells are present as homogeneous populations.
+    The model updates at discrete time intervals approximating the duration of a cell cycle.
+    During each cycle, the tumor cells are allowed one of four options: proliferation, apoptosis, immune clearance, and rest.
+    The probability of proliferation decreases with increasing tumor cell population, corresponding to competition for resources, and is also modified by cell intrinsic factors.
+    For apoptosis, the probability depends only on the EMT state of the tumor cell.
+    The probability of immune clearance relies on the number of mutated cells, cytotoxic cells, and regulatory cells as well as cell-intrinsic factors.
+    \newline
+    We assume that as the premalignant (and later the malignant) tumor cells proliferate, they become increasingly likely to acquire a driver mutation altering a key pathway that unlocks greater growth potential.
+    However, innate immune cells (Natural Killer cells) are present to identify and clear these tumor cells, a process which results in neoantigens being used to prime and activate T cells in local lymph nodes and subsequently infiltrate the tumor microenvironment.
+    Once there, these T cells move to the tumor site and begin to either lyse tumor cells (cytotoxic T lymphocytes or CTLs) or suppress cytotoxic activity (T regulatory cells or Tregs).
+    We use the following expression to determine the probability of a tumor cell being lysed by a CTL in one time step of the model:
+    $$\rho_{\text{CTL}} =\delta_{\text{MUT}} \frac{N_{\text{CTL}}}{N_C/K_{1}+N_{\text{CTL}}}  \frac{E_{\text{CTL}}}{1+N_{\text{Treg}}/K_2} (1-\delta_{\text{IE}}\Delta_{\text{IE}})(1-\zeta \Delta_{\text{MIE}})$$
+    Here, $\delta_\text{MUT}$ is 1 or 0 depending on whether the cell has a driver mutation or not, respectively.
+    The second term is a hill function similar to the de Pillis-Radunskaya law.
+    The third term is the regulatory effect of Tregs.
+    The final two terms are increased immune evasiveness due to a driver mutation in the immune evasion pathway and the immune evasiveness of mesenchymal cells (discussed below), respectively.
+    The inflammatory state of the tumor microenvironment affects the speed at which immune cells are recruited to the tumor.
+    \newline
+    Adding to the complexity of this situation, tumor cells can express heterogeneous cell types by their location along the epithelial-mesenchymal spectrum.
+    Depending on which end of this spectrum a tumor cell is on, its proliferation and its evasion of the immune system will change.
+    Specifically, if a cell is above a certain threshold, it will have its proliferation reduced by a fixed proportion and its immune evasiveness increased by a fixed proportion.
+    These cells also demonstrate cellular plasticity by being able to undergo epithelial-to-mesenchymal transition (EMT) and move along this spectrum as a result of intrinsic and extrinsic factors.
+    Here, we choose TGF-$\beta$ as a master regulator of EMT and connect the TGF-$\beta$ level in the tumor microenvironment (TME) with the propensity of tumor cells to undergo EMT.
+    As TGF-$\beta$ is produced by Tregs as well as by tumor cells, this creates an intriguing set of interlocking interactions that we show here to be capable of shaping tumor and patient outcomes.
+    To update the EMT state of a cell, we determine how much TGF-$\beta$ is present in the TME--based on the number of malignant cells and Tregs--and apply a Hill function to determine whether a cell undergoes EMT or MET (the reverse process).
+    For each cell, this quantity takes the following form:
+    $$
+    \tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{\tau/K_3}{1+\tau/K_3} + X_i, \quad X_i \sim N(0,\sigma^2)
+    $$
+    Here, $\tau$ is the concentration of TGF-$\beta$ in the TME and $\tau_\text{max}$ represents a limit on the amount of TGF-$\beta$ that can be absorbed by all cells.
+    A Gaussian noise term is added with standard deviation $\sigma$.
+    When this quantity is large enough relative to the current EMT score of the cell, the cell will transition towards the mesenchymal phenotype.
+    Otherwise, it transitions towards the epithelial phenotype.
+    \newline
+    What we desire to know is when do the malignant tumor cells reach a particular threshold proportion of all tumor cells.
+    Once the tumor reaches this threshold, we declare that the tumor has progressed and we note how long the process took to reach this stage.
+\end{framed}
+
 \section{Introduction}
-Cancer and the immune system interact in myriad ways.
+A key question in understanding cancer is how tumors transition from a premalignant state into an aggressive, malignant state.
+This process is commonly referred to as progression \tcr{(citation needed!)} and results in a much worse prognosis for the patient.
+There are many means \tcr{(citation needed!)} by which a tumor can transition from a premalignant state into a malignant state, but one of the main ones is that of driver mutations.
+Driver mutations are those that "directly or indirectly [confer] a selective growth advantage to the cell in which it occurs." \cite{ryan2016hallmarks}.
+There are several identifiable cell processes that can be affected by such driver mutations, including proliferation, apoptosis, and immunogenicity.
+\par
+When it comes to cancer and the immune system, they interact in myriad ways.
 Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME): responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}.
@@
 % Does the following sentence still fit with our new focus on progression? I'm looking specifically at the words "pre-cancerous" and "tumorigenesis".
-In addition, the immune system plays important regulatory roles in premalignant tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), including in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
+In addition, the immune system plays important regulatory roles in premalignant tissue, where inflammation can increase the probability of progression (and/or decrease the time to cancer), including in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
 % Similar concern to above as Guo's paper dealt with initiation, not progression.
@@
 % Should we even mention initiation here? Make sure to combine below end-of-sentence refs into single \cite command.
-While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation \cite{puisieux2014oncogenic} and progression \cite{nieto2016emt,peinado2007snail}.
+While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of tumor progression \cite{nieto2016emt,peinado2007snail}.
 TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
@@
 A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 In doing so, we shift the focus of the previous model from cancer initiation to cancer progression.
 We do this to reflect the fact that cancer progression hinges on escape from the immune system and the fact that EMT has a more well-defined role during progression and metastasis.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
@@
 We then study these in more depth, focusing on the competing effects of EMT and of the immune system on progression: we find that EMT intricately regulates progression and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.
@@
 %%%
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \begin{itemize}
@@
 \end{itemize}
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \\
@@
 We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
-% might want to define "driver mutation" in intro. See PreMalignant_Thoughts.txt for definition found in a source
+% might want to define "driver mutation" in intro. See PreMalignant_Thoughts.txt for definition found in a source. Will do this.
 The agents in the model are the tissue cells that have already formed a premalignant tumor yet lack key driver mutations.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-In the process of the simulation, these cells can acquire driver mutations in any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-If the pathway is mutated, the component is colored red and is marked with `X'.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+In the process of the simulation, these cells can acquire driver mutations altering any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
+If the pathway is altered, the component is colored red and is marked with `X'.
 % I'm not a fan of "malignant", but I'll use it for now so that a quick replace all can change it later.
 We distinguish between cells that have zero driver mutations and those with at least one as premalignant and malignant, respectively.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \par
-We model immune cells as continuous variables, 
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-effectively assuming that the TME is well-mixed with regards to the immune infiltrate.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-The cytokine TGF-$\beta$ is
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-also assumed to be well-mixed in the TME so that all tissue cells have equal access to it.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+We model immune cells as continuous variables, effectively assuming that the TME is well-mixed with regards to the immune infiltrate.
+The cytokine TGF-$\beta$ is also assumed to be well-mixed in the TME so that all tissue cells have equal access to it.
 Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+% Above line replaced
 In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the shade of blue of the cell.
@@
 The tissue cells at all times have associated with them two important quantities: which combination of driver mutations they carry and an EMT score.
-The model has three pathways that have the potential to mutate over the course of a simulation.
-There is a proliferation pathway that when mutated increases the probability of the cell proliferating.
-There is an apoptosis pathway that when mutated decreases the probability of the cell undergoing apoptosis.
-There is an immune evasion pathway that when mutated decreases the probability that an immune cell can effectively clear the mutated cell.
+The model has three pathways that have the potential to become altered over the course of a simulation.
+There is a proliferation pathway that when altered increases the probability of the cell proliferating.
+There is an apoptosis pathway that when altered decreases the probability of the cell undergoing apoptosis.
+There is an immune evasion pathway that when altered decreases the probability that an immune cell can effectively clear the mutated cell.
 The EMT score will be exposited in Section \ref{EMT}.
@@
 
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \subsection{Inflammation} % How did we not have a section on this before?
@@
 When the inflammation switches to high, then a number of the parameters change values to reflect the additional immune activity.
-See Table 2 in the supplement to see which parameters are changed by the inflammation score changing. % Table 2 or S2? Change label in supplement?
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+See Table S2 in the supplement to see which parameters are changed by the inflammation score changing.
 
@@
 \par
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
 In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
+\par
+For the cells that have undergone EMT and are currently labeled in the model as mesenchymal, they experience two modifications to their governing parameters.
+The first is a reduction in proliferation, which we call mesenchymal growth arrest (MGA).
+This causes to mesenchymal cells to be less likely to undergo proliferation and more likely to rest during a cell cycle.
+The second is a decrease to their likelihood of being cleared by the immune system, which we call mesenchymal immune evasion (MIE). 
+This decreases the likelihood of mesenchymal cells being cleared by both NKs and CTLs during a cell cycle.
+Though we have not found clinical data on which to base parameter values for either of these, by their very nature they are proportions on the interval $[0,1]$.
+Thus, it is possible to sample from the entire parameter and in fact we do present results from a large subset of this parameter space and our sensitivity analysis fully explores this space.
+
 \subsection{Model Simulation}
@@
 During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
@@
 \subsubsection{Tissue cell fate}
-During each cell cycle, every cell randomly is assigned a cell fate from the following options:
-\begin{itemize}
-\item proliferation
-\item apoptosis
-\item immune clearance (by NKs or CTLs)
-\item rest in $G_0$
-\end{itemize}
-
-For each cell, a weight is chosen for each option and these are normalized to probabilities which then are used to randomly determine what each cell does during the cell cycle.
-
-\paragraph{Proliferation}
-There are four factors that contribute to the weight of a cell to proliferate.
-The first is a base proliferation rate that all cells have, $p$.
-Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
-Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
-% Add: 
-This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.
-Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
-In total, the weight for proliferation is given by
-
-\begin{equation}\tag{2.1}
-\rho_P = p(1+\delta_{P}\Delta_P)(1-\zeta \Delta_{\text{MGA}})\frac{K_0}{K_0+N_C}
-\end{equation}
-
-\paragraph{Apoptosis}
-There are two factors that contribute to a cell's weight for undergoing apoptosis.
-There is a basal apoptosis rate that all cells experience, $d_C$ for death.
-Second, if the cell has a mutation in the apoptosis pathway ($\delta_A=1$), then the weight for undergoing apoptosis is proportionally decreased by $\Delta_A$.
-In total, the weight for apoptosis is given by
-
-\begin{equation}\tag{2.2}
-\rho_A = d_C(1-\delta_{A}\Delta_A)
-\end{equation}
-
-\paragraph{Immune Clearance}
-For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
-First of all, the cell needs to be malignant ($\delta_{\text{MUT}}=1$).
-Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
-Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
-Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
-Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
-In total, the weight of NK clearance is given by
-
-\begin{equation}\tag{2.3}
-\rho_{\text{NK}} =\delta_{\text{MUT}} \frac{N_{\text{NK}}}{N_C/K_{1}+N_{\text{NK}}}  \frac{E_{\text{NK}}}{1+N_{\text{Treg}}/K_2} (1-\delta_{\text{IE}}\Delta_{\text{IE}})(1-\zeta \Delta_{\text{MIE}})
-\end{equation}
-
-A similar formula holds for CTLs with only the number of CTLs and their efficacy being different from the above equation.
-
-\paragraph{Rest in $G_0$} 
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-The weight associated with rest is taken as 1 except in the case of mesenchymal cells.
-Recall that mesenchymal cells had their proliferation rate decreased by $1-\zeta\Delta_{\text{MGA}}$ (see Eq. 2.1).
-The biological assumption here is that mesenchymal cells instead of proliferating will instead rest, so this lost proliferation weight is added to the resting weight.
-Hence, the weight of rest is given by
-% Replace quiescence with rest and explain more above
-
-\begin{equation}\tag{2.4}
-\rho_R = 1 + \zeta p(1+\delta_P\Delta_P)\Delta_{\text{MGA}}\frac{K_0}{K_0+N_C}
-\end{equation}
-
-Again, the reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.
+During each cell cycle, every cell is randomly assigned one of four fates: proliferation, apoptosis, immune clearance, and rest in $G_0$.
+Depending on both cell intrinsic, e.g.
+EMT and mutational state, and cell extrinsic, e.g.
+immune infiltrate and total number of tissue cells, each of the four options is given a probability of occurring for each cell.
+For proliferation, we assume the total number of tissue cells exerts a negative feedback on the proliferation probability for each cell.
+Also, cells that have a driver mutation in the proliferation pathway have an increased probability of proliferating.
+Meanwhile, mesenchymal cells have a decreased probability of proliferating and an increased probability of resting.
+For apoptosis, the probability is constant except for those cells with a driver mutation in the apoptosis pathway which have a decreased probability of apoptosis.
+For immune clearance, we assume that only cells with newly acquired driver mutations are targeted by the immune system, and their probability of being cleared depends on the immune infiltrate.
+This probability can be decreased by a cell having a driver mutation in the immune evasion pathway or by a cell being mesenchymal.
+Further details and equations are given in the supplement.
 
 \subsubsection{Completing the Cell Cycle}
-After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
-First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
-This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
-Second, all proliferating cells have a cell-specific probability of undergoing a pathway mutation.
-If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
-If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases.
+Once all the cell fates are carried out, the rest of the components of the model update to account for these changes.
+First, the immune populations are updated in two steps.
+In step one, Immune cells are exhausted based on the number of malignant cells that were cleared by their type.
+So, each clearance by an NK cell results in the NK population decreasing by one, and similarly for CTLs.
+In step two, all three immune populations are updated deterministically based on a simple ODE model of their dynamics.
+NKs have a constant source and death rate while CTLs and Tregs have a source rate that is dependent on the number of malignant cells cleared that cycle and a constant death rate.
+In addition, the concentration of TGF-$\beta$ mediates recruitment of Tregs with higher concentrations of TGF-$\beta$ resulting in a greater influx of Tregs.
 \par
-Finally, the EMT values for each cell is updated.
-This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
-% The following sentence could be altered to reflect that the threshold value for $\tau$ is 0, so not only are negative values possible, but very common.
-So, this is not really so much a TGFB received as it is a change in TGFB signaling.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
-The saturation effect is to limit the amount of TGF-$\beta$ a cell can absorb in a given time interval.
-This quantity is then divided up randomly among the $N_C$ living cells via a normally distributed noise term to determine how much exogenous TGF-$\beta$ each cell receives during this cell cycle.
-Should this value, $\tau_i$ in Eq. 2.5, be negative, we interpret this as the cell losing TGF-$\beta$ to the TME and thus being more likely to undergo MET.
-
-\begin{equation}\tag{2.5}
-\tau_i = \frac{\tau_{\text{max}}}{N_C}\frac{\tau/K_3}{1+\tau/K_3} + X_i
-, \quad X_i \sim N(0,\sigma^2)
-\end{equation}
-
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-We then combine $\tau_i$ with the current EMT score of the cell, as a proxy for the endogenous TGF-$\beta$.
-Finally, if this quantity is large enough, the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% Need to add better explanation for Hill fn
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
-Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: concentration of TGF-$\beta$ in the system and the current EMT score of the cell.
-
-Next, the amount of TGF-$\beta$ for the next cell cycle is determined by the number of mutated cells, $N_{\text{MUT}}$, and the number of Treg cells, $N_{\text{Treg}}$, each one producing a fixed amount of TGF-$\beta$.
-It is given by
-
-\begin{equation}\tag{2.6}
-\tau = \tau_{\text{MUT}}N_{\text{MUT}} + \tau_{\text{Treg}}N_{\text{Treg}}
-\end{equation}
-
-Finally, the immune populations are updated.
-For the NKs, they obey the following differential equation:
-
-\begin{equation}\tag{2.7}
-N_{\text{NK}}' = \sigma_{\text{NK}} - d_{\text{NK}}N_{\text{NK}}
-\end{equation}
-
-which is discretized to
-
- \begin{equation}\tag{2.8}
-N_{\text{NK}}(k+1) = \left (N_{\text{NK}}(k)-\frac{\sigma_{\text{NK}}}{d_{\text{NK}}} \right )\exp(-d_{\text{NK}}\Delta t)+\frac{\sigma_{\text{NK}}}{d_{\text{NK}}}
-\end{equation}
-
-For CTLs and Tregs, they rely on malignant cells being cleared before they can be activated.
-Let $N_{\text{MUT}}^*(k)$ represent the number of malignant cells cleared by the immune system during cell cycle $k$.
-In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
-We choose the following differential equations to govern the CTL and Treg populations:
-
-\begin{equation}\tag{2.9}
-\begin{split}
-N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}N_\text{CTL} \\
-N_\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
-\end{split}
-\end{equation}
-
-Discretized, these are:
-
-\begin{equation}\tag{2.10}
-\begin{split}
-N_\text{CTL}(k+1) & =  \left (N_\text{CTL}(k)-\sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\\
-N_\text{Treg}(k+1) & =  \left (N_\text{Treg}(k)-\frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}\right )\exp(-d_{\text{Treg}}\Delta t)\\
-&+ \frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}
-\end{split}
-\end{equation}
+Also at this time, new driver mutations have a chance to occur.
+The cells that proliferated are the only ones that have a probability of acquiring a new driver mutation.
+This probability is cell-specific and is increased if no mutation occurs and reset to 0 if a mutation does occur.
+\par
+Finally, the concentration of TGF-$\beta$ is updated as well as the EMT score for each cell is updated.
+Malignant cells and Tregs are the two sources of TGF-$\beta$ each producing a constant amount.
+Tregs are assumed to produce much more TGF-$\beta$ than malignant cells.
+The resulting quantity of TGF-$\beta$ is divided up randomly among the tissue cells.
+This quantity is combined with the EMT score of each cell and the resulting quantity determines which cells undergo EMT and which undergo MET.
+The equations governing these processes can be found in the supplement.
 
@@
  \hline
-  $\Delta_A$ & mutant cells decreased apoptosis  \\ % should this and the others like it be more clear in stating this is not for all malignant cells, but only those with apoptosis pathway mutation?
+  $\Delta_A$ & decrease in apoptosis rate in cells with driver mutation in apoptosis pathway \\ % should this and the others like it be more clear in stating this is not for all malignant cells, but only those with apoptosis pathway mutation?
+Better to be longer and unambiguous "decrease in apoptosis rate in cells with driver mutation in apoptosis pathway"
   \hline
-  $\Delta_\text{IE}$ & mutant cells increased immune evasion  \\
+  $\Delta_\text{IE}$ & increase in immune evasion in cells with driver mutation in immune evasion pathway \\
   \hline
-  $\Delta_P$ & mutant cells increased proliferation  \\
+  $\Delta_P$ & increase in proliferation in cells with driver mutation in proliferation pathway\\
   \hline
@@
 \end{tabular}
-  \caption{The model parameter names and descriptions. Note that many of these values are affected by the inflammation state of the system.}
+  \caption{The model parameter names and descriptions. Note that some of these values are affected by the inflammation state of the system.}
 \end{center}
@@
 
-\subsubsection{Mutation and tumorigenesis}
+\subsubsection{Mutation and progression}
 % 3rd Referee did not like this terminology for a tumorigenic event. Should we change it or better explain it?
-At the end of each cell cycle, the proportion of tissue cells that are malignant is calculated, and if it is above a certain threshold, 
-the simulation ends and we declare the tumor to have progressed.
+At the end of each cell cycle, the proportion of tissue cells that are malignant is calculated, and if it is above a certain threshold, the simulation ends and we declare the tumor to have progressed.
 The Time to Cancer is given as the time from the start of the simulation, not including warmup, until this timepoint.
@@
 
-\subsection{A multiscale agent-based model of EMT-immune-tissue cell interactions to study tumorigenesis}\label{ExplModel}
+\subsection{A multiscale agent-based model of EMT-immune-tissue cell interactions to study tumor progression}\label{ExplModel}
 
@@
 {\includegraphics[width=0.9\textwidth]{Figure1/Figure1.pdf}}
-\caption{A. Schematic depiction of agent-based model components;
-each of the 10 columns represents a single tissue cell divided into three compartments representing the state (mutated or not) of the three pathways with mutagenic potential;
-red/blue denotes mutated/unmutated pathways. Black arrows depict regulation of the cell fate in each cell cycle. Inset depicts major interactions between the immune system and tissue cells.
+\caption{A.
+Schematic depiction of agent-based model components; each of the 10 columns represents a single tissue cell divided into three compartments representing the state (altered or not) of the three pathways with mutagenic potential; red/blue denotes altered/unaltered pathways.
+Black arrows depict regulation of the cell fate in each cell cycle.
+Inset depicts major interactions between the immune system and tissue cells.
  %The parameters used in this model can be found in the Supplementary material.
@@
 Once I have the table, I will use the name of that table.}
-B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
+B. A representative simulation of one patient. The model parameter values used can be found in Table S2. 
 The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: malignant cells; Mes: mesenchymal cells. 
-C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
+C. Survival curve for one cohort of patients with the parameter values given in Table S2.}
 \label{fig:ModelIntro}
@@
 \ref{fig:ModelIntro}A).
-For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
+For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a driver mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
 The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset).
@@
 For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 After about 700 cell cycles, the malignant cell population begins to grow steadily.
@@
 A peak in number of mesenchymal cells is observed shortly after this threshold is reached, and following this most cells transition back to epithelial.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \par 
@@
 The ILD and IHD parameters determine the duration of low and high inflammation, respectively.
-A cell with mutated pathways will be affected by the $\Delta_A$, $\Delta_P$, and $\Delta_{\text{IE}}$ parameters, depending on whether the apoptosis, proliferation, or immune evasion pathways are mutated.
+A cell with altered pathways will be affected by the $\Delta_A$, $\Delta_P$, and $\Delta_{\text{IE}}$ parameters, depending on whether the apoptosis, proliferation, or immune evasion pathways are altered.
 The mesenchymal cells are influenced by the MGA and MIE parameters.
@@
 These striking differences in the mean Time to Cancer -- extended by up to one {\it in silico} year by fine-grained optimization of mesenchymal growth arrest -- have clear therapeutic implications. 
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-Our model predicts that reducing the inflammatory state of the TME in three weeks out of every nine would be provide means to prolong cancer-free survival in combination with mesenchymal cell therapies to make the control of mesenchymal proliferation an effective anti-cancer strategy.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 Our model predicts that a patient suffering intermittent inflammatory attacks at the tumor site will benefit from controlling the proliferation of the mesenchymal cells.
-% commented to keep same line numbers as Paper.txt (as of 9/16/19)
 \par
@@
 
-To further assess the effects on tumorigenesis of the mesenchymal properties of immune evasion and growth arrest, we study the joint density plot for each of these parameters against the Time to Cancer (Fig.
+To further assess the effects on progression of the mesenchymal properties of immune evasion and growth arrest, we study the joint density plot for each of these parameters against the Time to Cancer (Fig.
 \ref{fig:MIEvsMGA}).
@@
 \par
-For each cohort of cancer patients, we cluster the patients via k-means (n = 2) either on gene ontologies relating only to EMT (linked figures) or on gene ontologies relating to EMT and inflammation.
+For each cohort of cancer patients, we cluster the patients via k-means (n = 2) on gene ontologies relating to only EMT (Figures \ref{fig:OV}C, \ref{fig:PAAD}C), only inflammation (Figures \ref{fig:OV}E, \ref{fig:PAAD}E), and both EMT and inflammation (Figures \ref{fig:OV}A, \ref{fig:PAAD}A).
 We also plot the corresponding survival curves for each of the clusters obtained (utilizing the originally reported OS endpoint).
@@
 This suggests that when studying interactions between cancer and the immune system, is it important to consider EMT; the effects of which may impact the dynamics and should not be overlooked.
-
-\par
 
@@
 We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$.
-We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).
+We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to driver mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).
 \par
@@
 We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model.
-There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in premalignant tissues is a pressing need to shed insight into cellular biomarkers of cancer.
+There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in pre-cancerous premalignant tissues is a pressing need to shed insight into cellular biomarkers of cancer.
 \par

### 2019-10-06 Paper/Supplement/CleanSupplement.tex
cleaned up main text and supplement

--- 2019-10-06 Paper/Supplement/Supplement_rewrite.tex
@@
 \renewcommand\Authands{ and }
-
-%\title{
-%The competing effects of the immune system and EMT on epithelial cancers
-%}
 
@@
 
-%\begin{table}[H]
-%\begin{center}
-% \begin{tabular}{|| c | p{5cm} | p{3cm} | p{3cm} ||} 
-% \hline
-% Name & Description & INFL Low Value & INFL High Value  \\ [0.5ex] 
-% \hline\hline
-%  $p$ & weight of proliferation for tissue cells & 0.28 & \\ 
-% \hline
-% $d_C$  & weight of apoptosis for tissue cells & 0.14 & \\
-% \hline
-%$\Delta_\text{MIE}$ & MIE & 0.6 & \\
-% \hline
-% $\Delta_\text{MGA}$ & MGA  & 0.2 &  \\
-% \hline
-%  $\Delta_A$ & proportional decrease to weight of apoptosis for cells with mutated apoptosis pathway & 0.3 &  \\
-%  \hline
-%  $\Delta_\text{IE}$ & proportional increase to weight of immune evasion for cells with mutated immune evasion pathway & 0.48 &  \\
-%  \hline
-%  $\Delta_P$ & proportional increase to weight of proliferation for cells with mutated proliferation pathway & 0.36 &  \\
-%  \hline
-% $K_0$ & EC50 term for negative feedback of tissue cells on own proliferation & 80 cells& \\
-% \hline
-% $K_1$ & EC50 term for probability of NK cell finding mutant cell & 8 cells& \\
-% \hline
-%  $K_2$ & EC50 term for Treg inhibition of cytotoxic functions & 5 cells / volume & 0.025 cells / volume \\
-%  \hline
-%  $K_3$ & EC50 term for cumulative absorption of TGF-$\beta$ & 200 amount / volume & \\
-%  \hline
-%  $K_4$ & EC50 term for TGF-$\beta$ activation of Tregs & 50 amount / volume & \\
-%  \hline
-% $E_\text{NK}$ & weight of NKs clearing mutants & 10 & 30  \\
-%  \hline
-%  $E_\text{CTL}$ & weight of CTLs clearing mutants & 200 & 600 \\
-%  \hline
-%  $\sigma_\text{NK}$ & NK source rate & 1.3 cells / cycle &  \\ 
-%  \hline
-%  $\sigma_\text{CTL}$ & CTL source rate per cleared mutant cell & 100 cells / (cleared mutants $\times$ cycles)& \\ 
-%  \hline
-%  $\sigma_\text{Treg}$ & Treg source rate per cleared mutant cell & 200 cells / (cleared mutants $\times$ concentration of TGF-$\beta$ $\times$ cycles)& \\ 
-%  \hline
-%  $d_\text{NK}$ & NK death rate & 0.13 / cycle &  \\ 
-%  \hline
-%  $d_\text{CTL}$ & CTL death rate & 0.0260 / cycle& \\ 
-%  \hline
-%  $d_\text{Treg}$ & Treg death rate & 0.0260 / cycle& \\ 
-%  \hline
-%  $k_\text{EMT}$ & EMT/MET rate & 0.01 / concentration of TGF-$\beta$ & \\
-%  \hline
-%  $\sigma$ & standard deviation of noise in TGF-$\beta$ each cell receives & 6 concentration of TGF-$\beta$ & \\
-%  \hline
-% $\tau_\text{max}$ & max amount of TGF-$\beta$ any cell can receive & 500 concentration of TGF-$\beta$ &\\
-%  \hline 
-% $\tau_\text{MUT}$ & rate of TGF-$\beta$ production by mutant cells & 0.05 concentration of TGF-$\beta$ / cell / cycle& \\
-%  \hline
-% $\tau_\text{Treg}$ & rate of TGF-$\beta$ production by Treg & 0.5 concentration of TGF-$\beta$ / cell / cycle& \\
-%  \hline
-%  & RP Cancer Line  & 0.5 &  \\ 
-% \hline
-%&  INFL High Duration & 30 cycles&   \\
-% \hline
-% & INFL Low Duration &30 cycles&   \\
-% \hline
-%& Mes Threshold & 0.7 &    \\
-% \hline
-%  & maximum initial mutation damage after warmup  & 0.01 &\\
-%  \hline
-%  & increase in probability to mutate for non-mutating proliferating cells & 0.0001 & \\
-%  \hline
-%\end{tabular}
-%  \caption{The model parameter names, descriptions, and values during both low and high inflammation. Parameters with only one value do not change with the inflammatory state.}
-%\end{center}
-%\end{table}
-
-\newpage
-
-\begin{table}[H]
-\begin{center}
- \begin{tabular}{|| c | p{7cm} | c | c ||} 
- \hline
- Name & Description & INFL Low Value & INFL High Value  \\ [0.5ex] 
- \hline\hline
-  $p$ & proliferation rate of tissue cells & 0.28 & \\ 
- \hline
- $d_C$  & death rate of tissue cells & 0.14 & \\
- \hline
-$\Delta_\text{MIE}$ &  MIE & 0.6 & \\
- \hline
- $\Delta_\text{MGA}$ & MGA  & 0.2 &  \\
- \hline
-  $\Delta_A$ & mutant cells decreased apoptosis & 0.3 &  \\
-  \hline
-  $\Delta_\text{IE}$ & mutant cells increased immune evasion & 0.48 &  \\
-  \hline
-  $\Delta_P$ & mutant cells increased proliferation & 0.36 &  \\
-  \hline
- $K_0$ & EC50 term for negative feedback of tissue cells on own proliferation & 80 & \\
- \hline
- $K_1$ & EC50 term for probability of NK cell finding mutant cell & 8 & \\
- \hline
-  $K_2$ & EC50 term for Treg inhibition of cytotoxic functions & 5 & 0.025  \\
-  \hline
-  $K_3$ & EC50 term for how much TGF-$\beta$ each cell has & 200 & \\
-  \hline
-  $K_4$ & EC50 term for TGF-$\beta$ activation of Tregs & 50 & \\
-  \hline
- $E_\text{NK}$ & rate of NKs clearing mutants & 10 & 30  \\
-  \hline
-  $E_\text{CTL}$ & rate of CTLs clearing mutants & 200 & 600 \\
-  \hline
-  $\sigma_\text{NK}$ & NK source rate & 1.3 &  \\ 
-  \hline
-  $\sigma_\text{CTL}$ & CTL source rate per cleared mutant cell & 100 & \\ 
-  \hline
-  $\sigma_\text{Treg}$ & Treg source rate per cleared mutant cell & 200 & \\ 
-  \hline
-  $d_\text{NK}$ & NK death rate & 0.13 &  \\ 
-  \hline
-  $d_\text{CTL}$ & CTL death rate & 0.0260 & \\ 
-  \hline
-  $d_\text{Treg}$ & Treg death rate & 0.0260 & \\ 
-  \hline
-  $k_\text{EMT}$ & EMT/MET rate & 0.01 & \\
-  \hline
-  $\sigma$ & standard deviation of noise in TGF-$\beta$ each cell receives & 6 & \\
-  \hline
- $\tau_\text{max}$ & max amount of TGF-$\beta$ any cell can receive & 500 &\\
-  \hline 
- $\tau_\text{MUT}$ & rate of TGF-$\beta$ production by mutant cells & 0.05 & \\
-  \hline
- $\tau_\text{Treg}$ & rate of TGF-$\beta$ production by Treg & 0.5 & \\
-  \hline
-  & RP Cancer Line  & 0.5 &  \\ 
- \hline
-&  INFL High Duration & 30&   \\
- \hline
- & INFL Low Duration &30 &   \\
- \hline
-& Mes Threshold & 0.7 &    \\
- \hline
-  & maximum initial mutation damage after warmup  & 0.01 &\\
-  \hline
-  & increase in probability to mutate for non-mutating proliferating cells & 0.0001 & \\
-  \hline
-\end{tabular}
-  \caption{The model parameter names, descriptions, and values during both low and high inflammation. Parameters with only one value do not change with the inflammatory state.}
-\end{center}
-\end{table}
-
 \begin{figure}
@@
 
-%\newpage
-%
-%
-% \begin{longtable}{||c | c | c | c | c | c | c | c | c | c ||} 
-% \hline
-% Name & Base Value & Single Patient & Survival Curve & Vary MIE & Vary MGA & Vary Treg TGFB & Vary MIE and INFL & Vary MGA and INFL & Heat Map\\ [0.5ex] 
-% \hline\hline
-% RP Cancer Line &  0.5 &  \\ 
-% \hline
-% MIE  & 0.6 & 0.9 & 0.4-0.9 & 0.7 & 0.5 & 0.9 \\
-% \hline
-% MGA & 0.2 & & 0.1-0.4 & 0 & 0.2 & 0 \\
-% \hline
-% INFL High Duration  & 30 & & & & 60 & & & & 60 \\
-% \hline
-%INFL Low Duration  & 30  \\
-% \hline
-%Mes Threshold & 0.7 \\
-% \hline
-% p  & 0.28 \\ 
-% \hline
-% d  & 0.14 \\
-% \hline
-% $N_{00}$  & 8\\
-% \hline
-% $N_0$  & 80 \\
-% \hline
-%  NK Efficacy Low  & 0.2 & 10 & & 10 & 10 & 10 & 10 & 10 & 10 \\
-%  \hline
-%  Treg EC50 Low & 5 & & & & & 2.5 \\
-%  \hline
-%  CTL Efficacy Low  & 4 & 200 & & 200 & 200 & 200 & 200 & 200 & 200\\
-%  \hline
-%  NK Efficacy Up & 2 & & & & 1.2 & & & &  1.2 \\
-%  \hline
-%  CTL Efficacy Up & 3 & & & & 3 & & & &  3\\
-%  \hline
-%  Treg Efficacy Up  & 200 & & & & 10& & & &  10\\
-%  \hline
-%  NK Source Low  & 1.3 \\ 
-%  \hline
-%  CTL Recruitment Low  & 100 \\ 
-%  \hline
-%  Treg Recruitment Low  & 200 \\
-%  \hline
-%  NK Recruitment Up & 1 & & & & & 3\\
-%  \hline
-%  CTL Recruitment Up & 1 & & & & & 3\\
-%  \hline
-%   Treg Recruitment Up & 1 & & & & & 5\\
-%  \hline
-%  $p_{mutation}$ Start & 1e-2 \\
-%  \hline
-%  $p_{mutation}$ & 1e-4 \\
-%  \hline
-%  Apoptosis Down & 0.3 \\
-%  \hline
-%  Immune Evasion & 0.48 \\
-%  \hline
-%  Proliferation Up & 0.36  \\
-%  \hline
-%  $k_{EMT}$  & .01 \\
-%  \hline
-%  c & 6 \\
-%  \hline
-%  TGFB Received EC50 & 200 \\
-%  \hline
-%  TGFB Max & 500 & 700 & & & & & &  & 700\\
-%  \hline 
-%  Mutant TGFB & 5e-2 & & & & & 0.5\\
-%  \hline
-%  Treg TGFB & 5e-1 \\
-%  \hline
-%\end{longtable}
-%
-%
-
 \end{document}

### 2019-10-07 Paper/CleanPaper.tex
addressed comments in paper and clean paper

@@
 The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
-% Does the following sentence still fit with our new focus on progression? I'm looking specifically at the words "pre-cancerous" and "tumorigenesis".
 In addition, the immune system plays important regulatory roles in premalignant tissue, where inflammation can increase the probability of progression (and/or decrease the time to cancer), including in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
-% Similar concern to above as Guo's paper dealt with initiation, not progression.
 Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.
@@
 \par 
-% Should we even mention initiation here? Make sure to combine below end-of-sentence refs into single \cite command.
 While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of tumor progression \cite{nieto2016emt,peinado2007snail}.
@@
 To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.
-% Consider rewriting above
 
@@
 We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
-% might want to define "driver mutation" in intro. See PreMalignant_Thoughts.txt for definition found in a source. Will do this.
 The agents in the model are the tissue cells that have already formed a premalignant tumor yet lack key driver mutations.
@@
 If the pathway is altered, the component is colored red and is marked with `X'.
-% I'm not a fan of "malignant", but I'll use it for now so that a quick replace all can change it later.
 We distinguish between cells that have zero driver mutations and those with at least one as premalignant and malignant, respectively.
@@
 While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
-% Above line replaced
 In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the shade of blue of the cell.
@@
 The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
-% We could justify the immune system not clearing "base" mutated cells by saying the immune surveillance provides a baseline control over these cells accounted for in their apoptosis.
-Or suggesting that we are interested in immune activity above and beyond the baseline requirement to keep this tumor in check.
 The NKs and CTLs act on the system by recognizing malignant cells and clearing them.
@@
 
-\subsection{Inflammation} % How did we not have a section on this before?
+\subsection{Inflammation} 
 We also include an inflammation score that follows a fixed cycling scheme between low and high inflammatory states.
@@
  \hline
-  $\Delta_A$ & decrease in apoptosis rate in cells with driver mutation in apoptosis pathway \\ % should this and the others like it be more clear in stating this is not for all malignant cells, but only those with apoptosis pathway mutation?
-Better to be longer and unambiguous "decrease in apoptosis rate in cells with driver mutation in apoptosis pathway"
+  $\Delta_A$ & decrease in apoptosis rate in cells with driver mutation in apoptosis pathway \\ 
     \hline
@@
 \subsubsection{Mutation and progression}
-% 3rd Referee did not like this terminology for a tumorigenic event. Should we change it or better explain it?
 At the end of each cell cycle, the proportion of tissue cells that are malignant is calculated, and if it is above a certain threshold, the simulation ends and we declare the tumor to have progressed.
@@
 \par
-% Referee 2 either missed the following line or doesn't understand what average of absolute change means.
 The algorithm computes the average of the absolute change of the output, which in our model is the area under the survival curve.
@@
 Inset depicts major interactions between the immune system and tissue cells.
- %The parameters used in this model can be found in the Supplementary material.
-Unless otherwise stated, parameter values are assumed to be those found in the Supplement.
-\tcr{Thoughts on the wording here?
-Once I have the table, I will use the name of that table.}
 B. A representative simulation of one patient. The model parameter values used can be found in Table S2. 

### 2019-10-07 Paper/Paper.tex
addressed comments in paper and clean paper

@@
 The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
-% Does the following sentence still fit with our new focus on progression? I'm looking specifically at the words "pre-cancerous" and "tumorigenesis".
 In addition, the immune system plays important regulatory roles in \sout{pre-cancerous} \tcb{premalignant} tissue, where inflammation can increase the probability of \sout{tumorigenesis} progression (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
-% Similar concern to above as Guo's paper dealt with initiation, not progression.
 \tcr{Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.}
@@
 \par 
-% Should we even mention initiation here? Make sure to combine below end-of-sentence refs into single \cite command.
 While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of tumor \sout{initiation} \tcb{\cite{puisieux2014oncogenic}} progression \cite{nieto2016emt}\tcb{\cite{peinado2007snail}}.
@@
 \tcb{To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.}
-% Consider rewriting above
 
@@
 We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
-% might want to define "driver mutation" in intro. See PreMalignant_Thoughts.txt for definition found in a source. Will do this.
 The agents in the model are the tissue cells \sout{with tumorigenic potential} \tcb{that have already formed a premalignant tumor yet lack key driver mutations}.
@@
 \tcb{
-% I'm not a fan of "malignant", but I'll use it for now so that a quick replace all can change it later.
 We distinguish between cells that have zero driver mutations and those with at least one as premalignant and malignant, respectively.
@@
 \tcb{While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
-% Above line replaced
 In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the \sout{color} \tcb{shade of blue} of the cell.
@@
 The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
-% We could justify the immune system not clearing "base" mutated cells by saying the immune surveillance provides a baseline control over these cells accounted for in their apoptosis.
-Or suggesting that we are interested in immune activity above and beyond the baseline requirement to keep this tumor in check.
 The NKs and CTLs act on the system by recognizing \sout{mutated} \tcb{malignant} cells and clearing them.
@@
 \tcb{
-\subsection{Inflammation} % How did we not have a section on this before?
+\subsection{Inflammation} 
 We also include an inflammation score that follows a fixed cycling scheme between low and high inflammatory states.
@@
 Further details and equations are given in the supplement.
+
 %%%%%% MOVED TO SUPPLEMENT %%%%%%%%%%%%%
+
 %During each cell cycle, every cell randomly is assigned a cell fate from the following options:
@@
 The equations governing these processes can be found in the supplement.
+
+%%%%%% MOVED TO SUPPLEMENT %%%%%%%%%%%%%
+
 %After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
@@
 
+%%%%%% END MOVED TO SUPPLEMENT %%%%%%%%%%%%%
+
 \begin{table}[H]
@@
  \hline
-  $\Delta_A$ & \tcb{decrease in apoptosis rate in cells with driver mutation in apoptosis pathway}  \\ % should this and the others like it be more clear in stating this is not for all malignant cells, but only those with apoptosis pathway mutation?
-Better to be longer and unambiguous "decrease in apoptosis rate in cells with driver mutation in apoptosis pathway"
+  $\Delta_A$ & \tcb{decrease in apoptosis rate in cells with driver mutation in apoptosis pathway}  \\ 
     \hline
@@
 \subsubsection{Mutation and \sout{tumorigenesis} \tcb{progression}}
-% 3rd Referee did not like this terminology for a tumorigenic event. Should we change it or better explain it?
 At the end of each cell cycle, the proportion of tissue cells that are \sout{mutated} \tcb{malignant} is calculated, and if it is above a certain threshold, \sout{a tumorigenic event is recorded and the patient is determined to have cancer.}
@@
 \par
-% Referee 2 either missed the following line or doesn't understand what average of absolute change means.
 \tcr{The algorithm computes the average of the absolute change of the output, which in our model is the area under the survival curve.}
@@
 Inset depicts major interactions between the immune system and tissue cells.
- %The parameters used in this model can be found in the Supplementary material.
-Unless otherwise stated, parameter values are assumed to be those found in the Supplement.
-\tcr{Thoughts on the wording here?
-Once I have the table, I will use the name of that table.}
 B. A representative simulation of one patient. The model parameter values used can be found in \sout{Supplementary} Table \tcb{S}2. 

### 2019-10-11 Paper/CleanPaper.tex
edits to CleanPaper

@@
 \section*{Abstract}
-During cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor prognosis.
-Inflammation also has significant effects on gastrointestinal and pancreatic tumors.
-Moreover, epithelial cell plasticity can significantly alter disease trajectories (for better or worse) via epithelial-to-mesenchymal transition (EMT).
+During cancer progression, the immune system is engaged in complex sets of interactions with various tumor cells that affect prognosis.
+Inflammation has particularly significant effects on gastrointestinal and pancreatic tumors.
+Moreover, epithelial cell plasticity can significantly alter disease trajectories via epithelial-to-mesenchymal transition (EMT).
 Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of these regulatory processes.
@@
 We performed analysis of pancreatic cancer data from The Cancer Genome Atlas to compare clinical measurements with model predictions.
-We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer.
+We found that association with EMT worsened the survival probabilities of inflammation-associated cancer.
 These results offer novel means to delay disease progression by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.
@@
 
+\section{Introduction}
+A key question in understanding cancer is how tumors transition from a premalignant state into an aggressive, malignant state.
+This process is commonly referred to as progression \tcr{(citation)} and results in a much worse prognosis for the patient.
+There are many means by which a tumor can transition from a premalignant state into a malignant state \tcr{(citation)}, but one of the main ones is that of driver mutations.
+Driver mutations are those that "directly or indirectly [confer] a selective growth advantage to the cell in which it occurs." \cite{ryan2016hallmarks}.
+There are several identifiable cell processes that can be affected by such driver mutations, including proliferation, apoptosis, and immunogenicity.
+\par
+When it comes to cancer and the immune system, they interact in myriad ways.
+Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME): responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}.
+Recent breakthroughs in the field of immunotherapy are beginning to produce therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}.
+\par 
+Tumor cells interact with the immune system by initiating a sequence of events that results in the activation of the adaptive immune response.
+This is accomplished by the presentation of antigens recognized by innate immune cells that are transported to lymph nodes where T cells (and other components) can be activated \cite{schreiber11_cancer}.
+The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via activation of T regulatory cells (Tregs) \cite{schreiber11_cancer}.
+\par
+The effects of the immune system on a tumor can be broadly summarized into two modules.
+The cytotoxic branch of the immune system, such as natural killer cells (NKs) and cytotoxic T cells (CTLs), seek out and lyse tumor cells.
+Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely \cite{finn12_immunooncology-1}.
+The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
+In addition, the immune system plays important regulatory roles in premalignant tissue, where inflammation can increase the probability of progression (and/or decrease the time to cancer), including in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
+Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.
+\par
+Epithelial-to-mesenchymal transition (EMT) describes a reversible process by which cells with an epithelial phenotype transition into cells with a mesenchymal phenotype.
+While epithelial cells are in part defined by their adhesion to neighbors, mesenchymal cells show much greater ranges of motility, and may display stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.
+Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}.
+Ongoing investigations into the plasticity and stability of EMT overlap with discussions elsewhere, e.g. of discrete vs. continuous processes during cell differentiation \cite{moris16_transition}.
+Intermediate states have emerged as a central mechanism by which cell fates (and the noise inherent within them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}.
+\par 
+While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of tumor progression \cite{nieto2016emt,peinado2007snail}.
+TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
+There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.
+Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
+It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.
+\par
+Two features of the mesenchymal phenotype are of particular relevance in the context of cancer and the immune system: i) mesenchymal cells are less susceptible to immune clearance \cite{terry2017new}; and ii) mesenchymal cells proliferate on average at a slower rate than epithelial cells.
+The immune evasiveness of mesenchymal cells is well-established: as a cell is targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be established before the lysing event.
+This immunological synapse is dependent upon the surface proteins, such as the T-cell receptors of the major histocompatibility complex, of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form \cite{terry2017new}.
+Below we refer to this phenotype as mesenchymal immune evasion (MIE).
+\par
+The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculate to exhibit a more ``stemness'' phenotype.
+While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects}.
+Cancer cells are of course characterized by some degree of unregulated proliferation, thus it is appealing to explore the effects of differential proliferation rates within the cancer cell population.
+In our model, we refer to the reduced proliferative capacity associated with a mesenchymal cell phenotype as mesenchymal growth arrest (MGA).
+
+Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including classic \cite{anderson98_continuous, sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{kim18_cell, gallaher14_bridging, gallaher18_spatial, an15_agentbased, serre16_mathematical, louzoun14_mathematical, briones-orta13_arkadia, lavi13_role, greene15_modeling, greene16_mathematical, cho17_modeling-1,  benzekry17_mathematical, owen11_mathematical, west18_multidrug}.
+These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximize the efficacy of cytotoxicity while minimizing risks to the patient.
+However, to our knowledge no models have addressed how the effects of EMT alter interactions between the immune system and cancer, and the subsequent implications for treatment.
+\par
+Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
+We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that premalignant tissue will progress and, if so, when.
+A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
+In doing so, we shift the focus of the previous model from cancer initiation to cancer progression.
+We do this to reflect the fact that cancer progression hinges on escape from the immune system and the fact that EMT has a more well-defined role during progression and metastasis.
+We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
+\par 
+In the next section we develop the model and give explanation of the intuition behind each component.
+We then go on to analyze its behavior.
+Global sensitivity analysis identifies parameters that are crucial for progression.
+We then study these in more depth, focusing on the competing effects of EMT and of the immune system on progression: we find that EMT intricately regulates progression and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
+To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.
+
 \begin{framed}
-    Quick Guide to Equations and Assumptions
+    {\bf Quick Guide to Equations and Assumptions}
 \newline
@@
   \end{framed}
-
-\section{Introduction}
-A key question in understanding cancer is how tumors transition from a premalignant state into an aggressive, malignant state.
-This process is commonly referred to as progression \tcr{(citation needed!)} and results in a much worse prognosis for the patient.
-There are many means \tcr{(citation needed!)} by which a tumor can transition from a premalignant state into a malignant state, but one of the main ones is that of driver mutations.
-Driver mutations are those that "directly or indirectly [confer] a selective growth advantage to the cell in which it occurs." \cite{ryan2016hallmarks}.
-There are several identifiable cell processes that can be affected by such driver mutations, including proliferation, apoptosis, and immunogenicity.
-\par
-When it comes to cancer and the immune system, they interact in myriad ways.
-Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME): responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}.
-Recent breakthroughs in the field of immunotherapy are beginning to produce therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}.
-\par 
-Tumor cells interact with the immune system by initiating a sequence of events that results in the activation of the adaptive immune response.
-This is accomplished by the presentation of antigens recognized by innate immune cells that are transported to lymph nodes where T cells (and other components) can be activated \cite{schreiber11_cancer}.
-The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via activation of T regulatory cells (Tregs) \cite{schreiber11_cancer}.
-\par
-The effects of the immune system on a tumor can be broadly summarized into two modules.
-The cytotoxic branch of the immune system, such as natural killer cells (NKs) and cytotoxic T cells (CTLs), seek out and lyse tumor cells.
-Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely \cite{finn12_immunooncology-1}.
-The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
-In addition, the immune system plays important regulatory roles in premalignant tissue, where inflammation can increase the probability of progression (and/or decrease the time to cancer), including in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
-Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.
-\par
-Epithelial-to-mesenchymal transition (EMT) describes a reversible process by which cells with an epithelial phenotype transition into cells with a mesenchymal phenotype.
-While epithelial cells are in part defined by their adhesion to neighbors, mesenchymal cells show much greater ranges of motility, and may display stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.
-Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}.
-Ongoing investigations into the plasticity and stability of EMT overlap with discussions elsewhere, e.g. of discrete vs. continuous processes during cell differentiation \cite{moris16_transition}.
-Intermediate states have emerged as a central mechanism by which cell fates (and the noise inherent within them) can be controlled \cite{maclean18_exploring, ta16_controlling, rackauckas18_meanindependent}.
-\par 
-While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of tumor progression \cite{nieto2016emt,peinado2007snail}.
-TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
-There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.
-Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
-It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.
-\par
-Two features of the mesenchymal phenotype are of particular relevance in the context of cancer and the immune system: i) mesenchymal cells are less susceptible to immune clearance \cite{terry2017new}; and ii) mesenchymal cells proliferate on average at a slower rate than epithelial cells.
-The immune evasiveness of mesenchymal cells is well-established: as a cell is targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be established before the lysing event.
-This immunological synapse is dependent upon the surface proteins, such as the T-cell receptors of the major histocompatibility complex, of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form \cite{terry2017new}.
-Below we refer to this phenotype as mesenchymal immune evasion (MIE).
-\par
-The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculate to exhibit a more ``stemness'' phenotype.
-While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects}.
-Cancer cells are of course characterized by some degree of unregulated proliferation, thus it is appealing to explore the effects of differential proliferation rates within the cancer cell population.
-In our model, we refer to the reduced proliferative capacity associated with a mesenchymal cell phenotype as mesenchymal growth arrest (MGA).
-\par
-Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including classic \cite{anderson98_continuous, sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{kim18_cell, gallaher14_bridging, gallaher18_spatial, an15_agentbased, serre16_mathematical, louzoun14_mathematical, briones-orta13_arkadia, lavi13_role, greene15_modeling, greene16_mathematical, cho17_modeling-1,  benzekry17_mathematical, owen11_mathematical, west18_multidrug}.
-These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximize the efficacy of cytotoxicity while minimizing risks to the patient.
-However, to our knowledge no models have addressed how the effects of EMT alter interactions between the immune system and cancer, and the subsequent implications for treatment.
-\par
-Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
-We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that premalignant tissue will progress and, if so, when.
-A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
-In doing so, we shift the focus of the previous model from cancer initiation to cancer progression.
-We do this to reflect the fact that cancer progression hinges on escape from the immune system and the fact that EMT has a more well-defined role during progression and metastasis.
-We seek to understand whether this more complex immune module will change our understanding of inflammatory effects on the tumor, and how the epithelial-mesenchymal axis influences these.
-\par 
-In the next section we develop the model and give explanation of the intuition behind each component.
-We then go on to analyze its behavior.
-Global sensitivity analysis identifies parameters that are crucial for progression.
-We then study these in more depth, focusing on the competing effects of EMT and of the immune system on progression: we find that EMT intricately regulates progression and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
-To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.
 
@@
 
-\begin{table}[H]
+\begin{table}
 \begin{center}
- \begin{tabular}{|| c | c ||} 
+ \begin{tabular}{|| c | l ||} 
  \hline
@@
 
-\subsection{Analysis of TCGA data in comparison with model predictions suggests that mesenchymal phenotypes reduce cancer-free survival probability}\label{tcga}
+\subsection{TCGA data analysis corroborates model predictions and shows that mesenchymal phenotypic properties reduce cancer-free survival probability}\label{tcga}
 
-To further assess the effects on progression of the mesenchymal properties of immune evasion and growth arrest, we study the joint density plot for each of these parameters against the Time to Cancer (Fig.
+The effects on disease progression of the mesenchymal properties of immune evasion and growth arrest can also be assessed via the joint density plot of these parameters against the Time to Cancer (Fig.
 \ref{fig:MIEvsMGA}).
@@
 To compare these model predictions with experimental studies, we analyzed data from The Cancer Genome Atlas (TCGA) database to study the effects of immune and EMT interactions on prognosis of cancers for which inflammation is known to play an important role, such as colonic or pancreatic cancers \cite{greten2019inflammation,hu2010inflammation}.
-We note that the comparison between simulation and data here is indirect since the model studies the progression of cancer from an arbitrary starting point, while the data address the progression of cancer following treatment, but analogous processes are at play during both the progression of  clonal dynamics addressed by the model, and the progression of disease after cancer treatment.
-In particular, the plasticity of tumor cells allows them to evade treatment by undergoing post-treatment processes resembling the de-novo appearance of cancer\cite{sanchez2018slow}.
-\par 
 The TCGA Pan-Cancer Clinical Data Resource (TCGA-CDR) provides multiple computed clinical endpoints for PAAD \cite{liu2018integrated}.
@@
 In turn, we expect the distance between DFI and OS to be relatively small for a tumor undergoing rapid progression following initial (post-treatment) detection.
-Therefore, recurrent tumors may be roughly categorized as either rapidly progressing or slowly progressing by some arbitrary threshold of DFI/OS divergence.
+Therefore, recurrent tumors may be roughly categorized as either rapidly progressing or slowly progressing by defining a threshold for the DFI/OS divergence.
 We selected three tumor types whose TCGA-CDR profile indicated slow progression (OV, SKCM, LIHC) and three tumor types whose profile indicated rapid progression (PAAD, LUAD, COAD).
@@
 This suggests that when studying interactions between cancer and the immune system, is it important to consider EMT; the effects of which may impact the dynamics and should not be overlooked.
+
+We note that the comparison between simulation and data here is indirect since the model studies the progression of cancer from an arbitrary starting point, while the data address the progression of cancer following treatment, but analogous processes are at play during both the progression of  clonal dynamics addressed by the model, and the progression of disease after cancer treatment.
+In particular, the plasticity of tumor cells allows them to evade treatment by undergoing post-treatment processes resembling the de-novo appearance of cancer\cite{sanchez2018slow}.
 

### 2019-10-18 Paper/CleanPaper.tex
edit main

@@
 Inflammation has particularly significant effects on gastrointestinal and pancreatic tumors.
-Moreover, epithelial cell plasticity can significantly alter disease trajectories via epithelial-to-mesenchymal transition (EMT).
+Moreover, tumor cell plasticity can significantly alter disease trajectories via epithelial-to-mesenchymal transition (EMT).
 Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of these regulatory processes.
@@
 
-\subsection{TCGA data analysis corroborates model predictions and shows that mesenchymal phenotypic properties reduce cancer-free survival probability}\label{tcga}
+\subsection{TCGA data analysis supports model predictions by showing that mesenchymal phenotypic properties reduce cancer-free survival }\label{tcga}
 
-The effects on disease progression of the mesenchymal properties of immune evasion and growth arrest can also be assessed via the joint density plot of these parameters against the Time to Cancer (Fig.
+The effects that mesenchymal phenotypic properties have on progression can also be assessed by studying mesenchymal immune evasion ($\Delta_\text{MIE}$) and mesenchymal growth arrest ($\Delta_\text{MGA}$) on a heatmap (Fig.
 \ref{fig:MIEvsMGA}).
@@
 \section{Discussion}\label{Discussion}
-Despite the intense interest in cancer and the immune system, and in the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these three components--cancer, the immune system, and EMT--during the progression from a premalignant to a malignant tumor.
+Despite the intense research focus on interactions between cancer and the immune system, and well as on the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these three components.
+Here we studied cancer, the immune system, and EMT, during the progression from a premalignant tumor state to a malignant, possibly invasive tumor.
 We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$.

### 2019-11-01 Paper/CleanPaper.tex
update intro/results

@@
 \author{Daniel R. Bergman$^{1}$,
-Matthew Karikomi\,$^{1}$,
+Matthew K. Karikomi\,$^{1}$,
+Min Yu\,$^{4}$,
 Qing Nie\,$^{1,2,*}$
-and Adam L. MacLean\,$^{3,*}$
+and Adam L. MacLean\,$^{1,3,*}$
 }
@@
   $^3$Department of Biological Sciences, University of Southern California, Los Angeles, CA 90089, USA \\
+  $^4$Department of Stem Cell \& Regenerative Medicine, Keck School of Medicine, University of Southern California, Los Angeles, CA 90033, USA \\
   $^*$Correspondence:  qnie@uci.edu (QN); macleana@usc.edu (ALM).
@@
 \section*{Abstract}
-During cancer progression, the immune system is engaged in complex sets of interactions with various tumor cells that affect prognosis.
+During progression from carcinoma in situ to an invasive tumor, the immune system is engaged in complex sets of interactions with various tumor cells.
 Inflammation has particularly significant effects on gastrointestinal and pancreatic tumors.
@@
 Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of these regulatory processes.
-Here we introduce a multiscale evolutionary model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial tissues.
-Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that maximize cancer-free survival.
-Delaying progression depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
+Here we introduce a multiscale evolutionary model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial cancers.
+Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that maximize invasion-free survival.
+Delaying progression depended crucially on properties of the mesenchymal tumor cell phenotype: its growth rate and its immune-evasiveness.
 We performed analysis of pancreatic cancer data from The Cancer Genome Atlas to compare clinical measurements with model predictions.
-We found that association with EMT worsened the survival probabilities of inflammation-associated cancer.
+We found that association with EMT worsened the invasion-free survival probabilities of inflammation-associated cancer.
 These results offer novel means to delay disease progression by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.
@@
 \section{Introduction}
-A key question in understanding cancer is how tumors transition from a premalignant state into an aggressive, malignant state.
-This process is commonly referred to as progression \tcr{(citation)} and results in a much worse prognosis for the patient.
-There are many means by which a tumor can transition from a premalignant state into a malignant state \tcr{(citation)}, but one of the main ones is that of driver mutations.
-Driver mutations are those that "directly or indirectly [confer] a selective growth advantage to the cell in which it occurs." \cite{ryan2016hallmarks}.
-There are several identifiable cell processes that can be affected by such driver mutations, including proliferation, apoptosis, and immunogenicity.
+The majority of deaths from cancer are due to metastasis of the disease \cite{dillekas}.
+It is thus of critical importance to understand better the progression from in situ to invasive disease.
+Underlying this progression are genetic and epigenetic events, including mutations in pathways critical to the success of the cancer cell (driver mutations) \cite{ryan2016hallmarks}.
+These pathways include cell proliferation, apoptosis, and immunogenicity.
 \par 
-When it comes to cancer and the immune system, they interact in myriad ways.
+Cancer and the immune system interact in myriad ways.
 Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME): responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}.
@@
 The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
-In addition, the immune system plays important regulatory roles in premalignant tissue, where inflammation can increase the probability of progression (and/or decrease the time to cancer), including in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
-Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.
+%In addition, the immune system plays important regulatory roles in premalignant tissue, where
+Inflammation can also increase the probability of progression (and/or decrease the time to invasion), with some of the most pronounced effects seen for tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
+Recent work has shown, contrary to these typical effects of inflammation that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.
 \par
@@
 Here we develop a model with the goal of studying interactions between the tumor, the immune system and EMT.
-We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that premalignant tissue will progress and, if so, when.
+We seek to describe a set of crucial molecular and cellular interactions in epithelial tumor cells, including effects due to DNA damage and mutation, to investigate the probability that in situ tumors will progress and, if so, when.
 A recent model of cancer-immune interactions \cite{guo17_multiscale} described the effects of the TME on the risk of cancer, and we build on the core cell cycle component of this model, adding significant new interactions to the immune component of the model (which was previously modeled by a single interaction), as well as adding the effects of EMT.
@@
 \newline
-    A premalignant tumor relies on driver mutations to alter key pathways that will allow the tumor to achieve greater growth potential and clinically progress to a malignant tumor.
+An in situ tumor relies on mutations to alter cellular signaling pathways that will ultimately allow the tumor to become invasive. The immune system
+
     At the same time, the immune system is recognizing and adapting to the resulting neoantigens to suppress the tumor.
@@
     \newline
-    We assume that as the premalignant (and later the malignant) tumor cells proliferate, they become increasingly likely to acquire a driver mutation altering a key pathway that unlocks greater growth potential.
-    However, innate immune cells (Natural Killer cells) are present to identify and clear these tumor cells, a process which results in neoantigens being used to prime and activate T cells in local lymph nodes and subsequently infiltrate the tumor microenvironment.
-    Once there, these T cells move to the tumor site and begin to either lyse tumor cells (cytotoxic T lymphocytes or CTLs) or suppress cytotoxic activity (T regulatory cells or Tregs).
-    We use the following expression to determine the probability of a tumor cell being lysed by a CTL in one time step of the model:
+    We assume that as the in situ (and later invasive) tumor cells proliferate, they become increasingly likely to acquire pathway mutations that can alter their death or proliferation propensities.
+Natural killer (NK) cells in the model can identify and clear these tumor cells, a process which results in neoantigens being used to prime and activate T cells in local lymph nodes and subsequently infiltrate the tumor microenvironment.
+At the tumor site, T cell populations can either lyse tumor cells (cytotoxic T lymphocytes; CTLs) or suppress cytotoxic activity (T regulatory cells; Tregs).
+    We use the following expression to determine the probability that during one time step of the model a tumor cell will be lysed by a CTL:
     $$\rho_{\text{CTL}} =\delta_{\text{MUT}} \frac{N_{\text{CTL}}}{N_C/K_{1}+N_{\text{CTL}}}  \frac{E_{\text{CTL}}}{1+N_{\text{Treg}}/K_2} (1-\delta_{\text{IE}}\Delta_{\text{IE}})(1-\zeta \Delta_{\text{MIE}})$$
@@
     \newline
-    What we desire to know is when do the malignant tumor cells reach a particular threshold proportion of all tumor cells.
+    What we desire to know is when do the invasive tumor cells reach a particular threshold proportion of all tumor cells.
     Once the tumor reaches this threshold, we declare that the tumor has progressed and we note how long the process took to reach this stage.
@@
 \item During a given cell cycle, exactly one event can happen to any one cell: proliferation, apoptosis, immune clearance, or rest
-\item All tissue cells (mutated and non-mutated) are equally likely to encounter immune cells
+\item All tumor cells (mutated and non-mutated) are equally likely to encounter immune cells
 \item Other spatial effects are ignored by the model, e.g. angiogenesis, hypoxia, invasive front, and chemotaxis
@@
 We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
-The agents in the model are the tissue cells that have already formed a premalignant tumor yet lack key driver mutations.
-In the process of the simulation, these cells can acquire driver mutations altering any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
+The agents in the model are the cells that have already formed an in situ tumor yet lack key pathway mutations to become invasive.
+In the process of the simulation, these cells can acquire mutations altering any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
 If the pathway is altered, the component is colored red and is marked with `X'.
-We distinguish between cells that have zero driver mutations and those with at least one as premalignant and malignant, respectively.
+%We distinguish between cells that have zero driver mutations and those with at least one as premalignant and malignant, respectively.
 \par
-We model immune cells as continuous variables, effectively assuming that the TME is well-mixed with regards to the immune infiltrate.
-The cytokine TGF-$\beta$ is also assumed to be well-mixed in the TME so that all tissue cells have equal access to it.
-Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
+We model immune cells as continuous variables, effectively assuming that the tumor microenvironment is well-mixed with regards to the infiltrating immune cells.
+The cytokine TGF-$\beta$ is also assumed to be well-mixed in the tumor microenvironment.
+Tumor cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
 While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
@@
 
-\subsection{Tissue Cells}\label{TissueCells}
-The tissue cells at all times have associated with them two important quantities: which combination of driver mutations they carry and an EMT score.
+\subsection{Tumor Cells}\label{TissueCells}
+The tumor cells at all times have associated with them two important quantities: which combination of driver mutations they carry and an EMT score.
 The model has three pathways that have the potential to become altered over the course of a simulation.
@@
 The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
-The NKs and CTLs act on the system by recognizing malignant cells and clearing them.
+The NKs and CTLs act on the system by recognizing invasive cells and clearing them.
 Upon successful clearance, they themselves are then deactivated and removed from the immune population.
@@
 \subsubsection{Initial conditions} 
-Every model run starts with an amount of premalignant tissue cells, $N_0$, determined by the set of parameters used for the run.
+Every model run starts with a number of in situ tumor cells, $N_0$, determined by the set of parameters used for the run.
 A set of warmup cycles are run to ensure arrival to this steady state.
 During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.
-After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
+After the warmup cycles are complete, a mutagenic event is simulated in which all the tumor cells experience a random increase in their probability to mutate.
 From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.
 
-\subsubsection{Tissue cell fate}
+\subsubsection{Tumor cell fate}
 During each cell cycle, every cell is randomly assigned one of four fates: proliferation, apoptosis, immune clearance, and rest in $G_0$.
@@
 EMT and mutational state, and cell extrinsic, e.g.
-immune infiltrate and total number of tissue cells, each of the four options is given a probability of occurring for each cell.
-For proliferation, we assume the total number of tissue cells exerts a negative feedback on the proliferation probability for each cell.
+immune infiltrate and total number of tumor cells, each of the four options is given a probability of occurring for each cell.
+For proliferation, we assume the total number of tumor cells exerts a negative feedback on the proliferation probability for each cell.
 Also, cells that have a driver mutation in the proliferation pathway have an increased probability of proliferating.
@@
 Tregs are assumed to produce much more TGF-$\beta$ than malignant cells.
-The resulting quantity of TGF-$\beta$ is divided up randomly among the tissue cells.
+The resulting quantity of TGF-$\beta$ is divided up randomly among the tumor cells.
 This quantity is combined with the EMT score of each cell and the resulting quantity determines which cells undergo EMT and which undergo MET.
@@
  \hline
- $p$ & proliferation rate of tissue cells \\ 
+ $p$ & proliferation rate of tumor cells \\ 
  \hline
- $d_C$  & death rate of tissue cells \\
+ $d_C$  & death rate of tumor cells \\
  \hline
@@
   \hline
- $K_0$ & EC50 term for negative feedback of tissue cells on own proliferation\\
+ $K_0$ & EC50 term for negative feedback of tumor cells on own proliferation\\
  \hline
@@
 \subsubsection{Mutation and progression}
-At the end of each cell cycle, the proportion of tissue cells that are malignant is calculated, and if it is above a certain threshold, the simulation ends and we declare the tumor to have progressed.
+At the end of each cell cycle, the proportion of tumor cells that are invasive is calculated, and if it is above a certain threshold, the simulation ends and we declare the tumor to have progressed.
 The Time to Cancer is given as the time from the start of the simulation, not including warmup, until this timepoint.
@@
 For example, many previous studies have established the parameters of immune population dynamics such as recruitment and death rates\cite{de2014modeling}.
-Because our model only works with on the order of $10^2$ tissue cells whereas an actual tumor would have on the order of $10^9$ cells\cite{de2014modeling}, we scaled the parameters so the immune populations would also be proportionately reduced.
+Because our model only works with on the order of $10^2$ tumor cells whereas an actual tumor would have on the order of $10^9$ cells\cite{de2014modeling}, we scaled the parameters so the immune populations would also be proportionately reduced.
 That is, for instance, how we ended up with NK cells having a steady state population of 10 cells in the model.
@@
 
-\subsection{A multiscale agent-based model of EMT-immune-tissue cell interactions to study tumor progression}\label{ExplModel}
+\subsection{A multiscale agent-based model of EMT-immune-tumor cell interactions to study tumor progression}\label{ExplModel}
 
@@
 \caption{A.
-Schematic depiction of agent-based model components; each of the 10 columns represents a single tissue cell divided into three compartments representing the state (altered or not) of the three pathways with mutagenic potential; red/blue denotes altered/unaltered pathways.
+Schematic depiction of agent-based model components; each of the 10 columns represents a single tumor cell divided into three compartments representing the state (altered or not) of the three pathways with mutagenic potential; red/blue denotes altered/unaltered pathways.
 Black arrows depict regulation of the cell fate in each cell cycle.
-Inset depicts major interactions between the immune system and tissue cells.
+Inset depicts major interactions between the immune system and tumor cells.
 B. A representative simulation of one patient. The model parameter values used can be found in Table S2. 
@@
 For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a driver mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
-The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset).
-NK cells and CTLs attempt to clear malignant tissue cells, and deactivate upon successfully carrying out their cytotoxic function.
+The different means by which the immune system acts on tumor cells are also shown (\ref{fig:ModelIntro}A Inset).
+NK cells and CTLs attempt to clear malignant tumor cells, and deactivate upon successfully carrying out their cytotoxic function.
 Tregs inhibit cytotoxic activity.
@@
 
-\subsection{Mesenchymal phenotypic properties dramatically change cancer-free survival times}\label{MesPars}
+\subsection{Mesenchymal phenotypic properties dramatically alter invasion-free survival times}\label{MesPars}
 
@@
 When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
-Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
-The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a malignant mesenchymal cell will be subject to immune clearance in a given cycle.
+Both parameters are defined proportionally, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
+The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that an invasive mesenchymal cell will be subject to immune clearance in a given cycle.
 The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
@@
 \par 
-As MIE is increased, the Time to Cancer decreases (Fig.
-\ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of malignant cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
+As MIE increases, the invasion-free survival decreases (Fig.
+\ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of invasive cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
 These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.
 \par
-Second, as MGA increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Fig \ref{fig:FirstSurvivalCurves}B).
-This is not immediately intuitive, since the decreased proliferation rate affects all mesenchymal cells, not just malignant cells.
+As MGA increases, invasion-free survival increases: lower proliferation rates for mesenchymal cells slow down cancer progression (Fig \ref{fig:FirstSurvivalCurves}B).
+This is not intuitive, since decreased proliferation rates affects all mesenchymal cells, not just the invasive cells.
 \par
 Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
-In Fig \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter Time to Cancer.
-The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tissue cells to a mesenchymal phenotype.
-Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease Time to Cancer.
-Mesenchymal cells are more likely to evade the immune system, so pushing the system towards an overall more mesenchymal phenotype will better protect the cancer and decrease the Time to Cancer.
+In Fig \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter invasion-free survival time.
+The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tumor cells to a mesenchymal phenotype.
+Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease the invasion-free survival.
+Mesenchymal cells are more likely to evade the immune system, so pushing the system towards an overall more mesenchymal state will lead to poorer outcomes, decreasing invasion-free survival times.
 
@@
 
-We explored the effects of varying the inflammation state of the patient on cancer-free survival, to investigate competing interactions within the TME and their effect on EMT.
+We explored the effects of varying the inflammation state of the patient on invasion-free survival, to investigate competing interactions within the tumor microenvironment and their effect on EMT.
 Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable inflammation.
-For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and the Time to Cancer is monotonic, i.e.
+For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and invasion-free survival is monotonic, i.e.
 increasing either MIE (Fig.
 \ref{fig:VaryINFL_and_MesPars}A-B) or MGA (Fig.
-\ref{fig:VaryINFL_and_MesPars}C-D) decreases the Time to Cancer.
-However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the Time to Cancer is found with respect to MGA ($\Delta_\text{MGA}= 0.3$).
+\ref{fig:VaryINFL_and_MesPars}C-D) decreases the invasion-free survival time.
+However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the invasion-free survival time is found with respect to MGA ($\Delta_\text{MGA}= 0.3$).
 This is seen both for intermittent or low inflammation (Fig.
@@
 \par
-These striking differences in the mean Time to Cancer -- extended by up to one {\it in silico} year by fine-grained optimization of mesenchymal growth arrest -- have clear therapeutic implications. 
+These striking differences in the mean invasion-free survival -- that can be extended by up to one year by optimization of the mesenchymal growth arrest -- have clear therapeutic implications. 
 Our model predicts that a patient suffering intermittent inflammatory attacks at the tumor site will benefit from controlling the proliferation of the mesenchymal cells.
 \par
-In contrast, when MIE is varied for different inflammation cycling schemes, regardless of the cycling scheme, increasing the MIE leads to decreases in the Time to Cancer (worse survival probabilities), although in the case where the inflammatory state is permanently high, MIE has little effect on the Time to Cancer.
-Thus under any inflammation regime with periods of low inflammation, any reductions in the rate of mesenchymal immune evasion will lead to improved patient outcomes.
+In contrast, when MIE is varied for different inflammation cycling schemes, regardless of the cycling scheme, increasing the MIE will decreases the invasion-free survival (worsen cancer progression and prognosis).
+In the case where the inflammatory state is permanently high, however, MIE has little effect on the invasion-free survival time.
+Thus under any inflammation regime with periods of low inflammation, reductions of any size in the rate of mesenchymal immune evasion lead to improved patient outcomes.
 
@@
 
-\subsection{TCGA data analysis supports model predictions by showing that mesenchymal phenotypic properties reduce cancer-free survival }\label{tcga}
+\subsection{TCGA data analysis supports model predictions by showing that mesenchymal phenotypic properties reduce invasion-free survival }\label{tcga}
 
@@
 \ref{fig:MIEvsMGA}).
-We found that over the full range of values of $\Delta_\text{MGA}$ considered, increasing $\Delta_\text{MIE}$ decreases the Time to Cancer.
-However, for any given value of mesenchymal immune evasion, there is a value of mesenchymal growth arrest that maximizes the Time to Cancer.
+We found that over the full range of values of $\Delta_\text{MGA}$ considered, increasing $\Delta_\text{MIE}$ decreases the invasion-free survival time.
+However, for any given value of mesenchymal immune evasion, there is a value of mesenchymal growth arrest that maximizes the invasion-free survival.
 Moreover, this optimal value increases with increasing MIE.
-Together these results show that while reducing the ability of a mutated cell to evade the immune system by any amount will improve the probability of cancer-free survival, an optimal value of MGA will prolong survival.
+Together these results show that while reducing the ability of a mutated cell to evade the immune system by any amount will improve the probability of invasion-free survival, an optimal value of MGA will prolong the tumor remaining in situ.
 \par
@@
 Despite the intense research focus on interactions between cancer and the immune system, and well as on the effects of EMT on cancer, there has not previously, to the best of our knowledge, been a model developed that combines these three components.
-Here we studied cancer, the immune system, and EMT, during the progression from a premalignant tumor state to a malignant, possibly invasive tumor.
+Here we studied cancer, the immune system, and EMT, during the progression from an in situ tumor to invasive disease.
 We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$.
@@
 \par
-We found that this model recapitulated cancer-free survival dynamics, and via global parameter sensitivity analysis, we identified parameters exerting key control over model behavior.
-Focusing on these led us to identify that: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, or increasing Treg TGF-$\beta$ production all lead to shorter Times to Cancer.
-However varying the level of inflammation leads to paradoxical effects: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the Time to Cancer.
+We found that this model recapitulated invasion-free survival dynamics, and via global parameter sensitivity analysis, we identified parameters exerting key control over model behavior.
+Focusing on these led us to identify that: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, or increasing Treg TGF-$\beta$ production all lead to shorter invasion-free survival times.
+However, varying the level of inflammation led to paradoxical effects: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the invasion-free survival.
 \par
 To test these predictions, we performed unsupervised analysis of pancreatic cancer data from The Cancer Genome Atlas, and looked at survival across two groups: we found that the presence of an EMT signature increased differences in survival between groups in combination with inflammation.
-We chose to summarize {\em in silico} patient simulations with a single parameter: the Time to Cancer, which we found captured the essential characteristics of the model.
-There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in pre-cancerous premalignant tissues is a pressing need to shed insight into cellular biomarkers of cancer.
+We chose to summarize {\em in silico} patient simulations with a single parameter: the invasion-free survival time, which we found captured the essential characteristics of the model.
+There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in premalignant tissues is a pressing need to shed insight into cellular biomarkers of cancer.
 \par
@@
 \ref{fig:MOAT} and \ref{fig:VaryINFL_and_MesPars}B, D).
-Currently, inflammation is modeled as independently cycling between high and low schemes, however many of the agents considered in the model actively contribute to the inflammatory state, thus it would be interesting to consider connecting these components (for example by letting the level of inflammation be conditional on the number of mutant cells in the tissue).
+Currently, inflammation is modeled as independently cycling between high and low schemes, however many of the agents considered in the model actively contribute to the inflammatory state, thus as an extension these components could be coupled, for example by assuming that the level of inflammation depends on the number of and the degree of mutations that cells in the tumor harbor).
+%% not sure about
 Another layer of complexity is revealed by the natural anti-inflammatory role of Tregs.
@@
 \section*{Acknowledgements}
-The work is partially supported by National Institutes of Health grants R01GM123731, U01AR073159, and U54-CA217378; National Science Foundation grants DMS1562176 and DMS1763272; The Simons Foundation grant (594598 to Q.N.) and a grant from the Jayne Koskinas Ted Giovanis Foundation for Health and Policy joint with the Breast Cancer Research Foundation.
+Q.N.
+would like to acknowledge partial support for this work from National Institutes of Health grants R01GM123731, U01AR073159, and U54-CA217378; National Science Foundation grants DMS1562176 and DMS1763272; Simons Foundation grant (594598); and the Jayne Koskinas Ted Giovanis Foundation for Health and Policy joint with the Breast Cancer Research Foundation.
+A.L.M.
+would like to acknowledge partial support for this work from the American Cancer Society grant \#IRG-16-181-57.
 

### 2019-11-27 Paper/CleanPaper.tex
Finished updating all figures with new runs
Updated fig2 and a little wording

@@
 \newline
-An in situ tumor relies on mutations to alter cellular signaling pathways that will ultimately allow the tumor to become invasive. The immune system
-
-    At the same time, the immune system is recognizing and adapting to the resulting neoantigens to suppress the tumor.
-    All of this occurs within an environment that is often under the influence of inflammatory signals that can vary over time.
+An in situ tumor relies on mutations to alter cellular signaling pathways that will ultimately allow the tumor to become invasive.
+The immune system responds to the tumor upon recognition of the neoantigens and infiltrates the tumor microenvironment (TME).
+This environment is under the influence of inflammatory signals that vary over time.
+%
+%    At the same time, the immune system is recognizing and adapting to the resulting neoantigens to suppress the tumor.
+%    All of this occurs within an environment that is often under the influence of inflammatory signals that can vary over time.
 \newline
@@
     Here, $\delta_\text{MUT}$ is 1 or 0 depending on whether the cell has a driver mutation or not, respectively.
-    The second term is a hill function similar to the de Pillis-Radunskaya law.
+    The second term is a hill function similar to the dePillis-Radunskaya law and it assumes that the immune cells are well-mixed within the TME.
     The third term is the regulatory effect of Tregs.
@@
     These cells also demonstrate cellular plasticity by being able to undergo epithelial-to-mesenchymal transition (EMT) and move along this spectrum as a result of intrinsic and extrinsic factors.
-    Here, we choose TGF-$\beta$ as a master regulator of EMT and connect the TGF-$\beta$ level in the tumor microenvironment (TME) with the propensity of tumor cells to undergo EMT.
+    Here, we choose TGF-$\beta$ as a master regulator of EMT and connect the well-mixed TGF-$\beta$ level in the TME with the propensity of tumor cells to undergo EMT.
     As TGF-$\beta$ is produced by Tregs as well as by tumor cells, this creates an intriguing set of interlocking interactions that we show here to be capable of shaping tumor and patient outcomes.
@@
 
-\begin{table}[H]
-\begin{tabular}{|p{\textwidth}|}
-\hline
-%%%
-\begin{itemize}
-\item TGF-$\beta$ is well-mixed within the TME
-\item All cells are on the same ``clock'', i.e. they all decide at the same time what to do next
-\item During a given cell cycle, exactly one event can happen to any one cell: proliferation, apoptosis, immune clearance, or rest
-\item All tumor cells (mutated and non-mutated) are equally likely to encounter immune cells
-\item Other spatial effects are ignored by the model, e.g. angiogenesis, hypoxia, invasive front, and chemotaxis
-\item The initial cells are mutated, but are not able to form an aggressive cancer without further mutations
-\end{itemize}
-\\
-\hline
-\end{tabular}
-\caption{Basic assumptions of the model}
-\label{table:model_assumptions}
-\end{table}
+%\begin{table}[H]
+%\begin{tabular}{|p{\textwidth}|}
+%\hline
+%%%%
+%\begin{itemize}
+%\item TGF-$\beta$ is well-mixed within the TME
+%\item All cells are on the same ``clock'', i.e. they all decide at the same time what to do next
+%\item During a given cell cycle, exactly one event can happen to any one cell: proliferation, apoptosis, immune clearance, or rest
+%\item All tumor cells (mutated and non-mutated) are equally likely to encounter immune cells
+%\item Other spatial effects are ignored by the model, e.g. angiogenesis, hypoxia, invasive front, and chemotaxis
+%\item The initial cells are mutated, but are not able to form an aggressive cancer without further mutations
+%\end{itemize}
+%\\
+%\hline
+%\end{tabular}
+%\caption{Basic assumptions of the model}
+%\label{table:model_assumptions}
+%\end{table}
 
@@
 \center
-{\includegraphics[width=0.85\textwidth]{Figure2/Figure2.jpg}}
+{\includegraphics[width=0.85\textwidth]{Figure2/MOAT.jpg}}
 \caption{Sensitivity analysis of model parameters, determined via the global Morris one-step-at-a-time method.
@@
 \center
-{\includegraphics[width=0.5\textwidth]{Figure5/MIEvsMGA_bigcbar.jpg}}
+{\includegraphics[width=0.5\textwidth]{Figure5/heatmap2.jpg}}
 \caption{Summary of the contrasting effects of MIE and MGA on the Time to Cancer.

### 2019-11-29 Paper/CleanPaper.tex
updated supplement with old figs

@@
 {\includegraphics[width=0.85\textwidth]{Figure3/Figure3.pdf}}
-\caption{Effects of mesencymal cell properites on the time to cancer.
+\caption{Effects of mesenchymal tumor cell properties on the Time to Cancer.
 Trajectories of one patient per cohort from warmup period ([-1000,0]) to 2000 cell cycles, for $\Delta_\text{MIE}$ (A); $\Delta_\text{MGA}$ (B); and $\tau_\text{Treg}$ (C).
@@
 {\includegraphics[width=0.95\textwidth]{Figure4/Figure4.pdf}}
-\caption{Effects of inflammation on the time to cancer under different cycling schemes.
+\caption{Effects of inflammation on the Time to Cancer under different cycling schemes.
 A-B.

### 2019-11-29 Paper/Supplement/CleanSupplement.tex
cleaned out comments from Supplement and clean supplement
updated supplement with old figs

@@
 \documentclass[11pt, a4paper, preprint]{article}
-%\usepackage{outline}
-%\usepackage{float}
-%\usepackage{graphicx}
-%\usepackage{color}
-%\usepackage{amsmath}
-%\usepackage{cite}
-%\graphicspath{ {Figs/} }
-%\usepackage{longtable}
 
@@
 \renewcommand{\thetable}{S\arabic{table}}
-
-%\input{../structure.tex} % Specifies the document structure and loads requires packages
 
@@
 Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
-% Add: 
 This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.
@@
 This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
-% The following sentence could be altered to reflect that the threshold value for $\tau$ is 0, so not only are negative values possible, but very common.
-So, this is not really so much a TGFB received as it is a change in TGFB signaling.
-Try calling it a gradient/change of TGFB in the cell.
 The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
@@
 Finally, if this quantity is large enough, the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.
-% Need to add better explanation for Hill fn. Make sure thoroughly explained
 Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
@@
 
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure1/SinglePatientBlue.jpg}}
+\caption{
+Single patient trajectory without immune cells targeting premalignant cells.
+}
+\label{fig:1b_first}
+\end{figure}
+
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure1/OneSurvivalCurve.jpg}}
+\caption{
+Sample cohort survival curve without immune cells targeting premalignant cells.
+}
+\label{fig:1c_first}
+\end{figure}
+
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure2/Figure2.jpg}}
+\caption{
+Morris-OAT global sensitivity without immune cells targeting premalignant cells.
+}
+\label{fig:2_first}
+\end{figure}
+
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure3/Figure3_first.pdf}}
+\caption{
+Effects of mesenchymal tumor cell properties on the Time to Cancer without immune cells targeting premalignant cells.
+}
+\label{fig:3_first}
+\end{figure}
+
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure4/Figure4_first.pdf}}
+\caption{
+Effects of inflammation on the Time to Cancer without immune cells targeting premalignant cells.
+}
+\label{fig:4_first}
+\end{figure}
+
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure5/MIEvsMGA_bigcbar.jpg}}
+\caption{
+Summary of the contrasting effects of MIE and MGA on Time to Cancer without immune cells targeting premalignant cells.
+}
+\label{fig:5_first}
+\end{figure}
+
 \end{document}

### 2019-12-02 Paper/CleanPaper.tex
finished this round of edits!

@@
 \section{Introduction}
-The majority of deaths from cancer are due to metastasis of the disease \cite{dillekas}.
+The majority of deaths from cancer are due to metastasis of the disease \cite{dillekaas201990}.
 It is thus of critical importance to understand better the progression from in situ to invasive disease.
@@
 \par 
+%
+%%%%%%% I think we should reorder these next two paragraphs
 While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of tumor progression \cite{nieto2016emt,peinado2007snail}.
@@
 Below we refer to this phenotype as mesenchymal immune evasion (MIE).
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
 \par
-The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculate to exhibit a more ``stemness'' phenotype.
+The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculated to exhibit a more ``stemness'' phenotype.
 While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects}.
@@
 In our model, we refer to the reduced proliferative capacity associated with a mesenchymal cell phenotype as mesenchymal growth arrest (MGA).
-
+\par
 Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including classic \cite{anderson98_continuous, sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{kim18_cell, gallaher14_bridging, gallaher18_spatial, an15_agentbased, serre16_mathematical, louzoun14_mathematical, briones-orta13_arkadia, lavi13_role, greene15_modeling, greene16_mathematical, cho17_modeling-1,  benzekry17_mathematical, owen11_mathematical, west18_multidrug}.
@@
 In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
-In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
+% Do we still want the table of assumptions with the quick guide box?
+%In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
+In addition, we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
 \par
@@
 For apoptosis, the probability is constant except for those cells with a driver mutation in the apoptosis pathway which have a decreased probability of apoptosis.
-For immune clearance, we assume that only cells with newly acquired driver mutations are targeted by the immune system, and their probability of being cleared depends on the immune infiltrate.
+%For immune clearance, we assume that only cells with newly acquired driver mutations are targeted by the immune system, and their probability of being cleared depends on the immune infiltrate.
+For immune clearance, the malignant tumor cells are assumed more immunogenic and thus have a higher probability of being cleared by the immune system.
 This probability can be decreased by a cell having a driver mutation in the immune evasion pathway or by a cell being mesenchymal.
@@
 First, the immune populations are updated in two steps.
-In step one, Immune cells are exhausted based on the number of malignant cells that were cleared by their type.
+In step one, immune cells are exhausted based on the number of malignant cells that were cleared by their type.
 So, each clearance by an NK cell results in the NK population decreasing by one, and similarly for CTLs.
@@
 The inflammation cycling scheme for a typical {\it in silico} patient consists of alternating high and low regimes (Fig.
-\ref{fig:ModelIntro}B); the inflammation schemes modeled will be discussed in detail in the next section.
+\ref{fig:ModelIntro}B); the inflammation schemes modeled will be discussed in detail below.
 For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor.
@@
 Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days.
-For this patient, we kept let the simulation go beyond this timepoint, and we se that the number of malignant cells grows rapidly and soon makes up 100\% of the cell population.
+For this patient, we let the simulation go beyond this timepoint, and we see that the number of malignant cells grows rapidly and soon makes up 100\% of the cell population.
 A peak in number of mesenchymal cells is observed shortly after this threshold is reached, and following this most cells transition back to epithelial.
@@
 Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of driver mutations.
-The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation.
+The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 30 cell cycles of low inflammation.
 Within each of these periods, adaptive immune populations increase or decrease rapidly in accordance with the inflammation state.
@@
 In Fig.
-\ref{fig:ModelIntro}C we simulate the survival curve for a cohort of $500$ patients: we see that all the patients survive for approximately 100 cell cycles (75 days).
+\ref{fig:ModelIntro}C we simulate the survival curve for a cohort of $500$ patients:
+%we see that all the patients survive for approximately 100 cell cycles (75 days).
 Subsequently, in the approximate range of $T= [100,300]$, the cancer onset rate is roughly constant, and after $600$ cell cycles, no patients remain cancer free.
+we see that after the mutagenic event that a large number of tumors quickly progress while others takes some time before progressing. By about 1200 cell cycles, all tumors have progressed.
 
@@
 \par
-To assess the sensitivity of the model parameters and thus identify those that are most important in determining cancer-free survival times, we performed Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods).
+To assess the sensitivity of the model parameters and thus identify those that are most important in determining progression-free survival times, we performed Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods).
 The results of Morris OAT on the 31 model parameters are shown in Figure \ref{fig:MOAT}. We see that a subset of parameters demonstrate much higher levels of sensitivity than others.
-The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
-This ties Treg cells to all three components of the model.
-Since we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for detailed analysis below.
+%The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
+The two most influential according to this analysis are the recruitment rates of Tregs and CTLs under low inflammation.
+%This ties Treg cells to all three components of the model.
+%Since we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for detailed analysis below.
+We also note that the parameters related to EMT have a moderate effect on Time to Cancer.
 \par 
@@
 When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
-Both parameters are defined proportionally, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
+Both parameters are defined proportionally and hence lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
 The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that an invasive mesenchymal cell will be subject to immune clearance in a given cycle.
@@
 \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of invasive cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
-These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}A.
+These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}D.
 \par
-As MGA increases, invasion-free survival increases: lower proliferation rates for mesenchymal cells slow down cancer progression (Fig \ref{fig:FirstSurvivalCurves}B).
-This is not intuitive, since decreased proliferation rates affects all mesenchymal cells, not just the invasive cells.
+The relationship between MGA and invasion-free survival is non-monotonic.
+For smaller values of MGA, an increase in MGA results in increased invasion-free survival (Fig \ref{fig:FirstSurvivalCurves}E).
+This result is explored in Section \ref{KeyEMT}.
+%As MGA increases, invasion-free survival increases: lower proliferation rates for mesenchymal cells slow down cancer progression (Fig \ref{fig:FirstSurvivalCurves}B).
+%This is not intuitive, since decreased proliferation rates affects all mesenchymal cells, not just the invasive cells.
 \par
-Third, TGF-$\beta$ can be varied in two ways: the production by mesenchymal cells and the production by Treg cells.
-In Fig \ref{fig:FirstSurvivalCurves}C, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ production leads to a shorter invasion-free survival time.
-The two main ways in which TGF-$\beta$ influences the system is in recruitment of Treg cells and in pushing tumor cells to a mesenchymal phenotype.
-Treg cells are modeled as tumor-protective and thus increasing their number will naturally decrease the invasion-free survival.
-Mesenchymal cells are more likely to evade the immune system, so pushing the system towards an overall more mesenchymal state will lead to poorer outcomes, decreasing invasion-free survival times.
+Third, TGF-$\beta$ can be varied in two ways: the production by malignant cells and the production by Treg cells.
+In Fig \ref{fig:FirstSurvivalCurves}F, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ does not have much effect on invasion-free survival time.
+The effect we do see is that lower Treg TGF-$\beta$ results in more rapid invasion or if it survives to a certain time point, a longer time to invasion.
+Here, we choose to focus on the results of the effect of MIE and MGA as the more direct link to the mesenchymal phenotype.
 
@@
 Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable inflammation.
-For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and invasion-free survival is monotonic, i.e.
+Compared to the other inflammation regimes, a permanently high inflammation state results in tumors that are relatively resistant to changes in mesenchymal parameters (Fig.
+\ref{fig:VaryINFL_and_MesPars}).
+However, when the inflammation either has temporary or permanent periods of low inflammation, different relationships emerge: invasion-free survival time becomes negatively correlated with MIE, and a local maximum for the invasion-free survival time is found with respect to MGA ($\Delta_\text{MGA}= 0.2$).
+
+%For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and invasion-free survival is monotonic, i.e.
 increasing either MIE (Fig.
@@
 \ref{fig:VaryINFL_and_MesPars}C-D) decreases the invasion-free survival time.
-However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the invasion-free survival time is found with respect to MGA ($\Delta_\text{MGA}= 0.3$).
+%However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the invasion-free survival time is found with respect to MGA ($\Delta_\text{MGA}= 0.3$).
 This is seen both for intermittent or low inflammation (Fig.
@@
 \par
-These striking differences in the mean invasion-free survival -- that can be extended by up to one year by optimization of the mesenchymal growth arrest -- have clear therapeutic implications. 
+These striking differences in the mean invasion-free survival -- that can be extended by up to double compared to other MGA values-- have clear therapeutic implications. 
 Our model predicts that a patient suffering intermittent inflammatory attacks at the tumor site will benefit from controlling the proliferation of the mesenchymal cells.
 \par
-In contrast, when MIE is varied for different inflammation cycling schemes, regardless of the cycling scheme, increasing the MIE will decreases the invasion-free survival (worsen cancer progression and prognosis).
-In the case where the inflammatory state is permanently high, however, MIE has little effect on the invasion-free survival time.
+In contrast, when MIE is varied for different inflammation cycling schemes, regardless of the cycling scheme, increasing the MIE will decreases the invasion-free survival (worsen cancer progression and prognosis) or have minimal effect.
 Thus under any inflammation regime with periods of low inflammation, reductions of any size in the rate of mesenchymal immune evasion lead to improved patient outcomes.
@@
 Therefore, recurrent tumors may be roughly categorized as either rapidly progressing or slowly progressing by defining a threshold for the DFI/OS divergence.
-We selected three tumor types whose TCGA-CDR profile indicated slow progression (OV, SKCM, LIHC) and three tumor types whose profile indicated rapid progression (PAAD, LUAD, COAD).
-See Figures \ref{fig:OV} and \ref{fig:PAAD} for OV and PAAD. The others will be in the supplement.
+We selected three tumor types whose TCGA-CDR profile indicated slow progression (OV, SKCM, LIHC) and two tumor types whose profile indicated rapid progression (PAAD, LUAD).
+See Figures \ref{fig:OV} and \ref{fig:PAAD} for OV and PAAD. The others are in the supplement.
 \par
@@
 We found that this model recapitulated invasion-free survival dynamics, and via global parameter sensitivity analysis, we identified parameters exerting key control over model behavior.
-Focusing on these led us to identify that: increasing mesenchymal immune evasion, decreasing mesenchymal growth arrest, or increasing Treg TGF-$\beta$ production all lead to shorter invasion-free survival times.
-However, varying the level of inflammation led to paradoxical effects: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the invasion-free survival.
+Focusing on these led us to identify that: increasing mesenchymal immune evasion and increasing Treg TGF-$\beta$ production both lead to shorter invasion-free survival times.
+However, varying the level of inflammation led to paradoxical effects with regards to mesenchymal growth arrest: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the invasion-free survival.
 \par
@@
 However, TGF-$\beta$ is a master regulator implicated in numerous cellular signaling processes, and changing the concentrations of TGF-$\beta$ even in a local tumor microenvironment could have large off-target effects.
-Indeed, it has been showns that TGF-$\beta$ promotes invasion and heterogeneity (although suppresses cell proliferation) in squamous cell carcinoma \cite{oshimori15_tgfv}.
+Indeed, it has been shown that TGF-$\beta$ promotes invasion and heterogeneity (although suppresses cell proliferation) in squamous cell carcinoma \cite{oshimori15_tgfv}.
 Future work thus ought to consider the effects of targeting signaling factors downstream of TGF-$\beta$ that still have the ability to modulate epithelial cell dynamics.

### 2019-12-02 Paper/Supplement/CleanSupplement.tex
finished this round of edits!

@@
 \center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure6/SKCM.pdf}}
+\caption{A. K-means clustering of SKCM using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
+B. Survival plots corresponding to the clustering on EMT and inflammation.
+C. K-means clustering of SKCM using gene ontology terms indicative of an EMT signature ($k=2$).
+D. Survival plots corresponding to the clustering on EMT.
+E. K-means clustering of SKCM using gene ontology terms indicative of inflammation ($k=2$).
+F. Survival plots corresponding to the clustering on inflammation.}
+\label{fig:SKCM}
+\end{figure}
+
+\begin{figure}
+\center
 {\includegraphics[width=0.65\textwidth]{../Figs/Figure6/LIHC.pdf}}
@@
 \center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure6/LUAD.pdf}}
+\caption{A. K-means clustering of LUAD using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
+B. Survival plots corresponding to the clustering on EMT and inflammation.
+C. K-means clustering of LUAD using gene ontology terms indicative of an EMT signature ($k=2$).
+D. Survival plots corresponding to the clustering on EMT.
+E. K-means clustering of LUAD using gene ontology terms indicative of inflammation ($k=2$).
+F. Survival plots corresponding to the clustering on inflammation.}
+\label{fig:LUAD}
+\end{figure}
+
+\begin{figure}
+\center
 {\includegraphics[width=0.65\textwidth]{../Figs/Figure1/SinglePatientBlue.jpg}}
 \caption{
-Single patient trajectory without immune cells targeting premalignant cells.
+Single patient trajectory without immune cells targeting premalignant cells. Compare to Fig. 1B.
 }
@@
 \caption{
-Sample cohort survival curve without immune cells targeting premalignant cells.
+Sample cohort survival curve without immune cells targeting premalignant cells. Compare to Fig. 1C.
 }
@@
 \caption{
-Morris-OAT global sensitivity without immune cells targeting premalignant cells.
+Morris-OAT global sensitivity without immune cells targeting premalignant cells. Compare to Fig. 2.
 }
@@
 \caption{
-Effects of mesenchymal tumor cell properties on the Time to Cancer without immune cells targeting premalignant cells.
+Effects of mesenchymal tumor cell properties on the Time to Cancer without immune cells targeting premalignant cells. Compare to Fig. 3.
 }
@@
 \caption{
-Effects of inflammation on the Time to Cancer without immune cells targeting premalignant cells.
+Effects of inflammation on the Time to Cancer without immune cells targeting premalignant cells. Compare to Fig. 4.
 }
@@
 \caption{
-Summary of the contrasting effects of MIE and MGA on Time to Cancer without immune cells targeting premalignant cells.
+Summary of the contrasting effects of MIE and MGA on Time to Cancer without immune cells targeting premalignant cells. Compare to Fig. 5.
 }

### 2019-12-09 Paper/CleanPaper.tex
update intro, model assumptions, methods

@@
 
-\author{Daniel R. Bergman$^{1}$,
-Matthew K. Karikomi\,$^{1}$,
-Min Yu\,$^{4}$,
-Qing Nie\,$^{1,2,*}$
-and Adam L. MacLean\,$^{1,3,*}$
-}
-\affil{
-  $^1$Department of Mathematics, University of California, Irvine,  Irvine, CA 92697, USA \\
-  $^2$Department of Cell and Developmental Biology, University of California, Irvine, Irvine, CA 92697, USA \\
-  $^3$Department of Biological Sciences, University of Southern California, Los Angeles, CA 90089, USA \\
-  $^4$Department of Stem Cell \& Regenerative Medicine, Keck School of Medicine, University of Southern California, Los Angeles, CA 90033, USA \\
-  $^*$Correspondence:  qnie@uci.edu (QN); macleana@usc.edu (ALM).
-}
+\author{ Daniel R. Bergman\textsuperscript{1},
+Matthew K. Karikomi\textsuperscript{1},
+Min Yu\textsuperscript{4},
+Qing Nie\textsuperscript{1,2,*},
+and Adam L. MacLean\textsuperscript{1,3,*}
+\newline\newline % Space before institutions
+	\textsuperscript{1}Department of Mathematics, University of California, Irvine,  Irvine, CA 92697 \\
+        \textsuperscript{2}Department of Cell and Developmental Biology, University of California, Irvine, Irvine, CA 92697 \\
+        \textsuperscript{3}Department of Biological Sciences, University of Southern California, Los Angeles, CA 90089 \\
+        \textsuperscript{4}Department of Stem Cell \& Regenerative Medicine, Keck School of Medicine, University of Southern California, Los Angeles, CA 90033 \\
+        \textsuperscript{*}Correspondence:  qnie@uci.edu (Q.N.); macleana@usc.edu (A.L.M.). }
 
@@
 
-\section*{Abstract}
-During progression from carcinoma in situ to an invasive tumor, the immune system is engaged in complex sets of interactions with various tumor cells.
-Inflammation has particularly significant effects on gastrointestinal and pancreatic tumors.
+\thispagestyle{firstpage} % Apply the page style for the first page (no headers and footers)
+
+%\section*{Abstract}
+\lettrineabstract{During progression from carcinoma in situ to an invasive tumor, the immune system is engaged in complex sets of interactions with various tumor cells.
 Moreover, tumor cell plasticity can significantly alter disease trajectories via epithelial-to-mesenchymal transition (EMT).
@@
 Here we introduce a multiscale evolutionary model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial cancers.
-Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that maximize invasion-free survival.
-Delaying progression depended crucially on properties of the mesenchymal tumor cell phenotype: its growth rate and its immune-evasiveness.
-We performed analysis of pancreatic cancer data from The Cancer Genome Atlas to compare clinical measurements with model predictions.
-We found that association with EMT worsened the invasion-free survival probabilities of inflammation-associated cancer.
+Through in silico analysis of large patient cohorts, we find controllable regimes that maximize invasion-free survival.
+We identify that delaying tumor progression depends crucially on properties of the mesenchymal tumor cell phenotype: its growth rate and its immune-evasiveness.
+Through analysis of pancreatic and ovarian cancers from The Cancer Genome Atlas to assess model predictions against clinical measurements, we find that association with EMT significantly worsened the invasion-free survival probabilities of inflammation-associated cancer.
 These results offer novel means to delay disease progression by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.
+}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      %                    INTRODUCTION                    %
-%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Cancer and the immune system interact in myriad ways.
-Beyond direct tumor-immune cell interactions, the immune system modulates the tumor microenvironment (TME): responding to local inflammatory signals, targeting mutated cells, and eradicating tumor cells to potentially restore homeostasis in a local context \cite{de2006paradoxical}.
-Recent breakthroughs in the field of immunotherapy are beginning to produce therapies that have had a significant impact on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}.
+The immune system modulates the tumor microenvironment (TME), since immune signals that affect the tumor can be amplified or repressed through feedback in response to local inflammatory signals.
+This complex cell signaling occurs alongside the targeting (and potential eradication) of the tumor by immune cells \cite{de2006paradoxical}.
 \par 
-Tumor cells interact with the immune system by initiating a sequence of events that results in the activation of the adaptive immune response.
-This is accomplished by the presentation of antigens recognized by innate immune cells that are transported to lymph nodes where T cells (and other components) can be activated \cite{schreiber11_cancer}.
-The tumor also engages in cellular processes that indirectly shape the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), thus shifting the TME towards a tumor-supportive environment by enhancing immunosuppression via activation of T regulatory cells (Tregs) \cite{schreiber11_cancer}.
+The effects of the immune system on a tumor can be broadly summarized into two branches.
+The cytotoxic branch of the immune system, such as natural killer cells (NKs) and cytotoxic T cells (CTLs), seek out and lyse tumor cells.
+Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate \cite{finn12_immunooncology-1}.
+The regulatory branch of the immune system (Tregs, and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
+Inflammation can increase the probability of cancer incidence and progression, with some of the most pronounced effects seen for tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
+Recent work has shown, contrary to the typical effects of inflammation on cancer, that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.
 \par
-The effects of the immune system on a tumor can be broadly summarized into two modules.
-The cytotoxic branch of the immune system, such as natural killer cells (NKs) and cytotoxic T cells (CTLs), seek out and lyse tumor cells.
-Upon carrying out their effector functions, these cytotoxic cells lose efficacy or deactivate entirely \cite{finn12_immunooncology-1}.
-The regulatory branch of the immune system (Tregs and other factors), inhibits the effective functioning of the cytotoxic branch \cite{ruffell2010lymphocytes}.
-%In addition, the immune system plays important regulatory roles in premalignant tissue, where
-Inflammation can also increase the probability of progression (and/or decrease the time to invasion), with some of the most pronounced effects seen for tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.
-Recent work has shown, contrary to these typical effects of inflammation that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.
+Immunotherapies are beginning to realize their potential, with significant impacts on patient health and survival \cite{pardoll2012blockade,restifo2012adoptive}, and may even provide a cure for certain hematopoietic cancers via anti-CD19 CAR-T cells \cite{safeandpotent}.
+The presentation of antigens on tumor cells is recognized by innate immune cells that are transported to lymph nodes where T cells (and other components) can be activated \cite{schreiber11_cancer}.
+The tumor also engages in processes that can indirectly modify the TME, for example by releasing transforming growth factor beta (TGF-$\beta$), which can shif the TME towards a tumor-supportive environment by enhancing immunosuppression via activation of Tregs \cite{schreiber11_cancer}.
 \par
-Epithelial-to-mesenchymal transition (EMT) describes a reversible process by which cells with an epithelial phenotype transition into cells with a mesenchymal phenotype.
-While epithelial cells are in part defined by their adhesion to neighbors, mesenchymal cells show much greater ranges of motility, and may display stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.
-Recent work has shown that -- rather than being a binary process -- at least two stable intermediate states exist on the EMT axis \cite{hong2015ovol2, jolly15_coupling}.
+Epithelial-to-mesenchymal transition (EMT) describes a reversible process by which cells displaying an epithelial phenotype transition into cells with a mesenchymal phenotype.
+Epithelial cells are -- in part -- defined by tight cell-cell adhesion.
+Mesenchymal cells exhibit less adhesion, greater ranges of motility, and may possess stem-like properties \cite{nieto2016emt}, although controversy regarding `stemness' and EMT remains \cite{nie18_stem, sha19_intermediate}.
+Recent work has shown that -- rather than being a binary process -- at least two stable intermediate EMT states exist \cite{hong2015ovol2, jolly15_coupling}.
 Ongoing investigations into the plasticity and stability of EMT overlap with discussions elsewhere, e.g. of discrete vs. continuous processes during cell differentiation \cite{moris16_transition}.
@@
 \par 
-%
-%%%%%%% I think we should reorder these next two paragraphs
-While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of tumor progression \cite{nieto2016emt,peinado2007snail}.
-TGF-$\beta$ is a master regulator of EMT \cite{lim2012epithelial}, linking it to the signaling networks involved in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
-There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.
-Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked.
-It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.
+Two features of the mesenchymal phenotype are of particular relevance in the context of cancer-immune interactions.
+i) mesenchymal tumor cells proliferate less than epithelial cells, we refer to this as mesenchymal growth arrest (MGA), and can be considered related to (in the sense of quiescence) the ``stemness'' phenotype of the mesenchymal tumor cells \cite{woods2014effects}.
+ii) mesenchymal cells are less susceptible to immune clearance \cite{terry2017new}.
+As a cell is targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be established.
+This immunological synapse -- mediated in part by T-cell receptors bound to antigens and the major histocompatibility complex on the target cell -- is down-regulated in mesenchymal cells, thus inhibiting formation of the synapse \cite{terry2017new}.
+We refer to this phenotype as mesenchymal immune evasion (MIE).
 \par
-Two features of the mesenchymal phenotype are of particular relevance in the context of cancer and the immune system: i) mesenchymal cells are less susceptible to immune clearance \cite{terry2017new}; and ii) mesenchymal cells proliferate on average at a slower rate than epithelial cells.
-The immune evasiveness of mesenchymal cells is well-established: as a cell is targeted by cytotoxic immune cells for clearance, a physical connection between the two cells must be established before the lysing event.
-This immunological synapse is dependent upon the surface proteins, such as the T-cell receptors of the major histocompatibility complex, of the target cell, and in the case of mesenchymal cells there is preliminary evidence that these surface markers are down-regulated in such a way that the immune synapse is more difficult to form \cite{terry2017new}.
-Below we refer to this phenotype as mesenchymal immune evasion (MIE).
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
+In addition to the prominent role it plays in metastasis, EMT has more recently been shown to also regulate other aspects of tumor progression \cite{nieto2016emt,peinado2007snail}.
+TGF-$\beta$, a master regulator of EMT \cite{lim2012epithelial}, is at once implicated heavily in tumor-mediated immune responses, since Tregs release TGF-$\beta$ upon arriving at the tumor site\cite{terry2017new}.
+In hepatocellular carcinoma, for example, there is direct evidence linking Treg-secreted TGF-$\beta$ with EMT \cite{shi2019cd4+}.
+Thus, even by considering only the TGF-$\beta$ pathway, we find compelling evidence that  these three core components (the tumor, the immune system, and EMT) all interact.
+It therefore strikes us as a priority to develop models to understand how the interactions between each of these three components affect cancer incidence and progression.
 \par
-The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculated to exhibit a more ``stemness'' phenotype.
+%The other facet of mesenchymal cells is their slower proliferation rate, and this is tied to the observation that mesenchymal cells have been speculated to exhibit a more ``stemness'' phenotype.
 While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects}.
-Cancer cells are of course characterized by some degree of unregulated proliferation, thus it is appealing to explore the effects of differential proliferation rates within the cancer cell population.
-In our model, we refer to the reduced proliferative capacity associated with a mesenchymal cell phenotype as mesenchymal growth arrest (MGA).
+%Cancer cells are of course characterized by some degree of unregulated proliferation, thus it is appealing to explore the effects of differential proliferation rates within the cancer cell population.
+%In our model, we refer to the reduced proliferative capacity associated with a mesenchymal cell phenotype as mesenchymal growth arrest (MGA).
 \par
-Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including classic \cite{anderson98_continuous, sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{kim18_cell, gallaher14_bridging, gallaher18_spatial, an15_agentbased, serre16_mathematical, louzoun14_mathematical, briones-orta13_arkadia, lavi13_role, greene15_modeling, greene16_mathematical, cho17_modeling-1,  benzekry17_mathematical, owen11_mathematical, west18_multidrug}.
+Mathematical oncology, that is, mathematical models of cancer incidence, progression, and treatment, has become a well-developed field; many models have offered insight into the cellular interactions underlying cancer and its interplay with the immune system, including older \cite{anderson98_continuous, sherrattjonathana.92_oncogenes, pillis05_validated} and more recent works \cite{kim18_cell, gallaher14_bridging, gallaher18_spatial, an15_agentbased, serre16_mathematical, louzoun14_mathematical, briones-orta13_arkadia, lavi13_role, greene15_modeling, greene16_mathematical, cho17_modeling-1,  benzekry17_mathematical, owen11_mathematical, west18_multidrug}.
 These studies have increased our understanding of how tumors grow in the presence of various immune components, and how treatment regimes can be designed to maximize the efficacy of cytotoxicity while minimizing risks to the patient.
@@
 \par 
-In the next section we develop the model and give explanation of the intuition behind each component.
-We then go on to analyze its behavior.
-Global sensitivity analysis identifies parameters that are crucial for progression.
-We then study these in more depth, focusing on the competing effects of EMT and of the immune system on progression: we find that EMT intricately regulates progression and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
-To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic cancer.
+In the next section we develop the model, explaining the intuition behind each of its components.
+We go on to analyze its behavior: global ``{\em one-at-a-time}'' sensitivity analysis identifies parameters that are crucial for progression.
+We study these in more depth, focusing on the competing effects of EMT and of the immune system on progression, and discover that EMT intricately regulates progression: under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong invasion-free survival.
+To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects of inflammation and EMT predicted by the model for patients with pancreatic or ovarian cancer.
 
@@
     \newline
-An in situ tumor relies on mutations to alter cellular signaling pathways that will ultimately allow the tumor to become invasive.
-The immune system responds to the tumor upon recognition of the neoantigens and infiltrates the tumor microenvironment (TME).
-This environment is under the influence of inflammatory signals that vary over time.
-%
-%    At the same time, the immune system is recognizing and adapting to the resulting neoantigens to suppress the tumor.
-%    All of this occurs within an environment that is often under the influence of inflammatory signals that can vary over time.
+    To become invasive, An in situ tumor relies on mutations to alter cellular signaling pathways that enable cancer progression.
+The immune system simultaneously responds to the tumor upon recognition of neoantigens, and shapes the TME through dynamic inflammatory and regulatory signals.
 \newline
-    To capture these dynamics, we built a non-spatial agent-based model where the tumor cells are the agents and immune cells are present as homogeneous populations.
-    The model updates at discrete time intervals approximating the duration of a cell cycle.
-    During each cycle, the tumor cells are allowed one of four options: proliferation, apoptosis, immune clearance, and rest.
-    The probability of proliferation decreases with increasing tumor cell population, corresponding to competition for resources, and is also modified by cell intrinsic factors.
-    For apoptosis, the probability depends only on the EMT state of the tumor cell.
-    The probability of immune clearance relies on the number of mutated cells, cytotoxic cells, and regulatory cells as well as cell-intrinsic factors.
+% \hspace{50mm}
+To capture these dynamics, we developed a non-spatial agent-based model.
+Tumor cells are modeled individually as agents and immune cell populations are described homogeneously by differential equations.
+We consider two tumor cell types: epithelial tumor cells (ETCs) and mesenchymal tumor cells (MTCs).
+Time is treated discretely in 18 hour steps; approximately the time of one cell cycle.
+During each cell cycle, tumor cells can either undergo division, apoptosis, immune clearance, or can rest.
+The likelihood that a cell will proliferate depends on the tumor size (competition for resources) and on cell-intrinsic factors.
+The likelihood that a cell will undergo apoptosis is constant but varies between cell types (ETC and MTC).
+The likelihood of immune clearance depends on the number and type of mutated cells in the tumor, cytotoxicity, regulatory cells, and cell-intrinsic factors.
 \newline
-    We assume that as the in situ (and later invasive) tumor cells proliferate, they become increasingly likely to acquire pathway mutations that can alter their death or proliferation propensities.
-Natural killer (NK) cells in the model can identify and clear these tumor cells, a process which results in neoantigens being used to prime and activate T cells in local lymph nodes and subsequently infiltrate the tumor microenvironment.
-At the tumor site, T cell populations can either lyse tumor cells (cytotoxic T lymphocytes; CTLs) or suppress cytotoxic activity (T regulatory cells; Tregs).
-    We use the following expression to determine the probability that during one time step of the model a tumor cell will be lysed by a CTL:
+As tumor cells proliferate, DNA damage can occur, and over time they become increasingly likely to acquire pathway mutations that change their propensities for proliferation or cell death.
+Natural killer (NK) cells identify and clear tumor cells, a process which results in neoantigens  priming and activating T cells in local lymph nodes.
+T cells can subsequently infiltrate the TME.
+At the tumor site, cytotoxic T cells (CTLs) lyse tumor cells, and T regulatory cells (Tregs) suppress cytotoxic activity.
+The following equation determine the per-cell cycle probability that a tumor cell will be lysed by a CTL:
     $$\rho_{\text{CTL}} =\delta_{\text{MUT}} \frac{N_{\text{CTL}}}{N_C/K_{1}+N_{\text{CTL}}}  \frac{E_{\text{CTL}}}{1+N_{\text{Treg}}/K_2} (1-\delta_{\text{IE}}\Delta_{\text{IE}})(1-\zeta \Delta_{\text{MIE}})$$
-    Here, $\delta_\text{MUT}$ is 1 or 0 depending on whether the cell has a driver mutation or not, respectively.
-    The second term is a hill function similar to the dePillis-Radunskaya law and it assumes that the immune cells are well-mixed within the TME.
-    The third term is the regulatory effect of Tregs.
-    The final two terms are increased immune evasiveness due to a driver mutation in the immune evasion pathway and the immune evasiveness of mesenchymal cells (discussed below), respectively.
-    The inflammatory state of the tumor microenvironment affects the speed at which immune cells are recruited to the tumor.
+    Here, $\delta_\text{MUT}$ is 1 or 0 depending on whether or not the cell has a mutation.
+The second term is a hill function modeled after \cite{de2014modeling} and the third term describes onco-protective effects of Tregs.
+The second-to-last term describes the increased immune-evasiveness that can occur following mutation in the immune evasion pathway, and the last term quantifies the additional immune evasiveness of associated with MTCs.
+The inflammatory state of the TME thus impacts (through multiple factors) immune recruitment and cytotoxicity at the tumor site.
 \newline
-    Adding to the complexity of this situation, tumor cells can express heterogeneous cell types by their location along the epithelial-mesenchymal spectrum.
-    Depending on which end of this spectrum a tumor cell is on, its proliferation and its evasion of the immune system will change.
-    Specifically, if a cell is above a certain threshold, it will have its proliferation reduced by a fixed proportion and its immune evasiveness increased by a fixed proportion.
-    These cells also demonstrate cellular plasticity by being able to undergo epithelial-to-mesenchymal transition (EMT) and move along this spectrum as a result of intrinsic and extrinsic factors.
-    Here, we choose TGF-$\beta$ as a master regulator of EMT and connect the well-mixed TGF-$\beta$ level in the TME with the propensity of tumor cells to undergo EMT.
-    As TGF-$\beta$ is produced by Tregs as well as by tumor cells, this creates an intriguing set of interlocking interactions that we show here to be capable of shaping tumor and patient outcomes.
-    To update the EMT state of a cell, we determine how much TGF-$\beta$ is present in the TME--based on the number of malignant cells and Tregs--and apply a Hill function to determine whether a cell undergoes EMT or MET (the reverse process).
-    For each cell, this quantity takes the following form:
+EMT impacts tumor cells through their proliferation and potential to evade the immune system change. MTCs have reduced proliferation %(by $\Delta_{MGA}$)
+and increased immune evasiveness %(by $\Delta_{MIE}$).
+%Cellular plasticity permits tumors cells by being able to undergo epithelial-to-mesenchymal transition (EMT) and move along this spectrum as a result of intrinsic and extrinsic factors.
+%TGF-$\beta$, an activator of EMT, connects the tumor-immune model with EMT via its concentration in the TME.
+TGF-$\beta$, an activator of EMT, is produced both by Tregs and tumor cells, thus connecting tumor-immune interactions with EMT, and as a result plays an important role in shaping tumor outcomes.
+EMT depends on TGF-$\beta$ concentration through a Hill-type function that determines the likelihood of EMT (or the reverse MET) through its magnitude relative to $\tau_{i-1}$.
+This is described by the equation:
     $$
@@
     $$
-    Here, $\tau$ is the concentration of TGF-$\beta$ in the TME and $\tau_\text{max}$ represents a limit on the amount of TGF-$\beta$ that can be absorbed by all cells.
-    A Gaussian noise term is added with standard deviation $\sigma$.
-    When this quantity is large enough relative to the current EMT score of the cell, the cell will transition towards the mesenchymal phenotype.
-    Otherwise, it transitions towards the epithelial phenotype.
+    Here, $\tau$ is the concentration of TGF-$\beta$ in the TME and $\tau_\text{max}$ represents a limit on the amount of TGF-$\beta$ that can be absorbed by all cells, with Gaussian noise added.
     \newline
-    What we desire to know is when do the invasive tumor cells reach a particular threshold proportion of all tumor cells.
-    Once the tumor reaches this threshold, we declare that the tumor has progressed and we note how long the process took to reach this stage.
+    Our goal is to determine when the cancer becomes invasive, determined through the proportion of tumor cells harboring mutations in pathways that permit escape from in situ, relative to the total tumor cell population.
   \end{framed}
@@
 \section{Methods}
-We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
+We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on the cell-cycle and tissue-cell components described in \cite{guo17_multiscale}. 
 The agents in the model are the cells that have already formed an in situ tumor yet lack key pathway mutations to become invasive.
-In the process of the simulation, these cells can acquire mutations altering any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
+In the process of the simulation, these cells can acquire mutations altering any of three key pathways (Fig. \ref{fig:ModelIntro}A).
 If the pathway is altered, the component is colored red and is marked with `X'.
@@
 \par
-We model immune cells as continuous variables, effectively assuming that the tumor microenvironment is well-mixed with regards to the infiltrating immune cells.
+We model immune cells as continuous variables, i.e. we assume that the tumor microenvironment is well-mixed with regards to the infiltrating immune cells.
 The cytokine TGF-$\beta$ is also assumed to be well-mixed in the tumor microenvironment.
 Tumor cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
-While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
-In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the shade of blue of the cell.
-Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.
+While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal (Fig. \ref{fig:ModelIntro}A).
+%Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.
 
-\subsection{Tumor Cells}\label{TissueCells}
-The tumor cells at all times have associated with them two important quantities: which combination of driver mutations they carry and an EMT score.
-The model has three pathways that have the potential to become altered over the course of a simulation.
-There is a proliferation pathway that when altered increases the probability of the cell proliferating.
-There is an apoptosis pathway that when altered decreases the probability of the cell undergoing apoptosis.
-There is an immune evasion pathway that when altered decreases the probability that an immune cell can effectively clear the mutated cell.
-The EMT score will be exposited in Section \ref{EMT}.
+\subsection{Tumor Evolution}\label{TissueCells}
+Associated with each tumor cells are two important quantities: their particular mutational signature and their EMT score.
+We consider three idealized pathways that can be mutated: a proliferation pathway that when altered increases the probability of the cell proliferating; an apoptosis pathway that when altered decreases the probability of the cell undergoing apoptosis; and an immune evasion pathway that when altered decreases the probability that an immune cell can effectively clear the mutated cell.
+%The EMT score will be exposited in Section \ref{EMT}.
 
-\subsection{The Immune System}\label{ImmuneSystem}
-The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
+\subsection{Immune Cell Population Dynamics}\label{ImmuneSystem}
+The immune system is modeled by three immune cell types: NKs, CTLs, and Tregs.
 The NKs and CTLs act on the system by recognizing invasive cells and clearing them.
-Upon successful clearance, they themselves are then deactivated and removed from the immune population.
-Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear malignant cells.
-In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.
+Upon clearance, they are deactivated and removed from the immune population.
+Tregs suppress the function of NKs and CTLs (reduce tumor cell clearance), and in addition, release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.
 
 \subsection{Inflammation} 
-We also include an inflammation score that follows a fixed cycling scheme between low and high inflammatory states.
-This is to simulate the various changes to the TME that can happen over long periods of time (days and weeks).
-For the purpose of the model, we consider low inflammation to be the default state.
-When the inflammation switches to high, then a number of the parameters change values to reflect the additional immune activity.
-See Table S2 in the supplement to see which parameters are changed by the inflammation score changing.
+Inflammation is modeled as a cycling scheme between low and high inflammatory states, with varying on/off durations and intensities.
+%This is to simulate the various changes to the TME that can happen over long periods of time (days and weeks).
+For the purpose of simulation we consider the default state to be low inflammation, and update the immune activity parameters whenever a switch to the high state occurs.
+%When the inflammation switches to high, then a number of the parameters change values to reflect the additional immune activity.
+In the SI (Table S2) we give full details of parameter settings during low and high inflammation.
 
-\subsection{EMT}\label{EMT}
-Each tissue cell has an EMT score between 0 and 1.
-When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
-Otherwise, it is below the threshold and labeled as epithelial.
-For the purpose of parameter values of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have a subset of its parameters updated.
-In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
-Note, that because the immune system can only clear malignant cells, the increased immune evasion only benefits malignant mesenchymal cells.
-\par
+\subsection{Epithelial-to-mesenchymal Transition}\label{EMT}
+Each tissue cell has an EMT score between 0 and 1 with is set according to the concentration of TGF-$\beta$ in the TME.
+Above a threshold, the cell acquires the phenotype of a mesenchymal tumor cell (MTC); otherwise, it is an epithelial tumor cell (ETC).
+For the purpose of simulation, ETCs are considered to be in the base state, and MTCs will have a subset of their parameters updated.
+%MTCs notable have increased rates of immune evasion and decreased rates of proliferation.
+% Note, that because the immune system can only clear malignant cells, the increased immune evasion only benefits malignant mesenchymal cells.
 In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
@@
 %In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
-In addition, we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
+%In addition, we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
+
 \par
-For the cells that have undergone EMT and are currently labeled in the model as mesenchymal, they experience two modifications to their governing parameters.
-The first is a reduction in proliferation, which we call mesenchymal growth arrest (MGA).
-This causes to mesenchymal cells to be less likely to undergo proliferation and more likely to rest during a cell cycle.
-The second is a decrease to their likelihood of being cleared by the immune system, which we call mesenchymal immune evasion (MIE). 
-This decreases the likelihood of mesenchymal cells being cleared by both NKs and CTLs during a cell cycle.
-Though we have not found clinical data on which to base parameter values for either of these, by their very nature they are proportions on the interval $[0,1]$.
-Thus, it is possible to sample from the entire parameter and in fact we do present results from a large subset of this parameter space and our sensitivity analysis fully explores this space.
+Cells that have undergone EMT (i.e.
+MTCs) experience a reduction in proliferation, referred to as mesenchymal growth arrest (MGA), and a decrease in the likelihood that they will be cleared by immune cells (NKs or CTLs), referred to as mesenchymal immune evasion (MIE).
+Both these parameters lie within the range $[0,1]$, thus we can sufficiently sample from their joint parameter space to explore it in depth with the need for informative priors to constrain their values.
 
 \subsection{Model Simulation}
-
 \subsubsection{Initial conditions} 
-Every model run starts with a number of in situ tumor cells, $N_0$, determined by the set of parameters used for the run.
-A set of warmup cycles are run to ensure arrival to this steady state.
-During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.
-After the warmup cycles are complete, a mutagenic event is simulated in which all the tumor cells experience a random increase in their probability to mutate.
-From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.
+Every model run starts with a number of in situ tumor cells, $N_0$, determined by the choice of parameter values.
+A set of warmup cycles are run to let the model reach steady state.
+During these warmup cycles, no mutations are allowed to occur, and the only immune cells present are NKs.
+After  warmup cycles, mutations are permitted. Cells that do not mutate undergo an increase in their probability of mutation in a later cell cycle.
+% a mutagenic event is simulated in which all the tumor cells experience a random increase in their probability to mutate.
+%From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.
 
@@
 A.L.M.
-would like to acknowledge partial support for this work from the American Cancer Society grant \#IRG-16-181-57.
+would like to acknowledge partial support for this work from an American Cancer Society grant \#IRG-16-181-57.
 

### 2019-12-09 Paper/CleanPaper.tex
update results

@@
 \section{Methods}
+Here we briefly describe to core components of the model.
+Full details and equations are provided in the Supplement.
 We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on the cell-cycle and tissue-cell components described in \cite{guo17_multiscale}.
@@
 In the process of the simulation, these cells can acquire mutations altering any of three key pathways (Fig. \ref{fig:ModelIntro}A).
-If the pathway is altered, the component is colored red and is marked with `X'.
+%If the pathway is altered, the component is colored red and is marked with `X'.
 %We distinguish between cells that have zero driver mutations and those with at least one as premalignant and malignant, respectively.
@@
 Tumor cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
-While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal (Fig. \ref{fig:ModelIntro}A).
+While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal (Fig. \ref{fig:ModelIntro}).
 %Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.
@@
 \subsection{Tumor Evolution}\label{TissueCells}
-Associated with each tumor cells are two important quantities: their particular mutational signature and their EMT score.
-We consider three idealized pathways that can be mutated: a proliferation pathway that when altered increases the probability of the cell proliferating; an apoptosis pathway that when altered decreases the probability of the cell undergoing apoptosis; and an immune evasion pathway that when altered decreases the probability that an immune cell can effectively clear the mutated cell.
+Associated with each tumor cells are two essential features: their mutational signature and their EMT score.
+We consider three idealized pathways that can be mutated: proliferation, when altered this increases the probability of the cell proliferating within each cell cycle; apoptosis, when altered this decreases the probability of a cell undergoing apoptosis; and immune evasion, when altered this decreases the probability that a mutated cell will be cleared by immune components.
 %The EMT score will be exposited in Section \ref{EMT}.
@@
 
-\subsection{Inflammation} 
+\subsection{Periodic Inflammation} 
 Inflammation is modeled as a cycling scheme between low and high inflammatory states, with varying on/off durations and intensities.
@@
 \subsection{Model Simulation}
-\subsubsection{Initial conditions} 
-Every model run starts with a number of in situ tumor cells, $N_0$, determined by the choice of parameter values.
-A set of warmup cycles are run to let the model reach steady state.
-During these warmup cycles, no mutations are allowed to occur, and the only immune cells present are NKs.
-After  warmup cycles, mutations are permitted. Cells that do not mutate undergo an increase in their probability of mutation in a later cell cycle.
+%\subsubsection{Initial conditions} 
+\paragraph{Initial conditions.}
+Simulations are initialized with $N_0$ in situ tumor cells, determined by the choice of parameter values. A number of warmup cycles are run so that the model reaches steady state.
+During warmup, no mutations occur, and the only immune cells present are NKs.
+After warmup, mutations are permitted. Cells that do not mutate undergo an increase in their probability of mutation in a later cell cycle.
 % a mutagenic event is simulated in which all the tumor cells experience a random increase in their probability to mutate.
@@
 
-\subsubsection{Tumor cell fate}
-During each cell cycle, every cell is randomly assigned one of four fates: proliferation, apoptosis, immune clearance, and rest in $G_0$.
-Depending on both cell intrinsic, e.g.
+%\subsubsection{Tumor cell fate}
+\paragraph{Tumor cell fate.}
+During each cell cycle, the fate of each cell is assigned: proliferation, apoptosis, immune clearance, and rest in $G_0$, according the model rules.
+The probability of proliferation is affected by mutations to the proliferation pathway (increased) and my cells in a mesenchymal state (decreased).
+Probabilities of immune clearance are affected by the number of mutations harbored: cells with more mutations are assumed to be more immunogenic and have a higher probability of being cleared by the immune system, unless the cell has a mutation in the immune evasion pathway.
+Cells in a mesenchymal state can exhibit greater capacity to evade immune clearance.
+
+%Depending on both cell intrinsic, e.g.
 EMT and mutational state, and cell extrinsic, e.g.
 immune infiltrate and total number of tumor cells, each of the four options is given a probability of occurring for each cell.
-For proliferation, we assume the total number of tumor cells exerts a negative feedback on the proliferation probability for each cell.
-Also, cells that have a driver mutation in the proliferation pathway have an increased probability of proliferating.
-Meanwhile, mesenchymal cells have a decreased probability of proliferating and an increased probability of resting.
-For apoptosis, the probability is constant except for those cells with a driver mutation in the apoptosis pathway which have a decreased probability of apoptosis.
+%For proliferation, we assume the total number of tumor cells exerts a negative feedback on the proliferation probability for each cell.
+%Also, cells that have a driver mutation in the proliferation pathway have an increased probability of proliferating.Meanwhile, mesenchymal cells have a decreased probability of proliferating and an increased probability of resting.
+%For apoptosis, the probability is constant except for those cells with a driver mutation in the apoptosis pathway which have a decreased probability of apoptosis.
 %For immune clearance, we assume that only cells with newly acquired driver mutations are targeted by the immune system, and their probability of being cleared depends on the immune infiltrate.
-For immune clearance, the malignant tumor cells are assumed more immunogenic and thus have a higher probability of being cleared by the immune system.
-This probability can be decreased by a cell having a driver mutation in the immune evasion pathway or by a cell being mesenchymal.
-Further details and equations are given in the supplement.
+%For immune clearance, the malignant tumor cells are assumed more immunogenic and thus have a higher probability of being cleared by the immune system.
+%This probability can be decreased by a cell having a driver mutation in the immune evasion pathway or by a cell being mesenchymal.
 
-\subsubsection{Completing the Cell Cycle}
-Once all the cell fates are carried out, the rest of the components of the model update to account for these changes.
-First, the immune populations are updated in two steps.
-In step one, immune cells are exhausted based on the number of malignant cells that were cleared by their type.
-So, each clearance by an NK cell results in the NK population decreasing by one, and similarly for CTLs.
-In step two, all three immune populations are updated deterministically based on a simple ODE model of their dynamics.
-NKs have a constant source and death rate while CTLs and Tregs have a source rate that is dependent on the number of malignant cells cleared that cycle and a constant death rate.
-In addition, the concentration of TGF-$\beta$ mediates recruitment of Tregs with higher concentrations of TGF-$\beta$ resulting in a greater influx of Tregs.
+%\subsubsection{Completing the Cell Cycle}
+\paragraph{Completing the Cell Cycle.}
+Once all tumor cells have been updated and fates chosen accordingly, non-tumor model components are updated.
+Immune cell populations are updated in two steps.
+First, immune cell exhaustion is calculated based on the number of tumor cells cleared, e.g.
+clearance of one tumor cell by an NK cells results in the NK cell population decreasing by one.
+Second, all immune cells (NKs, CTLs, Tregs) are updated according to a system of coupled ordinary differential equations that govern their population dynamics.
+CTL and Treg recruitment rates are dependent on the number of tumor cells; in addition  TGF-$\beta$ enhances the recruitment rate of Tregs.
 \par
-Also at this time, new driver mutations have a chance to occur.
-The cells that proliferated are the only ones that have a probability of acquiring a new driver mutation.
-This probability is cell-specific and is increased if no mutation occurs and reset to 0 if a mutation does occur.
-\par
-Finally, the concentration of TGF-$\beta$ is updated as well as the EMT score for each cell is updated.
-Malignant cells and Tregs are the two sources of TGF-$\beta$ each producing a constant amount.
-Tregs are assumed to produce much more TGF-$\beta$ than malignant cells.
-The resulting quantity of TGF-$\beta$ is divided up randomly among the tumor cells.
-This quantity is combined with the EMT score of each cell and the resulting quantity determines which cells undergo EMT and which undergo MET.
-The equations governing these processes can be found in the supplement.
+At the end of each cell cycle, new mutations can occur in cells that have undergone division, according to cell-specific probabilities that increase if no mutation occurs are reset to 0 in the event of a mutation.
+Finally, the concentration of TGF-$\beta$ and the EMT score for each cell are updated.
+Tregs and (to a lesser extent) invasive tumor cells are the sources of TGF-$\beta$; the total concentration per cell cycle is divided randomly among tumor cells.
+EMT is then assessed, depending on the EMT score of the cell and the local concentration of TGF-$\beta$.
+
+%\subsubsection{Mutation and progression}
+\paragraph{Mutational burden and progression to invasive disease.}
+At the end of each cell cycle, the proportion of tumor cells that are invasive is calculated based on their mutational burden, and if it is above a certain threshold, the tumor is declared to have progressed to an invasive state and the simulation ends.
+The time to invasion is calculated as the time from the start of the simulation, minus the warmup period, until the invasive state is reached.
+Simulations run until either this occurs or until the maximum number of cell cycles has been reached.
 
@@
 \begin{center}
- \begin{tabular}{|| c | l ||} 
+ \begin{tabular}{| c | l |} 
  \hline
@@
 \end{tabular}
-  \caption{The model parameter names and descriptions. Note that some of these values are affected by the inflammation state of the system.}
+  \caption{Description of key model parameters. Note that some are not constant as they can be affected by the inflammation state of the system.}
 \end{center}
@@
 
-\subsubsection{Mutation and progression}
-At the end of each cell cycle, the proportion of tumor cells that are invasive is calculated, and if it is above a certain threshold, the simulation ends and we declare the tumor to have progressed.
-The Time to Cancer is given as the time from the start of the simulation, not including warmup, until this timepoint.
-Simulations run until either a this occurs or until the predetermined maximum number of cycles has been reached.
-
-\subsection{Parameter Sensitivity Analysis}
-To study parameter sensitivity, we implemented the Morris method, a one-step-at-a-time global sensitivity analysis algorithm.
-The algorithm runs the model numerous times by picking several base points and then varying one parameter at a time from each base point and tracking the difference\cite{morris1991factorial, sohier2014improvement}.
-Each model run involved 1000 patients and the value returned was the area under the survival curve.
-We ran the algorithm by sampling at 30 points in parameter space and then changing each parameter one time at each point.
-In the implementation by Sohier, they recommend using at least 10 points\cite{sohier2014improvement}.
-\par
-To determine which points in parameter space to sample at, we had to create distributions on all parameter values.
-To do this, we derived many parameters from existing literature and the rest we made reasoned predictions.
-For example, many previous studies have established the parameters of immune population dynamics such as recruitment and death rates\cite{de2014modeling}.
-Because our model only works with on the order of $10^2$ tumor cells whereas an actual tumor would have on the order of $10^9$ cells\cite{de2014modeling}, we scaled the parameters so the immune populations would also be proportionately reduced.
-That is, for instance, how we ended up with NK cells having a steady state population of 10 cells in the model.
-\par
-Once we had these parameter values, we created Gaussian distributions centered at these values with large variances and truncated at 0.
-Some of the parameters were also truncated to the right because they represented a proportion, such as MIE and MGA, or because we simply wanted to keep the values within a reasonable range, such as the inflammation low and high duration parameters.
-The standard deviations were chosen as twice the mean to get a large spread of values.
-\par
-The algorithm computes the average of the absolute change of the output, which in our model is the area under the survival curve.
-This output of the algorithm, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.
+\subsection{Parameter Choice and Sensitivity Analysis}
+To study parameter sensitivity, we implemented Morris a one-step-at-a-time global sensitivity analysis.
+Parameters are varied one at a time from a set of sampled ``base'' points and the resulting simulations recorded \cite{morris1991factorial, sohier2014improvement}.
+For each run we simulated 1000 patients, and initialized the Morris sampling with 30 points in parameter space (at least 10 are recommended in \cite{sohier2014improvement}.
+Parameter sampling a choice of prior parameter distributions.
+For many parameters, such as for the immune population dynamics,  measurements or estimates were available from literature \cite{de2014modeling}.
+For parameters such as MIE and MGA related to the mesenchymal phenotype, little prior information was available, thus these was sampled across all possible values in $[0,1]$.
+Tumor size in the model was scaled from cell numbers on the order of $10^9$ cells\cite{de2014modeling} to the order of $10^2$, and parameter values were scaled accordingly.
+Where parameter estimates existed, the prior for parameter $\theta_i$ is given as $\theta_i \sim \textrm{N}(m_e, 2m_e)$, where $m_e$ is the previous estimate and we take twice this value as the variance to obtain a range of samples that does not rely too heavily on previous work.
+The Morris algorithm computes the sensitivity, $\mu^*$, as the average of the absolute change of the output, which in our model is the area under the survival curve (Fig.
+\ref{fig:MOAT}).
 
@@
 \end{enumerate}
-Below we analyze data gathered for pancreatic cancer, denoted by ``PAAD'' in TCGA.
+We apply this pipeline to several inflammation-associated cancers, denoted according to their annotations in TCGA, including ``PAAD'' for pancreatic cancer, ``OV'' for ovarian cancer, and ``LIHC'' for liver hepatocellular carcinoma.
 
@@
 {\includegraphics[width=0.9\textwidth]{Figure1/Figure1.pdf}}
-\caption{A.
-Schematic depiction of agent-based model components; each of the 10 columns represents a single tumor cell divided into three compartments representing the state (altered or not) of the three pathways with mutagenic potential; red/blue denotes altered/unaltered pathways.
+\caption{{\bf A.} Schematic depiction of agent-based model components; each of the 10 columns represents a single tumor cell divided into three compartments representing the state (altered or not) of the three pathways with mutagenic potential; red/blue denotes altered/unaltered pathways.
 Black arrows depict regulation of the cell fate in each cell cycle.
 Inset depicts major interactions between the immune system and tumor cells.
-B. A representative simulation of one patient. The model parameter values used can be found in Table S2. 
+{\bf B.} A representative simulation of one patient. The model parameter values used can be found in Table S2. 
 The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: malignant cells; Mes: mesenchymal cells. 
-C. Survival curve for one cohort of patients with the parameter values given in Table S2.}
+{\bf C.} Survival curve for one cohort of patients with the parameter values given in Table S2.}
 \label{fig:ModelIntro}
@@
 
-We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the probability of progression, and the Time to Cancer.
-Within the cell cycle, cell fate is determined via a set of rules that are influenced by EMT and immune interactions (Fig.
-\ref{fig:ModelIntro}A).
-For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a driver mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced.
-The different means by which the immune system acts on tumor cells are also shown (\ref{fig:ModelIntro}A Inset).
-NK cells and CTLs attempt to clear malignant tumor cells, and deactivate upon successfully carrying out their cytotoxic function.
-Tregs inhibit cytotoxic activity.
-Tregs also release TGF-$\beta$ which promotes the recruitment of Tregs, as well as increasing the rate of EMT.
+We begin by investigating general features of the model to establish baseline conditions and assess the impact of different model components on the key measured outcomes: the probability of progression, and the time to invasion.
+Within the cell cycle, cell fate is determined by rules that are influenced by EMT and immune interactions (Fig.
+\ref{fig:ModelIntro}A), e.g.
+if a cell undergoes EMT, its probability of proliferation is reduced; if it gains a  mutation in the apoptosis pathway, its probability of apoptosis is reduced.
+Meanwhile, NK cells and CTLs attempt to clear malignant tumor cells, and deactivate upon successful tumor cell clearance; Tregs inhibit this cytotoxic activity (\ref{fig:ModelIntro}A Inset).
+%Tregs also release TGF-$\beta$ which promotes the recruitment of Tregs, as well as increasing the rate of EMT.
 \par
-The inflammation cycling scheme for a typical {\it in silico} patient consists of alternating high and low regimes (Fig.
-\ref{fig:ModelIntro}B); the inflammation schemes modeled will be discussed in detail below.
-For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor.
-After about 700 cell cycles, the malignant cell population begins to grow steadily.
-At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being quickly recruited into the TME, and a peak in TGF-$\beta$ expression is observed.
-After 841 cell cycles, the proportion of malignant cells reaches 50\%: the threshold defining progression.
-Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days.
-For this patient, we let the simulation go beyond this timepoint, and we see that the number of malignant cells grows rapidly and soon makes up 100\% of the cell population.
-A peak in number of mesenchymal cells is observed shortly after this threshold is reached, and following this most cells transition back to epithelial.
-\par 
-Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of driver mutations.
-The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 30 cell cycles of low inflammation.
+The inflammation cycling scheme for a typical {\it in silico} patient consists of alternating high and low regimes with corresponding effects on the cell populations (Fig. \ref{fig:ModelIntro}B).
+For this patient, after warmup, mutations are observed at a rate low enough that they are cleared by cytotoxic cells for about 700 cell cycles, after which the mutated and thus invasive cell population begins to grow, leading to large recruitment of CTLs and Tregs and a peak in the concentration of TGF-$\beta$.
+After 841 cell cycles, the proportion of invasive cells reaches 50\%: the threshold defining progression, thus this patient has a time to invasion of 841 cell cycles, or 631 days.
+Beyond this timepoint, we see a rapid increase in the number of invasive cells until it comprises 100\% of the tumor population.
+Interesting EMT dynamics are also observed, the proportion of MTCs peaks shortly after the tumor becomes invasive, subsequently the majority of cells transition back to an epithelial state.
+We observe that while the NK population varies little over the simulation, CTLs and Tregs both undergo large expansions.
+CTLs and Tregs also appear to oscillate, however note that this is a direct result of the inflammation state, and is not immune cell-intrinsic.
+
+%Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of driver mutations.
+%The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 30 cell cycles of low inflammation.
 Within each of these periods, adaptive immune populations increase or decrease rapidly in accordance with the inflammation state.
 \par 
-The {\em in silico} patient described here is given for the purpose of illustrating features of the model.
-Given the multiple sources of noise in the model, in order to quantify patient dynamics and cancer-free survival rates, below we will simulate large cohorts of patients.
-In Fig.
-\ref{fig:ModelIntro}C we simulate the survival curve for a cohort of $500$ patients:
+In order to quantify patient dynamics and invasion-free survival as a population level, we simulate large cohorts of patients similar to the single patient shown in  Fig.
+\ref{fig:ModelIntro}B.
+For a cohort of $500$ patients, we simulate survival curves and see that a large number progress quickly to form invasive tumors, whereas a few lie in the tail of the distribution after the mutagenic event that a large number of tumors quickly progress while others takes some time before progressing Fig.
+\ref{fig:ModelIntro}C.
+By approximately 1200 cell cycles (2.5 years), all tumors have become invasive..
+
 %we see that all the patients survive for approximately 100 cell cycles (75 days).
 Subsequently, in the approximate range of $T= [100,300]$, the cancer onset rate is roughly constant, and after $600$ cell cycles, no patients remain cancer free.
-we see that after the mutagenic event that a large number of tumors quickly progress while others takes some time before progressing. By about 1200 cell cycles, all tumors have progressed.
 
-\subsection{Identifying regulatory parameters via Morris global sensitivity analysis}\label{SensAnalysis}
-Exploring the parameter spaces of models in systems biology is -- in general -- a hard problem.
-Performing Bayesian parameter inference to inform parameter values is advisable wherever possible \cite{kirk13_model}.
-Here, a lack of detailed molecular measurements  (i.e.
-data for tumor growth dynamics abound, but simultaneous data on the immune dynamics are lacking) preclude inference of the full model.
+\subsection{Identification of key model parameters via global sensitivity analysis}\label{SensAnalysis}
+Exploring the parameter spaces of systems biology models {\em adequately} is -- in general -- a hard problem.
+Fitting parameters via (Bayesian) parameter inference is advisable wherever possible \cite{kirk13_model}.
+Here, despite a wealth of data on tumor growth dynamics, a lack of sufficient molecular measurements (i.e.
+immune cell dynamics) precludes inference of the full model.
 In addition, while inference schemes for agent-based models are developing \cite{gallaher17_hybrid, warne19_simulation}, simulation times remain a hurdle \cite{lambert18_bayesian}.
-Parameters for aspects of this model that were studied previously can be constrained \cite{guo17_multiscale}, however even for these, the new additions to the model could push it into new behavioral regimes.
-Thus to adequately sample the parameter space of the model and identify those areas of parameter space that exert the most control, we use sensitivity analysis.
+Parameters for some components of the model studied previously can be constrained \cite{guo17_multiscale}.
+However, even here, new biological processes in the current system could push the model into new behavioral regimes.
+Thus to sample and characterize the parameter space of the model we use sensitivity analysis.
 \par
-To assess the sensitivity of the model parameters and thus identify those that are most important in determining progression-free survival times, we performed Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods).
-The results of Morris OAT on the 31 model parameters are shown in Figure \ref{fig:MOAT}. We see that a subset of parameters demonstrate much higher levels of sensitivity than others.
+The results of Morris one-step-at-a-time sensitivity analysis on the 31 model parameters (Fig.
+\ref{fig:MOAT}) find a subset of parameters with much higher levels of sensitivity than others.
+The two most influential by this analysis are the recruitment rates of Tregs and CTLs in the low inflammation state.
+The parameters influencing EMT are also identified as influencing model outcomes.
+Since one goal of our analysis is to assess the specific effects of EMT on immune-cancer dynamics, parameters MIE and MGA are of particular interest.
+In addition, inflammation parameters controlling the periodic high/low inflammation states are of interest because they strongly influence model outcomes and are capable of being targeted by therapeutic treatments.
+For immune cell dynamics, the secretion of TGF-$\beta$ by Tregs is found to be sensitive and thus will also be studied further below.
+
+%To assess the sensitivity of the model parameters and thus identify those that are most important in determining progression-free survival times, we performed Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods).
 %The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
-The two most influential according to this analysis are the recruitment rates of Tregs and CTLs under low inflammation.
 %This ties Treg cells to all three components of the model.
 %Since we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for detailed analysis below.
-We also note that the parameters related to EMT have a moderate effect on Time to Cancer.
-\par 
-Many parameters in the model change depending on the inflammation state.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%% PUT THIS SOMEWHERE ELSE  - Supplemental methods ? 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%Many parameters in the model change depending on the inflammation state.
 For the sake of a naming convention, any parameter which includes ``low'' in the name represents the parameter value when inflammation is low.
 For when the inflammation is high, these ``low'' parameters get scaled by parameters which include ``up'' in the name.
-The ILD and IHD parameters determine the duration of low and high inflammation, respectively.
+%The ILD and IHD parameters determine the duration of low and high inflammation, respectively.
 A cell with altered pathways will be affected by the $\Delta_A$, $\Delta_P$, and $\Delta_{\text{IE}}$ parameters, depending on whether the apoptosis, proliferation, or immune evasion pathways are altered.
-The mesenchymal cells are influenced by the MGA and MIE parameters.
-Finally, $\sigma$, $k_\text{EMT}$, $\tau_\text{max}$, and $K_3$ all control how cells transition between the epithelial and mesenchymal states.
-
-Since one goal of our analysis is to assess the specific effects of EMT on immune-cancer dynamics, the parameters MIE and MGA are of particular interest.
-In addition, inflammation parameters dictating the cycling scheme are of interest because they are both highly influential on Time to Cancer and capable of being targeted by therapeutic treatments.
-In terms of Treg cells, their secretion of TGF-$\beta$ is highly significant and we will also study it further.
+%The mesenchymal cells are influenced by the MGA and MIE parameters.
+%Finally, $\sigma$, $k_\text{EMT}$, $\tau_\text{max}$, and $K_3$ all control how cells transition between the epithelial and mesenchymal states.
 
@@
 {\includegraphics[width=0.85\textwidth]{Figure2/MOAT.jpg}}
-\caption{Sensitivity analysis of model parameters, determined via the global Morris one-step-at-a-time method.
-$\mu^*$ denotes the average absolute change in Time to Cancer when the parameter is varied.}
+\caption{Global sensitivity analysis of model parameters. The sensitivity ($\mu^*$) denotes the average absolute change in the time to invasion over the range of variation of the parameter.}
 \label{fig:MOAT}
@@
 
-\subsection{Mesenchymal phenotypic properties dramatically alter invasion-free survival times}\label{MesPars}
+\subsection{Mesenchymal properties dramatically alter invasion-free survival times}\label{MesPars}
 
@@
 {\includegraphics[width=0.85\textwidth]{Figure3/Figure3.pdf}}
-\caption{Effects of mesenchymal tumor cell properties on the Time to Cancer.
-Trajectories of one patient per cohort from warmup period ([-1000,0]) to 2000 cell cycles, for $\Delta_\text{MIE}$ (A); $\Delta_\text{MGA}$ (B); and $\tau_\text{Treg}$ (C).
-D. Survival curves corresponding to changes in MIE (A) for a patient cohort of size 1000; shaded region represents the 95\% confidence interval for the evaluated function. 
-E. Survival curves for patient cohort corresponding to changes in MGA (B).
-F. Survival curves for patient cohort corresponding to changes in Treg TGF-$\beta$ (C). 
-}
+\caption{Effects of mesenchymal tumor cell properties on the time to invasion.
+Trajectories of one patient per cohort including warmup and 2000 cell cycles for {\bf A.} mesenchymal immune evasion (MIE); {\bf A.} mesenchymal growth arrest (MGA); {\bf C.} Production of TGF-$\beta$ by Tregs.
+{\bf D.} Survival curve corresponding to changes in the parameter MIE (A) for a patient cohort of 1000. Shaded region represents the 95\% confidence interval over the cohort. 
+{\bf E.} Survival curve corresponding to changes in MGA.
+{\bf F.} Survival curve corresponding to changes in Treg production of TGF-$\beta$. }
 \label{fig:FirstSurvivalCurves}
@@
 
-When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
-Both parameters are defined proportionally and hence lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
-The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that an invasive mesenchymal cell will be subject to immune clearance in a given cycle.
-The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
-The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT).
+Mesenchymal tumor cells (MTCs) are characterized by changes in two parameters: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
+Here we assess the effects of each, alongside the effects of TGF-$\beta$ through its production by Tregs.
+%Both parameters are defined proportionally and hence lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
+% $\Delta_\text{MIE}$ quantifies the reduction in probability that a mesenchymal cell will be subject to immune clearance.
+% $\Delta_\text{MGA}$ quantifies the reduction in probability that a mesenchymal cell will proliferate, and is equivalent (in terms of its impact on the cell cycle) to an increased time spent in $G_0$.
+%The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT).
 All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
+As MIE increases, the invasion-free survival decreases (Fig.
+\ref{fig:FirstSurvivalCurves}A) for all sets of parameters studied: as the subpopulation of invasive cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus can grow faster (Fig.
+\ref{fig:FirstSurvivalCurves}D).
 \par
-As MIE increases, the invasion-free survival decreases (Fig.
-\ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of invasive cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
-These results are summarized by Fig. \ref{fig:FirstSurvivalCurves}D.
-\par
-The relationship between MGA and invasion-free survival is non-monotonic.
-For smaller values of MGA, an increase in MGA results in increased invasion-free survival (Fig \ref{fig:FirstSurvivalCurves}E).
-This result is explored in Section \ref{KeyEMT}.
+The relationship between MGA and invasion-free survival times displays a very different trend, and is non-monotonic with a local maximum appearing.
+For small values of MGA, increasing the MGA parameter results in increasing the invasion-free survival (Fig \ref{fig:FirstSurvivalCurves}B, E).
+However for large values of MGA, invasion-free survival times decrease.
+This is explored further below.
+%This result is explored in Section \ref{KeyEMT}.
 %As MGA increases, invasion-free survival increases: lower proliferation rates for mesenchymal cells slow down cancer progression (Fig \ref{fig:FirstSurvivalCurves}B).
@@
 \par
-Third, TGF-$\beta$ can be varied in two ways: the production by malignant cells and the production by Treg cells.
-In Fig \ref{fig:FirstSurvivalCurves}F, the results of varying Treg TGF-$\beta$ production are shown, indicating that an increased Treg TGF-$\beta$ does not have much effect on invasion-free survival time.
-The effect we do see is that lower Treg TGF-$\beta$ results in more rapid invasion or if it survives to a certain time point, a longer time to invasion.
-Here, we choose to focus on the results of the effect of MIE and MGA as the more direct link to the mesenchymal phenotype.
+TGF-$\beta$ varies according to its production by tumor cells and its production by Tregs.
+Here we assess the effects of varying the production of  TGF-$\beta$ by Tregs on invasion-free survival (Fig \ref{fig:FirstSurvivalCurves}B, E).
+We find, interestingly, that at lower production rates of TGF-$\beta$ lead to more rapid invasion or if it survives to a certain time point, a longer time to invasion.
+
+%The effect we do see is that lower Treg TGF-$\beta$ results in more rapid invasion or if it survives to a certain time point, a longer time to invasion.
+%Here, we choose to focus on the results of the effect of MIE and MGA as the more direct link to the mesenchymal phenotype.
+
+\subsection{A key EMT regime maximizes cancer-free survival time under chronic inflammation}\label{KeyEMT}
 
@@
 {\includegraphics[width=0.95\textwidth]{Figure4/Figure4.pdf}}
-\caption{Effects of inflammation on the Time to Cancer under different cycling schemes.
-A-B.
-As MIE varies, survival curves (each of 200 patients)  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.
-C-D.
-As MGA varies, survival curves  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.}
+\caption{Effects of inflammation on the time to invasion under different cycling schemes.
+{\bf A-B.} As MIE varies, survival curves (each of 200 patients)  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.
+{\bf C-D.} As MGA varies, survival curves  and corresponding bar plots to summarize the mean Time to Cancer for each cohort are shown.}
 \label{fig:VaryINFL_and_MesPars}
@@
 
-\subsection{A key EMT regime maximizes cancer-free survival time under chronic inflammation}\label{KeyEMT}
-
-We explored the effects of varying the inflammation state of the patient on invasion-free survival, to investigate competing interactions within the tumor microenvironment and their effect on EMT.
-Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable inflammation.
-Compared to the other inflammation regimes, a permanently high inflammation state results in tumors that are relatively resistant to changes in mesenchymal parameters (Fig.
-\ref{fig:VaryINFL_and_MesPars}).
-However, when the inflammation either has temporary or permanent periods of low inflammation, different relationships emerge: invasion-free survival time becomes negatively correlated with MIE, and a local maximum for the invasion-free survival time is found with respect to MGA ($\Delta_\text{MGA}= 0.2$).
+To investigate how competing interactions within the inflammatory tumor microenvironment affect EMT, we explored the effects of varying inflammation on invasion-free survival.
+Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable (periodic high/low) inflammation.
+Compared to the other inflammation states, permanently high inflammation results in outcomes that vary more subtly with changes in the mesenchymal parameters (Fig. \ref{fig:VaryINFL_and_MesPars}).
+When the inflammation state is either permanently or temporarily low, surprising trends emerge.
+In both these cases, invasion-free survival time is negatively correlated with MIE, and a local maximum for the invasion-free survival time is found with respect to MGA (close to $\Delta_\text{MGA}= 0.2$).
 
@@
 \ref{fig:VaryINFL_and_MesPars}D).
+
 \par
-These striking differences in the mean invasion-free survival -- that can be extended by up to double compared to other MGA values-- have clear therapeutic implications. 
-Our model predicts that a patient suffering intermittent inflammatory attacks at the tumor site will benefit from controlling the proliferation of the mesenchymal cells.
+These differences in the mean invasion-free survival lead to striking variation in outcomes: tumors can be contained in situ for up to twice as long as they would otherwise be simply by varying the rates of mesenchymal growth arrest.
+These predictions point to intriguing therapeutic outcomes: a patient suffering intermittent high inflammatory attacks will benefit directly from EMT-directed therapies, however patients for whom a relatively high inflammation state is observed continuously will not obtain this benefit.
 \par
-In contrast, when MIE is varied for different inflammation cycling schemes, regardless of the cycling scheme, increasing the MIE will decreases the invasion-free survival (worsen cancer progression and prognosis) or have minimal effect.
-Thus under any inflammation regime with periods of low inflammation, reductions of any size in the rate of mesenchymal immune evasion lead to improved patient outcomes.
+When MIE is varied under different inflammation cycling schemes, for all the periodic inflammation schemes studied, increasing MIE will decrease the invasion-free survival (i.e.
+worsen cancer progression and prognosis)  (Fig.
+\ref{fig:VaryINFL_and_MesPars}B).
+In the case of continuously high inflammation, the effects of MIE are minimal.
+Thus, under any inflammation regime with periods of low inflammation, as we might intuitively assume, any reduction in mesenchymal immune evasion will lead to improvements in patient outcomes.
+
+\subsection{Analysis of data from TCGA supports model predictions: EMT phenotypic properties worsen patient outcomes}\label{tcga}
 
@@
 {\includegraphics[width=0.5\textwidth]{Figure5/heatmap2.jpg}}
-\caption{Summary of the contrasting effects of MIE and MGA on the Time to Cancer.
+\caption{Summary of the effects of MIE and MGA on invasion-free survival.
 }
@@
 \end{figure}
-
-\subsection{TCGA data analysis supports model predictions by showing that mesenchymal phenotypic properties reduce invasion-free survival }\label{tcga}
-
-The effects that mesenchymal phenotypic properties have on progression can also be assessed by studying mesenchymal immune evasion ($\Delta_\text{MIE}$) and mesenchymal growth arrest ($\Delta_\text{MGA}$) on a heatmap (Fig.
-\ref{fig:MIEvsMGA}).
-We found that over the full range of values of $\Delta_\text{MGA}$ considered, increasing $\Delta_\text{MIE}$ decreases the invasion-free survival time.
-However, for any given value of mesenchymal immune evasion, there is a value of mesenchymal growth arrest that maximizes the invasion-free survival.
-Moreover, this optimal value increases with increasing MIE.
-Together these results show that while reducing the ability of a mutated cell to evade the immune system by any amount will improve the probability of invasion-free survival, an optimal value of MGA will prolong the tumor remaining in situ.
-\par
-To compare these model predictions with experimental studies, we analyzed data from The Cancer Genome Atlas (TCGA) database to study the effects of immune and EMT interactions on prognosis of cancers for which inflammation is known to play an important role, such as colonic or pancreatic cancers \cite{greten2019inflammation,hu2010inflammation}.
-The TCGA Pan-Cancer Clinical Data Resource (TCGA-CDR) provides multiple computed clinical endpoints for PAAD \cite{liu2018integrated}.
-Here, we focus on the disease-free interval (DFI) and the overall survival (OS) endpoints.
-In essence, a tumor where the DFI is short will undergo rapid (post-treatment) initiation.
-In turn, we expect the distance between DFI and OS to be relatively small for a tumor undergoing rapid progression following initial (post-treatment) detection.
-Therefore, recurrent tumors may be roughly categorized as either rapidly progressing or slowly progressing by defining a threshold for the DFI/OS divergence.
-We selected three tumor types whose TCGA-CDR profile indicated slow progression (OV, SKCM, LIHC) and two tumor types whose profile indicated rapid progression (PAAD, LUAD).
-See Figures \ref{fig:OV} and \ref{fig:PAAD} for OV and PAAD. The others are in the supplement.
-\par
-For each cohort of cancer patients, we cluster the patients via k-means (n = 2) on gene ontologies relating to only EMT (Figures \ref{fig:OV}C, \ref{fig:PAAD}C), only inflammation (Figures \ref{fig:OV}E, \ref{fig:PAAD}E), and both EMT and inflammation (Figures \ref{fig:OV}A, \ref{fig:PAAD}A).
-We also plot the corresponding survival curves for each of the clusters obtained (utilizing the originally reported OS endpoint).
-We see that in both cases, survival is affected by the gene ontology signature, but the presence of both EMT and inflammatory signatures has a greater impact on survival than the effects of EMT alone.
-This suggests that when studying interactions between cancer and the immune system, is it important to consider EMT; the effects of which may impact the dynamics and should not be overlooked.
-
-We note that the comparison between simulation and data here is indirect since the model studies the progression of cancer from an arbitrary starting point, while the data address the progression of cancer following treatment, but analogous processes are at play during both the progression of  clonal dynamics addressed by the model, and the progression of disease after cancer treatment.
-In particular, the plasticity of tumor cells allows them to evade treatment by undergoing post-treatment processes resembling the de-novo appearance of cancer\cite{sanchez2018slow}.
 
@@
 \end{figure}
+
+%The effects that mesenchymal phenotypic properties have on progression can also be assessed by studying mesenchymal immune evasion ($\Delta_\text{MIE}$) and mesenchymal growth arrest ($\Delta_\text{MGA}$) on a heatmap (Fig.
+\ref{fig:MIEvsMGA}).
+We found that over the full range of values of $\Delta_\text{MGA}$ considered, increasing $\Delta_\text{MIE}$ decreases the invasion-free survival time.
+The effects on cancer progression of mesenchymal phenotypic properties are summarized in Fig.
+\ref{fig:MIEvsMGA}.
+For all values of MGA, increasing MIE leads to a decrease in invasion-free survival times.
+For a given value of MIE however, there is an optimal value of MGA that maximizes invasion-free survival.
+Moreover, this optimum increases with increasing MIE.
+%These results show that while reducing the ability of a mutated cell to evade the immune system by any amount will improve the probability of invasion-free survival, an optimal value of MGA will prolong the tumor remaining in situ.
+\par
+To compare these model predictions with experimental studies, we analyzed data from The Cancer Genome Atlas (TCGA) \cite{Weinstein}.
+We studied the effects of immune interactions and EMT on cancer prognosis, especially on tumors for which inflammation is known to play an important role, such as colonic or pancreatic  \cite{greten2019inflammation,hu2010inflammation}.
+The TCGA Pan-Cancer Clinical Data Resource provides multiple computed clinical endpoints for pancreatic cancer (PAAD) \cite{liu2018integrated}.
+Here, we focus on the disease-free interval (DFI) and the overall survival (OS).
+A tumor with a short DFI will likely undergo rapid progression post-treatment.
+We thus expect the distance between DFI and OS to be small for a rapidly progressing tumor, following initial detection.
+We can therefore characterize tumors as either rapidly progressing or slowly progressing by defining a threshold for the ratio DFI/OS.
+We selected three tumors for study whose Clinical Data Resource profiles indicated slow progression (ovarian, OV; skin cutaneous melanoma, SKCM; liver hepatocellular carcinoma, LIHC) and two tumors whose profiles indicated rapid progression (pancreatic cancer, PAAD; lung adenocarcinoma, LUAD).
+%See Figures \ref{fig:OV} and \ref{fig:PAAD} for OV and PAAD. The others are in the supplement.
+\par
+For each cohort of patients for which we have clinical and expression data, we cluster patients via k-means (n = 2) against the gene ontologies relating to either: EMT signature alone, inflammatory signature alone, or the combination of both signatures.
+We then plot the corresponding survival curves (using the OS from TCGA) for each of the two groups (Fig.
+\ref{fig:OV} for ovarian cancer and Fig.
+\ref{fig:PAAD} for pancreatic cancer).
+%(Figures \ref{fig:OV}C, \ref{fig:PAAD}C), only inflammation (Figures \ref{fig:OV}E, \ref{fig:PAAD}E), and both EMT and inflammation (Figures \ref{fig:OV}A, \ref{fig:PAAD}A).
+We see that for both these tumor types, survival is affected by the gene ontology signature, and the combination of both EMT and inflammatory signatures has a greater impact on survival than the effects of either EMT or inflammation alone.
+This suggests that in considering interactions between cancer and the immune system, is it of critical importance to consider the effects of EMT, as these can significant impact outcomes and should not be overlooked.
+\par
+The model considered here studies tumor progression from in situ to invasive disease from a homogeneous initial point, whereas the data address how cancer may progress following treatment, thus comparisons between model and data should be made carefully.
+Nonetheless the core cellular tumor dynamics are at play
+%The comparison between simulation and data here is indirect since the model studies the progression of cancer from an arbitrary starting point, while the data address the progression of cancer following treatment, but analogous processes are at play
+ both during the tumor progression addressed by the model, and post-treatment progression described in data from TCGA.
+Of particular note, the plasticity of tumor cells allows them to evade treatment by undergoing post-treatment processes resembling the de-novo appearance of cancer\cite{sanchez2018slow}.
 
@@
 Here we studied cancer, the immune system, and EMT, during the progression from an in situ tumor to invasive disease.
-We saw this as an even more pressing need given the shared factors that influence all of these, such as TGF-$\beta$.
-We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to driver mutations in pathways that affect the fitness of the cell, which in turn affects the cell population dynamics, which are also influenced by the intrinsic state of the cell (EMT), and extrinsic immune factors (inflammation, cytotoxic cells, etc).
+We saw this as a particularly pressing need given the shared factors influencing all these components, such as TGF-$\beta$.
+We used an individual cell-based model framework to describe the multiscale processes that can lead to cancer: DNA damage occurs during the cell cycle and this can lead to mutations in pathways that affect cell fitness, which in turn affects the cell population dynamics.
+Population dynamics are also influenced by the intrinsic state of the cell (through EMT), and extrinsic immune factors.
 \par
-We found that this model recapitulated invasion-free survival dynamics, and via global parameter sensitivity analysis, we identified parameters exerting key control over model behavior.
-Focusing on these led us to identify that: increasing mesenchymal immune evasion and increasing Treg TGF-$\beta$ production both lead to shorter invasion-free survival times.
+We found that this model recapitulated invasion-free survival dynamics.
+Using global parameter sensitivity analysis, we identified parameters exerting key control over model behavior.
+Focusing on these led us to identify that increasing mesenchymal immune evasion and increasing Treg TGF-$\beta$ production both lead to shorter invasion-free survival times.
 However, varying the level of inflammation led to paradoxical effects with regards to mesenchymal growth arrest: under regimes with periods of low inflammation, an optimal level of mesenchymal growth arrest can improve outcomes and maximize the invasion-free survival.
+To test these predictions, we performed unsupervised analysis of pancreatic cancer data from The Cancer Genome Atlas, and looked at survival across groups with {\em rapidly progressing} or {\em slowly progressing}  tumors.
+We found that combinatorial effects of EMT + inflammation increased the differences in survival between groups.
 \par
-To test these predictions, we performed unsupervised analysis of pancreatic cancer data from The Cancer Genome Atlas, and looked at survival across two groups: we found that the presence of an EMT signature increased differences in survival between groups in combination with inflammation.
-We chose to summarize {\em in silico} patient simulations with a single parameter: the invasion-free survival time, which we found captured the essential characteristics of the model.
-There are, of course, many tissue trajectories that all result in cancer, and it is our opinion that analysis of the transient cell dynamics in premalignant tissues is a pressing need to shed insight into cellular biomarkers of cancer.
+To capture the essential characteristics of the model, we summarized {\em in silico} patient studies with a single parameter: the invasion-free survival time (see Fig.
+1B for a reminder of the full model heterogeneity).
+There are, of course, many trajectories that result in cancer progression.
+Analysis of the transient cell dynamics in cancer in situ and during progression is a pressing need to shed insight into cellular biomarkers of cancer.
 \par
-These results represent promising steps in understanding the competing roles of the immune system and EMT during progression of epithelial cancers, yet much remains to be done.
-Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on the epithelia and cancer-free survival (Figs.
+The predictions of this model offer promising leads in elucidating the competing roles of immune cells and EMT during tumor progression, but much remains to be done.
+Further development of the inflammation module of this model is important given the large and sometimes paradoxical roles that the inflammatory state exerts on tumor cells and invasion-free survival (Figs.
 \ref{fig:MOAT} and \ref{fig:VaryINFL_and_MesPars}B, D).
-Currently, inflammation is modeled as independently cycling between high and low schemes, however many of the agents considered in the model actively contribute to the inflammatory state, thus as an extension these components could be coupled, for example by assuming that the level of inflammation depends on the number of and the degree of mutations that cells in the tumor harbor).
+Currently, inflammation is modeled as independently cycling between high and low schemes, however several model components contribute to the inflammatory state.
+This can be modeled for example by assuming that the level of inflammation depends on the number of and the degree of mutations that tumor cells harbor.
 %% not sure about
-Another layer of complexity is revealed by the natural anti-inflammatory role of Tregs.
+%Another layer of complexity is revealed by the natural anti-inflammatory role of Tregs.
 One consequence of the current model is that decreasing the number of Tregs increases the cancer-free survival.
 Clearly, there exists a trade-off to be accounted for, and adding to the model the main effector function of Tregs could remedy this and add depth to our understanding of the various roles that Tregs play in and around the tumor.
-\par
-The roles that TGF-$\beta$ plays throughout the tumor and its microenvironment also warrant further investigation.
+The roles that TGF-$\beta$ play throughout the tumor and its microenvironment also warrant further investigation.
 We found that, below a certain threshold, reduction of TGF-$\beta$ increases the Time to Cancer  (Fig.

### 2019-12-10 Paper/CleanPaper.tex
update res/discussion

@@
 
-\linenumbers
+%\linenumbers
 
@@
 Clearly, there exists a trade-off to be accounted for, and adding to the model the main effector function of Tregs could remedy this and add depth to our understanding of the various roles that Tregs play in and around the tumor.
-The roles that TGF-$\beta$ play throughout the tumor and its microenvironment also warrant further investigation.
-We found that, below a certain threshold, reduction of TGF-$\beta$ increases the Time to Cancer  (Fig.
-\ref{fig:FirstSurvivalCurves}E), thus reducing expression levels of TGF-$\beta$ in the tumor microenvironment benefits survival.
-Intriguingly, recent experimental work demonstrated that TGF-$\beta$ drives tumor suppression in pancreatic cancer by promoting EMT \cite{david16_tgfv}.
-However, TGF-$\beta$ is a master regulator implicated in numerous cellular signaling processes, and changing the concentrations of TGF-$\beta$ even in a local tumor microenvironment could have large off-target effects.
-Indeed, it has been shown that TGF-$\beta$ promotes invasion and heterogeneity (although suppresses cell proliferation) in squamous cell carcinoma \cite{oshimori15_tgfv}.
-Future work thus ought to consider the effects of targeting signaling factors downstream of TGF-$\beta$ that still have the ability to modulate epithelial cell dynamics.
-Towards this end, we are currently developing a larger TGF-$\beta$ signaling pathway module with appropriate crosstalks to  epithelial/mesenchymal/immune cell functions to be incorporated into the model.
+The competing roles that TGF-$\beta$ play throughout the tumor and its microenvironment also warrant further investigation.
+We found that -- below a certain threshold -- reduction of TGF-$\beta$ increases the time to invasion (Fig.
+\ref{fig:FirstSurvivalCurves}E): reducing  TGF-$\beta$ in the TME can benefit survival.
+Recent experiment work in support of this demonstrated that TGF-$\beta$ drives tumor suppression in pancreatic cancer by promoting EMT \cite{david16_tgfv}.
+TGF-$\beta$, however, is implicated in numerous other cellular signaling processes, and changing TGF-$\beta$ concentration even in a local environment could have large off-target effects.
+Indeed, it has been shown that TGF-$\beta$ promotes invasion and heterogeneity while  suppressing cell proliferation in squamous cell carcinoma \cite{oshimori15_tgfv}.
+To help account for such complex circuitry, future work should incorporate the effects of signaling factors downstream of TGF-$\beta$ on tumor cell dynamics.
+We are currently developing a larger TGF-$\beta$ signaling pathway component within the full model that permits crosstalks between ETCs, MTCs, and immune cell populations through transcriptional signaling activity.
 \par
-A further goal for future work is to explore (and exploit) the heterogeneity of tumor evolution in greater depth: this heterogeneity aids the evasion of the tumor from immune effects.
-Studying the consequences of decanalization \cite{gibson09_decanalization} during cancer progression is too-often sidelined, despite evidence supporting its prominence \cite{cyll17_tumour, punt17_tumour, dagogo-jack18_tumour}.
-Yet despite these challenges, for which the complexity of the disease may be often in large part responsible, great progress has been and continues to be made.
-As we approach a new generation of immunotherapies, it is these very complexities that we must better understand in order to control or eradicate the disease.
+Further {\em in silico} work on this model and its extensions will explore (and exploit) the heterogeneity of tumor evolution in greater depth.
+Tumor heterogeneity greatly enhances the capabilities of the tumor to evade immune effects.
+Studying the consequences of this increased heterogeneity following disease incidence, i.e.
+decanalization \cite{gibson09_decanalization}, is too-often sidelined, despite mounting evidence in support of its prominent role in cancer evolution \cite{cyll17_tumour, punt17_tumour, dagogo-jack18_tumour}.
+Despite these challenges (for which disease complexity is often in part responsible), great progress has been and continues to be made.
+As we are rapidly approaching a new generation of immunotherapies, it is these very complexities that we must better understand in order to control or eradicate the disease.
 

### 2019-12-11 Paper/CleanPaper.tex

@@
 \title{
-Modeling the competing effects of the immune system and EMT on epithelial cancers
+Modeling the effects of the immune system and EMT on epithelial cancer progression
 }
@@
 Moreover, tumor cell plasticity can significantly alter disease trajectories via epithelial-to-mesenchymal transition (EMT).
-Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of these regulatory processes.
-Here we introduce a multiscale evolutionary model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial cancers.
-Through in silico analysis of large patient cohorts, we find controllable regimes that maximize invasion-free survival.
+Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of crosstalk between these regulatory processes.
+Here we introduce a multiscale evolutionary model to describe  tumor-immune-EMT interactions and their impact on epithelial cancer progression from in situ to invasive disease.
+Through in silico analyses of large patient cohorts, we find controllable regimes that maximize invasion-free survival.
 We identify that delaying tumor progression depends crucially on properties of the mesenchymal tumor cell phenotype: its growth rate and its immune-evasiveness.
-Through analysis of pancreatic and ovarian cancers from The Cancer Genome Atlas to assess model predictions against clinical measurements, we find that association with EMT significantly worsened the invasion-free survival probabilities of inflammation-associated cancer.
+Through analysis of inflammation-associated cancer data from The Cancer Genome Atlas, we find that -- in agreement with model predictions -- association with EMT significantly worsens invasion-free survival probabilities.
 These results offer novel means to delay disease progression by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.
@@
 
-%\linenumbers
+\linenumbers
 
@@
 
-\subsection{Tumor Evolution}\label{TissueCells}
+\subsection{Tumor evolution}\label{TissueCells}
 Associated with each tumor cells are two essential features: their mutational signature and their EMT score.
@@
 
-\subsection{Immune Cell Population Dynamics}\label{ImmuneSystem}
+\subsection{Immune population dynamics}\label{ImmuneSystem}
 The immune system is modeled by three immune cell types: NKs, CTLs, and Tregs.
@@
 
-\subsection{Periodic Inflammation} 
+\subsection{Periodic cycling inflammation states} 
 Inflammation is modeled as a cycling scheme between low and high inflammatory states, with varying on/off durations and intensities.
@@
 
-\subsection{Epithelial-to-mesenchymal Transition}\label{EMT}
+\subsection{Epithelial-to-mesenchymal transition}\label{EMT}
 Each tissue cell has an EMT score between 0 and 1 with is set according to the concentration of TGF-$\beta$ in the TME.
@@
 
-\subsection{Model Simulation}
+\subsection{Model simulation}
 %\subsubsection{Initial conditions} 
@@
 %\subsubsection{Completing the Cell Cycle}
-\paragraph{Completing the Cell Cycle.}
+\paragraph{Completing the cell cycle.}
 Once all tumor cells have been updated and fates chosen accordingly, non-tumor model components are updated.
@@
 
-\subsection{Parameter Choice and Sensitivity Analysis}
+\subsection{Parameter estimation and sensitivity analysis}
 To study parameter sensitivity, we implemented Morris a one-step-at-a-time global sensitivity analysis.
@@
 
-\subsection{Analysis of patient survival data from TCGA database}
+\subsection{Analysis of patient survival data from TCGA }
 We obtain primary tumor bulk mRNA sequencing and censored survival data for all individuals monitored in the relevant project from the Cancer Genome Atlas (TCGA) from the Genomic Data Commons portal in R, using the package TCGABiolinks.
@@
 \center
-{\includegraphics[width=0.5\textwidth]{Figure5/heatmap2.jpg}}
-\caption{Summary of the effects of MIE and MGA on invasion-free survival.
+{\includegraphics[width=0.85\textwidth]{Figure5/FigHeatmapLIHC.pdf}}
+\caption{{\bf A.} Summary of the effects of MIE and MGA on invasion-free survival.
+  {\bf B.} Bayesian classification model applied to liver hepatocellular carcinoma (LIHC).
+Patients projected into 2D by principle components analysis to determine most probable regions of disease-free interval (DFI).
+Two local maxima are identified, separated by \tcr{XXXX}.
 }
@@
 We found that over the full range of values of $\Delta_\text{MGA}$ considered, increasing $\Delta_\text{MIE}$ decreases the invasion-free survival time.
-The effects on cancer progression of mesenchymal phenotypic properties are summarized in Fig.
-\ref{fig:MIEvsMGA}.
-For all values of MGA, increasing MIE leads to a decrease in invasion-free survival times.
-For a given value of MIE however, there is an optimal value of MGA that maximizes invasion-free survival.
-Moreover, this optimum increases with increasing MIE.
-%These results show that while reducing the ability of a mutated cell to evade the immune system by any amount will improve the probability of invasion-free survival, an optimal value of MGA will prolong the tumor remaining in situ.
+
 \par
 To compare these model predictions with experimental studies, we analyzed data from The Cancer Genome Atlas (TCGA) \cite{Weinstein}.
-We studied the effects of immune interactions and EMT on cancer prognosis, especially on tumors for which inflammation is known to play an important role, such as colonic or pancreatic  \cite{greten2019inflammation,hu2010inflammation}.
+We studied the effects of immune interactions and EMT on cancer prognosis, especially on tumors for which inflammation is known to play an important role, such as those of the colon, or pancreas, or liver  \cite{greten2019inflammation,hu2010inflammation}.
 The TCGA Pan-Cancer Clinical Data Resource provides multiple computed clinical endpoints for pancreatic cancer (PAAD) \cite{liu2018integrated}.
@@
 We can therefore characterize tumors as either rapidly progressing or slowly progressing by defining a threshold for the ratio DFI/OS.
-We selected three tumors for study whose Clinical Data Resource profiles indicated slow progression (ovarian, OV; skin cutaneous melanoma, SKCM; liver hepatocellular carcinoma, LIHC) and two tumors whose profiles indicated rapid progression (pancreatic cancer, PAAD; lung adenocarcinoma, LUAD).
-%See Figures \ref{fig:OV} and \ref{fig:PAAD} for OV and PAAD. The others are in the supplement.
+We selected tumors for which Clinical Data Resource profiles indicated slow progression (ovarian, OV; skin cutaneous melanoma, SKCM; liver hepatocellular carcinoma, LIHC) and two tumors whose profiles indicated rapid progression (pancreatic cancer, PAAD; lung adenocarcinoma, LUAD).
 \par
-For each cohort of patients for which we have clinical and expression data, we cluster patients via k-means (n = 2) against the gene ontologies relating to either: EMT signature alone, inflammatory signature alone, or the combination of both signatures.
+The effects on cancer progression of mesenchymal phenotypic properties predicted by the model are summarized in Fig.
+\ref{fig:MIEvsMGA}A.
+Where, %For all values of MGA, increasing MIE leads to a decrease in invasion-free survival times.
+for a given value of MIE, there is an optimal value of MGA that maximizes invasion-free survival. %Moreover, this optimum increases with increasing MIE.
+We use two models to compare these predictions with TCGA data: a Bayesian supervised classification model, and survival analysis with k-means clustering (see Methods).
+The classification model applied to liver carcinoma identifies local maxima that promote survival, demonstrating intriguing similarities with model results (Fig.
+\ref{fig:MIEvsMGA}B).
+\tcr{Classification plots for OV/PAAD in SI}
+\par
+For each cohort of patients for which we have clinical and expression data, we cluster the patients via k-means (n = 2) against gene ontologies relating to either: EMT signature alone, inflammatory signature alone, or the combination of both signatures.
 We then plot the corresponding survival curves (using the OS from TCGA) for each of the two groups (Fig.

### 2019-12-11 Paper/structure.tex

@@
 
-%\newcommand{\HorRule}{\color{DarkGrey}\rule{\linewidth}{1pt}} % Defines the gold horizontal rule
-\newcommand{\HorRule}{\color{DarkGoldenrod}\rule{\linewidth}{1pt}} % Defines the gold horizontal rule
+\newcommand{\HorRule}{\color{DarkGrey}\rule{\linewidth}{1pt}} % Defines a grey horizontal rule
+%\newcommand{\HorRule}{\color{DarkGoldenrod}\rule{\linewidth}{1pt}} % Defines the gold horizontal rule
 
@@
 	\vspace{-30pt} % Move the entire title section up
-	\HorRule\vspace{10pt} % Horizontal rule before the title
+	\HorRule
+        \vspace{10pt} % Horizontal rule before the title
 	\fontsize{20}{18}\usefont{OT1}{phv}{b}{n}\selectfont % Helvetica
@@
 }
-
 \posttitle{\par\vskip 15pt} % Whitespace under the title
-
 \preauthor{} % Anything that will appear before \author is printed
-
 \postauthor{ % Anything that will appear after \author is printed
@@
 	\par\HorRule % Horizontal rule after the title
-	\vspace{20pt} % Space after the title section
+	\vspace{10pt} % Space after the title section
 }
@@
 		%\color{DarkRed}% Lettrine colour
-		\color{DarkGoldenrod}% Lettrine colour
+		\color{DarkRed}% Lettrine colour
 		{#1}% The letter
@@
 	\StrLeft{#1}{1}[\firstletter] % Capture the first letter of the abstract for the lettrine
-	\initial{\firstletter}\textbf{\StrGobbleLeft{#1}{1}} % Print the abstract with the first letter as a lettrine and the rest in bold
+	\initial{\firstletter}{\StrGobbleLeft{#1}{1}} % Print the abstract with the first letter as a lettrine and the rest in bold
 }

### 2019-12-12 Paper/CleanPaper.tex
responded to adam's suggestions

@@
 TGF-$\beta$, an activator of EMT, is produced both by Tregs and tumor cells, thus connecting tumor-immune interactions with EMT, and as a result plays an important role in shaping tumor outcomes.
-EMT depends on TGF-$\beta$ concentration through a Hill-type function that determines the likelihood of EMT (or the reverse MET) through its magnitude relative to $\tau_{i-1}$.
-This is described by the equation:
+To determine whether a cell undergoes EMT or MET, the TGF-$\beta$ is randomly divided among the cells so that associated with each cell $i$ is a value $\tau_i$.
+This value is given by the following equation:
 $$
@@
 Here, $\tau$ is the concentration of TGF-$\beta$ in the TME and $\tau_\text{max}$ represents a limit on the amount of TGF-$\beta$ that can be absorbed by all cells, with Gaussian noise added.
+\newline
+For each cell, $\tau_i$ is summed with the current EMT score for that cell and if the result is above a threshold value, the cell undergoes EMT, otherwise MET.
+This summation expresses the assumption that the epithelial and mesenchymal phenotypes are stable and cells will move along the EMT spectrum towards the equilibrium they identify as under typical conditions.
+%This is then combined with the cell's EMT score and a threshold value determines if the cell subsequently undergoes EMT or MET.
+%In doing this, we are assuming that the epithelial and mesenchymal phenotypes are stable and cells in a given type tend to move towards that 
+%This represents cell-extrinsic sources shaping the trajectory of the cell along the EMT spectrum.
+%Intrinsically, we assume that a cell is likely to keep its same phenotype, so we mesenchymal cells are more li
+%If the difference is large enough, the cell undergoes EMT, otherwise it undergoes MET.
+%This value $\tau_i$ is given by the following equation:
+%EMT depends on TGF-$\beta$ concentration through a Hill-type function that determines the likelihood of EMT (or the reverse MET) through its magnitude relative to $\tau_{i}$.
+This is described by the equation:
     \newline
@@
 \caption{Effects of mesenchymal tumor cell properties on the time to invasion.
-Trajectories of one patient per cohort including warmup and 2000 cell cycles for {\bf A.} mesenchymal immune evasion (MIE); {\bf A.} mesenchymal growth arrest (MGA); {\bf C.} Production of TGF-$\beta$ by Tregs.
+Trajectories of one patient per cohort including warmup and 2000 cell cycles for {\bf A.} mesenchymal immune evasion (MIE); {\bf B.} mesenchymal growth arrest (MGA); {\bf C.} Production of TGF-$\beta$ by Tregs.
 {\bf D.} Survival curve corresponding to changes in the parameter MIE (A) for a patient cohort of 1000. Shaded region represents the 95\% confidence interval over the cohort. 
@@
 \par
+\tcr{
 TGF-$\beta$ varies according to its production by tumor cells and its production by Tregs.
-Here we assess the effects of varying the production of  TGF-$\beta$ by Tregs on invasion-free survival (Fig \ref{fig:FirstSurvivalCurves}B, E).
-We find, interestingly, that at lower production rates of TGF-$\beta$ lead to more rapid invasion or if it survives to a certain time point, a longer time to invasion.
+Here we assess the effects of varying the production of  TGF-$\beta$ by Tregs on invasion-free survival (Fig \ref{fig:FirstSurvivalCurves}C, F).
+We find that at lower production rates of TGF-$\beta$, the survival curve initially declines faster whereas higher production rates result in a steeper drop off in survival later.
+This is due to the possibility that the immune system rapidly clears the tumor upon influx of adaptive immune cells.
+A higher production of TGF-$\beta$ by the Tregs mitigates this possibility.
+}
+
+% lead to more rapid invasion or if it survives to a certain time point, a longer time to invasion.
 

### 2019-12-12 Paper/Supplement/CleanSupplement.tex
responded to adam's suggestions

@@
 \tableofcontents
+\listoftables
+\listoffigures
 
@@
 \end{tabular}
-  \caption{The model parameter names and descriptions. Note that many of these values are affected by the inflammation state of the system.}
+  \caption[Parameter names and descriptions]]{The model parameter names and descriptions. Note that many of these values are affected by the inflammation state of the system.}
 \end{center}
@@
   \hline
-  \caption{The model parameter names, descriptions, and values during both low and high inflammation. Parameters with only one value do not change with the inflammatory state.}
+  \caption[Parameter values]{The model parameter names, descriptions, and values during both low and high inflammation. Parameters with only one value do not change with the inflammatory state.}
 \end{longtable}
+
+%%%%%%%%%%%%%%%%
+% Figures
+%%%%%%%%%%%%%%%%
+
+\begin{figure}
+\center
+{\includegraphics[width=1\textwidth]{../Figs/Figure1/SI_fig1.pdf}}
+\caption[Fig. 1BC without targeting premalignant cells]{
+A.
+Single patient trajectory without immune cells targeting premalignant cells.
+Compare to Fig.
+1B.
+B.
+Sample cohort survival curve without immune cells targeting premalignant cells.
+Compare to Fig.
+1C.
+}
+\label{fig:1_first}
+\end{figure}
+
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure2/Figure2.jpg}}
+\caption[Fig. 2 without targeting premalignant cells]{
+Morris-OAT global sensitivity without immune cells targeting premalignant cells. Compare to Fig. 2.
+}
+\label{fig:2_first}
+\end{figure}
+
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure3/Outdated/Figure3_first.pdf}}
+\caption[Fig. 3 without targeting premalignant cells]{
+Effects of mesenchymal tumor cell properties on the Time to Cancer without immune cells targeting premalignant cells. Compare to Fig. 3.
+}
+\label{fig:3_first}
+\end{figure}
+
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure4/Figure4_first.pdf}}
+\caption[Fig. 4 without targeting premalignant cells]{
+Effects of inflammation on the Time to Cancer without immune cells targeting premalignant cells. Compare to Fig. 4.
+}
+\label{fig:4_first}
+\end{figure}
+
+\begin{figure}
+\center
+{\includegraphics[width=0.65\textwidth]{../Figs/Figure5/MIEvsMGA_bigcbar.jpg}}
+\caption[Fig. 5 without targeting premalignant cells]{
+Summary of the contrasting effects of MIE and MGA on Time to Cancer without immune cells targeting premalignant cells. Compare to Fig. 5.
+}
+\label{fig:5_first}
+\end{figure}
 
@@
 {\includegraphics[width=0.65\textwidth]{../Figs/Figure6/SKCM.pdf}}
-\caption{A. K-means clustering of SKCM using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
+\caption[SKCM TCGA Data Analysis]{A. K-means clustering of SKCM using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
 B. Survival plots corresponding to the clustering on EMT and inflammation.
@@
 {\includegraphics[width=0.65\textwidth]{../Figs/Figure6/LIHC.pdf}}
-\caption{A. K-means clustering of LIHC using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
+\caption[LIHC TCGA Data Analysis]]{A. K-means clustering of LIHC using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
 B. Survival plots corresponding to the clustering on EMT and inflammation.
@@
 {\includegraphics[width=0.65\textwidth]{../Figs/Figure6/LUAD.pdf}}
-\caption{A. K-means clustering of LUAD using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
+\caption[LUAD TCGA Data Analysis]{A. K-means clustering of LUAD using gene ontology terms indicative of EMT and inflammation signatures ($k=2$).
 B. Survival plots corresponding to the clustering on EMT and inflammation.
@@
 
-\begin{figure}
-\center
-{\includegraphics[width=0.65\textwidth]{../Figs/Figure1/SinglePatientBlue.jpg}}
-\caption{
-Single patient trajectory without immune cells targeting premalignant cells. Compare to Fig. 1B.
-}
-\label{fig:1b_first}
-\end{figure}
-
-\begin{figure}
-\center
-{\includegraphics[width=0.65\textwidth]{../Figs/Figure1/OneSurvivalCurve.jpg}}
-\caption{
-Sample cohort survival curve without immune cells targeting premalignant cells. Compare to Fig. 1C.
-}
-\label{fig:1c_first}
-\end{figure}
-
-\begin{figure}
-\center
-{\includegraphics[width=0.65\textwidth]{../Figs/Figure2/Figure2.jpg}}
-\caption{
-Morris-OAT global sensitivity without immune cells targeting premalignant cells. Compare to Fig. 2.
-}
-\label{fig:2_first}
-\end{figure}
-
-\begin{figure}
-\center
-{\includegraphics[width=0.65\textwidth]{../Figs/Figure3/Figure3_first.pdf}}
-\caption{
-Effects of mesenchymal tumor cell properties on the Time to Cancer without immune cells targeting premalignant cells. Compare to Fig. 3.
-}
-\label{fig:3_first}
-\end{figure}
-
-\begin{figure}
-\center
-{\includegraphics[width=0.65\textwidth]{../Figs/Figure4/Figure4_first.pdf}}
-\caption{
-Effects of inflammation on the Time to Cancer without immune cells targeting premalignant cells. Compare to Fig. 4.
-}
-\label{fig:4_first}
-\end{figure}
-
-\begin{figure}
-\center
-{\includegraphics[width=0.65\textwidth]{../Figs/Figure5/MIEvsMGA_bigcbar.jpg}}
-\caption{
-Summary of the contrasting effects of MIE and MGA on Time to Cancer without immune cells targeting premalignant cells. Compare to Fig. 5.
-}
-\label{fig:5_first}
-\end{figure}
-
 \end{document}

### 2019-12-12 Paper/Figs/Figure3/ChangVaryMIEColors.m

@@ -1,20 +1,0 @@
-close all; clear all;
-
-openfig('VaryMIE_BigChange.fig');
-h=gca;
-
-for i = 1:length(h.Children)
-    ht = h.Children(i);
-    if endsWith(class(ht),'Patch')
-        col = ht.FaceColor;
-        col(1:2) = col(2:-1:1);
-        set(ht,'FaceColor',col);
-    elseif endsWith(class(ht),'Stair')
-        col = ht.Color;
-        col(1:2) = col(2:-1:1);
-        set(ht,'Color',col);
-    end
-end
-
-savefig('VaryMIE_BigChange.fig')
-print('VaryMIE_BigChange.jpg','-djpeg')
